### 2019-08-30 paper/paper.tex
0 version of paper

\documentclass[a4paper]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[dvipsnames]{xcolor} % for BrickRed color in abbrevs.tex
\usepackage{hyperref}
\usepackage{todonotes}

\input{abbrevs.tex}

\allowdisplaybreaks

\begin{document}

\title{Reducing finitary inductive-inductive types to inductive types}

\maketitle

\section{Introduction}
\label{sec:intro}

Note by Ambrus: to be beautified, I only wanted to add the line of
thought.

Most mutual inductive types (e.g.\ isEven-isOdd) can be reduced to
indexed inductive types where the index gives the information which
sort we meant. (Indexed inductive types can be reduced to inductive
types using a method similar to what we use here, preterms and
well-formedness predicate.) Inductive-inductive types (IITs
\cite{forsberg-phd}) allow the mutual definition of a type and a
family of types over the first one. The above reduction does not work
anymore.

Examples of IITs: Con-Ty is the minimal interesting one, then we have
James Chapman style \cite{chapman09eatitself} well-typed syntax for
type theory.

TODO: list examples here and explain. Also the elimination principles
which is recursive-recursive. Talk about connection to setoidified
CwFs (I think these are explained in some Dybjer paper) and maybe the
initiality conjecture.

Reduction of IITs to inductive types is an open problem. Forsberg
showed that IITs can be reduced to inductive types in the setting of
ETT, but he only constructed the nondependent (or simple? was this the
non-recursive-recursive?) elimination principle. Hugunin \cite{jasper}
reduced an example IIT to inductive types without UIP, but also only
constructing the non-recursive-recursive eliminator. Streicher
\cite{streichersemantics}. Brunerie \cite{} constructed the initial
category with extra structure (TODO: fix this) using UIP.

Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
inductive-inductive types (HIITs)
\cite{kaposi_et_al:LIPIcs:2018:9190}. This is a small type theory in
which every context encodes a valid IIT signature. For example, the
signature for natural numbers is simply the context
$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
$suc$ are simply variable names). Later they showed
\cite{Kaposi:2019:CQI:3302515.3290315} that if a subset of this theory
of signatures exists, then any finitary QIIT can be defined from
it. In this paper we further restrict the subset by not allowing
equality constructors (thus only covering IITs) but at the same time
we extend it to allow infinitary constructors. To derive the existence
of all IITs from the theory of IIT signatures we use a variant of the
CwF model \cite{Kaposi:2019:CQI:3302515.3290315} where types are
modelled not by displayed CwFs but by CwF isofibrations. We need this
extension to support infinitary constructors. Furthermore, using the
method introduced by Streicher \cite{streichersemantics} we show the
existence of the theory of IIT signatures. As opposed to Streicher, we
don't use quotients in this construction as the theory of IIT
signatures can be given only using neutral terms.

Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
we can translate our arguments to ITT+UIP+funext. Section
\ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
funext.

\subsection{Structure of the paper}

We first define a \emph{theory of IIT signatures} in which every
context is a valid IIT signature (Section
\ref{sec:theory_of_signatures}). We present this type theory using an
algebraic \cite{ttintt} syntax, that is, as a quotient
inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show
that this QIIT can be constructed only using inductive types.

\section{The theory of IIT signatures}
\label{sec:theory_of_signatures}

We use the naming convention: sorts, operators and equations. Sorts
and operators might have parameters.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \cdot && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\cdot \\
  & {\cdot\eta} && : \{\sigma : \Sub\,\Gamma\,\cdot\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
  & \rlap{(5) Infinitary parameters} \\[0.5em]
  & \Pii && : (T:\Set)\ra(T\ra\Tm\,\Gamma\,\U)\ra\Tm\,\Gamma\,\U \\
  & \blank\appi\blank && : \Tm\,\Gamma\,(\El\,(\Pii\,T\,b))\ra(\alpha:T)\ra\Tm\,\Gamma\,(\El\,(b\,\alpha)) \\
  & {\Pii[]} && : (\Pii\,T\,b)[\sigma] = \Pii\,T\,(\lambda \alpha.(b\,\alpha)[\sigma]) \\
  & {\appi[]} && : (t\appi\alpha)[\sigma] = (t[\sigma])\mathop{\appi}\alpha
\end{alignat*}
Abbreviation:
\[
  \sigma^\uparrow = (\sigma\circ\p,\q)
\]

TODO: add examples where all the different constructors are used,
especially infinitary ones.

\section{Constructing the theory of IIT signatures}
\label{sec:ambroise}

Show that the theory of IIT signatures as described in Section
\ref{sec:theory_of_signatures} exists, that is, it can be defined only
using indexed inductive types and a $\Prop$ sort.

Ambroise has done this in Agda with funext and UIP using an
extrinsic--intrinsic translation, but only for parts (1)--(5), and the
following parts are missing: $(\prop{1})$--$(\prop{5})$. We need
the $(\prop{1})$--$(\prop{5})$ parts to justify the next section.

Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.

In this formalization, there is a discrepancy for the typing rule of application
compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
\[
   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
\]
but
\[
   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
\]
Actually, I started with the first version, but at some point I
changed for the second one because it made some stuff easier (although
I don't remember exactly what it was).

I think this should be doable with only UIP (or a $\Prop$-valued
equality type?) and not funext. This would be important for a
computational justification of IITs.

Unfortunately, it seems that funext is needed (see the use of funext in
Syntax.agda) for the metatheoretic/infinitary
parameters part (4-5), for example in the case of the metatheoretic
$ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
investigated more carefully.

\paragraph{Construction}
\input{universaliit}

\section{Constructing all IITs from the theory of signatures}
\label{sec:andras}

Show using the CwF model that the existence of the theory of IIT
signatures (Section \ref{sec:theory_of_signatures}) entails the
existance of all IITs.

\bibliographystyle{plain}
\bibliography{b}

\end{document}

### 2019-08-30 paper/universaliit.tex
0 version of paper

We work in the setting of extensional type theory, although we occasionally indicate through $e_1,\dots,e_n \#
t$ that $t$ is well-typed thanks to the equalities $e_1$,\dots,$e_n$.
% We are also careful to explain where UIP or functional extensionality that is
% available in extensional type theory are used.

The steps consist of the following:
\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped
  syntax), and construction of a model of the theory of IIT
  signatures from well-formed terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the syntax and a given
    model (the idea is that given a syntactic context $\Gamma$ and a semantic
    context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
    \model{\Gamma}$ if and only if $\rec\,\Gamma = \model{\Gamma}$, and
    similarly for types, terms, and substitutions);
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item Showing that this relation is functional;
    \end{enumerate}
  \item Proving of the uniqueness of this morphism by showing that
    any morphism $f:S\to M$ satisfies the relation
    (in the sense, for contexts, that given any syntactic context $\Gamma$, we
    have $\Gamma\sim f\,\Gamma$).
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic model}

The goal is to define the syntactic model by packing the untyped syntax
with the typing judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\Gamma \Gamma\vdash
 \\
 & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
 \\
 & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
 \\
 & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
\end{alignat*}
The next sections are devoted to the definition of the untyped syntax and the
typing judgments.

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype:

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4) Metatheoretic and infinitary parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(5) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}
Note that we choose to share the constructor for untyped application to infinitary and
meta-theoretic parameters.
Variables are modelled as De Bruijn indices, i.e. as natural numbers pointing
to the relevant position in the context.

% The type of (untyped) substitution is then defined as $\Subu = \List\,\Tmu$.
% Note that $\Conu$ could also be defined as a list of types in a similar fashion.

\subsubsection{Untyped weakening}

% Substitutions of types and terms are defined recursively.

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,A[\sigma]\,B[\wk\,\sigma]$, and thus we need to define $\wk$, the weakening substitution.
The basic idea is to increment the de Bruijn indices of all the variables.
Actually this is not so simple because of the $\Piu$ type:
indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type of the weakening
of $A$ and $B$, but here, $B$ must be weakened with respect to the second last
variable of the context, rather than the last one.
For this reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

  \subsubsection{Untyped substitution}
  We define by induction the specific substitution of one variable of the context:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the well-typed judgments: indeed, the typing rule of application
  involves only a unary substitution.

  However, to construct the initial model of the QIIT, we need to define
  full substitution calculus:
  % First, we define

  % We then define by mutual recursion the full substitution of types and terms.
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
  These can be defined either by iterating unary substitutions, or
  by recursion on the syntax on the untyped syntax: the two ways yield provably
  equal definitions. In the following, we assume that it is defined by recursion.
 We use the additional default constructor
$\erru:\Tmu$ in case of error (for example, when the substitution is not
long enough).
This is an irrelevant constructor in the end, as the typing
judgments will not mention it.

We also make use of the following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. \varu\, 0 \,\consu\, \wkO\,\sigma
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma, A[\sigma]$
  and $\Delta,A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu (A[\sigma])(B[\keep \,\sigma])$.

  We can define the identity substitution on a context $\Gamma$ as follows,
  where $\length{\Gamma}$ is the length of the context $\Gamma$, and
  $\keep^{\length{\Gamma}}$ is $\keep$ iterated $\length{\Gamma}$ times:
\begin{alignat*}{5}
  & \length{\Gamma} && : \Conu\ra\N
  \\
  & \idu && : \Conu \ra \Subu \\
  & && := \lambda \Gamma.\keep^{\length{\Gamma}} \epsilonu
  \end{alignat*}

  \subsubsection{Exchange laws: weakening and substitution}
  Many lemmas are shown by recursion on the untyped syntax, for terms and types
  (below, $Z$ denotes either a term or a type):
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wk\mhyphen\wk && : \wk_{n+p+1} (\wk_n\,Z) = \wk_n(\wk_{n+p}\,Z)\\
  % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
  & \wk_n[n] && : (\wk_{n} Z)[ n := z] = Z \\
  % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
  & \wk_+[] && : (\wk_{n+p+1} Z)[ n := \wk_p u] = \wk_{n+p}(Z[n := u]) \\
  % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
  & \wk[+] && : (\wk_{n} Z)[ n+p+1 :=  u] = \wk_{n}(Z[n+p := u]) \\
  % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])]
  \end{alignat*}
  The following lemmas are shown by recursion on the syntax, for types and terms:
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  % liftV=wkS
  & [\keep^n\mhyphen \wkO] && : Z[\keep^n (\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
  \\
  % lifZₛV : ∀ n xp σp Zp → (lifZV n xp [ iZer n keep (Zp ∷ σp) ]V) ≡ (xp [ iZer n keep σp ]V)
  & \wk_n[\keep^n\mhyphen\cons]  & &
  :
  (\wk_n Z) [\keep^n (u \consu \sigma)] = Z [ \keep^n \sigma]
  \\
  % l-sub[]V : ∀ n x z σ
    % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
   & [:=][\keep] && : Z [ n:= u][\keep^n \sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
   \\
   & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau)
   \\
  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau
  \end{alignat*}
  where $\keep^n$ is $\keep$ iterated $n$ times. As a particular case
  for $n=0$, we get
\begin{alignat*}{5}
  & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
  & \wk_0[\cons] && : (\wk_0 Z)[u \consu \sigma] = Z[\sigma] \\
   & [0:=][] && : Z [ 0:= u][ \sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
  \end{alignat*}

Finally, we show the usual lemmas about substitution:
\begin{alignat*}{5}
  & [][] && : Z[\sigma][\tau] = Z[\sigma\circ\tau]
  \\
  & \ass && : (\sigma\circ \delta)\circ\tau = \sigma\circ (\delta\circ\tau)
  \end{alignat*}
  Other substitution lemmas are deferred after the definition of the typing
  judgments, as the proofs require that some input are well-typed.

  % The two ways yield provably equal definitions, but
  % The advantage of the second
  % one in a type theory with UIP and function extensionality is that we get
  % definitional equalities such as $(t\,\appu\,u)[\sigma]=\app\,t[\sigma]\,u[\sigma]$.

  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu

  \subsubsection{Typing judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)\ra
    \\
    & &&
    (\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \\
    & &&
    (\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
    \\
  & \rlap{(5) Infinitary parameters} \\[0.5em]
  & \Piiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in \Uu)
    \\ &&& \ra
    \Gamma \vdash \Pimu \,T\,A \in \Uu
  \\
  & \appiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in\Uu)
    \\
    & &&
    \ra
    (\Gamma \vdash t \in \Piiu \,T\, A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
Note that there is possibility of redundancy in the arguments of
the constructors. Here, we are rather ``{paranoid}'', so that we get more inductive
hypotheses when performing recursion.

% The substitution part could be moved after the mutual definition of terms, types,
% and contexts.

\subsubsection{Typed weakening}
Untyped weakening performs weakening anywhere in the context.
Stating a typing rule for this operation requires to have a way to express
the weakening occuring at the middle of a context.
To this end, we consider pairs of untyped contexts, which should be thought
of as a splitting at some place of a full context.
This full context is recovered by merging the two components:
% (which appears in the typing rule of a weakening occuring in the middle of a context),
% and the length $\length{\Gamma}$ of a
% context $\Gamma$:
\begin{alignat*}{5}
  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
  & \Gamma \merge \cdot && :=  \Gamma \\
  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
\end{alignat*}
We think of the second context as a telescope over the first context.
We need to define the define also the weakening of a telescope in order to
state the typing rule of weakening:
\begin{alignat*}{5}
  & \wkO && : \Conu\ra \Conu
  \\
  & \wkO\,\emptyu && := \emptyu
  \\
  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
  \end{alignat*}
Now we state and prove by mutual recursion that typing judgments are stable under
weakening, for contexts, types, terms, and substitutions:
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash) \ra
  \Gamma \extu A \merge \wkO\, \Delta \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, B  \\
  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash t\in B) \ra
  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, t \in \wk_{\length{\Delta}}\,B  \\
  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \vdash \sigma
  \Rightarrow \Delta) \ra
  \Gamma \extu A \vdash \wkO\, \sigma \Rightarrow \Delta
  \end{alignat*}
\subsubsection{Typed substitution}
  We show that judgments are stable under substitution, by mutual recursion:
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash t \in A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\Delta \vdash x \invar A)\ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \Gamma \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\Gamma \vdash) \ra
  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  (\Delta \vdash \tau \Rightarrow E) \ra
  \Gamma \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
  % (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
  % \Gamma \vdash A [ \sigma ]
  % \\
  \end{alignat*}
Then, we show that identity substitutions is neutral, for well-typed types,
terms, and substitutions:
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\Gamma \vdash A)\ra A [ \idu\,\Gamma]  \\
  & [\idu] && : (\Gamma \vdash x \invar A)\ra x [ \idu\,\Gamma] = V x \\
  & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma
  \end{alignat*}
  In a seperate and independant recursion, one shows that:
  \begin{alignat*}{5}
  & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
  \end{alignat*}
  Then, we show that the identity substitution itself is well typed:
  \begin{alignat*}{5}
  & \idw && : (\Gamma \vdash)\ra \Gamma \vdash \idu\,\Gamma \Rightarrow \Gamma
  \end{alignat*}

\subsubsection{Uniqueness of typing and uniqueness of types}
  We define what it means for a type $A$ to be a proposition as follows:

  \[
    \isaprop\,A := (a : A) \ra (a' : A) \ra a = a'
  \]
  It states that there is at most one inhabitant.
We show by recursion that each of these typing judgments is unique in the
following sense:
\begin{align*}
    \isp{\Conw}
  : \isaprop\, (\Gamma\vdash)
 & &
    \isp{\Tmw}
  : \isaprop\, (\Gamma\vdash t \in A)
    \\
    \isp{\Tyw}
  : \isaprop\, (\Gamma\vdash A)
    &&
    \isp{\wf\Var}
  : \isaprop\, (\Gamma\vdash x \invar A)
  \\
  &
        \isp{\wf\Sub}
  : \isaprop\, (\Gamma\vdash \sigma \in \Delta)
  \end{align*}
Actually, the application and successor cases require to show uniqueness of
types in a given context, before.
Let us consider for instance the application constructor $\appw$: it provides a
type $B$ such that the final type is $C=B[0 := u]$. Even if $C$ is known a
priori, there may be another $B$ for which $B[0 := u] = C$, possibly leading to
many proofs that $t\appu u$ has type $C$. Uniqueness of types
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$ of $t$.
More formally, we show by recursion:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\Gamma \vdash t \in A) \ra
  (\Gamma \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\Gamma \vdash x \invar A) \ra
  (\Gamma \vdash x \invar B) \ra A = B
  \end{alignat*}

\subsubsection{UIP and functional extensionality}
None of these constructions use UIP. Functional extensionality is
 necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a metatheoretic function
 as an argument. An example of induction step that uses it lies in the proof
 that identity substitutions do not change types (this is an equation satisfied
 by a model of the QIIT), in particular in the case  $(\Pim\,T\,
 A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
 equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction hypothesis
 states that $(t:T)\ra (A\,t)[\id]=A\,t$.
 % , which is not applyable without
 % functional extensionality.

\subsection{Functional relation enjoyed by initial morphism}
The previous section builds the syntactic model.
The challenge remains to show that it is initial.

Given a model $M$ of the QIIT, we define the functional relation enjoyed by the initial
morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
stage, $\rec$ is not available yet since the whole point of defining this
relation is to construct $\rec$ in the end. For a type $A$ in a context $\Gamma$,
it is natural to define the relation with a semantic type $\model{A}$ in a
semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality should
be used to transport the right hand side of the second one).
The discussion is similar for terms.

The ultimate goal is to prove that
$\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
$\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
are contractible. The first step consists in showing that these types are propositions.
% We explain below that this requires UIP.
Actually, we take a different approach for the relation on types
(and terms) which yields a more concise definition of the relation.
The adapted goal, for types, will be to prove that
$\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
is inhabited if if $\Gamma \sim \model{\Gamma}$ is.
 This means that in the definition of the relation, we assume
that contexts are already related, so that we don't need to enforce them to be,
as it is the case in the original approach. We give the two versions of the universe
case below, to explain the differences.

% Many of the terms below are well-typed because the target model satisfies some
% equalities that are reflected as definitional equalities.
% In the formalization, we even postulate rewrite rules for some of the equations that the
% model $M$ should satisfy (otherwise, we are faced with too many hellish transports).
% This is, in some sense, another place where UIP is needed, as it is known that
% extensional type theory can be translated in intensional type theory using UIP
% and functional extensionality. This requirement can be qualified: we
% could say that we only claim to eliminate to models that definitionally
% satisfies the equalities that we enforce, although we haven't checked that this is enough to
% construct any IIT from the universal one (and anyway, UIP and functional
% extensionality are needed to construct any IIT from the universal one).

Here we provide the definition of the relation by recursion on the typing judgments.
In the definitions, we abbreviate $\wf{A} \sim (\model{\Gamma}\vdash \model{A})$ by
$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ is clear, and similarly for terms.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \sim{ \blank} && :  \Gamma \vdash \, \ra \model{\Con} \ra \Set \\
  & \blank \sim (\model{\Gamma}\vdash  \blank) && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
  & \blank \sim{ (\model{\Gamma}\vdash  \blank \Rightarrow \model{\Delta})} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\
  \\
  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\cdot} \\
  & \epsilonw\sim
      ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&
 := (e_E : \model{E} = \model{\cdot})\times (\model{\delta} = e_E \#
 \model{\epsilon})
  \\
  & (\wf{\Gamma}\extw \wf{A})\sim \model{\Delta} &&
   :=
    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
    \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
    (\model{\Delta} = \model{\Gamma} \model{\ext} \model{A})
    \\
    & (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
    \\
    & \qquad
    ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&
   :=
    \sum_{\model{\Delta}} (\wf{\Delta}\sim\model{\Delta}) \times
    \sum_{\model{\sigma}} (\wf{\sigma}\sim\model{\sigma}) \times
    \\
    & && \qquad
    \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
    \sum_{\model{t}} (\wf{t}\sim\model{t}) \times
    \\
    & && \qquad
    (e_E : \model{E} = \model{\Delta} \model{\ext} \model{A}) \times
    (\delta = e_E \# \model\sigma \model{\cons} \model{t} )
    \\
  & \varw\,\wf{x} \sim \model{t}
   && := \wf{x}\sim \model{t} \\
  & \vzw\wf{\Gamma}\wf{A}
  \sim \\
  & \qquad (\model{\Delta}\vdash \model{t}\in \model{B} )
   && :=
    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
    \sum_{\model{A}} (\wf{A}\sim\model{A}) \times \\
    & &&
    \qquad
  (e_\Delta : \model{\Delta} = \model{\Gamma} \model{\ext} \model{A})
     \times
     (e_B : \model{B} = e_\Delta \# \model{\wk}\,\model{A}) \times
     \\
     & && \qquad \model{t} = e_\Delta,e_B \# \model{\vz}
  \\
  & \vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B} \sim
  \\
  & \qquad
     (\model{\Delta}\vdash \model{t}\in \model{C} )
   && :=
    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
    \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
    \sum_{\model{B}} (\wf{B}\sim\model{B}) \times
    \sum_{\model{n}} (\wf{n}\sim\model{n}) \times
    \\
    & &&
    \qquad
  (e_\Delta : \model{\Delta} = \model{\Gamma} \model{\ext} \model{B})
     \times
     (e_C : \model{C} = e_\Delta \# \model{\wk}\,\model{A}) \times
     \\
     & && \qquad
     \model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n}
  \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uw \wf{\Gamma} \wf{A} \sim \model{A} &&
   := \model{A} = \model{\U}
  \\
  & \Elw \wf{\Gamma} \wf{a} \sim \model{A}
  && :=
  \sum_{\model{a}} (\wf{a}\sim  \model{a}) \times
    (\model{A} = \model{\El}\,\model{a})
  \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piw \wf{\Gamma} \wf{a} \wf{B} \sim \model{C} &&
   :=
      \sum_{\model{a}} (\wf{a}\sim\model{a})
      \times
      \sum_{\model{B}} (\wf{B}\sim\model{B})
      \\
      &&& \qquad \times (\model{C}= \model{\Pi}\,\model{a}\,\model{B})
  \\
  & \appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
  & \qquad (\model{\Gamma}\vdash x\in \model{C})  &&
    :=
      \sum_{\model{a}} (\wf{a}\sim\model{a})
      \times
      \sum_{\model{B}} (\wf{B}\sim\model{B})
      \times
      \sum_{\model{t}} (\wf{t}\sim\model{t})
      \times
      \sum_{\model{u}} (\wf{u}\sim\model{u})
      \\
      & &&
     \qquad \times (e_C : \model{C} = \model{B}\model{[0 := \model{u}]})
      \times
      (\model{x} = e_C \# \model{t} \model{\oldapp}\model{u})
    \\
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pimw T\,A\,\wf{\Gamma}\wf{A} \sim \model{B} &&
    :=
      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
      \times
      (\model{B} = \model{\Pim}\,T\,\model{A})
    \\
    & \appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
    \\
    & \qquad
    (\model{\Gamma}\vdash x\in \model{B})
     &&
     :=
      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
      \times
      \sum_{\model{t}} (\wf{t}\sim\model{t})
      \times
      \\ & &&
      \qquad
      (e_B : \model{B} = \model{\Pim}\,T\,\model{A})
      \times
      (\model{x} = e_B \# \model{t}\model{\hat{\oldapp}}u)
    \\
  & \rlap{(5) Infinitary parameters} \\[0.5em]
    & \Piiw T\,A\,\wf{\Gamma}\wf{A} \sim \model{B} &&
    :=
      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
      \times
      (\model{B} = \model{\Pii}\,T\,\model{A})
    \\
    & \appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
    \\
    & \qquad
    (\model{\Gamma}\vdash x\in \model{B})
     &&
     :=
      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
      \times
      \sum_{\model{t}} (\wf{t}\sim\model{t})
      \times
      \\ & &&
      \qquad
      (e_B : \model{B} = \model{\Pii}\,T\,\model{A})
      \times
      (\model{x} = e_B \# \model{t}\model{\tilde{\oldapp}}u)
\end{alignat*}

As discussed above, when writing the definition of the type components, we assume that the input semantic context is already
related to the typing judgment of the syntactic context.
The first version
of the relation that we suggested at the beginning of the section
would rather enforce them to be related. In the case of $\U$, this would lead
to the definition
  $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
  instead of
  the current definition
  $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
  Our choice makes the definitions more concise, and similarly
 in the definition of the term components we assume that the input semantic
 context and type are already related to their associated well-formedness judgments.
 However, for some fields, we cannot avoid the first approach.
  % Unfortunately, we need sometimes to require some equalities that are actually
  % redundant with this external assumption.
  %, although we know that these equalities are
  % already satisfied by our assumption.
  A typical example is the case of empty substitution $\epsilon$: we require the equality
  $(e_C : \model{E} = \model{\cdot})$, although
  our claimed external assumption that $\model{E}$ is  related to the
  canonical proof $\epsilonw$ of the typing judgment $\epsilon \vdash$ should imply
  it.
  Another example is the first variable case, where we follow the first verbose
  approach:
  the relation gives a semantic context $\model\Gamma$ that must be related
  to the syntactic $\Gamma$, and a semantic
  type $\model{A}$ related to the syntactic $A$. The input semantic context is
  then required to equal $\model{\Gamma}\model{\ext}\model{A}$. These
  relations are needed to show that the relation is right unique as
  $\blank \model{\ext}\blank$ is not guaranteed to be injective.

  It is because of these redundancies that we use UIP to show that the
  types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
  $\sum_{\model{A}}A\sim\model{A}$ are propositions.
  One may believe that the use of UIP could be avoided by embracing fully the alternative verbose definition that we suggest above. Then, we expect that
  $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.
  However, there seems
  to be no way of defining $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion so that
  UIP can be avoided to prove this statement.

  \todo[inline]{András: Hugunin's paper has an example (which is
     likely generalizable) that a notion of redundancy in
     well-formedness relations implies that UIP is required.}

  After showing these facts, we prove that the relation is stable under weakening and substitution.
  The last step consists of giving a related semantic counterpart to any
  well-typed context, type or term.
  Everything is done by induction on the typing judgments.
  \subsubsection{Right uniqueness}
  We show by recursion that the relation is right unique in the following sense:
  \begin{alignat*}{5}
    &
    \isp{\Sigma{\sim}}
    && : (\wf\Gamma : \Gamma \vdash ) \ra
    \isaprop\, (\sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf A : \Gamma \vdash A) \ra
    \isaprop\, (\sum_{\model{A}} \wf{A} \sim \model{A})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf t : \Gamma \vdash t \in A) \ra
    \isaprop\, (\sum_{\model{t}} \wf{t} \sim \model{t})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf x : \Gamma \vdash x \invar A) \ra
    \isaprop\, (\sum_{\model{x}} \wf{x} \sim \model{x})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf \sigma : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
    \isaprop\, (\sum_{\model{\sigma}} \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  \subsubsection{Stability under weakening}
  Let us argue that it is necessary to show stability of the relation under
  weakening before showing that it is stable under substitution.
  Indeed, in the proof of stability under substitution, the $\Pi$ case
  requires to show that $\Pi \,A[\sigma] \,B[\keep\,\sigma]$ is related to
  $\model{\Pi}\,\model{A}\model{[\sigma]}
  \model{A}\model{[\model{\keep}\,\sigma]}$.
  We would like to apply the induction hypothesis, so we need to show that
  $\keep\,\sigma=\varu\,0\consu\wkO\,\sigma$ is related to
  $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
  related to $\model{\sigma}$.
  As $\keep\,\sigma=\varu 0\consu \wkO\,\sigma$, we are left with showing that
  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO \id$)
  relates to its semantic counterpart.

  To do that, we indeed show that $\wkO$ preserves the relation, for types and terms.
  This requires to generalize a bit and show that $\wk_n$ preserves the relation,
  as $\wk_0(\Pi\,A\,B)=\Pi\,(\wk_0\,A)(\wk_1\,B)$.
  But remember that $\wk_n$ performs a weakening in the middle of a context, so
  we first define the semantic counterpart of this:
%   {-

% Suppose that Γ ^^ Δ ⊢ and Γ ⊢ E
% The following function computes both Γ ▶ E ^^ wk_E Δ and a substitution
% from this context to Γ ^^ Δ.
% I don't see how to avoid constructing these two components simultaneously

% -}
% ΣwkTel⇒ᵐ :
%   ∀ {Γ}{Γw :  Γ ⊢}(Γm : ∃ (Con~ Γw) )
%     (Em : M.Ty (₁ Γm))
%         {Δ }{Δw : Γ ^^ Δ ⊢}(Δm : ∃ (Con~ Δw)) →
        % ∃ λ T → M.Sub T (₁ Δm)

\begin{alignat*}{5}
  & \model{\Sigma\wkO{\Rightarrow}}  && :
   % & && \qquad
  (\wf{\Gamma} : \Gamma \vdash) \ra
  (\wf\Gamma \sim \model{\Gamma}) \ra
  \\
  & && \qquad
  (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  (\wf\Delta \sim \model{\Delta}) \ra
  \\ &&& \qquad
  (\model{A} : \model\Ty\model\Gamma)\ra
  (\model{\Delta'} : \model{\Con}) \times (\model{\Sub}\model{\Delta'}{\model{\Delta}})
\end{alignat*}
Here, $\model{\Delta'}$ should be thought of as the context $\model\Delta$ where
the weakening has happened in the middle of the context, by inserting the type
$\model{A}$ after the prefix $\model{\Gamma}$. Indeed, we expect that
$\model{\Gamma}$ is a prefix of $\model{\Delta}$, as $\model{\Gamma}$ relates to
$\Gamma$ and $\model{\Delta}$ to $\Gamma\merge\Delta$.
% $\wf{\Gamma}\sim\model\Gamma$
% and $\wf\Delta\sim\model\Delta$, and $\Gamma$ is a prefix of $\Gamma\merge\Delta$.
The substitution from the weakened context to the original one must
be computed at the same time otherwise the recursion hypothesis is not strong enough.
Then, we seperate the two components under the same (implicit) hypotheses:
\begin{alignat*}{5}
  & \model{\wkO}\,\model{A}\,\model\Delta  && :
  % (\model{A} : \model\Ty\model\Gamma)\ra
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % \\
  % & && \qquad
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\ &&& \qquad
   \model{\Con}
   \\
  & \model{\wk{\Rightarrow}}\,\model{A}\,\model\Delta  && :
  % (\model{A} : \model\Ty\model\Gamma)\ra
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % \\
  % & && \qquad
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\ &&& \qquad
   \model{\Sub}(\model{\wkO}\model{A}\,\model\Delta)\model\Delta
\end{alignat*}
Now, we are ready to prove by mutual recursion on well-typed judgments that
weakening preserves typing. The following statements are all under the
hypotheses
  $(\wf{\Gamma} : \Gamma \vdash)$,
  $(\wf\Gamma \sim \model{\Gamma})$,
  $(\wf{\Delta} : \Gamma\merge\Delta \vdash)$,
  $(\wf\Delta \sim \model{\Delta})$,
  $(\wf{A} : \Gamma \vdash A)$,
  and
  $(\wf{A} \sim \model{A})$.
  % , except the last one about weakening substitutions,
  % which do not require that $\model\Delta$ is related to any well-formed context.
\begin{alignat*}{5}
  & \wkO{\sim}  &&
   :
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % \\
  %  & && \qquad
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\
  % & && \qquad
  % (\wf{A} : \Gamma \vdash A) \ra
  % (\wf{A} \sim \model{A}) \ra
  % \\
  % & && \qquad
  \wf{\wkO}\,\wf{A}\,\wf{\Delta}\sim\model{\wkO}\model{A}\model{\Delta}
  \\
  % wk
  & \wk{\sim} && :
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % \\
  %  & && \qquad
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\
  % & && \qquad
  % (\wf{A} : \Gamma \vdash A) \ra
  % (\wf{A} \sim \model{A}) \ra
  % \\
  % &&& \qquad
  (\wf{T} : \Gamma\merge\Delta \vdash T) \ra
  (\wf{T} \sim \model{T}) \ra
  % \\ &&& \qquad
  \wf{\wk}\,\wf{A}\,\wf{T}\sim\model{T}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
  \\
    % wk
  & \wk{\sim} && :
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % \\
  %  & && \qquad
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\
  % & && \qquad
  % (\wf{A} : \Gamma \vdash A) \ra
  % (\wf{A} \sim \model{A}) \ra
  % \\
  % &&& \qquad
  (\wf{t} : \Gamma\merge\Delta \vdash t\in T) \ra
  (\wf{t} \sim \model{t}) \ra
  % \\ &&& \qquad
  \wf{\wk}\,\wf{A}\,\wf{t}\sim\model{t}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
  \\
    % wk
  & \wk{\sim} && :
  % (\wf{\Gamma} : \Gamma \vdash) \ra
  % (\wf\Gamma \sim \model{\Gamma}) \ra
  % \\
  %  & && \qquad
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\
  % & && \qquad
  % (\wf{A} : \Gamma \vdash A) \ra
  % (\wf{A} \sim \model{A}) \ra
  % \\
  % &&& \qquad
  (\wf{x} : \Gamma\merge\Delta \vdash t\invar T) \ra
  (\wf{x} \sim \model{x}) \ra
  % \\ &&& \qquad
  \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
  % \\ &&& \qquad
  % \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
\end{alignat*}
Then we deduce, still by recursion, that weakening of substitution preserves the
relation:
\begin{alignat*}{5}
    % wk
  & \wkO{\sim} && :
  (\wf{\Gamma} : \Gamma \vdash) \ra
  (\wf\Gamma \sim \model{\Gamma}) \ra
  % \\
  %  & && \qquad
  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
  % (\wf\Delta \sim \model{\Delta}) \ra
  % \\
  % & && \qquad
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{A} \sim \model{A}) \ra
  \\
  &&& \qquad
  (\wf{\sigma} : \Gamma\vdash \sigma \Rightarrow \Delta) \ra
  (\wf{\sigma} \sim \model{\sigma}) \ra
  % \\
  % &&& \qquad
  (\wf{\wkO}\wf{A}\wf\sigma \sim \model{\sigma}\model{\circ}\model{\wk})
  \end{alignat*}

%   But to give a precise meaning to the statement that $\wk_n$ (which weakens at
%   the middle of context) preserve the
%   relation, we define (inductively) the notion of semantic telescopes.
%   First, we define the notion of being a prefix:
% \begin{alignat*}{5}
%   & \blank\leq\blank  && : \model\Con\ra \model\Con\ra\Set \\
%   & {\leq}\cdot  && : \model{\Gamma} \leq \model{\Gamma} \\
%   & \blank{\leq}{\ext}\blank  && : \model{\Gamma} \leq \model{\Delta} \ra (A :
%   \model{\Ty}\Delta) \ra \model{\Gamma}\leq\model{\Delta}\model{\ext}\model{A}
% \end{alignat*}
% Now, a telescope on a semantic context $\model{\Gamma}$ is a semantic context
% $\model{\Delta}$ such that $\model{\Gamma}\leq \model{\Delta}$.

% \begin{alignat*}{5}
%   & \Tel\,\model{\Gamma}  && := (\model{\Delta} : \model{\Con})\times (\model{\Gamma}\leq \model{\Delta})
% \end{alignat*}
% Similarly to the untyped case, we define the merging of a context and a telescope by recursion on the telescope:
% \begin{alignat*}{5}
%   & \blank\model{\merge}\blank  && := (\model{\Gamma} : \model{\Con}) \ra
%   (\model{\Delta} : \Tel\,\model\Gamma) \ra \model{\Con} \\
%   & \model\Gamma \merge (\model\Gamma,{\leq}\cdot) && :=  \model\Gamma \\
%   & \model\Gamma \merge (\model\Delta\model{\ext}\model{A}, \Delta_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\Gamma \model\merge \model\Delta)\model\ext \model{A}\\
% \end{alignat*}
% We then define a relevant relation between syntactic telescopes and semantic telescopes.
% Untyped syntactic telescopes are modeled by untyped contexts, because they are
% list of types.
% A syntactic telescope $\Delta$ is then well formed in a context $\Gamma$ if
% $\Gamma\merge\Delta$ is well formed.
% The relation is defined by recursion on the untyped syntactic telescope:
% \begin{alignat*}{5}
%   & \blank\sim\blank  && := ( \Gamma \merge \Delta \vdash) \ra (\model\Gamma\leq\model\Delta)
%   \ra \Set \\
%   & (\wf\Gamma : \Gamma \merge \emptyu \vdash)\sim\model\Delta  && :=
%   \model\Delta =  {\leq}\cdot\\
%   & (\wf\Delta \extw \wf{A} : \Gamma \merge (\Delta \extu A) \vdash)\sim\model{E}  && :=
%   \sum_{\model\Delta} (\wf\Delta\sim\model\Delta)
%   \times
%   \sum_{\model A} (\wf{A}\sim\model{A}) \times
%   \\
%   & && \qquad
%   \model{E} = \model\Delta {\leq}{\ext} \model{A}
% \end{alignat*}

\subsubsection{Stability under substitution}
Let us see an example of where stability under substitution is used in
the recursion building a related semantic counterpart to the syntax.
 We want to prove that given any well-typed
substitution from a context $\Gamma$ to a context $\Delta$, any semantic
contexts $\model{\Gamma}$ and $\model{\Delta}$ which are related to $\Gamma$ and
$\Delta$, there exist a semantic substitution which is related to the syntactic one.
In the extension case $\Gamma\vdash \sigma \consu t \Rightarrow \Delta \extu A$, the
induction hypothesis provides $\model{\sigma}$, $\model{\Delta}$, $\model{A}$ related to
their syntactic counterpart. However, the premises of the induction hypothesis
for getting a relevant $\model{t}$
require to show that the type $\model{A}\model{[\model{\sigma}]}$ is
related to the syntactic type $A[\sigma]$.

We first establish preservation of the relation by substitution for variables:
\begin{alignat*}{5}
    % []V~
  & []{\sim} && :
  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
  (\wf\sigma \sim \model{\sigma}) \ra
  (\wf{x} : \Delta \vdash x \invar A) \ra
  (\wf{x} \sim \model{x}) \ra
  \\ & && \qquad
  % \\
  % & && \qquad
   \wf{[]}\wf{x}\wf{\sigma} \sim \model{x}\model{[\model{\sigma}]}
\end{alignat*}
Then we show it for terms and types  by mutual recursion
under the common hypotheses
  $(\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta)$,
  $(\wf\sigma \sim \model{\sigma})$,
  $(\wf{\Gamma} : \Gamma \vdash )$,
  $(\wf{\Gamma} \sim \model{\Gamma})$,
  % \\
  % &&&\qquad
    $(\wf{\Delta} : \Delta \vdash )$,
    and
  $(\wf{\Delta} \sim \model{\Delta})$:
\begin{alignat*}{5}
     & []{\sim} && :
    (\wf{A} : \Delta \vdash A) \ra
  (\wf{A} \sim \model{A}) \ra
  % \\
  % & && \qquad
   \wf{[]}\wf{\Gamma}\wf{A}\wf{\sigma} \sim \model{A}\model{[\model{\sigma}]}
   \\
     & []{\sim} && :
    (\wf{t} : \Delta \vdash t \in A ) \ra
  (\wf{t} \sim \model{t}) \ra
  % \\
  % & && \qquad
   \wf{[]}\wf{\Gamma}\wf{t}\wf{\sigma} \sim \model{t}\model{[\model{\sigma}]}
  \end{alignat*}
  Eventually, we show under the same hypothese the following statement:
  \begin{alignat*}{5}
     & {\circ}{\sim} && :
    (\wf{E} : E \vdash ) \ra
  (\wf{E} \sim \model{E}) \ra
  % \\
  % &&&  \qquad
    (\wf{\delta} : \Delta \vdash \delta \Rightarrow E) \ra
  (\wf{\delta} \sim \model{\delta}) \ra
  \\
  & && \qquad
   \wf{{\circ}}\wf{\Gamma}\wf{\delta}\wf{\sigma} \sim \model{\delta}\model{\circ}\model{\sigma}
  \end{alignat*}
  and the fact that identity preserves the relation:
  \begin{alignat*}{5}
     & {\id}{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
  (\wf{\Gamma} \sim \model{\Gamma}) \ra
   \wf{{\id}}\wf{\Gamma} \sim {\id}_{\model{\Gamma}}
  \end{alignat*}
% $\model{t}\model\oldapp\model{u}$, but it has type
% $\model{B}\model{[0 := \model{u}]}$
% instead of type $\model{T}$. Fortunately, we know by hypothesis that $\model{T}$
% relates to $T=B[0:=u]$ so, by uniqueness of the related semantic counterpart,
% we can deduce that $\model{T}=\m$
% Thus, we need to show that $\model{T}=\model{B}\model{[0 := \model{u}]}$ if we
% know that this last type is related to $B[0 := u]$.

\subsection{Existence and uniqueness of the initial QIIT morphism}
Induction on the typing judgments shows that any QIIT morphism from the
syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
semantic counter-part. More formally, we show by recursion:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  Right uniqueness of the relation is used in this recursion.

It is then straightforward to show that the first projection of these
constructions yield a model morphism from the syntax to the model, using right
uniqueness of the relation.
Now, to show that it is unique, suppose that we have a morphism from the
syntax to the model, inducing the following maps:
\begin{alignat*}{5}
  &
  \mor{\Con}
  && :
   (\Gamma \vdash) \ra \model\Con
   \\
  &
  \mor{\Ty}
  && :
   (\wf\Gamma:\Gamma \vdash) \ra (\Gamma\vdash A)\ra\model\Ty\,(\mor\Con\wf\Gamma)
   \\
  &
  \mor{\Tm}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash t \in A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Var}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
  (\Gamma\vdash x \invar A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
  (\mor\Ty\wf\Gamma\,\wf{A})
   \\
  &
  \mor{\Sub}
  && :
  (\wf\Gamma:\Gamma \vdash) \ra
  (\wf\Delta:\Delta \vdash) \ra
  (\Gamma\vdash \sigma \Rightarrow \Delta)\ra\model\Sub\,(\mor\Con\wf\Gamma)\,
  (\mor\Con\wf\Delta)
\end{alignat*}
Then, we show by recursion on the typing judgments that
the image is related:
\begin{alignat*}{5}
  &
  {\sim}\mor{\Con}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra \wf\Gamma\sim \mor\Con\,\wf\Gamma
  \\
  &
  {\sim}\mor{\Ty}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  \wf\Gamma\sim \mor\Ty\,\wf\Gamma\,\wf{A}
  \\
  &
  {\sim}\mor{\Tm}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{t} : \Gamma \vdash t \in A) \ra
  \wf\Gamma\sim \mor\Tm\,\wf\Gamma\,\wf{A}\,\wf{t}
  \\
  &
  {\sim}\mor{\Var}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{A} : \Gamma \vdash A) \ra
  (\wf{x} : \Gamma \vdash x \invar A) \ra
  \wf\Gamma\sim \mor\Var\,\wf\Gamma\,\wf{A}\,\wf{x}
  \\
  &
  {\sim}\mor{\Sub}
  && :
  (\wf\Gamma : \Gamma \vdash) \ra
  (\wf{\Delta} : \Delta \vdash) \ra
  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
  \wf\Gamma\sim \mor\Sub\,\wf\Gamma\,\wf{\Delta}\,\wf{\sigma}
\end{alignat*}
This justifies the uniqueness of the morphism, by right uniqueness of the relation.

% \begin{description}
%   \item[untyped syntax and well typed judgments as an algebra]
%   \begin{enumerate}
%   \item define untyped syntax as an inductive datatype
%   \item define operations on the syntax, such as substitution, by recursion
%   \item define well-typed judgments as an inductive datatype indexed over the untyped syntax
%   \item show the dependant pair of untyped syntax with well-typed judgments
%     define an algebra
%   \end{enumerate}
%   \item[specification of the initial morphism as a functional relation]
%   \begin{enumerate}
%   \item given an algebra, define the functional relation enjoyed by the recursor
%     from the syntax to the algebra by recursion over the well-typed judgment
%   \item show right-uniqueness of the relation
%   \item show left-totality of the relation
%   \end{enumerate}
%   \item[existence and uniqueness of the morphism from the syntax to a model]
%     \begin{enumerate}
%   \item show uniqueness of such a morphism
%   \item extract an algebra morphism from the relation
%     \end{enumerate}
% \end{description}

### 2019-08-30 paper/universaliit.tex
some typos in the universal IIT part of the paper

@@
 
-\subsubsection{Untyped weakening}
+\subsubsection{Weakening preserves typing}
 
@@
 
-  \subsubsection{Untyped substitution}
+  \subsubsection{Substitution preserves typing}
   We define by induction the specific substitution of one variable of the context:
@@
 \end{alignat*}
-Note that there is possibility of redundancy in the arguments of
+There is possibility of redundancy in the arguments of
 the constructors. Here, we are rather ``{paranoid}'', so that we get more inductive
@@
   \end{alignat*}
-Then, we show that identity substitutions is neutral, for well-typed types,
+Then, we show that identity substitutions are neutral, for well-formed types,
 terms, and substitutions:
@@
 \subsubsection{Uniqueness of typing and uniqueness of types}
-  We define what it means for a type $A$ to be a proposition as follows:
+  We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:
 
   \[
-    \isaprop\,A := (a : A) \ra (a' : A) \ra a = a'
+    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
   \]
-  It states that there is at most one inhabitant.
+  Thus, a type is a proposition if there is at most
+  one inhabitant.
 We show by recursion that each of these typing judgments is unique in the
@@
         \isp{\wf\Sub}
-  : \isaprop\, (\Gamma\vdash \sigma \in \Delta)
+  : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
   \end{align*}
@@
 semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
-\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality should
-be used to transport the right hand side of the second one).
+\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
+required for the second one to be well-typed).
 The discussion is similar for terms.
@@
 $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
-are contractible. The first step consists in showing that these types are propositions.
+are contractible, i.e., have a unique inhabitant. The first step consists in showing that these types are propositions.
 % We explain below that this requires UIP.

### 2019-08-30 paper/paper.tex
lipics style

@@
-\documentclass[a4paper]{article}
+\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
+%This is a template for producing LIPIcs articles. 
+%See lipics-manual.pdf for further information.
+%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+%for section-numbered lemmas etc., use "numberwithinsect"
+%for enabling cleveref support, use "cleveref"
+%for enabling cleveref support, use "autoref"
 \usepackage{amssymb}
 \usepackage{amsmath}
-\usepackage[utf8]{inputenc}
-\usepackage[dvipsnames]{xcolor} % for BrickRed color in abbrevs.tex
 \usepackage{hyperref}
 \usepackage{todonotes}
+
+%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\title{For induction-induction, induction is enough} %TODO Please add
+
+\titlerunning{For induction-induction, induction is enough}%optional, please use if title is longer than one line
+
+\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by Thematic Excellence Programme, Industry and Digitization Subprogramme (NRDI Office, 2019) and by the European Union, co-financed by the European Social Fund (EFOP-3.6.2-16-2017-00013, Thematic Fundamental Research Collaborations Grounding Innovation in Informatics and Infocommunication).}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional
+\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
+\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}
+
+\authorrunning{A. Kaposi, A. Kov{\'a}cs and A. Lafont}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{A. Kaposi, A. Kov{\'a}cs and A. Lafont}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[500]{Theory of computation~Logic~Type theory}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{type theory, inductive types, inductive-inductive types}%TODO mandatory; please add comma-separated list of keywords
+
+\category{}%optional, e.g. invited paper
+
+\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+%\relatedversion{A full version of the paper is available at \url{...}.}
+
+\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+\acknowledgements{The authors would like to thank Thorsten Altenkirch, Simon Boulier, Fredrik Nordvall-Forsberg and Jakob von Raumer for discussions on the topics of this paper.}%optional
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\title{Reducing finitary inductive-inductive types to inductive types}
+\maketitle
 
-\maketitle
+%TODO mandatory: add short abstract of the document
+\begin{abstract}
+  Inductive-inductive types (IITs) are a generalisation of inductive
+  types in type theory. They allow the mutual definition of types with
+  multiple sorts depending on each other. An example is the James
+  Chapman-style syntax of type theory with conversion relations for
+  each sort where e.g.\ the sort of types is indexed by contexts.  In
+  previous work it was shown that all IITs can be constructed from a
+  quotient inductive-inductive type (QIIT), namely the theory of IIT
+  signatures. This theory is a small domain-specific type theory where
+  a context is a signature for an IIT. In this paper we construct this
+  theory of signatures only using inductive types. All of the
+  constructions of this paper were formalised in the proof assistant
+  Agda.
+\end{abstract}
 
@@
 Ambroise has done this in Agda with funext and UIP using an
-extrinsic--intrinsic translation, but only for parts (1)--(5), and the
-following parts are missing: $(\prop{1})$--$(\prop{5})$. We need
-the $(\prop{1})$--$(\prop{5})$ parts to justify the next section.
+extrinsic--intrinsic translation.
 
@@
 
-\bibliographystyle{plain}
+%%
+%% Bibliography
+%%
+
+%% Please use bibtex,
+
 \bibliography{b}
 
+\appendix
+
+\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
+
+List of different predefined enumeration styles:
+
+\begin{itemize}
+\item \verb|\begin{itemize}...\end{itemize}|
+\item \dots
+\item \dots
+%\item \dots
+\end{itemize}
+
+\begin{enumerate}
+\item \verb|\begin{enumerate}...\end{enumerate}|
+\item \dots
+\item \dots
+%\item \dots
+\end{enumerate}
+
+\begin{alphaenumerate}
+\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
+\item \dots
+\item \dots
+%\item \dots
+\end{alphaenumerate}
+
+\begin{romanenumerate}
+\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
+\item \dots
+\item \dots
+%\item \dots
+\end{romanenumerate}
+
+\begin{bracketenumerate}
+\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
+\item \dots
+\item \dots
+%\item \dots
+\end{bracketenumerate}
+
+\begin{description}
+\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
+\item[Description 2] Fusce eu leo nisi.
+Cras eget orci neque, eleifend dapibus felis.
+Duis et leo dui.
+Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
+\item[Description 3]  \dots
+%\item \dots
+\end{description}
+
 \end{document}

### 2019-08-30 paper/paper.tex
related work

@@
 
+This is the paper version of our TYPES 2019 talk.
+
+\subsection{Related work}
+
+Reduction of classes of inductive types: reduce indexed W-types to
+W-types. IR types to indexed types. IITs were introduced by
+Forsberg. What he did.  QIITs by Forsberg-lots of people. Our QIIT
+paper. Jakob TYPES 2019 abstract.  Jasper Huguin: he didn't do
+recursion-recursion. Without UIP.
+
+Similar techique for other type theories: Streicher. Initiality
+project, Brunerie. These all use quotients. Closest: Lafont TYPES
+abstract 2017. Does not use quotients.
+
+Constructing intrinsic syntax but different techniques: Chapman,
+Danielsson, Nuo. These use induction-induction which is our goal.
+
 \subsection{Structure of the paper}
@@
 that this QIIT can be constructed only using inductive types.
+
+\subsection{Formalisation}
+
+UIP, funext, rewrite rules, version of Agda.
 
@@
 
-\section{Constructing all IITs from the theory of signatures}
+\section{Constructing all IITs from the theory IIT of signatures}
 \label{sec:andras}
 
-Show using the CwF model that the existence of the theory of IIT
-signatures (Section \ref{sec:theory_of_signatures}) entails the
-existance of all IITs.
+Explain that the previous paper was modular.
+
+\section{Further work}
+
+Infinitary, explain why it is difficult. Without UIP, without funext,
+for QIITs (in the setoid model).
 

### 2019-09-13 paper/universaliit.tex
Rks Ambrus

@@
   \begin{enumerate}
-  \item defining a relation $\blank \sim \blank$ between the syntax and a given
+  \item defining a relation $\blank \sim \blank$ between the (well typed) syntax and a given
     model (the idea is that given a syntactic context $\Gamma$ and a semantic
@@
     % syntax to a given model;
-  \item Showing that this relation is functional;
+  \item showing that this relation is functional.
     \end{enumerate}
@@
 
-The goal is to define the syntactic model by packing the untyped syntax
-with the typing judgments:
-% % \subsection{QIIT model}
-% Then, we show that they are stable by weakening and substitution.
-% Packing the untyped syntax with well-formed typed judgments yields
-% a model $S$ of the QIIT. Let us give the most important components:
-\begin{alignat*}{5}
- & \syn\Con && := \sum_\Gamma \Gamma\vdash
- \\
- & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
- \\
- & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
- \\
- & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
-\end{alignat*}
-The next sections are devoted to the definition of the untyped syntax and the
+The goal is to define the syntactic model where contexts are pairs of an untyped 
+context together with a proof that it is well formed, and similarly for types,
+terms and substitutions.
+The next sections are devoted to the definition of the untyped syntax and their associated
 typing judgments.
@@
   & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
-  & \rlap{(4) Metatheoretic and infinitary parameters} \\[0.5em]
+  & \rlap{(4-5) Metatheoretic and infinitary parameters} \\[0.5em]
   & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
@@
   & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
-  & \rlap{(5) Default value} \\[0.5em]
+  & \rlap{(6) Default value} \\[0.5em]
   & \erru && : \Tmu
@@
 
-\subsubsection{Weakening preserves typing}
+\subsubsection{Untyped weakening}
 
@@
 Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
-\,A[\sigma]\,B[\wk\,\sigma]$, and thus we need to define $\wk$, the weakening substitution.
+\,A[\sigma]\,B[\varu\, 0 \,\consu\, \wkO\,\sigma]$, and thus we need to define
+$\wk0$, the weakening of substitutions.
 The basic idea is to increment the de Bruijn indices of all the variables.
@@
 
-  \subsubsection{Substitution preserves typing}
+  Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves typing.
+  But as we just argued, weakening may occur anywhere in the context.
+Stating a typing rule for this operation requires to have a way to express
+the weakening occuring at the middle of a context.
+To this end, we consider pairs of untyped contexts, which should be thought
+of as a splitting at some place of a full context.
+This full context is recovered by merging the two components:
+% (which appears in the typing rule of a weakening occuring in the middle of a context),
+% and the length $\length{\Gamma}$ of a
+% context $\Gamma$:
+\begin{alignat*}{5}
+  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
+  & \Gamma \merge \cdot && :=  \Gamma \\
+  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
+\end{alignat*}
+We think of the second context as a telescope over the first context.
+We need to define the define also the weakening of a telescope in order to
+state the typing rule of weakening:
+\begin{alignat*}{5}
+  & \wkO && : \Conu\ra \Conu
+  \\
+  & \wkO\,\emptyu && := \emptyu
+  \\
+  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
+  \end{alignat*}
+  \subsubsection{Untyped substitution}
   We define by induction the specific substitution of one variable of the context:
@@
 
-\subsubsection{Typed weakening}
-Untyped weakening performs weakening anywhere in the context.
-Stating a typing rule for this operation requires to have a way to express
-the weakening occuring at the middle of a context.
-To this end, we consider pairs of untyped contexts, which should be thought
-of as a splitting at some place of a full context.
-This full context is recovered by merging the two components:
-% (which appears in the typing rule of a weakening occuring in the middle of a context),
-% and the length $\length{\Gamma}$ of a
-% context $\Gamma$:
-\begin{alignat*}{5}
-  & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
-  & \Gamma \merge \cdot && :=  \Gamma \\
-  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
-\end{alignat*}
-We think of the second context as a telescope over the first context.
-We need to define the define also the weakening of a telescope in order to
-state the typing rule of weakening:
-\begin{alignat*}{5}
-  & \wkO && : \Conu\ra \Conu
-  \\
-  & \wkO\,\emptyu && := \emptyu
-  \\
-  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
-  \end{alignat*}
-Now we state and prove by mutual recursion that typing judgments are stable under
+\subsubsection{Weakening preserves typing}
+\label{ss:weaken-typing}
+We state and prove by mutual recursion that typing judgments are stable under
 weakening, for contexts, types, terms, and substitutions:
@@
   \end{alignat*}
-\subsubsection{Typed substitution}
+\subsubsection{Substitution preserves typing}
   We show that judgments are stable under substitution, by mutual recursion:
@@
 \subsubsection{Uniqueness of typing and uniqueness of types}
+\label{ss:uniq-typing-types}
   We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:
@@
 
-\subsubsection{UIP and functional extensionality}
-None of these constructions use UIP. Functional extensionality is
+% \subsubsection{UIP and functional extensionality}
+ % , which is not applyable without
+ % functional extensionality.
+
+ \subsubsection{Syntactic model}
+ The syntactic model is obtained 
+ by packing the untyped syntax
+with the typing judgments:
+% % \subsection{QIIT model}
+% Then, we show that they are stable by weakening and substitution.
+% Packing the untyped syntax with well-formed typed judgments yields
+% a model $S$ of the QIIT. Let us give the most important components:
+\begin{alignat*}{5}
+ & \syn\Con && := \sum_\Gamma \Gamma\vdash
+ \\
+ & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
+ \\
+ & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
+ \\
+ & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
+\end{alignat*}
+The other fields are given straightforwardly.
+Regarding the equations, it is enough to prove them only for the untyped syntactic
+part: as we argued in Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are automatically
+equal.
+
+None of the constructions in this section uses UIP. Functional extensionality is
  necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a metatheoretic function
@@
  states that $(t:T)\ra (A\,t)[\id]=A\,t$.
- % , which is not applyable without
- % functional extensionality.
-
-\subsection{Functional relation enjoyed by initial morphism}
+
+\subsection{A functional relation}
 The previous section builds the syntactic model.
@@
 
-Given a model $M$ of the QIIT, we define the functional relation enjoyed by the initial
+Given a model $M$ of the QIIT, we define the functional relation satisfied by the initial
 morphism $\rec : S\to M$ by recursion on the typing judgments.
@@
 stage, $\rec$ is not available yet since the whole point of defining this
-relation is to construct $\rec$ in the end. For a type $A$ in a context $\Gamma$,
-it is natural to define the relation with a semantic type $\model{A}$ in a
-semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
-\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
-required for the second one to be well-typed).
-The discussion is similar for terms.
-
-The ultimate goal is to prove that
-$\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
-$\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
-are contractible, i.e., have a unique inhabitant. The first step consists in showing that these types are propositions.
-% We explain below that this requires UIP.
-Actually, we take a different approach for the relation on types
-(and terms) which yields a more concise definition of the relation.
-The adapted goal, for types, will be to prove that
-$\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
-is inhabited if if $\Gamma \sim \model{\Gamma}$ is.
- This means that in the definition of the relation, we assume
-that contexts are already related, so that we don't need to enforce them to be,
-as it is the case in the original approach. We give the two versions of the universe
-case below, to explain the differences.
+relation is to construct $\rec$ in the end.
+
+For a type $A$ in a context $\Gamma$, we want to define a relation
+$A\sim \model{A}$ that is equivalent to $\rec(A)=\model{A}$.
+% where $\model{A}$ is a
+% semantic type in a semantic context $\model{\Gamma}$.
+For this equality to make sense, the semantic type $\model{A}$ must live
+in the semantic context $\rec\ \Gamma$. But again, $\rec$ is not
+yet available at this stage. Exploiting the expected equivalence between
+$\Gamma\sim\model\Gamma$ and $\rec(\Gamma)=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
+hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
+related to $\Gamma$. Then, the type of the relation for types is
+\[
+  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
+  (\model\Gamma : \model\Con)
+  \ra
+  (\Gamma \sim \model\Gamma)
+  \ra
+  (\model{A}:\model\Ty\model\Gamma)
+  \ra
+  \Set
+\]
+Note that the relation on contexts
+must be defined mutually with the relation on types (see for example the case of
+context extension), but here, the relation on contexts appears as the type of an
+argument of the relation on types.
+% We leave this possibility aside,
+As we do not assume that our metatheory
+supports this kind of recursive-recursive definitions, we refuse to further consider
+ such a possiblity, and instead just remove the 
+hypothesis $\Gamma\sim\model\Gamma$ from the list of arguments.
+We proceed similarly for terms and substitutions. Actually, 
+this removal is not without harm. Consider for example the case of relating the
+empty substitution $\Gamma\vdash \epsilonu \Rightarrow \emptyu $
+with a semantic substitution $\model\sigma$ between semantic contexts
+$\model\Gamma$ and $\model\Delta$. We would like to assert that $\model\sigma$
+equals the empty semantic substitution $\model\epsilon$, but this is not
+possible because typechecking requires that $\model\Delta$ is 
+the empty semantic context. This is precisely what ensured the removed hypothesis
+$\syn\cdot \sim \model\Delta$.
+Our way out here is to state that $\model\sigma$ relates to the empty
+substitution if the target semantic context is empty, and, acknowledging
+this equality, if $\model\sigma$ is
+the empty substitution.
+% The idea is that we will first prove that there is atmost one semantic type
+% which relates to the syntactic one, and then later we will provide such a
+% semantic type under the hypothesis that the contexts are related.
+
+Let us mention another possible solution avoiding the recursive-recursive
+requirement: it consists in defining $A\sim\model{A}$ so that it is equivalent to
+$(e:\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
+% whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
+% For a substitution $\sigma$ between contexts $\Gamma$ and $\Delta$,
+% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \Gamma
+% \sim \model\Gamma \to \Set $ where $\model{A}$ is a semantic type in the
+% semantic context $\model{\Gamma}$. We define the relation on types under the hypothesis
+% that the contexts are already related.
+% it is natural to define the relation with a semantic type $\model{A}$ in a
+% semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
+% required for the second one to be well-typed).
+% The discussion is similar for terms.
+% Then, the ultimate goal is to prove that
+% $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
+% $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
+% are contractible, i.e., have a unique inhabitant.
+%The first step consists in showing that these types are propositions.
+Our different approach
+% Actually, we take a different approach for the relation on types
+% (and terms) which
+yields a more concise definition of the relation.
+For example, in the case of the universe, this would lead
+to the definition
+  $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
+  instead of
+  our definition
+  $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
+% Recall that our adapted goal, for types, is to prove that
+% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
+% is inhabited if $\Gamma \sim \model{\Gamma}$ is.
+%  This means that in the definition of the relation, we assume
+% that contexts are already related, so that we don't need to enforce them to be,
+% as it is the case in the original approach.
+% We give the two versions of the universe
+% case below, to explain the differences.
 
@@
 Here we provide the definition of the relation by recursion on the typing judgments.
-In the definitions, we abbreviate $\wf{A} \sim (\model{\Gamma}\vdash \model{A})$ by
-$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ is clear, and similarly for terms.
+In the definitions, we abbreviate $\relT{\wf{A}}{\model\Gamma}{\model{A}}$ by
+$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ can be inferred, and similarly for
+terms and substitutions.
 \begin{alignat*}{5}
@@
   & \blank \sim{ \blank} && :  \Gamma \vdash \, \ra \model{\Con} \ra \Set \\
-  & \blank \sim (\model{\Gamma}\vdash  \blank) && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
-  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
-  & \blank \sim{ (\model{\Gamma}\vdash  \blank \in \model{A})} && :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
-  & \blank \sim{ (\model{\Gamma}\vdash  \blank \Rightarrow \model{\Delta})} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\
+  & \relT{\blank}{\model{\Gamma}}{\blank} && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
+  & \relt{\blank}{\model{\Gamma}}{\model{A}}{\blank} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
+  & \relV{\blank}{\model{\Gamma}}{\model{A}}{\blank}&& :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
+  & \relS{\blank}{\model{\Gamma}}{\model{\Delta}}{ \blank} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\
   \\
   & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\cdot} \\
-  & \epsilonw\sim
-      ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&
+  & \relS{\epsilonw}{\model{\Gamma}}{\model{E}}{\model{\delta}}
+  % (  \vdash \model{\delta}\Rightarrow \model{E})
+  &&
  := (e_E : \model{E} = \model{\cdot})\times (\model{\delta} = e_E \#
@@
     \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
+    \\
+    & && \qquad
     (\model{\Delta} = \model{\Gamma} \model{\ext} \model{A})
     \\
-    & (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
-    \\
-    & \qquad
-    ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})&&
+    &
+    \relS{(\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})}{\model{\Gamma}}{\model{E}}{\model{\delta}}
+    % (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
+    % \\
+    % & \qquad
+    % ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})
+    &&
    :=
@@
     (e_E : \model{E} = \model{\Delta} \model{\ext} \model{A}) \times
+    \\
+    & && \qquad
     (\delta = e_E \# \model\sigma \model{\cons} \model{t} )
@@
    && := \wf{x}\sim \model{t} \\
-  & \vzw\wf{\Gamma}\wf{A}
-  \sim \\
-  & \qquad (\model{\Delta}\vdash \model{t}\in \model{B} )
+   & \relV{\vzw\wf{\Gamma}\wf{A}}{\model{\Delta}}{\model{B}}{\model{t}}
+  %  \vzw\wf{\Gamma}\wf{A}
+  % \sim \\
+  % & \qquad (\model{\Delta}\vdash \model{t}\in \model{B} )
    && :=
@@
      \times
+    \\
+    & && \qquad
      (e_B : \model{B} = e_\Delta \# \model{\wk}\,\model{A}) \times
-     \\
-     & && \qquad \model{t} = e_\Delta,e_B \# \model{\vz}
-  \\
-  & \vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B} \sim
-  \\
-  & \qquad
-     (\model{\Delta}\vdash \model{t}\in \model{C} )
+     % \\
+     % & && \qquad
+     (\model{t} = e_\Delta,e_B \# \model{\vz})
+  \\
+  &
+   \relV{\vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B}}{\model\Delta}{\model{C}}{\model{t}}
+  % \vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B} \sim
+  % \\
+  % & \qquad
+  %    (\model{\Delta}\vdash \model{t}\in \model{C} )
    && :=
@@
     \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
+    \\
+    & &&\qquad
     \sum_{\model{B}} (\wf{B}\sim\model{B}) \times
@@
      \times
+    \\
+    & && \qquad
      (e_C : \model{C} = e_\Delta \# \model{\wk}\,\model{A}) \times
-     \\
-     & && \qquad
-     \model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n}
+     % \\
+     % & && \qquad
+     (\model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n})
   \\
@@
   \\
-  & \appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
-  & \qquad (\model{\Gamma}\vdash x\in \model{C})  &&
+  &
+  \relt{\appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u}}{\model{\Gamma}}{\model{C}}{\model{x}}  &&
+  % & \appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
+  % & \qquad (\model{\Gamma}\vdash x\in \model{C})  &&
     :=
@@
       \times
+      \\
+      & && \qquad
       \sum_{\model{t}} (\wf{t}\sim\model{t})
@@
       \sum_{\model{u}} (\wf{u}\sim\model{u})
+      \times
       \\
       & &&
-     \qquad \times (e_C : \model{C} = \model{B}\model{[0 := \model{u}]})
-      \times
+     \qquad  (e_C : \model{C} = \model{B}\model{[0 := \model{u}]})
+      \times
+      \\
+      & &&
+      \qquad
       (\model{x} = e_C \# \model{t} \model{\oldapp}\model{u})
@@
     \\
-    & \appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
-    \\
-    & \qquad
-    (\model{\Gamma}\vdash x\in \model{B})
+    & \relt{\appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u}{\model{\Gamma}}{\model{B}}{\model{x}}
+    % & \appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
+    % \\
+    % & \qquad
+    % (\model{\Gamma}\vdash x\in \model{B})
      &&
@@
     \\
-    & \appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
-    \\
-    & \qquad
-    (\model{\Gamma}\vdash x\in \model{B})
+    & \relt{\appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u}{\model{\Gamma}}{\model{B}}{\model{x}}
+    % & \appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
+    % \\
+    % & \qquad
+    % (\model{\Gamma}\vdash x\in \model{B})
      &&
@@
 
-As discussed above, when writing the definition of the type components, we assume that the input semantic context is already
-related to the typing judgment of the syntactic context.
-The first version
-of the relation that we suggested at the beginning of the section
-would rather enforce them to be related. In the case of $\U$, this would lead
-to the definition
-  $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
-  instead of
-  the current definition
-  $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
-  Our choice makes the definitions more concise, and similarly
- in the definition of the term components we assume that the input semantic
- context and type are already related to their associated well-formedness judgments.
- However, for some fields, we cannot avoid the first approach.
+% As discussed above, when writing the definition of the type components, we assume that the input semantic context is already
+% related to the typing judgment of the syntactic context.
+% The first version
+% of the relation that we suggested at the beginning of the section
+% would rather enforce them to be related.
+% In the case of $\U$, this would lead
+% to the definition
+%   $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
+%   instead of
+%   the current definition
+%   $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
+%   Our choice makes the definitions more concise, and similarly
+%  in the definition of the term components we assume that the input semantic
+%  context and type are already related to their associated well-formedness judgments.
+ % However, for some fields, we cannot avoid the first approach.
   % Unfortunately, we need sometimes to require some equalities that are actually
@@
   % already satisfied by our assumption.
-  A typical example is the case of empty substitution $\epsilon$: we require the equality
-  $(e_C : \model{E} = \model{\cdot})$, although
-  our claimed external assumption that $\model{E}$ is  related to the
-  canonical proof $\epsilonw$ of the typing judgment $\epsilon \vdash$ should imply
-  it.
-  Another example is the first variable case, where we follow the first verbose
-  approach:
-  the relation gives a semantic context $\model\Gamma$ that must be related
-  to the syntactic $\Gamma$, and a semantic
-  type $\model{A}$ related to the syntactic $A$. The input semantic context is
-  then required to equal $\model{\Gamma}\model{\ext}\model{A}$. These
-  relations are needed to show that the relation is right unique as
-  $\blank \model{\ext}\blank$ is not guaranteed to be injective.
-
-  It is because of these redundancies that we use UIP to show that the
-  types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
-  $\sum_{\model{A}}A\sim\model{A}$ are propositions.
-  One may believe that the use of UIP could be avoided by embracing fully the alternative verbose definition that we suggest above. Then, we expect that
-  $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
-  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.
-  However, there seems
-  to be no way of defining $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion so that
-  UIP can be avoided to prove this statement.
-
-  \todo[inline]{András: Hugunin's paper has an example (which is
-     likely generalizable) that a notion of redundancy in
-     well-formedness relations implies that UIP is required.}
-
-  After showing these facts, we prove that the relation is stable under weakening and substitution.
+  % As we argued before, in the  of empty substitution $\epsilon$: we require the equality
+  % $(e_C : \model{E} = \model{\cdot})$, although
+  % our claimed external assumption that $\model{E}$ is  related to the
+  % canonical proof $\epsilonw$ of the typing judgment $\epsilon \vdash$ should imply
+  % it.
+% Let us comment the first variable case:
+  % Another example is the first variable case, where we follow the first verbose
+  % approach:
+  % the relation gives a semantic context $\model\Gamma$ that must be related
+  % to the syntactic $\Gamma$, and a semantic
+  % type $\model{A}$ related to the syntactic $A$. The input semantic context is
+  % then required to equal $\model{\Gamma}\model{\ext}\model{A}$.   As
+  % $\blank \model{\ext}\blank$ is not guaranteed to be injective, these
+  % relations are required to show that the relation is right unique.
+Next, we prove that this relation is right unique.
+Then, we show that the relation is stable under weakening and substitution.
   The last step consists of giving a related semantic counterpart to any
@@
   \end{alignat*}
+  We mentionned that in order to avoid a recursive-recursive definition,
+  we removed some hypothesis in the list of arguments of the relation. Such
+  hypotheses are sometimes missed, for example in the case of the empty
+  substitution as we explained, or in the case of variables, leading us to state additional equalities.
+  % some redundancies that we could not avoid when defining
+  % the relation, for example in the case of the empty substitution.
+  Because of this,
+  % It is because of the redundancies which we mentionned above
+  we need UIP to show that the
+  types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
+  $\sum_{\model{A}}A\sim\model{A}$ are propositions.
+  One may believe that the use of UIP could be avoided by embracing fully the
+  alternative verbose definition that we suggested before, expecting that
+  $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
+  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.
+  However, this not so obvious. For example, we did not find
+  a way of defining $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion so that
+  UIP can be avoided to prove this statement.
+
+  \todo[inline]{András: Hugunin's paper has an example (which is
+     likely generalizable) that a notion of redundancy in
+     well-formedness relations implies that UIP is required.}
   \subsubsection{Stability under weakening}
@@
 \end{alignat*}
+Note that if recursion-recursion is available in the metatheory, $\model\wkO$
+and $\wk\model{\Rightarrow}$ can be defined directly without introducing this
+intermediate $\model{\Sigma\wkO\Rightarrow}$.
+
 Now, we are ready to prove by mutual recursion on well-typed judgments that
@@
 contexts $\model{\Gamma}$ and $\model{\Delta}$ which are related to $\Gamma$ and
-$\Delta$, there exist a semantic substitution which is related to the syntactic one.
+$\Delta$, there exists a semantic substitution which is related to the syntactic one.
 In the extension case $\Gamma\vdash \sigma \consu t \Rightarrow \Delta \extu A$, the
@@
 This justifies the uniqueness of the morphism, by right uniqueness of the relation.
+
+\subsection{Recursion}
+\subsection{Uniqueness}
 

### 2019-09-25 paper/universaliit.tex
Subsections on recursion and uniqueness of the initial morphism

@@
 
-  \todo[inline]{András: Hugunin's paper has an example (which is
-     likely generalizable) that a notion of redundancy in
-     well-formedness relations implies that UIP is required.}
+  % \todo[inline]{András: Hugunin's paper has an example (which is
+  %    likely generalizable) that a notion of redundancy in
+  %    well-formedness relations implies that UIP is required.}
   \subsubsection{Stability under weakening}
-  Let us argue that it is necessary to show stability of the relation under
-  weakening before showing that it is stable under substitution.
+  % Let us argue that it is necessary to show s
+  Stability of the relation under
+  weakening must be proved before stability under substitution.
   Indeed, in the proof of stability under substitution, the $\Pi$ case
@@
 
-\subsection{Existence and uniqueness of the initial QIIT morphism}
-Induction on the typing judgments shows that any QIIT morphism from the
-syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
-semantic counter-part. More formally, we show by recursion:
+\subsection{Recursion}
+Recursion consists in building a morphism from the syntactic model to the
+semantic one.
+The image of a syntactic context is going to be a semantic context which relates
+to it (which is unique thanks to right uniqueness of the relation), and
+similarly for types and terms.
+% We construct it as the semantic
+% Induction on the typing judgments shows that any QIIT morphism from the
+% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
+% semantic counter-part.
+Thus, as a first step, we construct by induction on well-typed judgments the
+semantic counter-parts of the syntax:
   \begin{alignat*}{5}
@@
   \end{alignat*}
-  Right uniqueness of the relation is used in this recursion.
-
-It is then straightforward to show that the first projection of these
-constructions yield a model morphism from the syntax to the model, using right
+  Right uniqueness of the relation is used in this induction.
+
+It is then straightforward to show (without induction) that the first projection of these
+constructions yield a model morphism from the syntax to the model, again using right
 uniqueness of the relation.
-Now, to show that it is unique, suppose that we have a morphism from the
+\subsection{Uniqueness}
+Our goal is to show that the syntactic model is initial. What remains thus is to
+show that the morphism constructed in the previous section is actually the
+unique model morphism between the syntax and the arbitray model.
+To this end, we exploit right uniqueness of the relation: it is then enough to show
+that any such morphism maps a syntactic context to a related semantic context,
+and similarly for types and terms.
+
+More formally, we assume an arbitrary morphism from the
 syntax to the model, inducing the following maps:
@@
 This justifies the uniqueness of the morphism, by right uniqueness of the relation.
-
-\subsection{Recursion}
-\subsection{Uniqueness}
 

### 2019-09-25 paper/universaliit.tex
small rephrasings

@@
 Our way out here is to state that $\model\sigma$ relates to the empty
-substitution if the target semantic context is empty, and, acknowledging
+substitution if the target semantic context $\model\Delta$ is empty, and, acknowledging
 this equality, if $\model\sigma$ is
@@
 Thus, as a first step, we construct by induction on well-typed judgments the
-semantic counter-parts of the syntax:
+related semantic counter-parts:
   \begin{alignat*}{5}

### 2019-11-10 paper/paper.tex
abstrac
Rks Ambrus
Streicher, Hugunin reference, edits etc

@@
 \usepackage{todonotes}
+\presetkeys{todonotes}{inline}{}
 
@@
 
-\title{For induction-induction, induction is enough} %TODO Please add
+\title{For Induction-Induction, Induction is Enough} %TODO Please add
 
-\titlerunning{For induction-induction, induction is enough}%optional, please use if title is longer than one line
+%% \titlerunning{For Induction-Induction, Induction is Enough}%optional, please use if title is longer than one line
 
@@
 \begin{abstract}
-  Inductive-inductive types (IITs) are a generalisation of inductive
-  types in type theory. They allow the mutual definition of types with
-  multiple sorts depending on each other. An example is the James
-  Chapman-style syntax of type theory with conversion relations for
-  each sort where e.g.\ the sort of types is indexed by contexts.  In
-  previous work it was shown that all IITs can be constructed from a
-  quotient inductive-inductive type (QIIT), namely the theory of IIT
-  signatures. This theory is a small domain-specific type theory where
-  a context is a signature for an IIT. In this paper we construct this
-  theory of signatures only using inductive types. All of the
-  constructions of this paper were formalised in the proof assistant
-  Agda.
+  Inductive-inductive types (IITs) are a generalisation of inductive types in
+  type theory. They allow the mutual definition of types with multiple sorts
+  where later sorts can be indexed by previous ones. An example is the James
+  Chapman-style syntax of type theory with conversion relations for each sort
+  where e.g.\ the sort of types is indexed by contexts.  In previous work it was
+  shown that all finitary IITs can be constructed from a quotient
+  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
+  a small domain-specific type theory where a context is a signature for an
+  IIT. In this paper we construct this theory of signatures only using inductive
+  types. We rely on function extensionality and uniqueness of identity
+  proofs. All of the constructions of this paper were formalised in the proof
+  assistant Agda.
 \end{abstract}
@@
 
-Note by Ambrus: to be beautified, I only wanted to add the line of
-thought.
+\todo{Note by Ambrus: to be beautified, I only wanted to add the line of
+  thought.}
 
-Most mutual inductive types (e.g.\ isEven-isOdd) can be reduced to
-indexed inductive types where the index gives the information which
-sort we meant. (Indexed inductive types can be reduced to inductive
-types using a method similar to what we use here, preterms and
-well-formedness predicate.) Inductive-inductive types (IITs
-\cite{forsberg-phd}) allow the mutual definition of a type and a
-family of types over the first one. The above reduction does not work
-anymore.
+\todo{
+  András: construction of infinitary IITs doesn't (yet) work. Should we drop
+  infinitary functions from signatures?
+  }
+
+Most mutual inductive types (e.g.\ isEven-isOdd) can be reduced to indexed
+inductive types where the index gives the information which sort we
+meant. (Indexed inductive types can be reduced to inductive types using a method
+similar to what we use here, preterms and well-formedness predicate.)
+Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual definition
+of a type and a family of types over the first one. The above reduction does not
+work anymore.
 
@@
 
-\subsection{Related work}
+\subsection{Related Work}
 
@@
 
-\subsection{Structure of the paper}
+\subsection{Structure of the Paper}
 
-We first define a \emph{theory of IIT signatures} in which every
-context is a valid IIT signature (Section
-\ref{sec:theory_of_signatures}). We present this type theory using an
-algebraic \cite{ttintt} syntax, that is, as a quotient
-inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show
-that this QIIT can be constructed only using inductive types.
+We first define a \emph{theory of IIT signatures} in which every context is a
+valid IIT signature (Section \ref{sec:theory_of_signatures}). We present this
+type theory as an algebraic \cite{ttintt} theory, specifically as a quotient
+inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
+QIIT can be constructed only using inductive types.
 
@@
 
-UIP, funext, rewrite rules, version of Agda.
+\todo{UIP, funext, rewrite rules, version of Agda.}
 
-\section{The theory of IIT signatures}
+\section{The Theory of IIT Signatures}
 \label{sec:theory_of_signatures}
@@
 
-TODO: add examples where all the different constructors are used,
-especially infinitary ones.
+\todo[inline]{add examples where all the different constructors are used}
 
-\section{Constructing the theory of IIT signatures}
+\section{Constructing the Theory of IIT Signatures}
 \label{sec:ambroise}
+
+\todo{Write this section}
 
@@
 
-\paragraph{Construction}
+\paragraph*{Construction}
 \input{universaliit}
 
-\section{Constructing all IITs from the theory IIT of signatures}
+\section{Constructing IITs from the Theory IIT of Signatures}
 \label{sec:andras}
@@
 
-\section{Further work}
+\section{Further Work}
 
@@
 
-%%
-%% Bibliography
-%%
-
-%% Please use bibtex,
-
 \bibliography{b}
 
-\appendix
-
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
-
-List of different predefined enumeration styles:
-
-\begin{itemize}
-\item \verb|\begin{itemize}...\end{itemize}|
-\item \dots
-\item \dots
-%\item \dots
-\end{itemize}
-
-\begin{enumerate}
-\item \verb|\begin{enumerate}...\end{enumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{enumerate}
-
-\begin{alphaenumerate}
-\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{alphaenumerate}
-
-\begin{romanenumerate}
-\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{romanenumerate}
-
-\begin{bracketenumerate}
-\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{bracketenumerate}
-
-\begin{description}
-\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
-\item[Description 2] Fusce eu leo nisi.
-Cras eget orci neque, eleifend dapibus felis.
-Duis et leo dui.
-Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\item[Description 3]  \dots
-%\item \dots
-\end{description}
-
 \end{document}

### 2019-11-10 paper/universaliit.tex
Streicher, Hugunin reference, edits etc

--- 2019-09-25 paper/universaliit.tex
@@
 
-The steps consist of the following:
+The construction consists of the following steps:
 \begin{enumerate}
@@
   \begin{enumerate}
-  \item defining a relation $\blank \sim \blank$ between the (well typed) syntax and a given
-    model (the idea is that given a syntactic context $\Gamma$ and a semantic
+  \item defining a relation $\blank \sim \blank$ between the (well-typed) syntax and a given
+    model. The idea is that given a syntactic context $\Gamma$ and a semantic
     context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
-    \model{\Gamma}$ if and only if $\rec\,\Gamma = \model{\Gamma}$, and
-    similarly for types, terms, and substitutions);
+    \model{\Gamma}$ if and only if $\rec\,\,\Gamma = \model{\Gamma}$, and
+    similarly for types, terms, and substitutions;
     % semantic context $\model{\Gamma}$ of
@@
     \end{enumerate}
-  \item Proving of the uniqueness of this morphism by showing that
-    any morphism $f:S\to M$ satisfies the relation
-    (in the sense, for contexts, that given any syntactic context $\Gamma$, we
-    have $\Gamma\sim f\,\Gamma$).
+  \item Proving the uniqueness of this morphism by showing that any morphism
+    $f:S\to M$ satisfies the relation. For example, for any syntactic context
+    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
       % TODO donner composantes sur les contextes
@@
 The next sections detail each of these steps.
-\subsection{Syntactic model}
-
-The goal is to define the syntactic model where contexts are pairs of an untyped 
-context together with a proof that it is well formed, and similarly for types,
-terms and substitutions.
-The next sections are devoted to the definition of the untyped syntax and their associated
-typing judgments.
+\subsection{Syntactic Model}
+
+The goal is to define the syntactic model where contexts are pairs of a
+precontext together with a well-formedness proof, and similarly for types, terms
+and substitutions.
+
+Crucially, we do not have conversion relations for typed syntax, nor do we need
+to use quotients when giving the syntactic model. This is possible because there
+are no $\beta$-rules in the theory of signatures. Hence, we consider only normal
+terms in the untyped syntax, and define weakening and substitution by recursion.
+Avoiding quotients is important for two reasons. First, it greatly simplifies
+formalisation. Second, we aim to reduce IITs to a minimal feature set, and we
+get a stronger result if we do not use quotients.
+
+The next sections present the definition of the untyped syntax and the
+associated typing judgments.
 
@@
 The untyped syntax is defined as the following inductive datatype:
-
+\todo{András: two columns below perhaps}
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Note that we choose to share the constructor for untyped application to infinitary and
-meta-theoretic parameters.
-Variables are modelled as De Bruijn indices, i.e. as natural numbers pointing
-to the relevant position in the context.
+Note that we overload the constructor for untyped application for infinitary and
+meta-theoretic parameters. Variables are modeled as de Bruijn indices, i.e.\ as
+natural numbers pointing to a position in the context.
 
@@
 \,A[\sigma]\,B[\varu\, 0 \,\consu\, \wkO\,\sigma]$, and thus we need to define
-$\wk0$, the weakening of substitutions.
-The basic idea is to increment the de Bruijn indices of all the variables.
-Actually this is not so simple because of the $\Piu$ type:
-indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type of the weakening
-of $A$ and $B$, but here, $B$ must be weakened with respect to the second last
-variable of the context, rather than the last one.
-For this reason, we need to generalize the weakening as occuring anywhere in the context.
+$\wk0$, the weakening of substitutions.  The basic idea is to increment the de
+Bruijn indices of all the variables.  Actually this is not so simple because of
+the $\Piu$ type: indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type
+of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
+the second last variable of the context, rather than the last one.  For this
+reason, we need to generalize the weakening as occuring anywhere in the context.
 \begin{alignat*}{5}
@@
 
-  Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves typing.
-  But as we just argued, weakening may occur anywhere in the context.
-Stating a typing rule for this operation requires to have a way to express
-the weakening occuring at the middle of a context.
-To this end, we consider pairs of untyped contexts, which should be thought
-of as a splitting at some place of a full context.
-This full context is recovered by merging the two components:
+Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
+typing. Stating a typing rule for this operation requires to have a way to
+express the weakening occuring at the middle of a context. We consider pairs of
+untyped contexts, which should be thought of as a splitting of a context at some
+position. The full context is recovered by merging the two components:
 % (which appears in the typing rule of a weakening occuring in the middle of a context),
@@
 \end{alignat*}
-We think of the second context as a telescope over the first context.
-We need to define the define also the weakening of a telescope in order to
-state the typing rule of weakening:
+We think of the second context as a telescope over the first one. We also define
+weakening for telescopes, which will be used to give typing rules for telescopes
+in Section \ref{sec:typing_judgments}:
 \begin{alignat*}{5}
@@
   \\
-  & \wkO\,(\Delta \extu A) && := \wkO\,\Delta \extu \wk_{\length{\Delta}}\,A
+  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
   \end{alignat*}
   \subsubsection{Untyped substitution}
-  We define by induction the specific substitution of one variable of the context:
+  We define single substitution by recursion on presyntax:
   % & \rlap{(1) Substitution calculus} \\[0.5em]
@@
   \end{alignat*}
-  This is enough to define the well-typed judgments: indeed, the typing rule of application
+  This is enough to define the typing judgments: indeed, the typing rule for application
   involves only a unary substitution.
@@
   However, to construct the initial model of the QIIT, we need to define
-  full substitution calculus:
-  % First, we define
-
-  % We then define by mutual recursion the full substitution of types and terms.
+  the full substitution calculus:
 \begin{alignat*}{5}
@@
   \end{alignat*}
-  These can be defined either by iterating unary substitutions, or
-  by recursion on the syntax on the untyped syntax: the two ways yield provably
-  equal definitions. In the following, we assume that it is defined by recursion.
- We use the additional default constructor
-$\erru:\Tmu$ in case of error (for example, when the substitution is not
-long enough).
-This is an irrelevant constructor in the end, as the typing
-judgments will not mention it.
-
-We also make use of the following definition:
+  These can be defined either by iterating unary substitutions, or by recursion
+  on untyped syntax: the two ways yield provably equal definitions. In the
+  following, we assume that it is defined by recursion.
+ We use the additional default constructor $\erru:\Tmu$ in case of error (for
+example, when the substitution is not long enough).  This is an irrelevant
+constructor in the end, as the typing judgments will not mention it. We also
+make use of the following definition:
   \begin{alignat*}{5}
@@
   \subsubsection{Exchange laws: weakening and substitution}
-  Many lemmas are shown by recursion on the untyped syntax, for terms and types
-  (below, $Z$ denotes either a term or a type):
+  Many lemmas for types and terms are shown by induction on the untyped syntax.
+  Below, $Z$ denotes either a term or a type.
 \begin{alignat*}{5}
@@
   \end{alignat*}
-  The following lemmas are shown by recursion on the syntax, for types and terms:
+  Below, $\keep^n$ denotes $\keep$ iterated $n$ times.
 \begin{alignat*}{5}
@@
    & [:=][\keep] && : Z [ n:= u][\keep^n \sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
-   \\
-   & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau)
-   \\
-  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau
-  \end{alignat*}
-  where $\keep^n$ is $\keep$ iterated $n$ times. As a particular case
-  for $n=0$, we get
-\begin{alignat*}{5}
+  \end{alignat*}
+As particular cases for $n=0$, we get
+\begin{alignat*}{5}
+  & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
+  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau \\
   & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
@@
 \end{alignat*}
-
-Finally, we show the usual lemmas about substitution:
+Finally, we show the following:
 \begin{alignat*}{5}
@@
   \end{alignat*}
-  Other substitution lemmas are deferred after the definition of the typing
-  judgments, as the proofs require that some input are well-typed.
+We defer laws for identity substitutions to after the definition of the typing
+judgments, as the proofs require that some inputs are well-typed.
 
@@
   \subsubsection{Typing judgments}
+  \label{sec:typing_judgments}
   The typing judgments are defined as the following inductive datatype indexed over the
@@
     (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
-    \Elu\,a \vdash B)\ra
+    \Elu\,a \vdash B)
     \\
     & &&
-    (\Gamma \vdash t \in \Piu \, a \, B )
+    \ra(\Gamma \vdash t \in \Piu \, a \, B )
     \ra
@@
     (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
-    ((t : T)\ra\Gamma\vdash A\,t) \ra
+    ((t : T)\ra\Gamma\vdash A\,t)
     \\
     & &&
-    (\Gamma \vdash t \in \Pimu \,T\,A)
+    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
     \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
@@
 \end{alignat*}
-There is possibility of redundancy in the arguments of
-the constructors. Here, we are rather ``{paranoid}'', so that we get more inductive
-hypotheses when performing recursion.
+There is possibility of redundancy in the arguments of the constructors. Here,
+we are ``{paranoid}'', so that we get more inductive hypotheses when performing
+recursion.
 
@@
 \label{ss:weaken-typing}
-We state and prove by mutual recursion that typing judgments are stable under
-weakening, for contexts, types, terms, and substitutions:
+We prove by mutual induction that typing judgments are stable under weakening,
+for contexts, types, terms, and substitutions:
 % wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
@@
 \subsubsection{Substitution preserves typing}
-  We show that judgments are stable under substitution, by mutual recursion:
+  First, we show that judgments are stable under substitution.
 \begin{alignat*}{5}
@@
   \end{alignat*}
-Then, we show that identity substitutions are neutral, for well-formed types,
-terms, and substitutions:
+\subsubsection{Laws for identity substitutions}
+We show category and functor laws involving identity substitution, for
+well-formed types, terms and substitutions.
 \begin{alignat*}{5}
@@
   & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
-  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma
-  \end{alignat*}
-  In a seperate and independant recursion, one shows that:
-  \begin{alignat*}{5}
+  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma \\
   & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
   \end{alignat*}
-  Then, we show that the identity substitution itself is well typed:
+Finally, we show that the identity substitution itself is well typed:
   \begin{alignat*}{5}
@@
 
-\subsubsection{Uniqueness of typing and uniqueness of types}
+\subsubsection{Proof irrelevance and unicity of typing}
 \label{ss:uniq-typing-types}
-  We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:
-
+A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
   \[
@@
   \]
-  Thus, a type is a proposition if there is at most
-  one inhabitant.
-We show by recursion that each of these typing judgments is unique in the
+We show that each of the typing judgments is unique in the
 following sense:
+
+\todo{Prettify table.}
 \begin{align*}
@@
   \end{align*}
-Actually, the application and successor cases require to show uniqueness of
-types in a given context, before.
-Let us consider for instance the application constructor $\appw$: it provides a
-type $B$ such that the final type is $C=B[0 := u]$. Even if $C$ is known a
-priori, there may be another $B$ for which $B[0 := u] = C$, possibly leading to
-many proofs that $t\appu u$ has type $C$. Uniqueness of types
-solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$ of $t$.
-More formally, we show by recursion:
+We also show unicity of typing:
 \begin{alignat*}{5}
@@
   \end{alignat*}
+Let us consider for instance the application constructor $\appw$: for a codomain
+type $B$ it yields an overall type $C=B[0 := u]$ for an application. Even
+if $C$ is known a priori, there may be another $B$ for which $B[0 := u] = C$,
+possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
+solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$
+of $t$.
+
+%% Actually, the application and successor cases require to show uniqueness of
+%% types in a given context, before.
+%% Let us consider for instance the application constructor $\appw$: it provides a
+%% type $B$ such that the final type is $C=B[0 := u]$. Even if $C$ is known a
+%% priori, there may be another $B$ for which $B[0 := u] = C$, possibly leading to
+%% many proofs that $t\appu u$ has type $C$. Uniqueness of types
+%% solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$ of $t$.
 
@@
  \subsubsection{Syntactic model}
- The syntactic model is obtained 
- by packing the untyped syntax
-with the typing judgments:
+ The syntactic model is obtained by packing the untyped syntax with the typing
+ judgments:
 % % \subsection{QIIT model}
@@
 \end{alignat*}
-The other fields are given straightforwardly.
-Regarding the equations, it is enough to prove them only for the untyped syntactic
-part: as we argued in Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are automatically
-equal.
-
-None of the constructions in this section uses UIP. Functional extensionality is
- necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a metatheoretic function
- as an argument. An example of induction step that uses it lies in the proof
- that identity substitutions do not change types (this is an equation satisfied
- by a model of the QIIT), in particular in the case  $(\Pim\,T\,
- A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
- equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction hypothesis
- states that $(t:T)\ra (A\,t)[\id]=A\,t$.
-
-\subsection{A functional relation}
-The previous section builds the syntactic model.
-The challenge remains to show that it is initial.
-
-Given a model $M$ of the QIIT, we define the functional relation satisfied by the initial
-morphism $\rec : S\to M$ by recursion on the typing judgments.
+The other fields are given straightforwardly.  Regarding the equations, it is
+enough to prove them only for the untyped syntactic part: as we argued in
+Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are
+automatically equal.
+
+None of the constructions in this section use UIP. Functional extensionality is
+necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a
+metatheoretic function as an argument. An example induction step that uses it
+is in the type preservation proof for identity substitutions (this is an
+equation satisfied by a model of the QIIT), in particular in the case
+$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
+equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
+hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.
+
+\subsection{Relating the Syntax to a Model}
+It remains to show that the constructed syntactic model is initial. To achieve
+this, we first define a relation between the syntactic model and an arbitrary
+model, then show that the relation is functional, which lets us extract a
+homomorphism from the relation.
+
+This approach is an alternative version of Streicher's method for interpreting
+preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
+a family of partial maps from the presyntax to a model, then shows that the maps
+are total on well-formed input. We have found that our approach is significantly
+easier to formalise. Too see why, note that the right notion of partial map in
+type theory, which does not presume decidable definedness, is fairly
+heavyweight:
+\[
+  \mathsf{PartialMap}\,A\,B := A\ra (P : \mathsf{hProp})\times(P \ra B)
+\]
+Here $\mathsf{hProp}$ denotes a type which is propositional. In the above
+definition, we notice an opportunity for converting a fibered definition of a
+type family into an indexed one; if we drop the propositionality for $P$ for the
+time being, we may equivalently return a family indexed over $B$, which is
+exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we recover
+uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
+relation, and totality by already assuming well-formedness in the syntax. In type
+theory, using indexed families instead of display maps is a common convenience,
+since the former are natively supported in type theory, while the latter require
+carrying around auxiliary propositional equalities.
+
+\subsubsection{The functional relation}
+
+Given a model $M$ of the QIIT, we define the functional relation satisfied by
+the initial morphism $\rec : S\to M$ by recursion on the typing judgments.
 % As these judgments are propositions, we will sometimes
-If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
-context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
-equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
-stage, $\rec$ is not available yet since the whole point of defining this
-relation is to construct $\rec$ in the end.
+If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic context
+(i.e.\ a context of the model $M$), we want to define a type
+$\Gamma\sim \model{\Gamma}$ equivalent to $\rec\,\,\Gamma=\model{\Gamma}$. Of
+course, at this stage, $\rec$ is not available yet since the point of
+defining this relation is to construct $\rec$ in the end.
 
 For a type $A$ in a context $\Gamma$, we want to define a relation
-$A\sim \model{A}$ that is equivalent to $\rec(A)=\model{A}$.
+$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
 % where $\model{A}$ is a
@@
 For this equality to make sense, the semantic type $\model{A}$ must live
-in the semantic context $\rec\ \Gamma$. But again, $\rec$ is not
+in the semantic context $\rec\,\,\Gamma$. But again, $\rec$ is not
 yet available at this stage. Exploiting the expected equivalence between
-$\Gamma\sim\model\Gamma$ and $\rec(\Gamma)=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
+$\Gamma\sim\model\Gamma$ and $\rec\,\,\Gamma=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
 hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
@@
 % We leave this possibility aside,
-As we do not assume that our metatheory
-supports this kind of recursive-recursive definitions, we refuse to further consider
- such a possiblity, and instead just remove the 
-hypothesis $\Gamma\sim\model\Gamma$ from the list of arguments.
-We proceed similarly for terms and substitutions. Actually, 
-this removal is not without harm. Consider for example the case of relating the
-empty substitution $\Gamma\vdash \epsilonu \Rightarrow \emptyu $
-with a semantic substitution $\model\sigma$ between semantic contexts
-$\model\Gamma$ and $\model\Delta$. We would like to assert that $\model\sigma$
-equals the empty semantic substitution $\model\epsilon$, but this is not
-possible because typechecking requires that $\model\Delta$ is 
-the empty semantic context. This is precisely what ensured the removed hypothesis
-$\syn\cdot \sim \model\Delta$.
-Our way out here is to state that $\model\sigma$ relates to the empty
-substitution if the target semantic context is empty, and, acknowledging
-this equality, if $\model\sigma$ is
-the empty substitution.
+Our metatheory does not support this kind of recursive-recursive
+definitions\footnote{And to our knowledge there is no established semantics for
+it in existing literature.}, so we instead just remove the hypothesis
+$\Gamma\sim\model\Gamma$ from the list of arguments.  We proceed similarly for
+terms and substitutions. Actually, this removal is not without harm. For
+example, consider relating the empty substitution
+$\Gamma\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
+$\model\sigma : \model\Sub\,\model\Gamma\,\model\Delta$. We would like to assert
+that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
+this is not possible because typechecking requires that $\model\Delta$ is the
+empty semantic context. This is precisely what ensured the removed hypothesis
+$\syn\cdot \sim \model\Delta$.  Our way out here is to state that $\model\sigma$
+is related to the empty substitution if the target semantic context $\model\Delta$
+is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
+substitution.
 % The idea is that we will first prove that there is atmost one semantic type
@@
 
-Let us mention another possible solution avoiding the recursive-recursive
-requirement: it consists in defining $A\sim\model{A}$ so that it is equivalent to
-$(e:\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
+Let us mention another possible solution for avoiding recursion-recursion:
+defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\Gamma
+= \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
 % whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
@@
 %The first step consists in showing that these types are propositions.
-Our different approach
+In contrast to this, our approach
 % Actually, we take a different approach for the relation on types
@@
   \end{alignat*}
-  We mentionned that in order to avoid a recursive-recursive definition,
-  we removed some hypothesis in the list of arguments of the relation. Such
+  We mentionned that in order to avoid a recursive-recursive definition, we
+  removed some hypotheses in the list of arguments of the relation. Such
   hypotheses are sometimes missed, for example in the case of the empty
-  substitution as we explained, or in the case of variables, leading us to state additional equalities.
-  % some redundancies that we could not avoid when defining
-  % the relation, for example in the case of the empty substitution.
-  Because of this,
-  % It is because of the redundancies which we mentionned above
-  we need UIP to show that the
-  types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
-  $\sum_{\model{A}}A\sim\model{A}$ are propositions.
-  One may believe that the use of UIP could be avoided by embracing fully the
-  alternative verbose definition that we suggested before, expecting that
+  substitution, or in the case of variables, requiring us to state additional
+  equalities.  % some redundancies that we could not avoid when defining % the
+  relation, for example in the case of the empty substitution.  Because of
+  this, % It is because of the redundancies which we mentionned above we need
+  UIP to show that the types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
+  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may believe that the
+  use of UIP could be avoided by embracing the alternative verbose definition
+  that we suggested before, expecting that
   $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
-  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.
-  However, this not so obvious. For example, we did not find
-  a way of defining $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion so that
-  UIP can be avoided to prove this statement.
+  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this not so
+  obvious. For example, we were not able to define
+  $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion. In related work,
+  Hugunin investigated constructing inductive-inductive types without
+  UIP\cite{hugunin2019constructing}, and demonstrated that well-formedness
+  predicates used in syntactic models can be subtly incompatible with UIP. Also,
+  while Hugunin does not use UIP, he only shows a weak version of dependent
+  elimination for the constructed IITs. Hence, the question whether IITs are
+  reducible to inductive types in a UIP-free setting remains open.
+
+%% in this fashion so that
+%%   UIP can be avoided to prove this statement.
 
@@
 
-  To do that, we indeed show that $\wkO$ preserves the relation, for types and terms.
-  This requires to generalize a bit and show that $\wk_n$ preserves the relation,
-  as $\wk_0(\Pi\,A\,B)=\Pi\,(\wk_0\,A)(\wk_1\,B)$.
-  But remember that $\wk_n$ performs a weakening in the middle of a context, so
-  we first define the semantic counterpart of this:
 %   {-
@@
 
+  To do that, we show that $\wkO$ preserves the relation, for types and terms.
+  This requires to generalise a bit and show that $\wk_n$ preserves the relation,
+  as $\wk_0(\Pi\,A\,B)=\Pi\,(\wk_0\,A)(\wk_1\,B)$.
+  But remember that $\wk_n$ performs a weakening in the middle of a context, so
+  we first define the semantic counterpart of this:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Then we deduce, still by recursion, that weakening of substitution preserves the
+Then we deduce, still by induction, that weakening of substitution preserves the
 relation:
@@
 \subsubsection{Stability under substitution}
-Let us see an example of where stability under substitution is used in
-the recursion building a related semantic counterpart to the syntax.
- We want to prove that given any well-typed
-substitution from a context $\Gamma$ to a context $\Delta$, any semantic
-contexts $\model{\Gamma}$ and $\model{\Delta}$ which are related to $\Gamma$ and
-$\Delta$, there exists a semantic substitution which is related to the syntactic one.
-In the extension case $\Gamma\vdash \sigma \consu t \Rightarrow \Delta \extu A$, the
-induction hypothesis provides $\model{\sigma}$, $\model{\Delta}$, $\model{A}$ related to
+
+ We want to prove that given any well-typed substitution $\sigma
+: \Sub\,\Gamma\,\Delta$ and semantic contexts $\model{\Gamma}$ and
+$\model{\Delta}$, respectively related to $\Gamma$ and $\Delta$, there exists a
+semantic substitution which is related to $\sigma$.  In the extension case
+$\Gamma\vdash \sigma \consu\,t\Rightarrow \Delta\,\extu\,A$, the induction
+hypothesis provides $\model{\sigma}$, $\model{\Delta}$, $\model{A}$ related to
 their syntactic counterpart. However, the premises of the induction hypothesis
-for getting a relevant $\model{t}$
-require to show that the type $\model{A}\model{[\model{\sigma}]}$ is
-related to the syntactic type $A[\sigma]$.
-
-We first establish preservation of the relation by substitution for variables:
+for getting a relevant $\model{t}$ require showing that the type
+$\model{A}\model{[\model{\sigma}]}$ is related to the syntactic type
+$A[\sigma]$. We first establish preservation of the relation by substitution for
+variables:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Then we show it for terms and types  by mutual recursion
+Then we show it for terms and types  by mutual induction
 under the common hypotheses
@@
   \end{alignat*}
-  Eventually, we show under the same hypothese the following statement:
+  Eventually, we show under the same hypotheses the following statement
   \begin{alignat*}{5}
@@
 
-\subsection{Recursion}
-Recursion consists in building a morphism from the syntactic model to the
-semantic one.
-The image of a syntactic context is going to be a semantic context which relates
-to it (which is unique thanks to right uniqueness of the relation), and
-similarly for types and terms.
+\subsection{The Recursor}
+For the recursor, we build a morphism from the syntactic model to the semantic
+one. The image of a syntactic context is a unique semantic context which is
+related to it, and similarly for types and terms.
 % We construct it as the semantic
@@
 % semantic counter-part.
-Thus, as a first step, we construct by induction on well-typed judgments the
-semantic counter-parts of the syntax:
+Thus, as a first step, we use induction on well-formedness judgments to construct
+semantic counterparts:
   \begin{alignat*}{5}
@@
   \end{alignat*}
-  Right uniqueness of the relation is used in this induction.
-
-It is then straightforward to show (without induction) that the first projection of these
-constructions yield a model morphism from the syntax to the model, again using right
-uniqueness of the relation.
+  The right uniqueness of the relation is used in this induction. It is then
+straightforward to show (without induction) that the first projection of these
+constructions yield a model morphism from the syntax to the model, again using
+right uniqueness.
 \subsection{Uniqueness}
-Our goal is to show that the syntactic model is initial. What remains thus is to
-show that the morphism constructed in the previous section is actually the
-unique model morphism between the syntax and the arbitray model.
-To this end, we exploit right uniqueness of the relation: it is then enough to show
-that any such morphism maps a syntactic context to a related semantic context,
-and similarly for types and terms.
+Our goal is to show that the syntactic model is initial. Thus, what remains is
+to show that the morphism constructed in the previous section is unique. We
+exploit right uniqueness of the relation: it is enough to show that any such
+morphism maps a syntactic context to a related semantic context, and similarly
+for types and terms.
 
@@
 \end{alignat*}
-Then, we show by recursion on the typing judgments that
+Then, we show by induction on typing judgments that
 the image is related:
@@
 \end{alignat*}
-This justifies the uniqueness of the morphism, by right uniqueness of the relation.
+This justifies the uniqueness of the morphism, by right uniqueness of
+$\blank\sim\blank$.
 

### 2019-11-10 paper/paper.tex
section 4

@@
 \begin{document}
-
 \maketitle
 
-%TODO mandatory: add short abstract of the document
 \begin{abstract}
@@
   a small domain-specific type theory where a context is a signature for an
-  IIT. In this paper we construct this theory of signatures only using inductive
-  types. We rely on function extensionality and uniqueness of identity
-  proofs. All of the constructions of this paper were formalised in the proof
-  assistant Agda.
+  IIT. In this paper we construct this theory of signatures using only inductive
+  types, thereby showing a reduction of all finitary IITs to inductive types.
+  We rely on function extensionality and uniqueness of identity proofs. All
+  constructions in this paper were formalised in the proof assistant Agda.
 \end{abstract}
@@
 
-\section{Constructing IITs from the Theory IIT of Signatures}
+\section{Constructing IITs from the Theory of IIT Signatures}
 \label{sec:andras}
 
-Explain that the previous paper was modular.
+We construct all IITs describable by the theory of IIT signatures through the
+term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
+shown that from the syntax of the theory of signatures for finitary QIITs, one
+can construct any particular finitary QIIT. The idea is that for a signature
+$\Gamma$, the initial algebra can be built from sets of terms of the form
+$\Tm\,\Gamma\,(\El\,a)$.
+
+In this section, we only need to check that the signatures in
+\cite{Kaposi:2019:CQI:3302515.3290315} and the term model construction restricts
+to IITs, i.e.\ that from a syntax for the theory of IIT signatures all finitary
+IITs are constructible.
+
+\begin{itemize}
+\item
+  Finitary IIT signatures are obtained exactly by dropping equality constructors from
+  finitary QIIT signatures. It follows that any model for the theory of QIIT signatures
+  restricts to IITs, hence we inherit the categorical semantics of QIITs.
+\item
+  The term model construction also restrict in a straightforward way. For QIITs,
+  term algebras are constructed by induction on the syntax of signatures, and
+  then another induction constructs the eliminator (i.e.\ dependent induction
+  principle). In both cases, simply dropping equality constructors from the
+  induction yields restriction to IITs.
+\end{itemize}
+
+However, we shall mention that the Agda formalisation for the current paper
+cannot be directly plugged into the Agda code for
+\cite{Kaposi:2019:CQI:3302515.3290315} (which includes most of the term model
+construction). One reason is that the QIIT formalisation uses strict computation
+rules for induction over signatures, while here we show propositional
+computation. This mismatch can be in principle remedied by noting that both
+formalisations use UIP and function extensionality, hence by the known
+translations between extensional and intensional type
+theories\cite{hofmann95extensional,winterhalter2019eliminating}, we can switch
+between strict and propositional equalities.
+
+Also, the current paper proves initiality, i.e.\ unique recursion for
+signatures, while \cite{Kaposi:2019:CQI:3302515.3290315} uses dependent
+elimination. We expect that the two notions are equivalent. An extension of
+\cite{Kaposi:2019:CQI:3302515.3290315} to large infinitary QIITs would entail
+this equivalence, since that would cover the (large, infinitary) theory of IIT
+signatures.
 

### 2019-11-11 paper/paper.tex
stuff

@@
   a small domain-specific type theory where a context is a signature for an
-  IIT. In this paper we construct this theory of signatures using only inductive
-  types, thereby showing a reduction of all finitary IITs to inductive types.
-  We rely on function extensionality and uniqueness of identity proofs. All
-  constructions in this paper were formalised in the proof assistant Agda.
+  IIT. In this paper we construct this theory using only inductive types,
+  thereby showing a reduction of all finitary IITs to inductive types.  We rely
+  on function extensionality and uniqueness of identity proofs. We formalised
+  the construction of the theory of IIT signatures in Agda.
+
 \end{abstract}
@@
 \item
-  The term model construction also restrict in a straightforward way. For QIITs,
+  The term model construction also restricts in a straightforward way. For QIITs,
   term algebras are constructed by induction on the syntax of signatures, and
@@
 computation. This mismatch can be in principle remedied by noting that both
-formalisations use UIP and function extensionality, hence by the known
-translations between extensional and intensional type
-theories\cite{hofmann95extensional,winterhalter2019eliminating}, we can switch
-between strict and propositional equalities.
+formalisations use UIP and function extensionality, hence we can switch between
+strict and propositional equalities, via the known translations between
+extensional and intensional type
+theories\cite{hofmann95extensional,winterhalter2019eliminating}.
 

### 2019-11-11 paper/paper.tex
remove infinitary from universaliit.tex

@@
   Chapman-style syntax of type theory with conversion relations for each sort
-  where e.g.\ the sort of types is indexed by contexts.  In previous work it was
-  shown that all finitary IITs can be constructed from a quotient
+  where e.g.\ the sort of types is indexed by contexts. It follows from previous
+  work that all finitary IITs can be constructed from a quotient
   inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
@@
   thought.}
+
+\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)n}
 
@@
   & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
-  & \rlap{(5) Infinitary parameters} \\[0.5em]
-  & \Pii && : (T:\Set)\ra(T\ra\Tm\,\Gamma\,\U)\ra\Tm\,\Gamma\,\U \\
-  & \blank\appi\blank && : \Tm\,\Gamma\,(\El\,(\Pii\,T\,b))\ra(\alpha:T)\ra\Tm\,\Gamma\,(\El\,(b\,\alpha)) \\
-  & {\Pii[]} && : (\Pii\,T\,b)[\sigma] = \Pii\,T\,(\lambda \alpha.(b\,\alpha)[\sigma]) \\
-  & {\appi[]} && : (t\appi\alpha)[\sigma] = (t[\sigma])\mathop{\appi}\alpha
 \end{alignat*}
+\todo{We may need more abbrevs here}
 Abbreviation:
@@
 
-\todo[inline]{add examples where all the different constructors are used}
+\todo{add examples where all the different constructors are used}
 

### 2019-11-11 paper/universaliit.tex
section 4
remove infinitary from universaliit.tex

@@
+\todo{Section on metatheoretic notation missing?}
+
 We work in the setting of extensional type theory, although we occasionally indicate through $e_1,\dots,e_n \#
@@
 \subsubsection{Untyped syntax}
-The untyped syntax is defined as the following inductive datatype:
-\todo{András: two columns below perhaps}
+The untyped syntax is defined as the following inductive datatype. Variables are
+modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
+the context. We use the additional default constructor $\erru:\Tmu$ in case of
+error (ill-scoped substitution). The typing judgments will not mention $\erru$.
+\todo{András: prettify below maybe}
 \begin{alignat*}{5}
@@
   & \Elu && : \Tmu\ra\Tyu \\
-  & \rlap{(3) Parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\[0.5em]
   & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
   & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
-  & \rlap{(4-5) Metatheoretic and infinitary parameters} \\[0.5em]
+  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
   & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
@@
 \end{alignat*}
-Note that we overload the constructor for untyped application for infinitary and
-meta-theoretic parameters. Variables are modeled as de Bruijn indices, i.e.\ as
-natural numbers pointing to a position in the context.
 
@@
 on untyped syntax: the two ways yield provably equal definitions. In the
-  following, we assume that it is defined by recursion.
- We use the additional default constructor $\erru:\Tmu$ in case of error (for
-example, when the substitution is not long enough).  This is an irrelevant
-constructor in the end, as the typing judgments will not mention it. We also
-make use of the following definition:
+following, we assume that it is defined by recursion. We also make use of the
+following definition:
   \begin{alignat*}{5}
@@
   \end{alignat*}
-We defer laws for identity substitutions to after the definition of the typing
+We defer laws for identity substitutions after the definition of the typing
 judgments, as the proofs require that some inputs are well-typed.
@@
   & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
-  & \rlap{(3) Parameters} \\[0.5em]
+  & \rlap{(3) Inudctive parameters} \\[0.5em]
   & \Piw && :
@@
     \Gamma \vdash t \appu u \in  B [0 := u] \\
-  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
+  & \rlap{(4) Inductive parameters} \\[0.5em]
   & \Pimw && :
@@
     \ra(\Gamma \vdash t \in \Pimu \,T\,A)
-    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
-    \\
-  & \rlap{(5) Infinitary parameters} \\[0.5em]
-  & \Piiw && :
-    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
-    ((t : T)\ra\Gamma\vdash A\,t\in \Uu)
-    \\ &&& \ra
-    \Gamma \vdash \Pimu \,T\,A \in \Uu
-  \\
-  & \appiw && :
-    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
-    ((t : T)\ra\Gamma\vdash A\,t\in\Uu)
-    \\
-    & &&
-    \ra
-    (\Gamma \vdash t \in \Piiu \,T\, A)
     \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
@@
   \end{alignat*}
-Finally, we show that the identity substitution itself is well typed:
+Finally, we show that the identity substitution itself is well-typed:
   \begin{alignat*}{5}
@@
 None of the constructions in this section use UIP. Functional extensionality is
-necessary because the metatheoretic or infinitary (untyped) $\Pi$ takes a
-metatheoretic function as an argument. An example induction step that uses it
-is in the type preservation proof for identity substitutions (this is an
-equation satisfied by a model of the QIIT), in particular in the case
-$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
-equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
+necessary because the untyped metatheoretic $\Pi$ takes a metatheoretic function
+as an argument. An example induction step that uses it is in the type
+preservation proof for identity substitutions (this is an equation satisfied by
+a model of the QIIT), in particular in the case $(\Pim\,T\,
+A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is equal to
+$\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
 hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.
@@
       (\model{x} = e_B \# \model{t}\model{\hat{\oldapp}}u)
-    \\
-  & \rlap{(5) Infinitary parameters} \\[0.5em]
-    & \Piiw T\,A\,\wf{\Gamma}\wf{A} \sim \model{B} &&
-    :=
-      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
-      \times
-      (\model{B} = \model{\Pii}\,T\,\model{A})
-    \\
-    & \relt{\appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u}{\model{\Gamma}}{\model{B}}{\model{x}}
-    % & \appiw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
-    % \\
-    % & \qquad
-    % (\model{\Gamma}\vdash x\in \model{B})
-     &&
-     :=
-      \sum_{\model{A}} ((t : T) \ra \wf{A}\sim\model{A}\, t)
-      \times
-      \sum_{\model{t}} (\wf{t}\sim\model{t})
-      \times
-      \\ & &&
-      \qquad
-      (e_B : \model{B} = \model{\Pii}\,T\,\model{A})
-      \times
-      (\model{x} = e_B \# \model{t}\model{\tilde{\oldapp}}u)
 \end{alignat*}
@@
   \end{alignat*}
-  We mentionned that in order to avoid a recursive-recursive definition, we
+  We mentioned that in order to avoid a recursive-recursive definition, we
   removed some hypotheses in the list of arguments of the relation. Such
   hypotheses are sometimes missed, for example in the case of the empty
-  substitution, or in the case of variables, requiring us to state additional
-  equalities.  % some redundancies that we could not avoid when defining % the
-  relation, for example in the case of the empty substitution.  Because of
-  this, % It is because of the redundancies which we mentionned above we need
-  UIP to show that the types $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
-  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may believe that the
-  use of UIP could be avoided by embracing the alternative verbose definition
-  that we suggested before, expecting that
+  substitution or in the case of variables, requiring us to state additional
+  equalities. Because of this, we need UIP to show that
+  $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
+  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
+  of UIP could be avoided by using the alternative verbose definition that we
+  suggested before, expecting that
   $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
-  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this not so
+  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
   obvious. For example, we were not able to define
   $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion. In related work,
-  Hugunin investigated constructing inductive-inductive types without
+  Hugunin investigated constructing IITs without
   UIP\cite{hugunin2019constructing}, and demonstrated that well-formedness
@@
 \subsection{Uniqueness}
-Our goal is to show that the syntactic model is initial. Thus, what remains is
-to show that the morphism constructed in the previous section is unique. We
-exploit right uniqueness of the relation: it is enough to show that any such
-morphism maps a syntactic context to a related semantic context, and similarly
-for types and terms.
+Our goal is to show that the syntactic model is initial. Thus, it remains to
+show that the morphism constructed in the previous section is unique. We exploit
+right uniqueness of the relation: it is enough to show that any such morphism
+maps a syntactic context to a related semantic context, and similarly for types
+and terms.
 

### 2019-11-11 paper/paper.tex
infinitary IIT future work

@@
 
-\begin{itemize}
+\begin{enumerate}
 \item
@@
   induction yields restriction to IITs.
-\end{itemize}
+\end{enumerate}
 
@@
 
-Infinitary, explain why it is difficult. Without UIP, without funext,
-for QIITs (in the setoid model).
+\subsection{Infinitary IITs}.
+
+The current work only concerns finitary IITs. An extension would be to also
+allow infinitely branching inductive types such as W-types. This would first
+require giving semantics for infinitary IITs (to our knowledge there is no
+previously published semantics that we can borrow), and also giving a term model
+construction analogously to finitary QIITs. These steps seem feasible. However,
+it seems to be more difficult to construct the syntax of infinitary IIT
+signatures without using quotients. The reason is that such syntax would not be
+strictly restricted to neutral terms: we would need $\lambda$-abstraction and
+$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
+construction for infinitary IITs. A definition for normal preterms and typing
+judgments on them may still be possible, but it appears to be much more
+complicated than before (the current authors attempted this without conclusive
+success).
+
+\todo{Without UIP, without funext,
+for QIITs (in the setoid model).}
 

### 2019-11-11 paper/paper.tex
stuff

@@
 
-\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)n}
+\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)n}
 
@@
 
+\subsection{Construction without UIP and function extensionality}
 \todo{Without UIP, without funext,

### 2019-11-11 paper/paper.tex

@@
 inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
-QIIT can be constructed only using inductive types.
+QIIT can be constructed only using inductive types. In Section
+\ref{sec:constructingiits} we describe the construction of IITs from the syntax
+of IIT signatures.
 
@@
 \section{Constructing IITs from the Theory of IIT Signatures}
-\label{sec:andras}
+\label{sec:constructingiits}
 
@@
 $\Gamma$, the initial algebra can be built from sets of terms of the form
-$\Tm\,\Gamma\,(\El\,a)$.
+$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
+\[
+  \Gamma := \cdot\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
+\]
+we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
+since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.
 
-In this section, we only need to check that the signatures in
-\cite{Kaposi:2019:CQI:3302515.3290315} and the term model construction restricts
-to IITs, i.e.\ that from a syntax for the theory of IIT signatures all finitary
-IITs are constructible.
+In this section, we only need to check that the signatures and term model
+construction in \cite{Kaposi:2019:CQI:3302515.3290315} restrict to IITs,
+i.e.\ that from a syntax for the theory of IIT signatures all finitary IITs are
+constructible.
 
@@
 judgments on them may still be possible, but it appears to be much more
-complicated than before (the current authors attempted this without conclusive
-success).
+complicated than before (the current authors have attempted this without
+conclusive success).
 

### 2019-11-11 paper/universaliit.tex

@@
 uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
-relation, and totality by already assuming well-formedness in the syntax. In type
+relation, and totality by already assuming well-formedness of $A$. In type
 theory, using indexed families instead of display maps is a common convenience,

### 2019-11-11 paper/paper.tex
stuff

@@
 
-\subsection{Infinitary IITs}.
-
 The current work only concerns finitary IITs. An extension would be to also
@@
 $\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
-construction for infinitary IITs. A definition for normal preterms and typing
+construction for infinitary IITs. A definition of normal preterms and typing
 judgments on them may still be possible, but it appears to be much more
@@
 
-\subsection{Construction without UIP and function extensionality}
-\todo{Without UIP, without funext,
-for QIITs (in the setoid model).}
+As mentioned in Section \ref{sec:right_uniqueness}, it also remains an open
+problem whether IITs are reducible to inductive types in a UIP-free setting. To
+show this, we would need to construct the syntax of signatures without UIP, and
+also reproduce the semantics and term model construction for IITs without UIP.
 

### 2019-11-11 paper/paper.tex
notation section

@@
 
+\subsection{Metatheory and Notation}
+
+We work in intensional Martin-Löf type theory, equipped with the following features:
+\begin{itemize}
+  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
+  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when non-dependent.
+  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
+        when we want to leave the type of the first projection implicit. We use $A
+        \times B$ for non-dependent pairs.
+  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
+    this paper, we use extensional notation, and do not write explicit
+    transports over equalities. However, we occasionally indicate by
+    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
+    $e_1$,\dots,$e_n$.
+  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
+  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.
+
+\end{itemize}
+
 \subsection{Formalisation}

### 2019-11-11 paper/paper.tex
formalisation section

@@
 
-\todo{UIP, funext, rewrite rules, version of Agda.}
+Agda formalisation of Sections \ref{sec:theory_of_signatures} and
+\ref{sec:ambroise} can be found at
+\url{https://github.com/amblafont/UniversalII}. In addition to the features in
+the previous section, this formalisation also uses equality reflection for some equalities, through rewriting pragmas\cite{cockxsprinkles}.
+This is purely for convenience, and in principle we could eliminate the instances of equality reflection via UIP and function extensionality.
 

### 2019-11-11 paper/paper.tex
part of the intro

--- 2019-11-11 paper/paper.tex
@@
   a small domain-specific type theory where a context is a signature for an
-  IIT. In this paper we construct this theory using only inductive types,
+  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
   thereby showing a reduction of all finitary IITs to inductive types.  We rely
@@
   the construction of the theory of IIT signatures in Agda.
-
 \end{abstract}
@@
 
-\todo{Note by Ambrus: to be beautified, I only wanted to add the line of
-  thought.}
+\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)}
 
-\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)n}
+Most mutual inductive types can be reduced to indexed inductive types
+where the index gives the information which sort we meant. An example
+is the predicates $\mathsf{isEven}$ and $\mathsf{isOdd}$ defined
+mutually by the following constructors.
+\begin{alignat*}{5}
+  & \mathsf{isEven} && : \N\ra\Set \\
+  & \mathsf{isOdd} && : \N\ra \Set \\
+  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,n \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,n
+\end{alignat*}
+These can be reduced to the following single inductive family.
+\begin{alignat*}{5}
+  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
+  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,n \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,n
+\end{alignat*}
 
-\todo{
-  András: construction of infinitary IITs doesn't (yet) work. Should we drop
-  infinitary functions from signatures?
-  }
+Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
+definition of a type and a family of types over the first one. IITs
+were originally introduced to represent the well-typed syntax of type
+theory itself, and the main example is still the James Chapman style
+\cite{chapman09eatitself} syntax for type theory. A minimal use case
+is the IIT of contexts and types given by the following constructors.
+\begin{alignat*}{5}
+  & \Con && : \Set \\
+  & \Ty && : \Con\ra\Set \\
+  & \mathsf{empty} && : \Con \\
+  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
+  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
+  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
+%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
+\end{alignat*}
+This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
+constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
+$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
+at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
+precludes the above reduction which worked for
+$\mathsf{isEven}$--$\mathsf{isOdd}$ as we would get a type indexed
+over itself. Another unique feature of IITs (which also shows up in
+higher inductive types \cite{HoTTbook}) is that later constructors can
+refer to previous constructors: in our case, $\mathsf{El}$ mentions
+$\mathsf{ext}$.
+% \[
+%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
+% \]
 
-Most mutual inductive types (e.g.\ isEven-isOdd) can be reduced to indexed
-inductive types where the index gives the information which sort we
-meant. (Indexed inductive types can be reduced to inductive types using a method
-similar to what we use here, preterms and well-formedness predicate.)
-Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual definition
-of a type and a family of types over the first one. The above reduction does not
-work anymore.
+The elimination principle for an IIT has two motives and one method
+for each constructor. It is an example of a \emph{recursive-recursive}
+function (following the nomenclature of \cite{forsberg-phd}) which
+means that there are two mutually defined functions where the type of
+the second function depends on the first function. The proof assistant
+Agda \cite{norell07thesis} allows the definition of such functions
+(even from non-IITs) and it is also currently the only proof assistant
+supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
+given the following components,
+\begin{alignat*}{5}
+  & Con^M && : \Con\ra\Set \\
+  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & empty^M && : Con^M\,\mathsf{empty} \\
+  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
+  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
+  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
+%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
+\end{alignat*}
+we obtain two functions
+\begin{alignat*}{5}
+  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
+  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
+\end{alignat*}
+with the following computation rules.
+\begin{alignat*}{5}
+  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
+\end{alignat*}
 
-Examples of IITs: Con-Ty is the minimal interesting one, then we have
-James Chapman style \cite{chapman09eatitself} well-typed syntax for
-type theory.
+As opposed to mutual inductive types, reducing IITs to inductive types
+is an open problem. Forsberg \cite{forsberg-phd} showed a reduction
+which only supports a simpler, non recursive-recursive elimination
+principle (\todo{check}). Hugunin \cite{jasper} reduced an example IIT
+to inductive types without UIP, but he also only constructed the
+simpler non recursive-recursive eliminator.
 
-TODO: list examples here and explain. Also the elimination principles
-which is recursive-recursive. Talk about connection to setoidified
-CwFs (I think these are explained in some Dybjer paper) and maybe the
-initiality conjecture.
+The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is
+a small type theory in which every context encodes a valid quotient
+inductive-inductive type (QIIT) signature. For example, the signature
+for natural numbers is simply the context
+$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
+$suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315}
+showed that if this theory of signatures exists, then any finitary
+QIIT can be defined from it. If we omit the type former $\mathsf{Id}$
+from the theory of signatures, we get a theory of finitary IIT
+signatures, and the construction of
+\cite{Kaposi:2019:CQI:3302515.3290315} still works: constructing an
+arbitrary IIT from the theory of IIT signatures does not need the
+$\mathsf{Id}$ type former to be present.
 
-Reduction of IITs to inductive types is an open problem. Forsberg
-showed that IITs can be reduced to inductive types in the setting of
-ETT, but he only constructed the nondependent (or simple? was this the
-non-recursive-recursive?) elimination principle. Hugunin \cite{jasper}
-reduced an example IIT to inductive types without UIP, but also only
-constructing the non-recursive-recursive eliminator. Streicher
-\cite{streichersemantics}. Brunerie \cite{} constructed the initial
-category with extra structure (TODO: fix this) using UIP.
+In this paper, we construct the theory of IIT signatures only using
+indexed inductive types, thus \todo{rewrite from here}
 
-Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
-inductive-inductive types (HIITs)
-\cite{kaposi_et_al:LIPIcs:2018:9190}. This is a small type theory in
-which every context encodes a valid IIT signature. For example, the
-signature for natural numbers is simply the context
-$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
-$suc$ are simply variable names). Later they showed
-\cite{Kaposi:2019:CQI:3302515.3290315} that if a subset of this theory
-of signatures exists, then any finitary QIIT can be defined from
-it. In this paper we further restrict the subset by not allowing
-equality constructors (thus only covering IITs) but at the same time
-we extend it to allow infinitary constructors. To derive the existence
-of all IITs from the theory of IIT signatures we use a variant of the
-CwF model \cite{Kaposi:2019:CQI:3302515.3290315} where types are
-modelled not by displayed CwFs but by CwF isofibrations. We need this
-extension to support infinitary constructors. Furthermore, using the
-method introduced by Streicher \cite{streichersemantics} we show the
+To derive the existence of all IITs from the theory of IIT signatures
+we use a variant of the CwF model
+\cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
+displayed CwFs but by CwF isofibrations. We need this extension to
+support infinitary constructors. Furthermore, using the method
+introduced by Streicher \cite{streichersemantics} we show the
 existence of the theory of IIT signatures. As opposed to Streicher, we
@@
 
-Similar techique for other type theories: Streicher. Initiality
-project, Brunerie. These all use quotients. Closest: Lafont TYPES
-abstract 2017. Does not use quotients.
+(Indexed inductive types can be reduced to inductive types using a method
+similar to what we use here, preterms and well-formedness predicate.)
+
+Talk about connection to setoidified CwFs (I think these are explained
+in some Dybjer paper) and maybe the initiality conjecture.
+
+Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
+inductive-inductive types (HIITs)
+\cite{kaposi_et_al:LIPIcs:2018:9190}.
+
+Similar techique for other type theories: Streicher
+\cite{streichersemantics}. Initiality project, Brunerie \cite{}. These
+all use quotients. Closest: Lafont TYPES abstract 2017. Does not use
+quotients.
 
@@
 
+\subsection{Infinitary IITs}.
+
 The current work only concerns finitary IITs. An extension would be to also
@@
 $\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
-construction for infinitary IITs. A definition of normal preterms and typing
+construction for infinitary IITs. A definition for normal preterms and typing
 judgments on them may still be possible, but it appears to be much more
@@
 
-As mentioned in Section \ref{sec:right_uniqueness}, it also remains an open
-problem whether IITs are reducible to inductive types in a UIP-free setting. To
-show this, we would need to construct the syntax of signatures without UIP, and
-also reproduce the semantics and term model construction for IITs without UIP.
+\subsection{Construction without UIP and function extensionality}
+\todo{Without UIP, without funext,
+for QIITs (in the setoid model).}
 

### 2019-11-11 paper/paper.tex

@@
 
+\subsection{Metatheory and Notation}
+
+We work in intensional Martin-Löf type theory, equipped with the following features:
+\begin{itemize}
+  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
+  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when non-dependent.
+  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
+        when we want to leave the type of the first projection implicit. We use $A
+        \times B$ for non-dependent pairs.
+  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
+    this paper, we use extensional notation, and do not write explicit
+    transports over equalities. However, we occasionally indicate by
+    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
+    $e_1$,\dots,$e_n$.
+  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
+  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.
+
+\end{itemize}
+
 \subsection{Formalisation}
 
-\todo{UIP, funext, rewrite rules, version of Agda.}
+Agda formalisation of Sections \ref{sec:theory_of_signatures} and
+\ref{sec:ambroise} can be found at
+\url{https://github.com/amblafont/UniversalII}. In addition to the features in
+the previous section, this formalisation also uses equality reflection for some equalities, through rewriting pragmas\cite{cockxsprinkles}.
+This is purely for convenience, and in principle we could eliminate the instances of equality reflection via UIP and function extensionality.
 

### 2019-11-11 paper/paper.tex
stuff

--- 2019-11-11 paper/paper.tex
@@
   \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
-  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when non-dependent.
+  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
+    non-dependent. We sometimes omit function arguments, by implicitly
+    generalising over variables.
   \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
@@
 
-We use the naming convention: sorts, operators and equations. Sorts
-and operators might have parameters.
+The theory of IIT signatures is given as an initial category with families
+(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
+part constitutes the substitution calculus, while the type formers allow
+declaring new sorts and operators in a signature.
 \begin{alignat*}{5}
@@
   & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
-  & \cdot && : \Con \\
-  & \epsilon && : \Sub\,\Gamma\,\cdot \\
-  & {\cdot\eta} && : \{\sigma : \Sub\,\Gamma\,\cdot\} \ra \sigma = \epsilon \\
+  & \boldsymbol{\cdot} && : \Con \\
+  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
+  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
   & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
@@
   & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
-  & \rlap{(2) Sorts and operators} \\[0.5em]
+  & \rlap{(2) Universe} \\[0.5em]
   & \U && : \Ty\,\Gamma \\
@@
   & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
-  & \rlap{(3) Parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\[0.5em]
   & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
@@
 \end{alignat*}
-\todo{We may need more abbrevs here}
-Abbreviation:
-\[
-  \sigma^\uparrow = (\sigma\circ\p,\q)
-\]
+\subsubsection{Notation and abbreviations}
 
-\todo{add examples where all the different constructors are used}
+We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
+for inductive parameters, we have $\app$ as a categorical version of function
+application. We recover the usual type-theoretic application as follows:
+\begin{alignat*}{5}
+  & \blank\appm\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
+  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
+  & t \appm u := \app\,t\,[\id\,u]
+\end{alignat*}
+We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
+$p$ composed with itself $n$ times. However, for informal examples we instead
+use a nameful notation, where we introduce names in context extension and in the
+$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,(A : \U)\,\ext\,(a :
+\El\,A)$ in a signature for pointed sets.
+
+\subsubsection{Example signatures}
 
@@
 \[
-  \Gamma := \cdot\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
+  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
 \]

### 2019-11-11 paper/universaliit.tex
stuff

@@
-\todo{Section on metatheoretic notation missing?}
-
-We work in the setting of extensional type theory, although we occasionally indicate through $e_1,\dots,e_n \#
-t$ that $t$ is well-typed thanks to the equalities $e_1$,\dots,$e_n$.
 % We are also careful to explain where UIP or functional extensionality that is
@@
   & \varu  && : \N \ra \Tmu \\
-  & \rlap{(2) Sorts and operators} \\[0.5em]
+  & \rlap{(2) Universe} \\[0.5em]
   & \Uu && : \Tyu \\
@@
   \vdash S\,n \invar \wku \, A \\
-  & \rlap{(2) Sorts and operators} \\[0.5em]
+  & \rlap{(2) Universe} \\[0.5em]
   & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
@@
   \\
-  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\cdot} \\
+  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\boldsymbol{\cdot}} \\
   & \relS{\epsilonw}{\model{\Gamma}}{\model{E}}{\model{\delta}}
@@
   &&
- := (e_E : \model{E} = \model{\cdot})\times (\model{\delta} = e_E \#
+ := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = e_E \#
  \model{\epsilon})
@@
   \\
-  & \rlap{(2) Sorts and operators} \\[0.5em]
+  & \rlap{(2) Universe} \\[0.5em]
   & \Uw \wf{\Gamma} \wf{A} \sim \model{A} &&
@@
   % As we argued before, in the  of empty substitution $\epsilon$: we require the equality
-  % $(e_C : \model{E} = \model{\cdot})$, although
+  % $(e_C : \model{E} = \model{\boldsymbol{\cdot}})$, although
   % our claimed external assumption that $\model{E}$ is  related to the
@@
   \subsubsection{Right uniqueness}
+  \label{sec:right_uniqueness}
   We show by recursion that the relation is right unique in the following sense:
@@
 %   & \blank\leq\blank  && : \model\Con\ra \model\Con\ra\Set \\
-%   & {\leq}\cdot  && : \model{\Gamma} \leq \model{\Gamma} \\
+%   & {\leq}\boldsymbol{\cdot}  && : \model{\Gamma} \leq \model{\Gamma} \\
 %   & \blank{\leq}{\ext}\blank  && : \model{\Gamma} \leq \model{\Delta} \ra (A :
@@
 %   (\model{\Delta} : \Tel\,\model\Gamma) \ra \model{\Con} \\
-%   & \model\Gamma \merge (\model\Gamma,{\leq}\cdot) && :=  \model\Gamma \\
+%   & \model\Gamma \merge (\model\Gamma,{\leq}\boldsymbol{\cdot}) && :=  \model\Gamma \\
 %   & \model\Gamma \merge (\model\Delta\model{\ext}\model{A}, \Delta_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\Gamma \model\merge \model\Delta)\model\ext \model{A}\\
@@
 %   & (\wf\Gamma : \Gamma \merge \emptyu \vdash)\sim\model\Delta  && :=
-%   \model\Delta =  {\leq}\cdot\\
+%   \model\Delta =  {\leq}\boldsymbol{\cdot}\\
 %   & (\wf\Delta \extw \wf{A} : \Gamma \merge (\Delta \extu A) \vdash)\sim\model{E}  && :=

### 2019-11-11 paper/paper.tex
example sigs and notation

@@
 
-\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)}
-
-Most mutual inductive types can be reduced to indexed inductive types
-where the index gives the information which sort we meant. An example
-is the predicates $\mathsf{isEven}$ and $\mathsf{isOdd}$ defined
-mutually by the following constructors.
+Many mutual inductive types can be reduced to indexed inductive types, where the
+index disambiguates different sorts. An example is the predicates
+$\mathsf{isEven}$ and $\mathsf{isOdd}$ defined mutually by the following
+constructors.
 \begin{alignat*}{5}
@@
 were originally introduced to represent the well-typed syntax of type
-theory itself, and the main example is still the James Chapman style
-\cite{chapman09eatitself} syntax for type theory. A minimal use case
+theory itself, and the main example is still the Chapman-style
+\cite{chapman09eatitself} syntax for a type theory. A minimal use case
 is the IIT of contexts and types given by the following constructors.
@@
   & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
-  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
+  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
   & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
@@
 precludes the above reduction which worked for
-$\mathsf{isEven}$--$\mathsf{isOdd}$ as we would get a type indexed
+$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed
 over itself. Another unique feature of IITs (which also shows up in
@@
 
-The elimination principle for an IIT has two motives and one method
+The elimination principle for the above IIT has two motives and one method
 for each constructor. It is an example of a \emph{recursive-recursive}
 function (following the nomenclature of \cite{forsberg-phd}) which
-means that there are two mutually defined functions where the type of
+means that there are two mutually defined functions, where the type of
 the second function depends on the first function. The proof assistant
@@
 
-As opposed to mutual inductive types, reducing IITs to inductive types
-is an open problem. Forsberg \cite{forsberg-phd} showed a reduction
-which only supports a simpler, non recursive-recursive elimination
-principle (\todo{check}). Hugunin \cite{jasper} reduced an example IIT
-to inductive types without UIP, but he also only constructed the
-simpler non recursive-recursive eliminator.
+As opposed to mutual inductive types, reducing IITs to inductive types is an
+open problem. Forsberg \cite{forsberg-phd} showed a reduction which only
+supports a simpler, non-recursive-recursive elimination principle
+(\todo{check}). Hugunin \cite{jasper} reduced several IITs to inductive types,
+working inside a UIP-free cubical type theory, but he also only constructed the
+simpler non-recursive-recursive eliminator.
 
@@
   \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
-    this paper, we use extensional notation, and do not write explicit
-    transports over equalities. However, we occasionally indicate by
+    this paper, we informally use extensional notation, and do not write
+    explicit transports over equalities. However, we occasionally indicate by
     $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
@@
 \begin{alignat*}{5}
-  & \blank\appm\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
+  & \blank\oldapp\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
   \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
-  & t \appm u := \app\,t\,[\id\,u]
+  & t \oldapp u := \app\,t\,[\id\,u]
 \end{alignat*}
+We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[p])$ as $a\ra B$.
+
 We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
@@
 use a nameful notation, where we introduce names in context extension and in the
-$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,(A : \U)\,\ext\,(a :
-\El\,A)$ in a signature for pointed sets.
+$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,A : \U\,\ext\,a :
+\El\,A$ in a signature for pointed sets.
 
 \subsubsection{Example signatures}
+
+Natural numbers:
+\[
+\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
+\]
+Lists with elements of a given $A : \Set$ type. Here we use the $\Pim$ function
+to include an external type into the signature.
+\[
+\boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
+      cons : \Pim\,A\,(\lambda \_.\, L\ra L)
+\]
+The inductive-inductive Con-Ty example from Section \ref{sec:intro}:
+\begin{alignat*}{3}
+  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \ra Con) \\
+  & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
+    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp U)))
+\end{alignat*}
+
+  %% & \Ty && : \Con\ra\Set \\
+  %% & \mathsf{empty} && : \Con \\
+  %% & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
+  %% & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
+  %% & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
 
@@
 $\Gamma$, the initial algebra can be built from sets of terms of the form
-$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
+$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
+numbers
 \[

### 2019-11-11 paper/universaliit.tex
example sigs and notation

@@
   Hugunin investigated constructing IITs without
-  UIP\cite{hugunin2019constructing}, and demonstrated that well-formedness
-  predicates used in syntactic models can be subtly incompatible with UIP. Also,
-  while Hugunin does not use UIP, he only shows a weak version of dependent
-  elimination for the constructed IITs. Hence, the question whether IITs are
-  reducible to inductive types in a UIP-free setting remains open.
+  UIP\cite{hugunin2019constructing} in cubical type theory, and demonstrated
+  that well-formedness predicates used in syntactic models can subtly break
+  in that setting. Also, while Hugunin does not use UIP, he only shows a
+  weak version of dependent elimination for the constructed IITs. Hence, the
+  question whether IITs are reducible to inductive types in a UIP-free setting
+  remains open.
 

### 2019-11-11 paper/paper.tex
stuff

@@
 \end{alignat*}
-This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
-constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
-$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
-at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
-precludes the above reduction which worked for
-$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed
-over itself. Another unique feature of IITs (which also shows up in
-higher inductive types \cite{HoTTbook}) is that later constructors can
-refer to previous constructors: in our case, $\mathsf{El}$ mentions
+This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
+$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
+hence the two sorts have to be defined at the same time. Moreover, $\Ty$ is
+indexed over $\Con$. This precludes a reduction analogous to the reduction of
+$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
+itself. Another unique feature of IITs (which also shows up in higher inductive
+types \cite{HoTTbook}) is that later constructors can refer to previous
+constructors: in our case, $\mathsf{El}$ mentions
 $\mathsf{ext}$.
@@
 
-As opposed to mutual inductive types, reducing IITs to inductive types is an
-open problem. Forsberg \cite{forsberg-phd} showed a reduction which only
-supports a simpler, non-recursive-recursive elimination principle
-(\todo{check}). Hugunin \cite{jasper} reduced several IITs to inductive types,
-working inside a UIP-free cubical type theory, but he also only constructed the
-simpler non-recursive-recursive eliminator.
+Reducing IITs to inductive types is an open problem. Forsberg
+\cite{forsberg-phd} showed a reduction which only supports a simpler,
+non-recursive-recursive elimination principle. Hugunin \cite{jasper} reduced
+several IITs to inductive types, working inside a cubical type theory, but he
+also only constructed a weaker eliminator.
 
-The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is
-a small type theory in which every context encodes a valid quotient
-inductive-inductive type (QIIT) signature. For example, the signature
-for natural numbers is simply the context
-$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
-$suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315}
-showed that if this theory of signatures exists, then any finitary
-QIIT can be defined from it. If we omit the type former $\mathsf{Id}$
-from the theory of signatures, we get a theory of finitary IIT
-signatures, and the construction of
-\cite{Kaposi:2019:CQI:3302515.3290315} still works: constructing an
-arbitrary IIT from the theory of IIT signatures does not need the
-$\mathsf{Id}$ type former to be present.
+The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
+type theory in which every context encodes a valid quotient inductive-inductive
+type (QIIT) signature. For example, the signature for natural numbers is simply
+the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
+and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
+that if this theory of signatures exists, then any finitary QIIT can be derived
+from it via a term model construction. If we omit the type former for equality
+constructors from the theory of signatures, we get a theory of finitary IIT
+signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
+works: constructing an arbitrary IIT from the theory of IIT signatures does not
+need the $\mathsf{Id}$ type former to be present.
 
-In this paper, we construct the theory of IIT signatures only using
-indexed inductive types, thus \todo{rewrite from here}
+However, the QIIT term model construction assumes that we already have the
+syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
+that all finitary QIITs are derivable from a particular fairly complicated QIIT.
+Likewise, restriced to finitary IITs, we only know that they are constructible
+from a particular QIIT.
 
-To derive the existence of all IITs from the theory of IIT signatures
-we use a variant of the CwF model
-\cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
-displayed CwFs but by CwF isofibrations. We need this extension to
-support infinitary constructors. Furthermore, using the method
-introduced by Streicher \cite{streichersemantics} we show the
-existence of the theory of IIT signatures. As opposed to Streicher, we
-don't use quotients in this construction as the theory of IIT
-signatures can be given only using neutral terms.
+In this paper, we construct the theory of IIT signatures, which is a large
+infinitary QIIT, using only indexed inductive types, uniqueness of identity
+types (UIP) and function extensionality. Thus, we show that all finitary IITs are
+reducible to the same feature set.
 
-Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
-we can translate our arguments to ITT+UIP+funext. Section
-\ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
-funext.
+\subsection{Structure of the Paper}
 
-This is the paper version of our TYPES 2019 talk.
+We first define a theory of IIT signatures in which every context is a signature
+(Section \ref{sec:theory_of_signatures}). We present this as an algebraic
+\cite{ttintt} theory, specifically as a quotient inductive-inductive type
+(QIIT). In Section \ref{sec:ambroise} we show that this QIIT can be constructed
+using inductive types, UIP and function extensionality. In Section
+\ref{sec:constructingiits} we describe the construction of IITs from the syntax
+of IIT signatures.
+
+%% To derive the existence of all IITs from the theory of IIT signatures
+%% we use a variant of the CwF model
+%% \cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
+%% displayed CwFs but by CwF isofibrations. We need this extension to
+%% support infinitary constructors. Furthermore, using the method
+%% introduced by Streicher \cite{streichersemantics} we show the
+%% existence of the theory of IIT signatures. As opposed to Streicher, we
+%% don't use quotients in this construction as the theory of IIT
+%% signatures can be given only using neutral terms.
+
+%% Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
+%% we can translate our arguments to ITT+UIP+funext. Section
+%% \ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
+%% funext.
+
+%% This is the paper version of our TYPES 2019 talk.
 
@@
 
-\subsection{Structure of the Paper}
-
-We first define a \emph{theory of IIT signatures} in which every context is a
-valid IIT signature (Section \ref{sec:theory_of_signatures}). We present this
-type theory as an algebraic \cite{ttintt} theory, specifically as a quotient
-inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
-QIIT can be constructed only using inductive types. In Section
-\ref{sec:constructingiits} we describe the construction of IITs from the syntax
-of IIT signatures.
-
 \subsection{Metatheory and Notation}
@@
 \url{https://github.com/amblafont/UniversalII}. In addition to the features in
-the previous section, this formalisation also uses equality reflection for some equalities, through rewriting pragmas\cite{cockxsprinkles}.
-This is purely for convenience, and in principle we could eliminate the instances of equality reflection via UIP and function extensionality.
+the previous section, this formalisation also uses equality reflection for some
+equalities, through rewriting pragmas\cite{cockxsprinkles}. This is purely for
+convenience, and in principle we could eliminate the instances of equality
+reflection via UIP and function extensionality.
 
@@
 
-The theory of IIT signatures is given as an initial category with families
+The theory of IIT signatures is given as a category with families
 (CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
 part constitutes the substitution calculus, while the type formers allow
-declaring new sorts and operators in a signature.
+declaring new sorts and operators in a signature. Below, we list all components
+for the theory of signatures; this can be viewed as a notion of model for the
+theory, which is an iterated $\Sigma$ type. In this paper, we construct the
+\emph{initial} such model, i.e.\ a model equipped with a unique recursion principle.
+
+Note that initiality refers to a notion of homomorphism between models. As
+of now, there is no general metatheory for (large, infinitary) QIITs from which
+this notion would fall out. Hence, we give an ad-hoc definition for
+homomorphisms, which can be summarized as a functor between underlying
+categories which strictly preserves all additional family structure and type
+formers.
+
 \begin{alignat*}{5}
@@
 
-  %% & \Ty && : \Con\ra\Set \\
-  %% & \mathsf{empty} && : \Con \\
-  %% & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
-  %% & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
-  %% & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
-
 \section{Constructing the Theory of IIT Signatures}
@@
 
-\todo{Write this section}
+%% \todo{Write this section}
 
-Show that the theory of IIT signatures as described in Section
-\ref{sec:theory_of_signatures} exists, that is, it can be defined only
-using indexed inductive types and a $\Prop$ sort.
+%% Show that the theory of IIT signatures as described in Section
+%% \ref{sec:theory_of_signatures} exists, that is, it can be defined only
+%% using indexed inductive types and a $\Prop$ sort.
 
-Ambroise has done this in Agda with funext and UIP using an
-extrinsic--intrinsic translation.
+%% Ambroise has done this in Agda with funext and UIP using an
+%% extrinsic--intrinsic translation.
 
-Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.
+%% Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.
 
-In this formalization, there is a discrepancy for the typing rule of application
-compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
-\[
-   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
-\]
-but
-\[
-   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
-\]
-Actually, I started with the first version, but at some point I
-changed for the second one because it made some stuff easier (although
-I don't remember exactly what it was).
+%% In this formalization, there is a discrepancy for the typing rule of application
+%% compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
+%% \[
+%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
+%% \]
+%% but
+%% \[
+%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
+%% \]
+%% Actually, I started with the first version, but at some point I
+%% changed for the second one because it made some stuff easier (although
+%% I don't remember exactly what it was).
 
-I think this should be doable with only UIP (or a $\Prop$-valued
-equality type?) and not funext. This would be important for a
-computational justification of IITs.
+%% I think this should be doable with only UIP (or a $\Prop$-valued
+%% equality type?) and not funext. This would be important for a
+%% computational justification of IITs.
 
-Unfortunately, it seems that funext is needed (see the use of funext in
-Syntax.agda) for the metatheoretic/infinitary
-parameters part (4-5), for example in the case of the metatheoretic
-$ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
-investigated more carefully.
+%% Unfortunately, it seems that funext is needed (see the use of funext in
+%% Syntax.agda) for the metatheoretic/infinitary
+%% parameters part (4-5), for example in the case of the metatheoretic
+%% $ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
+%% investigated more carefully.
 
-\paragraph*{Construction}
 \input{universaliit}

### 2019-11-11 paper/universaliit.tex
stuff

@@
 The construction consists of the following steps:
+
 \begin{enumerate}
 \item Definition of untyped syntax (as a family of inductive datatypes) together
-  with typing judgments (as inductive relations on the untyped
-  syntax), and construction of a model of the theory of IIT
-  signatures from well-formed terms, denoted $S$.
+  with typing judgments (as inductive relations on the untyped syntax), and
+  construction of a model of the theory of IIT signatures from well-formed
+  terms, denoted $S$.
 \item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
   \begin{enumerate}
-  \item defining a relation $\blank \sim \blank$ between the (well-typed) syntax and a given
-    model. The idea is that given a syntactic context $\Gamma$ and a semantic
-    context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
-    \model{\Gamma}$ if and only if $\rec\,\,\Gamma = \model{\Gamma}$, and
-    similarly for types, terms, and substitutions;
+  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
+    and a given model. The idea is that given a syntactic context $\Gamma$ and a
+    semantic context $\model{\Gamma}$ of the model $M$, we have
+    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
+    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
     % semantic context $\model{\Gamma}$ of
@@
 
-None of the constructions in this section use UIP. Functional extensionality is
-necessary because the untyped metatheoretic $\Pi$ takes a metatheoretic function
-as an argument. An example induction step that uses it is in the type
-preservation proof for identity substitutions (this is an equation satisfied by
-a model of the QIIT), in particular in the case $(\Pim\,T\,
-A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is equal to
-$\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
+Up until the construction of the syntactic model, UIP is not used. Function
+extensionality on the other hand is necessary because the untyped metatheoretic
+$\Pi$ takes a metatheoretic function as an argument. An example induction step
+that uses it is in the type preservation proof for identity substitutions (this
+is an equation satisfied by a model of the QIIT), in particular in the case
+$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
+equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
 hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.
@@
 \[
-  \mathsf{PartialMap}\,A\,B := A\ra (P : \mathsf{hProp})\times(P \ra B)
+  \mathsf{PartialMap}\,A\,B := A\ra (P : \Set)\times\isaprop\,P\times (P \ra B)
 \]
-Here $\mathsf{hProp}$ denotes a type which is propositional. In the above
-definition, we notice an opportunity for converting a fibered definition of a
-type family into an indexed one; if we drop the propositionality for $P$ for the
-time being, we may equivalently return a family indexed over $B$, which is
-exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we recover
-uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
-relation, and totality by already assuming well-formedness of $A$. In type
-theory, using indexed families instead of display maps is a common convenience,
-since the former are natively supported in type theory, while the latter require
-carrying around auxiliary propositional equalities.
+In the above definition, we notice an opportunity for converting a fibered
+definition of a type family into an indexed one; if we drop the propositionality
+for $P$ for the time being, we may equivalently return a family indexed over
+$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
+recover uniqueness of $P$ through the functionality requirement on the $A\ra
+B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
+type theory, using indexed families instead of display maps is a common
+convenience, since the former are natively supported in type theory, while the
+latter require carrying around auxiliary propositional equalities.
 
@@
   \\
-  & \rlap{(3) Parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\[0.5em]
   & \Piw \wf{\Gamma} \wf{a} \wf{B} \sim \model{C} &&
@@
   % relations are required to show that the relation is right unique.
-Next, we prove that this relation is right unique.
-Then, we show that the relation is stable under weakening and substitution.
-  The last step consists of giving a related semantic counterpart to any
-  well-typed context, type or term.
-  Everything is done by induction on the typing judgments.
+Next, we prove that this relation is right unique.  Then, we show that the
+relation is stable under weakening and substitution.  The last step consists of
+showing left-totality, i.e.\ giving a related semantic counterpart to any
+well-typed context, type or term.  Everything is done by induction on the typing
+judgments.
   \subsubsection{Right uniqueness}
   \label{sec:right_uniqueness}
-  We show by recursion that the relation is right unique in the following sense:
+  We show by induction that the relation is right unique in the following sense:
   \begin{alignat*}{5}
@@
 
-\subsection{The Recursor}
+\subsection{Left-Totality and the Recursor}
 For the recursor, we build a morphism from the syntactic model to the semantic
@@
   The right uniqueness of the relation is used in this induction. It is then
-straightforward to show (without induction) that the first projection of these
-constructions yield a model morphism from the syntax to the model, again using
-right uniqueness.
+straightforward to show (without induction, but using right uniqueness) that the
+first projection of these constructions yield a model morphism from the syntax
+to the model.
+
 \subsection{Uniqueness}

### 2019-11-11 paper/paper.tex
stuff

@@
 
+The current work builds heavily on the work of Kaposi et al.\ on quotient
+inductive-inductive types\cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
+QIIT syntax and semantics by restricting to IITs, and we reuse the term model
+construction of QIITs as well.
+
+Forsberg\cite{forsberg-phd} specifies inductive-inductive types and develops a
+semantics based on dialgebras. He also provides a reduction of IITs to indexed
+inductive types, but only constructs a weak version of elimination.
+
+Hugunin\cite{jasper} constructs several IITs in cubical Agda from inductive
+types. In this setting, the lack of UIP makes constructions significantly more
+involved, and essentially involves coinductive-coinductive well-formedness
+predicates defined as homotopy limits. Hugunin does not consider a generic
+syntax of IITs and only works on specific examples (although the examples vary
+greatly). He also does not construct full recursive-recursive eliminators.
+
+Streicher\cite{streichersemantics} presents an interpretation of the well-formed
+presyntax of a type theory into a categorical model, which is an important
+ingredient in constructing an initial model, although he does not present
+details on the construction of the term model or its initiality proof.
+
+Voevodsky was interested in constructing initial models of type theories
+from presyntaxes. There is a communal project \cite{
+
 Reduction of classes of inductive types: reduce indexed W-types to
@@
 
-Similar techique for other type theories: Streicher
+Similar technique for other type theories: Streicher
 \cite{streichersemantics}. Initiality project, Brunerie \cite{}. These
@@
 \]
-The inductive-inductive Con-Ty example from Section \ref{sec:intro}:
+The inductive-inductive example from Section \ref{sec:intro}:
 \begin{alignat*}{3}
@@
   & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
-    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp U)))
+    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
 \end{alignat*}
@@
 construction). One reason is that the QIIT formalisation uses strict computation
-rules for induction over signatures, while here we show propositional
+rules (given by Agda rewrite rules) for induction over signatures, while here we show propositional
 computation. This mismatch can be in principle remedied by noting that both

### 2019-11-11 paper/paper.tex
small fixes, lot more to come

@@
 The current work builds heavily on the work of Kaposi et al.\ on quotient
-inductive-inductive types\cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
+inductive-inductive types \cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
 QIIT syntax and semantics by restricting to IITs, and we reuse the term model
@@
 
-Forsberg\cite{forsberg-phd} specifies inductive-inductive types and develops a
+Forsberg \cite{forsberg-phd} specifies inductive-inductive types and develops a
 semantics based on dialgebras. He also provides a reduction of IITs to indexed
@@
 
-Hugunin\cite{jasper} constructs several IITs in cubical Agda from inductive
+Hugunin \cite{jasper} constructs several IITs in cubical Agda from inductive
 types. In this setting, the lack of UIP makes constructions significantly more
@@
 
-Streicher\cite{streichersemantics} presents an interpretation of the well-formed
+Streicher \cite{streichersemantics} presents an interpretation of the well-formed
 presyntax of a type theory into a categorical model, which is an important
@@
 Voevodsky was interested in constructing initial models of type theories
-from presyntaxes. There is a communal project \cite{
+from presyntaxes. There is a communal project \cite{}
 
@@
 the previous section, this formalisation also uses equality reflection for some
-equalities, through rewriting pragmas\cite{cockxsprinkles}. This is purely for
+equalities, through rewriting pragmas \cite{cockxsprinkles}. This is purely for
 convenience, and in principle we could eliminate the instances of equality
@@
 The theory of IIT signatures is given as a category with families
-(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
+(CwF) \cite{Dybjer96internaltype} with certain additional type formers. The CwF
 part constitutes the substitution calculus, while the type formers allow
@@
 \end{alignat*}
-\subsubsection{Notation and abbreviations}
+\subsection{Notation and abbreviations}
 
@@
   \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
-  & t \oldapp u := \app\,t\,[\id\,u]
+  & t \oldapp u := (\app\,t)[\id\,u]
 \end{alignat*}
-We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[p])$ as $a\ra B$.
+We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\p])$ as $a\Ra B$.
 
-We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
-$p$ composed with itself $n$ times. However, for informal examples we instead
+We can also recover de Bruijn indices by having $n := \q[\p^n]$, where $\p^n$ is
+$\p$ composed with itself $n$ times. However, for informal examples we instead
 use a nameful notation, where we introduce names in context extension and in the
@@
 
-\subsubsection{Example signatures}
+\subsection{Example signatures}
 
@@
 \[
-\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
+\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
 \]
@@
 \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
-      cons : \Pim\,A\,(\lambda \_.\, L\ra L)
+      cons : \Pim\,A\,(\lambda \_.\, L\Ra L)
 \]
@@
 \begin{alignat*}{3}
-  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \ra Con) \\
+  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\Ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra Con) \\
   & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
@@
 extensional and intensional type
-theories\cite{hofmann95extensional,winterhalter2019eliminating}.
+theories \cite{hofmann95extensional,winterhalter2019eliminating}.
 

### 2019-11-11 paper/universaliit.tex
small fixes, lot more to come

@@
 Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
-\,A[\sigma]\,B[\varu\, 0 \,\consu\, \wkO\,\sigma]$, and thus we need to define
-$\wk0$, the weakening of substitutions.  The basic idea is to increment the de
+\,(A[\sigma])\,(B[\varu\, 0 \,\consu\, \wkO\,\sigma])$, and thus we need to define
+$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
 Bruijn indices of all the variables.  Actually this is not so simple because of
-the $\Piu$ type: indeed, we want to define $\wk(\Piu\,A\,B)$ as the $\Pi$ type
+the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
 of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
@@
   \end{alignat*}
+$\length{\Gamma}$ denotes the length of a the context $\Gamma$.
+
 \subsubsection{Untyped substitution}
@@
   & \keep && : \Subu \ra \Subu \\
-   &&& := \lambda \sigma. \varu\, 0 \,\consu\, \wkO\,\sigma
+   &&& := \lambda \sigma. (\varu\, 0 \,\consu\, \wkO\,\sigma)
   \end{alignat*}
   The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
-  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma, A[\sigma]$
-  and $\Delta,A$ for any type $A$. This occurs when defining
-  $(\Piu\,A\,B)[\sigma]$ as $\Piu (A[\sigma])(B[\keep \,\sigma])$.
-
-  We can define the identity substitution on a context $\Gamma$ as follows,
-  where $\length{\Gamma}$ is the length of the context $\Gamma$, and
-  $\keep^{\length{\Gamma}}$ is $\keep$ iterated $\length{\Gamma}$ times:
-\begin{alignat*}{5}
-  & \length{\Gamma} && : \Conu\ra\N
-  \\
+  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
+  and $\Delta\ext A$ for any type $A$. This occurs when defining
+  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.
+
+  We can define the identity substitution on a context $\Gamma$ as
+  follows, where $\keep^{\length{\Gamma}}$ is $\keep$ iterated
+  $\length{\Gamma}$ times:
+\begin{alignat*}{5}
   & \idu && : \Conu \ra \Subu \\
@@
   & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
-  & \rlap{(3) Inudctive parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\[0.5em]
   & \Piw && :

### 2019-11-11 paper/universaliit.tex
more small fixes, now looking at related work

@@
   construction of a model of the theory of IIT signatures from well-formed
-  terms, denoted $S$.
-\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
+  terms, denoted $\mathsf{S}$.
+\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
   \begin{enumerate}
@@
   \item Proving the uniqueness of this morphism by showing that any morphism
-    $f:S\to M$ satisfies the relation. For example, for any syntactic context
+    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
     $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
@@
 error (ill-scoped substitution). The typing judgments will not mention $\erru$.
-\todo{András: prettify below maybe}
 \begin{alignat*}{5}
@@
 Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
-\,(A[\sigma])\,(B[\varu\, 0 \,\consu\, \wkO\,\sigma])$, and thus we need to define
+\,(A[\sigma])\,(B[\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma])$, and thus we need to define
 $\wkO$, the weakening of substitutions.  The basic idea is to increment the de
@@
   & \keep && : \Subu \ra \Subu \\
-   &&& := \lambda \sigma. (\varu\, 0 \,\consu\, \wkO\,\sigma)
+   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
   \end{alignat*}
@@
   % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
-  & \wk_n[n] && : (\wk_{n} Z)[ n := z] = Z \\
+  & \wk_n[n] && : (\wk_{n}\,Z)[ n := z] = Z \\
   % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
-  & \wk_+[] && : (\wk_{n+p+1} Z)[ n := \wk_p u] = \wk_{n+p}(Z[n := u]) \\
+  & \wk_+[] && : (\wk_{n+p+1}\,Z)[ n := \wk_p\,u] = \wk_{n+p}\,(Z[n := u]) \\
   % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
-  & \wk[+] && : (\wk_{n} Z)[ n+p+1 :=  u] = \wk_{n}(Z[n+p := u]) \\
+  & \wk[+] && : (\wk_{n}\,Z)[ n+p+1 :=  u] = \wk_{n}\,(Z[n+p := u]) \\
   % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
@@
   % liftV=wkS
-  & [\keep^n\mhyphen \wkO] && : Z[\keep^n (\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
+  & [\keep^n\mhyphen \wkO] && : Z[\keep^n\,(\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
   \\
@@
   :
-  (\wk_n Z) [\keep^n (u \consu \sigma)] = Z [ \keep^n \sigma]
+  (\wk_n\,Z) [\keep^n\, (u \consu \sigma)] = Z [ \keep^n \sigma]
   \\
@@
     % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
-   & [:=][\keep] && : Z [ n:= u][\keep^n \sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
+   & [:=][\keep] && : Z [ n:= u][\keep^n\,\sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
   \end{alignat*}
@@
   & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
+  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
   & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
-  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
   & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
@@
   & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
-  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
+  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash \mathsf{0} \invar \wku \, A \\
   & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
-  \vdash S\,n \invar \wku \, A \\
+  \vdash \mathsf{S}\,n \invar \wku \, A \\
   & \rlap{(2) Universe} \\[0.5em]
@@
 We prove by mutual induction that typing judgments are stable under weakening,
-for contexts, types, terms, and substitutions:
+for contexts, types, terms, and substitutions.
 % wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
@@
 \subsubsection{Substitution preserves typing}
-  First, we show that judgments are stable under substitution.
+  We show that judgments are stable under substitution.
 \begin{alignat*}{5}
@@
 \subsubsection{Laws for identity substitutions}
-We show category and functor laws involving identity substitution, for
+We show category and functor laws involving identity substitution for
 well-formed types, terms and substitutions.
@@
 following sense:
-
-\todo{Prettify table.}
-\begin{align*}
-    \isp{\Conw}
-  : \isaprop\, (\Gamma\vdash)
- & &
-    \isp{\Tmw}
-  : \isaprop\, (\Gamma\vdash t \in A)
-    \\
-    \isp{\Tyw}
-  : \isaprop\, (\Gamma\vdash A)
-    &&
-    \isp{\wf\Var}
-  : \isaprop\, (\Gamma\vdash x \invar A)
-  \\
-  &
-        \isp{\wf\Sub}
-  : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
-  \end{align*}
+\begin{alignat*}{5}
+ & \isp{\Conw} && : \isaprop\, (\Gamma\vdash) \\
+ & \isp{\Tyw} && : \isaprop\, (\Gamma\vdash A) \\
+ & \isp{\wf\Var} && : \isaprop\, (\Gamma\vdash x \invar A) \\
+ & \isp{\Tmw} && : \isaprop\, (\Gamma\vdash t \in A) \\
+ & \isp{\wf\Sub} && : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
+\end{alignat*}
 We also show unicity of typing:
@@
 Let us consider for instance the application constructor $\appw$: for a codomain
-type $B$ it yields an overall type $C=B[0 := u]$ for an application. Even
-if $C$ is known a priori, there may be another $B$ for which $B[0 := u] = C$,
+type $B$ it yields an overall type $C=B[\mathsf{0} := u]$ for an application. Even
+if $C$ is known a priori, there may be another $B$ for which $B[\mathsf{0} := u] = C$,
 possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
@@
 $\Pi$ takes a metatheoretic function as an argument. An example induction step
-that uses it is in the type preservation proof for identity substitutions (this
+that uses function extensionality is in the type preservation proof for identity substitutions (this
 is an equation satisfied by a model of the QIIT), in particular in the case
@@
 This approach is an alternative version of Streicher's method for interpreting
-preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
+preterms in an arbitrary model \cite{streichersemantics}. Streicher first defines
 a family of partial maps from the presyntax to a model, then shows that the maps
@@
 \[
-  \mathsf{PartialMap}\,A\,B := A\ra (P : \Set)\times\isaprop\,P\times (P \ra B)
+  \mathsf{PartialMap}\,A\,B := A\ra \big((P : \Set)\times\isaprop\,P\times (P \ra B)\big)
 \]
@@
 type theory, using indexed families instead of display maps is a common
-convenience, since the former are natively supported in type theory, while the
+convenience, since the former are natively supported, while the
 latter require carrying around auxiliary propositional equalities.
@@
 Given a model $M$ of the QIIT, we define the functional relation satisfied by
-the initial morphism $\rec : S\to M$ by recursion on the typing judgments.
+the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
 % As these judgments are propositions, we will sometimes
@@
 \[
-  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
+  (\Gamma : \syn\Con)\ra (A : \syn\Ty\,\syn\Gamma) \ra
   (\model\Gamma : \model\Con)
@@
   \ra
-  (\model{A}:\model\Ty\model\Gamma)
+  (\model{A}:\model\Ty\,\model\Gamma)
   \ra
@@
 % We leave this possibility aside,
-Our metatheory does not support this kind of recursive-recursive
-definitions\footnote{And to our knowledge there is no established semantics for
-it in existing literature.}, so we instead just remove the hypothesis
+We want to avoid using such recursive-recursive
+definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
 $\Gamma\sim\model\Gamma$ from the list of arguments.  We proceed similarly for
@@
 this is not possible because typechecking requires that $\model\Delta$ is the
-empty semantic context. This is precisely what ensured the removed hypothesis
-$\syn\cdot \sim \model\Delta$.  Our way out here is to state that $\model\sigma$
+empty semantic context. This is precisely what was ensured by the hypothesis
+$\syn\cdot \sim \model\Delta$ we removed.  Our way out here is to state that $\model\sigma$
 is related to the empty substitution if the target semantic context $\model\Delta$
@@
 %The first step consists in showing that these types are propositions.
-In contrast to this, our approach
+In comparison, our approach
 % Actually, we take a different approach for the relation on types
@@
 to the definition
-  $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
+  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
   instead of
   our definition
-  $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
+  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
 % Recall that our adapted goal, for types, is to prove that
@@
 
-Here we provide the definition of the relation by recursion on the typing judgments.
-In the definitions, we abbreviate $\relT{\wf{A}}{\model\Gamma}{\model{A}}$ by
+We define the relation by recursion on the typing judgments.
+In the following, we abbreviate $\relT{\wf{A}}{\model\Gamma}{\model{A}}$ by
 $\wf{A}\sim\model{A}$ when $\model{\Gamma}$ can be inferred, and similarly for
@@
     &
-    \relS{(\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})}{\model{\Gamma}}{\model{E}}{\model{\delta}}
+    \relS{(\consw\wf{\Delta}\wf{\sigma}\wf{A}\wf{t})}{\model{\Gamma}}{\model{E}}{\model{\delta}}
     % (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
@@
      (e_C : \model{C} = e_\Delta \# \model{\wk}\,\model{A}) \times
-     % \\
-     % & && \qquad
+     \\
+     & && \qquad
      (\model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n})
@@
   & \rlap{(2) Universe} \\[0.5em]
-  & \Uw \wf{\Gamma} \wf{A} \sim \model{A} &&
+  & \Uw\, \wf{\Gamma} \wf{A} \sim \model{A} &&
    := \model{A} = \model{\U}
   \\
-  & \Elw \wf{\Gamma} \wf{a} \sim \model{A}
+  & \Elw\, \wf{\Gamma} \wf{a} \sim \model{A}
   && :=
@@
   & \rlap{(3) Inductive parameters} \\[0.5em]
-  & \Piw \wf{\Gamma} \wf{a} \wf{B} \sim \model{C} &&
+  & \Piw\,\wf{\Gamma} \wf{a} \wf{B} \sim \model{C} &&
    :=
@@
   &
-  \relt{\appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u}}{\model{\Gamma}}{\model{C}}{\model{x}}  &&
+  \relt{\appw\, \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u}}{\model{\Gamma}}{\model{C}}{\model{x}}  &&
   % & \appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
@@
 showing left-totality, i.e.\ giving a related semantic counterpart to any
-well-typed context, type or term.  Everything is done by induction on the typing
+well-typed context, type or term.  Everything is proved by induction on the typing
 judgments.
@@
     \isp{\Sigma{\sim}}
-    && : (\wf\Gamma : \Gamma \vdash ) \ra
+    && : (\wf\Gamma : \Gamma \vdash ) && \ra
     \isaprop\, (\sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf A : \Gamma \vdash A) \ra
+    && : (\wf A : \Gamma \vdash A) && \ra
     \isaprop\, (\sum_{\model{A}} \wf{A} \sim \model{A})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf t : \Gamma \vdash t \in A) \ra
+    && : (\wf t : \Gamma \vdash t \in A) && \ra
     \isaprop\, (\sum_{\model{t}} \wf{t} \sim \model{t})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf x : \Gamma \vdash x \invar A) \ra
+    && : (\wf x : \Gamma \vdash x \invar A) && \ra
     \isaprop\, (\sum_{\model{x}} \wf{x} \sim \model{x})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf \sigma : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
+    && : (\wf \sigma : \Gamma \vdash \sigma \Rightarrow \Delta) && \ra
     \isaprop\, (\sum_{\model{\sigma}} \wf{\sigma} \sim \model{\sigma})
@@
   obvious. For example, we were not able to define
-  $\Elw\wf{\Gamma}\wf{a}\sim \model{A}$ in this fashion. In related work,
+  $\Elw\,\wf{\Gamma}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
   Hugunin investigated constructing IITs without
-  UIP\cite{hugunin2019constructing} in cubical type theory, and demonstrated
+  UIP \cite{hugunin2019constructing} in cubical type theory, and demonstrated
   that well-formedness predicates used in syntactic models can subtly break
@@
   Indeed, in the proof of stability under substitution, the $\Pi$ case
-  requires to show that $\Pi \,A[\sigma] \,B[\keep\,\sigma]$ is related to
-  $\model{\Pi}\,\model{A}\model{[\sigma]}
-  \model{A}\model{[\model{\keep}\,\sigma]}$.
+  requires to show that $\Pi \,(A[\sigma]) \,(B[\keep\,\sigma])$ is related to
+  $\model{\Pi}\,(\model{A}\model{[\sigma]})\,
+  (\model{B}\model{[\model{\keep}\,\sigma]})$.
   We would like to apply the induction hypothesis, so we need to show that
-  $\keep\,\sigma=\varu\,0\consu\wkO\,\sigma$ is related to
+  $\keep\,\sigma=\varu\,\mathsf{0}\consu\wkO\,\sigma$ is related to
   $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
   related to $\model{\sigma}$.
-  As $\keep\,\sigma=\varu 0\consu \wkO\,\sigma$, we are left with showing that
-  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO \id$)
+  As $\keep\,\sigma=\varu\, \mathsf{0}\consu \wkO\,\sigma$, we are left with showing that
+  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO\,\id$)
   relates to its semantic counterpart.
@@
 
-  To do that, we show that $\wkO$ preserves the relation, for types and terms.
+  To achieve that, we show that $\wkO$ preserves the relation, for types and terms.
   This requires to generalise a bit and show that $\wk_n$ preserves the relation,
-  as $\wk_0(\Pi\,A\,B)=\Pi\,(\wk_0\,A)(\wk_1\,B)$.
+  as $\wk_0\,(\Pi\,A\,B)=\Pi\,(\wk_0\,A)\,(\wk_1\,B)$.
   But remember that $\wk_n$ performs a weakening in the middle of a context, so
@@
   & && \qquad
-   \wf{{\circ}}\wf{\Gamma}\wf{\delta}\wf{\sigma} \sim \model{\delta}\model{\circ}\model{\sigma}
+   \wf{{\circ}}\,\wf{\Gamma}\,\wf{\delta}\,\wf{\sigma} \sim \model{\delta}\mathbin{\model{\circ}}\model{\sigma}
   \end{alignat*}
@@
   (\wf{\Gamma} \sim \model{\Gamma}) \ra
-   \wf{{\id}}\wf{\Gamma} \sim {\id}_{\model{\Gamma}}
+   \wf{{\id}}\,\wf{\Gamma} \sim {\id}_{\model{\Gamma}}
   \end{alignat*}
@@
 
-More formally, we assume an arbitrary morphism from the
+More formally, we assume an arbitrary morphism $f$ from the
 syntax to the model, inducing the following maps:

### 2019-11-12 paper/paper.tex
more small fixes, now looking at related work
finished related work

@@
 \begin{alignat*}{5}
-  & Con^M && : \Con\ra\Set \\
-  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
-  & empty^M && : Con^M\,\mathsf{empty} \\
-  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
-  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
-  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
-%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
+  & Con^D && : \Con\ra\Set \\
+  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & empty^D && : Con^D\,\mathsf{empty} \\
+  & ext^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
+  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
+  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,(ext^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
+%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:Ty^D\,\Gamma^D\,A)\ra Ty^D\,(ext^D\,\Gamma^D\,A^D)\ra\,Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
-  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
+  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
-  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
-  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
-  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon}\,\mathsf{empty} && = empty^D \\
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
 \end{alignat*}
@@
 reducible to the same feature set.
+
+The contents of this paper were presented at the TYPES 2019 conference
+in Oslo \cite{types}.
 
@@
 ingredient in constructing an initial model, although he does not present
-details on the construction of the term model or its initiality proof.
+details on the construction of the term model or its initiality proof. Our paper an indexed variant of his construction (see Subsection \ref{sec:streicher} for a comparison).
 
-Voevodsky was interested in constructing initial models of type theories
-from presyntaxes. There is a communal project \cite{}
+Voevodsky was interested in constructing initial models of type
+theories from presyntaxes. Inspired by this, Brunerie \cite{brunerie}
+formalised Streicher's proof in Agda for a type theory with $\Pi$,
+$\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
+universes. He used UIP, function extensionality and quotient types in
+his formalisation. Our proof uses a similar approach, but we construct
+a much simpler type theory which does not have any computation rules,
+hence we can avoid using quotients.
 
-Reduction of classes of inductive types: reduce indexed W-types to
-W-types. IR types to indexed types. IITs were introduced by
-Forsberg. What he did.  QIITs by Forsberg-lots of people. Our QIIT
-paper. Jakob TYPES 2019 abstract.  Jasper Huguin: he didn't do
-recursion-recursion. Without UIP.
+Intrinsic (well-typed) syntax for type theories were constructed
+before using IITs \cite{chapman09eatitself}, inductive-recursive types
+\cite{nisse,Altenkirch:2014:CO:2631172.2631176} and QIITs
+\cite{ttintt}. In this paper we avoid using such general classes of
+inductive types, as our goal is to reduce IITs to indexed inductive
+types.
 
-(Indexed inductive types can be reduced to inductive types using a method
-similar to what we use here, preterms and well-formedness predicate.)
+Reducing general classes of inductive types to simpler classes has a
+long tradition in type theory. Indexed W-types were reduced to W-types
+\cite{indexedcont} (using the essentially Streicher's idea of preterms
+and a typing predicate), small inductive-recursive types to indexed
+W-types \cite{malatasta13smallir}, mutual inductive types to indexed
+W-types \cite{mutual}. Quotient inductive-inductive types (and
+inductive-inductive types) can be reduced to quotient inductive types
+using the reduction of generalised algebraic theories to essentially
+algebraic theories \cite{gat}.
 
-Talk about connection to setoidified CwFs (I think these are explained
-in some Dybjer paper) and maybe the initiality conjecture.
-
-Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
-inductive-inductive types (HIITs)
-\cite{kaposi_et_al:LIPIcs:2018:9190}.
-
-Similar technique for other type theories: Streicher
-\cite{streichersemantics}. Initiality project, Brunerie \cite{}. These
-all use quotients. Closest: Lafont TYPES abstract 2017. Does not use
-quotients.
-
-Constructing intrinsic syntax but different techniques: Chapman,
-Danielsson, Nuo. These use induction-induction which is our goal.
+Our reduction of IITs to indexed inductive types goes through two
+steps: first we construct a concrete QIIT using inductive types, then
+we construct all IITs from this particular QIIT. A more direct
+approach is proposed by \cite{erasure}: here the initial algebra would
+be constructed directly for any IIT signature without going through an
+intermediate step.
 
@@
   \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.
+  \item Natural numbers $\mathbb{N}$ with constructors $\mathsf{0}:\mathbb{N}$ and $\mathsf{S}:\mathbb{N}\ra\mathbb{N}$.
 
@@
 \[
-  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
+  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N\big),
 \]

### 2019-11-12 paper/paper.tex
more related work

@@
 semantics based on dialgebras. He also provides a reduction of IITs to indexed
-inductive types, but only constructs a weak version of elimination.
+inductive types, but only constructs a weak version of the eliminator.
 
@@
 ingredient in constructing an initial model, although he does not present
-details on the construction of the term model or its initiality proof. Our paper an indexed variant of his construction (see Subsection \ref{sec:streicher} for a comparison).
+details on the construction of the term model or its initiality proof. Our initiality proof can be seen as an indexed variant of his construction (see Subsection \ref{sec:streicher} for a comparison).
 
@@
 universes. He used UIP, function extensionality and quotient types in
-his formalisation. Our proof uses a similar approach, but we construct
-a much simpler type theory which does not have any computation rules,
-hence we can avoid using quotients.
+his formalisation. In this paper we construct a type theory without
+computation rules, hence we avoid using quotients.
 
-Intrinsic (well-typed) syntax for type theories were constructed
-before using IITs \cite{chapman09eatitself}, inductive-recursive types
+Intrinsic (well-typed) syntax for type theories were constructed using
+IITs \cite{chapman09eatitself}, inductive-recursive types
 \cite{nisse,Altenkirch:2014:CO:2631172.2631176} and QIITs
 \cite{ttintt}. In this paper we avoid using such general classes of
-inductive types, as our goal is to reduce IITs to indexed inductive
+inductive types as our goal is to reduce IITs to indexed inductive
 types.
@@
 W-types \cite{malatasta13smallir}, mutual inductive types to indexed
-W-types \cite{mutual}. Quotient inductive-inductive types (and
+W-types \cite{mutual}, W-types to natural numbers and quotients
+\cite{Ahrens2019}. Quotient inductive-inductive types (and
 inductive-inductive types) can be reduced to quotient inductive types

### 2019-11-12 paper/paper.tex
Some version of Coq supports IITs

@@
 (even from non-IITs) and it is also currently the only proof assistant
-supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
+supporting IITs\footnote{An experimental version of Coq with IITs is also
+  available on github.}. The elimination principle for $\Con$--$\Ty$ says that
 given the following components,

### 2019-11-12 paper/paper.tex
mathsf font for _^D constructors in the intro

@@
 \begin{alignat*}{5}
-  & Con^D && : \Con\ra\Set \\
-  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
-  & empty^D && : Con^D\,\mathsf{empty} \\
-  & ext^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
-  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
-  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,(ext^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
-%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:Ty^D\,\Gamma^D\,A)\ra Ty^D\,(ext^D\,\Gamma^D\,A^D)\ra\,Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
+  & \Con^D && : \Con\ra\Set \\
+  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
+  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
+  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
+  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
+%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
-  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon} && : (\Gamma:\Con)\ra \Con^D\,\Gamma \\
+  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra \Ty^D\,(\mathsf{elimCon}\,\Gamma)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon}\,\mathsf{empty} && = empty^D \\
-  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
-  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
-  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = \U^D\,(\mathsf{elimCon}\,\Gamma) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = \El^D\,(\mathsf{elimCon}\,\Gamma)
 \end{alignat*}

### 2019-11-12 paper/paper.tex
Type theoretic application rather than the categorical one

--- 2019-11-12 paper/paper.tex
@@
   & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
-  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
+  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
+  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
   & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
-  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
+  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\[0.5em]
   & \rlap{(4) Metatheoretic parameters} \\[0.5em]
@@
 
-We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
-for inductive parameters, we have $\app$ as a categorical version of function
-application. We recover the usual type-theoretic application as follows:
-\begin{alignat*}{5}
-  & \blank\oldapp\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
-  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
-  & t \oldapp u := (\app\,t)[\id\,u]
-\end{alignat*}
+We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$.
 We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\p])$ as $a\Ra B$.

### 2019-11-12 paper/paper.tex
π₁ and  π₂ rather than p and q in the definition of models

@@
   & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
-  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
-  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
-  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
-  & {\ext\beta_2} && : \q[\sigma, t] = t \\
-  & {\ext\eta} && : (\p, \q) = \id \\
+  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
+  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
+  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
+  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
+  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
   & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]

### 2019-11-12 paper/paper.tex
update notations and abbreviations

@@
 
-We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$.
-We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\p])$ as $a\Ra B$.
-
-We can also recover de Bruijn indices by having $n := \q[\p^n]$, where $\p^n$ is
-$\p$ composed with itself $n$ times. However, for informal examples we instead
+We define $\wk : \Sub\,(\Gamma\ext A)\ra \Gamma$ as $\pi_1\,\id$.
+We can recover de Bruijn indices by having $0= \pi_2 \,\id$ and $n+1 = n[\wk]$.
+However, for informal examples we instead
 use a nameful notation, where we introduce names in context extension and in the
@@
 \El\,A$ in a signature for pointed sets.
+
+We define $\sigma^\uparrow$ as $(\sigma\circ\wk,0)$.
+We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\wk])$ as $a\Ra B$.
 

### 2019-11-12 paper/paper.tex

@@
 (even from non-IITs) and it is also currently the only proof assistant
-supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
+supporting IITs\footnote{An experimental version of Coq with IITs is also
+  available on github.}. The elimination principle for $\Con$--$\Ty$ says that
 given the following components,
 \begin{alignat*}{5}
-  & Con^D && : \Con\ra\Set \\
-  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
-  & empty^D && : Con^D\,\mathsf{empty} \\
-  & ext^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
-  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
-  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,(ext^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
-%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:Ty^D\,\Gamma^D\,A)\ra Ty^D\,(ext^D\,\Gamma^D\,A^D)\ra\,Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
+  & \Con^D && : \Con\ra\Set \\
+  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
+  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
+  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
+  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
+%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
 \end{alignat*}

### 2019-11-12 paper/universaliit.tex
finished related work
small typo in sec 3.
grammar and typos

@@
 Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
-\,(A[\sigma])\,(B[\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma])$, and thus we need to define
+\,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$, and thus we need to define
 $\wkO$, the weakening of substitutions.  The basic idea is to increment the de
-Bruijn indices of all the variables.  Actually this is not so simple because of
-the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
+Bruijn indices of all the variables.  Actually, this is not so simple because of
+the $\Piu$ type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
 of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
 the second last variable of the context, rather than the last one.  For this
-reason, we need to generalize the weakening as occuring anywhere in the context.
+reason, we need to generalise the weakening as occuring anywhere in the context.
 \begin{alignat*}{5}
@@
 Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
-typing. Stating a typing rule for this operation requires to have a way to
-express the weakening occuring at the middle of a context. We consider pairs of
-untyped contexts, which should be thought of as a splitting of a context at some
-position. The full context is recovered by merging the two components:
+typing. Stating a typing rule for this operation requires weakening at the
+middle of a context. We consider pairs of untyped contexts, which should be
+thought of as a splitting of a context at some position. The full context is
+recovered by merging the two components:
 % (which appears in the typing rule of a weakening occuring in the middle of a context),
@@
 \subsection{Relating the Syntax to a Model}
+\label{sec:streicher}
 It remains to show that the constructed syntactic model is initial. To achieve
@@
 straightforward to show (without induction, but using right uniqueness) that the
-first projection of these constructions yield a model morphism from the syntax
+first projection of these constructions yields a model morphism from the syntax
 to the model.

### 2019-11-12 paper/paper.tex
grammar and typos
Brunerie et al

@@
 Voevodsky was interested in constructing initial models of type
-theories from presyntaxes. Inspired by this, Brunerie \cite{brunerie}
+theories from presyntaxes. Inspired by this, Brunerie et al.\ \cite{brunerie}
 formalised Streicher's proof in Agda for a type theory with $\Pi$,
@@
 \[
-  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N\big),
+  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N\big),
 \]

### 2020-04-08 paper/paper.tex
some fixes suggested by Ambrus and the reviewers

@@
 
-\title{For Induction-Induction, Induction is Enough} %TODO Please add
+\title{For finitary induction-Induction, Induction is Enough} %TODO Please add
 
@@
 Many mutual inductive types can be reduced to indexed inductive types, where the
-index disambiguates different sorts. An example is the predicates
+index disambiguates different sorts. An example is the predicates (or \emph{indexed sorts})
 $\mathsf{isEven}$ and $\mathsf{isOdd}$ defined mutually by the following
@@
   & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
-  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,n \\
-  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,n
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(n+1) \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(n+1)
 \end{alignat*}
@@
   & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
-  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,n \\
-  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,n
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(n+1) \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(n+1)
 \end{alignat*}
@@
 $\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
-hence the two sorts have to be defined at the same time. Moreover, $\Ty$ is
+hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
 indexed over $\Con$. This precludes a reduction analogous to the reduction of
@@
 given the following components,
+where the variable $\Gamma$ is implicitly universally quantified when required
+(we indeed sometimes omit universal quantification in the following, for readability),
 \begin{alignat*}{5}
@@
 of now, there is no general metatheory for (large, infinitary) QIITs from which
-this notion would fall out. Hence, we give an ad-hoc definition for
+this notion would be deduced. Hence, we give an ad-hoc definition for
 homomorphisms, which can be summarized as a functor between underlying
@@
 \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
-      cons : \Pim\,A\,(\lambda \_.\, L\Ra L)
+      cons : \Pim\,A\,(\lambda \_.\, L\Ra \El\,L)
 \]

### 2020-04-08 paper/universaliit.tex
some fixes suggested by Ambrus and the reviewers

@@
 error (ill-scoped substitution). The typing judgments will not mention $\erru$.
+The main interest of $\erru$ is that it behaves like a closed term (which the
+theory of signatures lacks), in the sense that it is
+invariant under substitution. This makes expected equalities about 
+substitution (see Section~\ref{ss:exch-laws}) true even in the ill-typed case, thus reducing the amount of hypotheses.
 \begin{alignat*}{5}
@@
   \subsubsection{Exchange laws: weakening and substitution}
+  \label{ss:exch-laws}
   Many lemmas for types and terms are shown by induction on the untyped syntax.
@@
 are total on well-formed input. We have found that our approach is significantly
-easier to formalise. Too see why, note that the right notion of partial map in
+easier to formalise. To see why, note that the right notion of partial map in
 type theory, which does not presume decidable definedness, is fairly
@@
   Hugunin investigated constructing IITs without
-  UIP \cite{hugunin2019constructing} in cubical type theory, and demonstrated
+  UIP \cite{jasper} in cubical type theory, and demonstrated
   that well-formedness predicates used in syntactic models can subtly break

### 2020-04-09 paper/paper.tex
other fixes

@@
   where e.g.\ the sort of types is indexed by contexts. It follows from previous
-  work that all finitary IITs can be constructed from a quotient
+  work by T. Altenkirch, A. Kaposi, and A. Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
   inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
@@
 Many mutual inductive types can be reduced to indexed inductive types, where the
-index disambiguates different sorts. An example is the predicates (or \emph{indexed sorts})
-$\mathsf{isEven}$ and $\mathsf{isOdd}$ defined mutually by the following
+index disambiguates different sorts. An example is the mutual inductive datatype
+with two sorts
+$\mathsf{isEven}$ and $\mathsf{isOdd}$, defined by the following
 constructors.
@@
 supporting IITs\footnote{An experimental version of Coq with IITs is also
-  available on github.}. The elimination principle for $\Con$--$\Ty$ says that
-given the following components,
-where the variable $\Gamma$ is implicitly universally quantified when required
-(we indeed sometimes omit universal quantification in the following, for readability),
+  available on github.}.
+
+We now state  
+the elimination principle for $\Con$--$\Ty$.
+We implicitly universally quantify over the variables $\Gamma$ of type $\Con$
+and $A$ of type $\Ty$ when required (later in the development, we may similarly omit universal quantification).
+Given the following components,
 \begin{alignat*}{5}
@@
 that if this theory of signatures exists, then any finitary QIIT can be derived
-from it via a term model construction. If we omit the type former for equality
+from it via a term model construction.
+Here, by finitary QIIT, we mean that infinitary constructors, such as the following
+one, are not allowed:
+\[ TODO \]
+If we omit the type former for equality
 constructors from the theory of signatures, we get a theory of finitary IIT
@@
 that all finitary QIITs are derivable from a particular fairly complicated QIIT.
-Likewise, restriced to finitary IITs, we only know that they are constructible
+Likewise, restricted to finitary IITs, we only know that they are constructible
 from a particular QIIT.

### 2020-04-09 paper/universaliit.tex
other fixes

@@
 invariant under substitution. This makes expected equalities about 
-substitution (see Section~\ref{ss:exch-laws}) true even in the ill-typed case, thus reducing the amount of hypotheses.
+substitution true even in the ill-typed case,
+thus reducing the number of hypotheses for the corresponding lemmas
+(see Section~\ref{ss:exch-laws}).
 \begin{alignat*}{5}

### 2020-04-09 paper/paper.tex
cite paper about large inf qiit for morphism between models

@@
 
-Note that initiality refers to a notion of homomorphism between models. As
-of now, there is no general metatheory for (large, infinitary) QIITs from which
-this notion would be deduced. Hence, we give an ad-hoc definition for
-homomorphisms, which can be summarized as a functor between underlying
+Here, initiality refers to the notion of homomorphism between models deduced
+from the general theory of large infinitary QIITs developed in \cite{large_inf_qiit}.
+Roughly, a homomorphism is a functor between underlying
 categories which strictly preserves all additional family structure and type

### 2020-04-13 paper/paper.tex
example: infinitary Pi

@@
 Here, by finitary QIIT, we mean that infinitary constructors, such as the following
-one, are not allowed:
-\[ TODO \]
+infinitary $\Pi$, are not allowed:
+\[
+  \Pi_\infty : (\Gamma : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\Gamma) \rightarrow \Ty\,\Gamma
+   \]
 If we omit the type former for equality

### 2020-04-23 paper/paper.tex
some intro

@@
 \presetkeys{todonotes}{inline}{}
+\usepackage{csquotes} %for \llbracket and \rrbracket
+\usepackage{stmaryrd}
 
@@
 
-\title{For finitary induction-Induction, Induction is Enough} %TODO Please add
+\title{For Finitary Induction-Induction, \\ Induction is Enough} %TODO Please add
 
@@
   type theory. They allow the mutual definition of types with multiple sorts
-  where later sorts can be indexed by previous ones. An example is the James
+  where later sorts can be indexed by previous ones. An example is the
   Chapman-style syntax of type theory with conversion relations for each sort
   where e.g.\ the sort of types is indexed by contexts. It follows from previous
-  work by T. Altenkirch, A. Kaposi, and A. Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
+  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
   inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
@@
 Many mutual inductive types can be reduced to indexed inductive types, where the
-index disambiguates different sorts. An example is the mutual inductive datatype
+index disambiguates different sorts. For example, consider the mutual inductive datatype
 with two sorts
@@
   & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
-  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(n+1) \\
-  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(n+1)
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(\mathsf{suc}\,n) \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(\mathsf{suc}\,n)
 \end{alignat*}
-These can be reduced to the following single inductive family.
+This can be reduced to the following single inductive family where
+$\mathsf{isEven?}\,\mathsf{true}$ represents $\mathsf{isEven}$ and
+$\mathsf{isEven?}\,\mathsf{false}$ represent $\mathsf{isOdd}$.
 \begin{alignat*}{5}
@@
   & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
-  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(n+1) \\
-  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(n+1)
+  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(\mathsf{suc}\,n) \\
+  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(\mathsf{suc}\,n)
 \end{alignat*}
@@
 theory itself, and the main example is still the Chapman-style
-\cite{chapman09eatitself} syntax for a type theory. A minimal use case
+\cite{chapman09eatitself} syntax for a type theory. A minimsed version
 is the IIT of contexts and types given by the following constructors.
@@
 
-The elimination principle for the above IIT has two motives and one method
-for each constructor. It is an example of a \emph{recursive-recursive}
-function (following the nomenclature of \cite{forsberg-phd}) which
-means that there are two mutually defined functions, where the type of
-the second function depends on the first function. The proof assistant
-Agda \cite{norell07thesis} allows the definition of such functions
-(even from non-IITs) and it is also currently the only proof assistant
-supporting IITs\footnote{An experimental version of Coq with IITs is also
-  available on github.}.
+The elimination principle for the above IIT has the following two
+motives (one for each sort) and four methods (one for each
+constructor).
+\begin{alignat*}{5}
+  & Con^D && : \Con\ra\Set \\
+  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & {empty}^D && : Con^D\,\mathsf{empty} \\
+  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
+  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
+  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
+%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
+\end{alignat*}
+Above we used implicit quantifications for $\Gamma:\Con$ and
+$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
+parameter $\Gamma$ before its explicit parameter of type
+$Con^D\,\Gamma$.
 
-We now state  
-the elimination principle for $\Con$--$\Ty$.
-We implicitly universally quantify over the variables $\Gamma$ of type $\Con$
-and $A$ of type $\Ty$ when required (later in the development, we may similarly omit universal quantification).
-Given the following components,
+Given the above motives and methods the elimination principle provides
+two functions
 \begin{alignat*}{5}
-  & \Con^D && : \Con\ra\Set \\
-  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
-  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
-  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
-  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
-  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
-%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
-\end{alignat*}
-we obtain two functions
-\begin{alignat*}{5}
-  & \mathsf{elimCon} && : (\Gamma:\Con)\ra \Con^D\,\Gamma \\
-  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra \Ty^D\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
+  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)\,A
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
-  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
-  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = \U^D\,(\mathsf{elimCon}\,\Gamma) \\
-  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = \El^D\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = {ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
 \end{alignat*}
+The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
+of a \emph{recursive-recursive} definition (using nomenclature from
+\cite{forsberg-phd}). This means two mutually defined functions where
+the type of the second function depends on the first function. The
+proof assistant Agda \cite{norell07thesis} allows defining such
+functions (even from non-IITs) and it is currently the only proof
+assistant supporting IITs\footnote{An experimental version of Coq with
+  IITs is also available on GitHub.}.
 
-Reducing IITs to inductive types is an open problem. Forsberg
-\cite{forsberg-phd} showed a reduction which only supports a simpler,
-non-recursive-recursive elimination principle. Hugunin \cite{jasper} reduced
-several IITs to inductive types, working inside a cubical type theory, but he
-also only constructed a weaker eliminator.
+Reducing IITs to inductive types (more precisely, to indexed W-types)
+is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
+extensional type theory however this only provides a simpler,
+non-recursive-recursive elimination principle. Hugunin \cite{jasper}
+reduced several IITs to inductive types, working inside a cubical type
+theory, but he also only constructed the simple eliminator. To
+illustrate the difference, we list the motives and methods and the
+simple elimination principle for the $\Con$--$\Ty$ example. Again, we
+use implicit quantifications.
+\begin{alignat*}{5}
+  & Con^S && : \Con\ra\Set \\
+  & Ty^S && : \Ty\,\Gamma\ra\Set \\
+  & {empty}^S && : Con^S\,\mathsf{empty} \\
+  & {ext}^S && : Con^S\,\Gamma\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\Gamma\,A) \\
+  & U^S && : Con^S\,\Gamma\ra Ty^S\,(\mathsf{U}\,\Gamma) \\
+  & El^S && : Con^S\,\Gamma\ra\,Ty^S\,(\mathsf{El}\,\Gamma) \\
+  & \mathsf{selimCon} && : (\Gamma:\Con)\ra Con^S\,\Gamma \\
+  & \mathsf{selimTy} && : (A:\Ty\,\Gamma)\ra Ty^S\,A
+\end{alignat*}
+This simple elimination principle is too weak to define e.g.\ the
+standard (metacircular) interpretation of our small syntax
+\cite{ttintt}. Using pattern matching notation, this interpretation is
+the following:
+\begin{alignat*}{5}
+  & \ll\blank\rr && : \Con\ra\Set_1 \\
+  & \ll\blank\rr && : \ll\Gamma\rr\ra\Set_1 \\
+  & \ll\mathsf{empty}\rr && := \top \\
+  & \ll\mathsf{ext}\,\Gamma\,A\rr && := (\gamma:\ll\Gamma\rr)\times\ll A\rr\,\gamma \\
+  & \ll\mathsf{U}\,\Gamma\rr\,\gamma && := \Set \\
+  & \ll\mathsf{El}\,\Gamma\rr\,(\gamma,X) && := X
+\end{alignat*}
+The reason that we need the general elimination principle is that
+$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
+this function is recursive-recursive.
 
-The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
-type theory in which every context encodes a valid quotient inductive-inductive
-type (QIIT) signature. For example, the signature for natural numbers is simply
-the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
-and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
-that if this theory of signatures exists, then any finitary QIIT can be derived
-from it via a term model construction.
-Here, by finitary QIIT, we mean that infinitary constructors, such as the following
-infinitary $\Pi$, are not allowed:
+Kaposi, Kov{\'a}cs, and Altenkirch
+\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory
+called the theory of signatures to describe quotient
+inductive-inductive types (QIIT). QIITs are generalisations of IITs
+where equality constructors are also allowed. A QIIT signature is a
+context in the theory of QIIT signatures, for example natural numbers
+are specified by the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of
+length three ($Nat$, $zero$ and $suc$ are variable names). The theory
+of QIIT signatures is itself a QIIT. In ibid., it is proved that if a
+model of extensional type theory supports the theory of QIIT
+signatures, then it supports all QIITs.
+
+By omitting the identity type former from the theory of QIIT
+signatures, we obtain a theory of IIT signatures and the construction
+is still valid. It follows that if a model of extensional type theory
+supports the theory of IIT signatures, it supports all IITs. In this
+paper we show that any model of extensional type theory with indexed
+W-types supports the theory of signatures, hence all IITs.
+
+Just as ibid., we only consider finitary IITs, that is, constructors
+can only have a finite number of recursive arguments. An example
+constructor for $\Con$--$\Ty$ which is not allowed is the following:
 \[
@@
 \]
-If we omit the type former for equality
-constructors from the theory of signatures, we get a theory of finitary IIT
-signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
-works: constructing an arbitrary IIT from the theory of IIT signatures does not
-need the $\mathsf{Id}$ type former to be present.
 
-However, the QIIT term model construction assumes that we already have the
-syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
-that all finitary QIITs are derivable from a particular fairly complicated QIIT.
-Likewise, restricted to finitary IITs, we only know that they are constructible
-from a particular QIIT.
-
-In this paper, we construct the theory of IIT signatures, which is a large
-infinitary QIIT, using only indexed inductive types, uniqueness of identity
-types (UIP) and function extensionality. Thus, we show that all finitary IITs are
-reducible to the same feature set.
+TODO: list of contributions here.
 
@@
 %% funext.
+
+ (later in the development, we may similarly omit universal quantification)
 

### 2020-04-24 paper/paper.tex
stuff

@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{hyperref}
@@
 
-%% \titlerunning{For Induction-Induction, Induction is Enough}%optional, please use if title is longer than one line
+\titlerunning{For Finitary Induction-Induction, Induction is Enough}%optional, please use if title is longer than one line
 
@@
   on function extensionality and uniqueness of identity proofs. We formalised
-  the construction of the theory of IIT signatures in Agda.
+  the construction of the theory of IIT signatures in Agda. TODO: rewrite
 \end{abstract}
@@
 \end{alignat*}
-This simple elimination principle is too weak to define e.g.\ the
+This simple elimination principle is too weak e.g.\ to define the
 standard (metacircular) interpretation of our small syntax
@@
 
-By omitting the identity type former from the theory of QIIT
+By omitting the equality type former from the theory of QIIT
 signatures, we obtain a theory of IIT signatures and the construction
 is still valid. It follows that if a model of extensional type theory
-supports the theory of IIT signatures, it supports all IITs. In this
-paper we show that any model of extensional type theory with indexed
-W-types supports the theory of signatures, hence all IITs.
+supports the theory of IIT signatures, it supports all IITs.
+
+In this paper we show that any model of extensional type theory with
+indexed W-types supports the theory of IIT signatures, hence all
+IITs. The difficulty is that the theory of IIT signatures is a QIIT,
+it is both inductive-inductive and has equality constructors. However
+it can be seen as the well-typed syntax of a small type theory without
+any computation rules. Hence we can represent the syntax of normal
+forms without quotienting. We construct this well-typed normal syntax
+using preterms and typing relations from indexed W-types. Finally, we
+prove the elimination principle in the style of the initiality proof
+of Streicher \cite{streichersemantics}.
 
@@
 
-%% To derive the existence of all IITs from the theory of IIT signatures
-%% we use a variant of the CwF model
-%% \cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
-%% displayed CwFs but by CwF isofibrations. We need this extension to
-%% support infinitary constructors. Furthermore, using the method
-%% introduced by Streicher \cite{streichersemantics} we show the
-%% existence of the theory of IIT signatures. As opposed to Streicher, we
-%% don't use quotients in this construction as the theory of IIT
-%% signatures can be given only using neutral terms.
-
-%% Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
-%% we can translate our arguments to ITT+UIP+funext. Section
-%% \ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
-%% funext.
-
- (later in the development, we may similarly omit universal quantification)
-
-%% This is the paper version of our TYPES 2019 talk.
-
 \subsection{Related Work}
 
-The current work builds heavily on the work of Kaposi et al.\ on quotient
-inductive-inductive types \cite{Kaposi:2019:CQI:3302515.3290315}; we reuse both
-QIIT syntax and semantics by restricting to IITs, and we reuse the term model
-construction of QIITs as well.
+The current work builds heavily on the work of Kaposi et al.\
+\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
+inductive-inductive types (QIITs); we reuse both QIIT syntax and
+semantics by restricting to IITs, and we reuse the term model
+construction of QIITs as well. We also make use of the extesion to
+infinitary QIITs \cite{large_inf_qiit} to derive the specification of
+the elimination principle for the theory of IIT signatures.
 
-Forsberg \cite{forsberg-phd} specifies inductive-inductive types and develops a
-semantics based on dialgebras. He also provides a reduction of IITs to indexed
-inductive types, but only constructs a weak version of the eliminator.
+IITs (however not by this name) were first used to describe the
+well-typed syntax of type theory \cite{nisse,chapman09eatitself}. Agda
+supported these general inductive definitions even before they were
+named IITs and given semantics by Nordvall Forsberg and Setzer
+\cite{nordvallforsbergSetzer2010inductiveinductive}. Nordvall
+Forsberg's thesis \cite{forsberg-phd} contains a specification similar
+in style to Dybjer and Setzer's codes for inductive-recursive types
+\cite{Dybjer99afinite}. He also develops a categorical semantics based
+on dialgebras and provides a reduction of IITs to indexed inductive
+types, however only constructs the simple elimination principle as
+opposed to the general one. Altenkirch et al.\ \cite{gabe} define
+signatures for QIITs (thus IITs as well) and their categorical
+semantics, however without proving existence of initial
+algebras. Their notion of signature, just as Nordvall Forsberg's
+involves more encoding overhead than ours.
 
-Hugunin \cite{jasper} constructs several IITs in cubical Agda from inductive
-types. In this setting, the lack of UIP makes constructions significantly more
-involved, and essentially involves coinductive-coinductive well-formedness
-predicates defined as homotopy limits. Hugunin does not consider a generic
-syntax of IITs and only works on specific examples (although the examples vary
-greatly). He also does not construct full recursive-recursive eliminators.
+Cartmell \cite{gat} introduced generalised algebraic theories using a
+type-theoretic syntax. Removing equations from his signatures, we
+obtain finitary IIT signatures similar to ours. He does not consider
+constructing initial algebras using simpler classes of inductive
+types.
+
+Hugunin \cite{jasper} constructs several IITs in cubical Agda from
+inductive types. In this setting, the lack of UIP makes constructions
+significantly more involved, and essentially involves
+coinductive-coinductive well-formedness predicates defined as homotopy
+limits. Hugunin does not consider a generic syntax of IITs and only
+works on specific examples (although the examples vary greatly). He
+also only constructs the simple elimination principle.
 
@@
 ingredient in constructing an initial model, although he does not present
-details on the construction of the term model or its initiality proof. Our initiality proof can be seen as an indexed variant of his construction (see Subsection \ref{sec:streicher} for a comparison).
+details on the construction of the term model or its initiality proof. Our
+initiality proof can be seen as an indexed variant of his construction
+(see Subsection \ref{sec:streicher} for a comparison).
 
@@
 W-types \cite{mutual}, W-types to natural numbers and quotients
-\cite{Ahrens2019}. Quotient inductive-inductive types (and
-inductive-inductive types) can be reduced to quotient inductive types
+\cite{Ahrens2019}. (Q)IITs can be reduced to quotient inductive types
 using the reduction of generalised algebraic theories to essentially
-algebraic theories \cite{gat}.
+algebraic theories \cite{gat}. Using the same reduction as mutual
+inductive types to indexed inductive types, (Q)IITs with more than two
+sorts can be reduced to (Q)IITs with only two sorts \cite{szumiemail}.
 
@@
 
-\subsection{Metatheory and Notation}
+\subsection{Notation and Formalisation}
 
-We work in intensional Martin-Löf type theory, equipped with the following features:
+\begin{definition}[Model of extensional type theory (ETT)]
+  By a model of ETT we mean a category with families (CwF)
+  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
+  predicative hierarchy of universes closed under the following type
+  formers: $\Pi$, $\Sigma$, $0$, $1$, $2$, indexed W-types
+  \cite{indexedcont} and an identity type with uniqueness of identity
+  proofs and equality reflection.
+\end{definition}
+
+We will use Agda-like type theoretic syntax to work in the internal
+language of models of ETT:
 \begin{itemize}
-  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
-  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
+\item Universes are written $\Set_i$. We usually omit level indices in this paper.
+\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
   non-dependent. We sometimes omit function arguments, by implicitly
   generalising over variables.
-  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
-        when we want to leave the type of the first projection implicit. We use $A
-        \times B$ for non-dependent pairs.
-  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
-    this paper, we informally use extensional notation, and do not write
-    explicit transports over equalities. However, we occasionally indicate by
-    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
-    $e_1$,\dots,$e_n$.
-  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
-  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.
-  \item Natural numbers $\mathbb{N}$ with constructors $\mathsf{0}:\mathbb{N}$ and $\mathsf{S}:\mathbb{N}\ra\mathbb{N}$.
-
+\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
+  $\sum\limits_{x} B$ when we want to leave the type of the first
+  projection implicit. We use $A \times B$ for non-dependent pairs.
+\item We will define indexed W-types \cite{indexedcont} by listing
+  their constructors and using pattern matching to eliminate from
+  them.
+\item The equality (identity) type is written $t = u$, it has a
+  constructor $\refl : t = t$, and equality reflection, hence we use
+  the same $=$ sign for definitional equality. We occasionally
+  indicate by $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the
+  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
+  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
+  function extensionality, expressing
+  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
 \end{itemize}
 
-\subsection{Formalisation}
-
-Agda formalisation of Sections \ref{sec:theory_of_signatures} and
-\ref{sec:ambroise} can be found at
-\url{https://github.com/amblafont/UniversalII}. In addition to the features in
-the previous section, this formalisation also uses equality reflection for some
-equalities, through rewriting pragmas \cite{cockxsprinkles}. This is purely for
-convenience, and in principle we could eliminate the instances of equality
-reflection via UIP and function extensionality.
+The contents of Sections \ref{sec:theory_of_signatures} and
+\ref{sec:ambroise} were formalised in Agda, the formalisation is
+available at \url{https://github.com/amblafont/UniversalII}. Agda's
+pattern matching mechanism implies uniqueness of identity proofs, we
+assumed function extensionality as an axiom and used rewrite rules
+\cite{cockxsprinkles} to obtain limited equality reflection.
 
@@
 
-The theory of IIT signatures is given as a category with families
-(CwF) \cite{Dybjer96internaltype} with certain additional type formers. The CwF
-part constitutes the substitution calculus, while the type formers allow
-declaring new sorts and operators in a signature. Below, we list all components
-for the theory of signatures; this can be viewed as a notion of model for the
-theory, which is an iterated $\Sigma$ type. In this paper, we construct the
-\emph{initial} such model, i.e.\ a model equipped with a unique recursion principle.
+In this section we describe signatures for IITs. We will say what it
+means that a model of ETT supports the theory of IIT signatures. The
+theory of IIT signatures is a particular QIIT, and supporting a QIIT
+means that we have a notion of algebra and morphism, and there is an
+initial algebra with a unique morphism to every other algebra (called
+recursor or non-dependent eliminator).
 
-Here, initiality refers to the notion of homomorphism between models deduced
-from the general theory of large infinitary QIITs developed in \cite{large_inf_qiit}.
-Roughly, a homomorphism is a functor between underlying
-categories which strictly preserves all additional family structure and type
-formers.
+This description is analogous to saying that a model of ETT supports
+natural numbers as follows: Using the internal language of the model,
+a natural number algebra is given by a $\mathsf{Nat}:\Set$,
+$\mathsf{zero}:\mathsf{Nat}$ and
+$\mathsf{suc}:\mathsf{Nat}\ra \mathsf{Nat}$, while a morphism between
+algebras $M$ and $N$ is given by a function
+$f:\mathsf{Nat}^M\ra\mathsf{Nat}^N$ such that
+$f\,\mathsf{zero}^M=\mathsf{zero}^N$ and
+$f\,(\mathsf{suc}^M\,m)=\mathsf{suc}^N\,(f\,m)$. Natural numbers
+themselves are given by an algebra equipped with a unique morphism
+into any other algebra (the recursor).
 
+We will call \emph{theory of IIT signature algebras} simply
+\emph{signature algebras}. A signature algebra can be seen as a model
+of a small type theory which consists of (1) a substitution calculus
+(CwF \cite{Dybjer96internaltype}) equipped with (2) a universe, (3) a
+function space where the domain is small and (3) another function
+space with external domain. We explain the usage of these type
+formers through examples after the definitions.
+
+\begin{definition}[Signature algebra]\label{def:algebra}
+  In a model of ETT, a signature algebra consists of the following
+  four (families of) sets, 17 operations and 18 equalities written in
+  the internal language.
 \begin{alignat*}{5}
-  & \rlap{(1) Substitution calculus} \\[0.5em]
+  & \rlap{(1) Substitution calculus} \\
   & \Con && : \Set \\
@@
   & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
-  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
+  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
   & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
@@
   & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
-  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
-  & \rlap{(2) Universe} \\[0.5em]
+  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
+  & \rlap{(2) Universe} \\
   & \U && : \Ty\,\Gamma \\
@@
   & {\U[]} && : \U[\sigma] = \U \\
-  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
-  & \rlap{(3) Inductive parameters} \\[0.5em]
+  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
+  & \rlap{(3) Inductive parameters} \\
   & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
@@
   \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
-  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
-  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\[0.5em]
-  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
+  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
+  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
+  & \rlap{(4) External parameters} \\
   & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
@@
   & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
-  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
+  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
 \end{alignat*}
-\subsection{Notation and abbreviations}
+\end{definition}
 
-We define $\wk : \Sub\,(\Gamma\ext A)\ra \Gamma$ as $\pi_1\,\id$.
-We can recover de Bruijn indices by having $0= \pi_2 \,\id$ and $n+1 = n[\wk]$.
-However, for informal examples we instead
-use a nameful notation, where we introduce names in context extension and in the
-$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,A : \U\,\ext\,a :
-\El\,A$ in a signature for pointed sets.
+The definition of morphism is determined by the notion of algebra, it
+can be computed following \cite{large_inf_qiit}, but we include it
+here for completeness.
+\begin{definition}[Signature morphism]\label{def:morphism}
+  A morphism between signature algebras $M$ and $N$ consists of four
+  functions and 17 equalities expressing that the functions preserve
+  the operations of the two algebras. We use the same naming as in
+  \Cref{def:algebra} and use superscripts to denote which algebra is
+  meant.
+\begin{alignat*}{5}
+  & \rlap{(1) Substitution calculus} \\
+  & \Con && : \Con^M && \ra\Con^N \\
+  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
+  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
+  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
+  & \id && : \Sub\,\id^M && = \id^N \\
+  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
+  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
+  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
+  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
+  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
+  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
+  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
+  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
+  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
+  & \rlap{(2) Universe} \\
+  & \U && : \Ty\,\U^M && = \U^N \\
+  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
+  & \rlap{(3) Inductive parameters} \\
+  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
+  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
+  & \rlap{(4) External parameters} \\
+  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
+  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
+\end{alignat*}
+\end{definition}
+\begin{definition}\label{def:theoryofsignatures}
+  A model of ETT supports the theory of IIT signatures if there is a
+  signature algebra equipped with a unique morphism into any
+  algebra. We call this algebra the syntax or initial algebra, the
+  morphism is called recursor.
+\end{definition}
+\begin{definition}[Abbreviations]\label{def:abbrevs}
+  For any signature algebra, we use
+  $\wk : \Sub\,(\Gamma\ext A)\ra \Gamma$ to mean $\pi_1\,\id$. We
+  recover de Bruijn indices by having $0= \pi_2 \,\id$ and
+  $1+n = n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
+  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
+\end{definition}
 
-We define $\sigma^\uparrow$ as $(\sigma\circ\wk,0)$.
-We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[\wk])$ as $a\Ra B$.
-
-\subsection{Example signatures}
-
-Natural numbers:
+In a model supporting the theory of IIT signatures, we can write down
+signatures as elements of $\Con^\S$ (the $\Con$ component in the
+syntax). For example, natural numbers are given by the following
+signature. For readability, on the right we list the version of the
+signature using variable names instead of CwF-combinators.
 \[
+\boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
 \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
 \]
-Lists with elements of a given $A : \Set$ type. Here we use the $\Pim$ function
-to include an external type into the signature.
-\[
-\boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
-      cons : \Pim\,A\,(\lambda \_.\, L\Ra \El\,L)
-\]
-The inductive-inductive example from Section \ref{sec:intro}:
-\begin{alignat*}{3}
-  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\Ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra Con) \\
-  & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
-    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
+Natural numbers are given by a context of length 3, starting with the
+empty context $\boldsymbol{\cdot}$ and declaring a sort by $\U$,
+saying that we have an operator producing an element of $\El\,0$ where
+$0$ is a de Bruijn index referring to the sort $\U$. Finally we have
+an operator which takes as input an element of the sort $1$ and
+produces an element of the same sort. Note the assymetry of the
+function type $\Ra$: on the left there needs to be an element of $\U$,
+while on the right any type can appear (including another function
+type). This ensures strict positivity of the operators.
+
+Lists with elements of a given $T : \Set$ type are given by the
+following signature. Here we use the function space with external domain $\Ram$ to include
+an $T$ in the signature.
+\begin{alignat*}{10}
+ & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
+\end{alignat*}
+The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
+follows.
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
+  & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
+  & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
+  & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
+  & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
+  & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
+  & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
 \end{alignat*}

### 2020-04-29 paper/paper.tex
stuff

@@
 \usepackage{stmaryrd}
+% \usepackage{pdflscape} % if we want \begin{landscape} ... \end{landscape}
 
@@
 \end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-TODO: list of contributions here.
+TODO: list of contributions here. Notion of signature free from encoding.
 
@@
 infinitary QIITs \cite{large_inf_qiit} to derive the specification of
-the elimination principle for the theory of IIT signatures.
+the elimination principle for the theory of IIT signatures. TODO: also
+for specifying signatures.
 
@@
   predicative hierarchy of universes closed under the following type
-  formers: $\Pi$, $\Sigma$, $0$, $1$, $2$, indexed W-types
+  formers: $\Pi$, $\Sigma$, $\bot$, $\top$, $\Bool$, indexed W-types
   \cite{indexedcont} and an identity type with uniqueness of identity
-  proofs and equality reflection.
+  proofs and equality reflection. TODO: remove indexed W-types, bot
+  and bool.
 \end{definition}
@@
   $\sum\limits_{x} B$ when we want to leave the type of the first
-  projection implicit. We use $A \times B$ for non-dependent pairs.
+  projection implicit. Projections are either named or given by
+  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
+  pairs.
 \item We will define indexed W-types \cite{indexedcont} by listing
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{The Theory of IIT Signatures}
@@
 
-In this section we describe signatures for IITs. We will say what it
-means that a model of ETT supports the theory of IIT signatures. The
-theory of IIT signatures is a particular QIIT, and supporting a QIIT
-means that we have a notion of algebra and morphism, and there is an
-initial algebra with a unique morphism to every other algebra (called
-recursor or non-dependent eliminator).
+TODO: rewrite.
+
+In this section we show how to describe signatures for IITs using a
+QIIT, the theory of IIT signatures. We will say what it means for a
+model of ETT to support the theory of IIT signatures. First we need to
+say what a \emph{theory of IIT signature algebra} is and what
+morphisms between such algebras are. Then supporting the theory of IIT
+signatures means that there exists an algebra with a unique morphism
+to every other algebra.
 
@@
 
-\begin{definition}[Signature algebra]\label{def:algebra}
-  In a model of ETT, a signature algebra consists of the following
-  four (families of) sets, 17 operations and 18 equalities written in
-  the internal language.
+\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
+  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
+  consisting of the following components: four (families of) sets, 17
+  operations and 18 equalities.
 \begin{alignat*}{5}
@@
 \end{alignat*}
+Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
+$\Sub^M$, $\Tm^M$, $\id^M$, and so on.
 \end{definition}
+
+\begin{definition}[Abbreviations]\label{def:abbrevs}
+  For any signature algebra, we use
+  $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$ to mean $\pi_1\,\id$. We
+  recover de Bruijn indices by setting $0:= \pi_2 \,\id$ and
+  $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
+  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
+\end{definition}
+
+\begin{definition}[IIT signature]
+  An IIT signature is a context in an arbitrary signature algebra,
+  that is, a function $(M:\SignAlg)\ra\Con^M$.
+\end{definition}
+Note that the notion of IIT signature is relative to a model of ETT:
+it is expressed as a term (of a function type) in the model. This is
+necessary because of the function space $\Pim$ which has as domain an
+arbitrary type in the model. We make use of $\Pim$ in signatures with
+external parameters, like the type of the elements in lists.
+
+\begin{example}[Example signatures]\label{ex:signatures}
+  The signature for natural numbers is given by the function
+  \[
+    \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
+  \]
+  For readability we write the same signature without the $\lambda$
+  and the superscripts below on the left and we also write an informal
+  version with variable names on the right for comparison.
+  \[
+  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
+  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
+  \]
+  This signature is given by a context of length three. We start with
+  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
+  $\U$, then we declare an operator producing an element of $\El\,0$
+  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
+  we declare an operator which takes as input an element of the sort
+  $1$ and produces an element of the same sort. Note the assymetry of
+  the function type $\Ra$: on the left there needs to be an element of
+  $\U$, while on the right any type can appear (including another
+  function type). This ensures strict positivity of the operators.
+
+  Lists with elements of a given $T : \Set$ type are given by the
+  following signature. Here we use the function space with external
+  domain $\Ram$ to include a $T$ in the signature. Again, we omit the
+  $\lambda$ and the superscripts for readability.
+  \begin{alignat*}{10}
+    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
+  \end{alignat*}
+
+  The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
+  follows.
+  \begin{alignat*}{5}
+    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
+    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
+    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
+    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
+    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
+    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
+    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
+  \end{alignat*}
+\end{example}
+
+Given a model of ETT and an IIT signature in it, we would like to say
+what it means that the model supports the given IIT. For this we
+define the signature algebra $\ADS$ which will provide notions of
+algebras, displayed algebras and sections for each signature. This is
+the same as the $\blank^\A$, $\blank^\D$ and $\blank^\S$ operations in
+\cite{Kaposi:2019:CQI:3302515.3290315}. Before defining $\ADS$, we
+illustrate its usage by an example.
+
+\begin{example}[Algebras, displayed algebras and sections for natural
+  numbers]
+  For the signature of natural numbers as given in
+  \Cref{ex:signatures}, algebras are given by the $\Sigma$-type
+  $(N:\Set)\times N\times(N\ra N)$. A displayed algebra over $(N,z,s)$
+  is given by the $\Sigma$-type
+  \[
+    (N^D:N\ra\Set)\times N^D\,z \times((n:N)\ra N^D\,n\ra N^D\,(s\,n)),
+  \]
+  and a section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$
+  is given by the $\Sigma$-type
+  \[
+    (N^S:(n:N)\ra N^D\,n)\times (N^S\,z = z^D) \times((n:N)\ra N^S\,(s\,n) = s^D\,n\,(N^S\,n)).
+  \]
+  Displayed algebras over the initial algebra are called motives and
+  methods of the eliminator, while a section of a displayed algebra
+  over the initial algebra is the eliminator together with its
+  computation rules.
+\end{example}
+
+\begin{definition}[The signature algebra $\ADS$]\label{def:ads}
+  We define an element of $\SignAlg$ by listing all its components
+  $\Con$, $\Ty$, $\Sub$, and so on, one per line. Each such components
+  has three parts denoted by $^\A$, $^\D$ and $^\S$. The equality
+  components of $\SignAlg$ are omitted as they are all reflexivity.
+  \begin{alignat*}{10}
+    & (\Gamma^\A:\Set) && \hspace{-0.3em}\times && (\Gamma^\D:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set) \\
+    & (A^\A:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) && \hspace{-0.3em}\times && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra \\
+    & && && && && \hspace{1em} A^\D\,\gamma^D\,\alpha\ra\Set) \\
+    & (\sigma^\A:\Gamma^\A\ra\Delta^\A) && \hspace{-0.3em}\times && (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) && \hspace{-0.3em}\times && (\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra \\
+    & && && && && \hspace{1em} \Delta^\S\,(\sigma^\A\,\gamma)\,(\sigma^\D\,\gamma^D)) \\
+    & (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) && \hspace{-0.3em}\times && (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra && \hspace{-0.3em}\times && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra \\
+    & && && \hspace{1em} A^\D\,\gamma^D\,(t^\A\,\gamma)) && && \hspace{1em}  A^\S\,(t^\A\,\gamma)\,(t^\D\,\gamma^D)) \\
+    & \id^\A\,\gamma := \gamma && && \id^\D\,\gamma^D:=\gamma^D && && \id^\S\,\gamma^S := \gamma^S \\
+    & (\sigma\circ\delta)^\A\,\gamma := \sigma^\A\,(\delta^\A\,\gamma) && && (\sigma\circ\delta)^\D\,\gamma^D := \sigma^\D\,(\delta^\D\,\gamma^D) && && (\sigma\circ\delta)^\S\,\gamma^S := \sigma^\S\,(\delta^\S\,\gamma^S) \\
+    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^\D\,\gamma^D := A^\D\,(\sigma^\D\,\gamma^D) && && (A[\sigma])^\S\,\gamma^S := A^\S\,(\sigma^\S\,\gamma^S) \\
+    & (t[\sigma])^\A\,\gamma := t^\A\,(\sigma^\A\,\gamma) && && (t[\sigma])^\D\,\gamma^D := t^\D\,(\sigma^\D\,\gamma^D) && && (t[\sigma])^\S\,\gamma^S := t^\S\,(\sigma^\S\,\gamma^S) \\
+    & \cdot^\A := \top && && \cdot^\D\,\_ := \top && && \cdot^\S\,\_\,\_ := \top \\
+    & \epsilon^\A\,\_ := \tt && && \epsilon^\D\,\_ := \tt && && \epsilon^\S\,\_ := \tt \\
+    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^\D\,(\gamma,\alpha) :=  && && (\Gamma\rhd A)^\S\,(\gamma,\alpha)\,(\gamma^D,\alpha^D) := \\
+    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} (\gamma^D:\Gamma^\D\,\gamma)\times A^\D\,\gamma^D\,\alpha && && \hspace{1em} (\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\times A^\S\,\gamma^S\,\alpha\,\alpha^D \\
+    & (\sigma,t)^\A\,\gamma := (\sigma^\A\,\gamma,t^\A\,\gamma) && && (\sigma,t)^\D\,\gamma^D := (\sigma^\D\,\gamma^D,t^\D\,\gamma^D) && && (\sigma,t)^\S\,\gamma^S := (\sigma^\S\,\gamma^S,t^\S\,\gamma^S) \\
+    & (\pi_1\,\sigma)^\A\,\gamma := \proj_1\,(\sigma^\A\,\gamma) && && (\pi_1\,\sigma)^\D\,\gamma^D := \proj_1\,(\sigma^\D\,\gamma^D) && && (\pi_1\,\sigma)^\S\,\gamma^S := \proj_1\,(\sigma^\S\,\gamma^S) \\
+    & (\pi_2\,\sigma)^\A\,\gamma := \proj_2\,(\sigma^\A\,\gamma) && && (\pi_2\,\sigma)^\D\,\gamma^D := \proj_2\,(\sigma^\D\,\gamma^D) && && (\pi_2\,\sigma)^\S\,\gamma^S := \proj_2\,(\sigma^\S\,\gamma^S) \\
+    & \U^\A\,\gamma := \Set && && \U^\D\,\gamma^D\,T := T\ra\Set && && \U^\S\,\gamma^S\,T\,T^D := (\alpha:T)\ra T^D\,\alpha \\
+    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^\D\,\gamma^D\,\alpha := a^\D\,\gamma^D\,\alpha && && (\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D := (a^\S\,\gamma^S\,\alpha = \alpha^D) \\
+    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^\D\,\gamma^D\,f := && && (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)\ra \\
+    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em}  (\alpha^D:a^\D\,\gamma^D\,\alpha)\ra && && \hspace{1em} B^\S\,(\gamma^S,\refl_{a^\S\,\gamma^S\,\alpha})\,(f\,\alpha)\,\\
+    & && && \hspace{1em}  B^\D\,(\gamma^D,\alpha^D)\,(f\,\alpha) && && \hspace{1em}\hphantom{B^\S\,}(f^D\,(a^\S\,\gamma^S\,\alpha)) \\
+    & (t\oldapp u)^\A\,\gamma := t^\A\,\gamma\,(u^\A\,\gamma) && &&  (t\oldapp u)^\D\,\gamma^D := t^\D\,\gamma^D\,(u^\D\,\gamma^D) && &&  (t\oldapp u)^\S\,\gamma^S := u^\S\,\gamma^S \# t^\S\,\gamma^\S\,(u^\A\,\gamma) \\
+    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^\D\,\gamma^D\,f := && && (\Pim\,T\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:T)\ra \\
+    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\D\,\gamma^D\,(f\,\alpha) && && \hspace{1em} (B\,\alpha)^\S\,\gamma^S\,(f\,\alpha)\,(f^D\,\alpha) \\
+    & (t\appm \alpha)^\A\,\gamma := t^\A\,\gamma\,\alpha && && (t\appm \alpha)^\D\,\gamma^D := t^\D\,\gamma^D\,\alpha && && (t\appm \alpha)^\S\,\gamma^S := t^\S\,\gamma^S\,\alpha
+  \end{alignat*}
+\end{definition}
+\Cref{def:ads} can be explained by columns (see
+\cite[Sections 4 and 6]{Kaposi:2019:CQI:3302515.3290315} for more details) or by rows
+(see \cite[Section 7.4]{Kaposi:2019:CQI:3302515.3290315}).
+
+We first explain it by columns: the first column ($^A$ components)
+corresponds to the standard model (set model, metacircular
+interpretation \cite{ttintt}): contexts are sets, types are families,
+terms are functions, the universe $\U$ is given by $\Set$, function
+spaces are given by the external function space. The $^D$ column is a
+logical predicate interpretation, $^\A$ and $^\D$ together are a unary
+version of the parametric model for dependent types
+\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
+families of predicates, terms say that the $^\A$ interpretation
+respects the predicates (this is ususally called fundamental lemma of
+the logical predicate). $\U$ is given by predicate space, the
+predicate at a $\Pi$ type holds for a function if it respects the
+predicates. For $\Pim$, the predicate is defined pointwise. The last
+column ($^\S$) is a modified dependent logical relation which refers
+to both $^A$ and $^\D$. Contexts are binary relations where the second
+parameter depends on the first one, types are dependent variants of
+this, terms say that the relation is respected by $^\A$ and $^\D$,
+respectively. $\U$ is however not relation space, but a function and
+$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
+function again says that the function respects the relation, however
+we don't simply say
+\[
+  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
+\]
+as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
+$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
+contract. So we omit $\alpha^D$ and this equality as an input and
+replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.
+
+When viewing $\ADS$ by rows, we can see that it is a part of the CwF
+model of type theory \cite[Section
+7.4]{Kaposi:2019:CQI:3302515.3290315}. In the CwF model, a context is
+given by a CwF. From the category, we only have objects ($\Gamma^\A$),
+and from the families, we have the families for types $\Gamma^\D$ and
+terms $\Gamma^\S$. Types are the corresponding parts of displayed
+CwFs, substitutions are parts of CwF morphisms, terms are parts of CwF
+sections. $\U$ is part of the CwF of sets, $\El\,a$ is the part of the
+discrete displayed CwF coming from $a$ (which is a CwF-morphism from
+$\Gamma$ to the CwF of sets). $\Pi$ is given as by a dependent product
+of displayed CwFs where it is essential that the domain is discrete,
+$\Pim$ is the pointwise direct product.
+
+\begin{definition}[A model of ETT supports IITs]
+  A model of ETT supports IITs if for any signature signature
+  $\Omega:(M:\SignAlg)\ra\Con^M$ we have an $$
+\end{definition}
+
+Now, given a signature $f:(M:\SignAlg)\ra\Con^M$, we obtain
+$(f\,\ADS)^\A$
+
+  We describe the most important
+  components, a full definition is given in the appendix (TODO). A
+  $\Gamma:\Con$ is given by a $\Sigma$ type with three elements:
+  \[
+    (\Gamma^\A:\Set) \times (\Gamma^\D:\Gamma^\A\ra\Set) \times (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set)
+  \]
+  As a signature is a context in an arbitrary signature algebra, using
+  $\ADS$ a signature provides an algebra $\Gamma^\A$, a displayed
+  algebra $\Gamma^\D$ and a section $\Gamma^\S$.
+
+  A type in $\ADS$ is given by dependent versions of $\Con$:
+  \begin{alignat*}{6}
+    & (A:\Ty\,\Gamma) := \,\,&& (A^\A:\Gamma^A\ra\Set) \times (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) \times \\
+    & && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra A^\D\,\gamma^D\,\alpha\ra\Set)
+  \end{alignat*}
+  Substitutions (and terms) are given by (dependent) functions which
+  respect the predicates $^\D$ and relations $^\S$.
+  \begin{alignat*}{10}
+    & (\sigma:\Sub\,\Gamma\,\Delta) && := \,\, && (\sigma^\A:\Gamma^\A\ra\Delta^\A) \times (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) \times \\
+    & && && \rlap{$(\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra\Delta^\S\,(\sigma^A\,\gamma)\,(\sigma^\D\,\gamma^D))$} \\
+    & (t:\Tm\,\Gamma\,A) && := && (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) \times (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)) \times \\
+    & && && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra A^\S\,(t^A\,\gamma)\,(t^\D\,\gamma^D))
+  \end{alignat*}
+
+  xxx
 
@@
 \end{definition}
-\begin{definition}[Abbreviations]\label{def:abbrevs}
-  For any signature algebra, we use
-  $\wk : \Sub\,(\Gamma\ext A)\ra \Gamma$ to mean $\pi_1\,\id$. We
-  recover de Bruijn indices by having $0= \pi_2 \,\id$ and
-  $1+n = n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
-  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
-\end{definition}
 
-In a model supporting the theory of IIT signatures, we can write down
-signatures as elements of $\Con^\S$ (the $\Con$ component in the
-syntax). For example, natural numbers are given by the following
-signature. For readability, on the right we list the version of the
-signature using variable names instead of CwF-combinators.
-\[
-\boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
-\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
-\]
-Natural numbers are given by a context of length 3, starting with the
-empty context $\boldsymbol{\cdot}$ and declaring a sort by $\U$,
-saying that we have an operator producing an element of $\El\,0$ where
-$0$ is a de Bruijn index referring to the sort $\U$. Finally we have
-an operator which takes as input an element of the sort $1$ and
-produces an element of the same sort. Note the assymetry of the
-function type $\Ra$: on the left there needs to be an element of $\U$,
-while on the right any type can appear (including another function
-type). This ensures strict positivity of the operators.
-
-Lists with elements of a given $T : \Set$ type are given by the
-following signature. Here we use the function space with external domain $\Ram$ to include
-an $T$ in the signature.
-\begin{alignat*}{10}
- & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
-\end{alignat*}
-The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
-follows.
-\begin{alignat*}{5}
-  & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
-  & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
-  & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
-  & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
-  & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
-  & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
-  & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
-\end{alignat*}
-
-\section{Constructing the Theory of IIT Signatures}
-\label{sec:ambroise}
-
-%% \todo{Write this section}
-
-%% Show that the theory of IIT signatures as described in Section
-%% \ref{sec:theory_of_signatures} exists, that is, it can be defined only
-%% using indexed inductive types and a $\Prop$ sort.
-
-%% Ambroise has done this in Agda with funext and UIP using an
-%% extrinsic--intrinsic translation.
-
-%% Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.
-
-%% In this formalization, there is a discrepancy for the typing rule of application
-%% compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
-%% \[
-%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
-%% \]
-%% but
-%% \[
-%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
-%% \]
-%% Actually, I started with the first version, but at some point I
-%% changed for the second one because it made some stuff easier (although
-%% I don't remember exactly what it was).
-
-%% I think this should be doable with only UIP (or a $\Prop$-valued
-%% equality type?) and not funext. This would be important for a
-%% computational justification of IITs.
-
-%% Unfortunately, it seems that funext is needed (see the use of funext in
-%% Syntax.agda) for the metatheoretic/infinitary
-%% parameters part (4-5), for example in the case of the metatheoretic
-%% $ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
-%% investigated more carefully.
-
-\input{universaliit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \label{sec:constructingiits}
+
+In this section, we first specify what it means that a model of ETT
+supports all IITs. In order to even say this, the model already has to
+support the theory of IIT signatures as given in
+\Cref{def:theoryofsignatures}.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Constructing the Theory of IIT Signatures}
+\label{sec:ambroise}
+
+\input{universaliit}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Further Work}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \bibliography{b}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+
+\appendix
+
 \end{document}

### 2020-04-29 paper/paper.tex
stuff

@@
   \[
-    \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
+    nat := \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
   \]
@@
   We define an element of $\SignAlg$ by listing all its components
-  $\Con$, $\Ty$, $\Sub$, and so on, one per line. Each such components
+  $\Con$, $\Ty$, $\Sub$, and so on, one per row. Each such components
   has three parts denoted by $^\A$, $^\D$ and $^\S$. The equality
@@
 
-\begin{definition}[A model of ETT supports IITs]
+\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
   A model of ETT supports IITs if for any signature signature
-  $\Omega:(M:\SignAlg)\ra\Con^M$ we have an $$
+  $f:(M:\SignAlg)\ra\Con^M$ we have a $\con_f: (f\,\ADS)^\A$ and an
+  $\elim_f : (\gamma^D:(f\,\ADS)^\D\,\con_f)\ra (f\,\ADS)^\S\,\con_f\,\gamma^D$.
 \end{definition}
+In other words, for any signature, we have an algebra called $\con$
+(constructors) and for any displayed algebra over the constructors, we
+have a section.
 
-Now, given a signature $f:(M:\SignAlg)\ra\Con^M$, we obtain
-$(f\,\ADS)^\A$
-
-  We describe the most important
-  components, a full definition is given in the appendix (TODO). A
-  $\Gamma:\Con$ is given by a $\Sigma$ type with three elements:
+One can check that \Cref{sec:hasiits} gives the right
+notion of constructors and elimination principle for the signatures in
+\Cref{ex:signatures}.
+\begin{example}[An IIT supports natural numbers]
+  For the signature $nat$ of natural numbers in \Cref{ex:signatures},
+  the type of $\con_{nat}$ is
+  \begin{alignat*}{5}
+    & ({nat}\,\ADS)^\A = \\
+    & (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\A = \\
+    & \Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,(\pi_2\,(\pi_1\,\id)) \Ra \El\,(\pi_2\,(\pi_1\,\id))\Big)^\A = \\
+    & \Big(\gamma'':\big(\gamma':((\gamma:\top)\times\Set)\big)\times(\proj_2\,\gamma')\Big)\times\big(\proj_2\,(\proj_1\,\gamma'')\ra\proj_2\,(\proj_1\,\gamma'')\big),
+  \end{alignat*}
+  which is the left-associated $\Sigma$ type isomorphic to its
+  right-associated counterpart
   \[
-    (\Gamma^\A:\Set) \times (\Gamma^\D:\Gamma^\A\ra\Set) \times (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set)
+    (N:\Set)\times \big(N \times (N\ra N)\big).
   \]
-  As a signature is a context in an arbitrary signature algebra, using
-  $\ADS$ a signature provides an algebra $\Gamma^\A$, a displayed
-  algebra $\Gamma^\D$ and a section $\Gamma^\S$.
-
-  A type in $\ADS$ is given by dependent versions of $\Con$:
-  \begin{alignat*}{6}
-    & (A:\Ty\,\Gamma) := \,\,&& (A^\A:\Gamma^A\ra\Set) \times (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) \times \\
-    & && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra A^\D\,\gamma^D\,\alpha\ra\Set)
+  Pretending that we always had the right-associated variant, the type
+  of $\elim_{nat}$ computes as follows.
+  \begin{alignat*}{5}
+    & (\gamma^D:\,\, && ({nat}\,\ADS)^\D\,\con_{nat})\ra ({nat}\,\ADS)^\S\,\con_{nat}\,\gamma^D = \\
+    & (\gamma^D:\,\, && (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\D\,\con_{nat})\ra \\
+    & && (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\S\,\con_{nat}\,\gamma^D = \\
+    & (\gamma^D:\,\, && (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\D\,\con_{nat})\ra \\
+    & && (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\S\,\con_{nat}\,\gamma^D = \\
   \end{alignat*}
-  Substitutions (and terms) are given by (dependent) functions which
-  respect the predicates $^\D$ and relations $^\S$.
-  \begin{alignat*}{10}
-    & (\sigma:\Sub\,\Gamma\,\Delta) && := \,\, && (\sigma^\A:\Gamma^\A\ra\Delta^\A) \times (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) \times \\
-    & && && \rlap{$(\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra\Delta^\S\,(\sigma^A\,\gamma)\,(\sigma^\D\,\gamma^D))$} \\
-    & (t:\Tm\,\Gamma\,A) && := && (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) \times (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)) \times \\
-    & && && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra A^\S\,(t^A\,\gamma)\,(t^\D\,\gamma^D))
-  \end{alignat*}
+\end{example}
 

### 2020-04-29 paper/paper.tex
stuff

--- 2020-04-29 paper/paper.tex
@@
 
-\section{The Theory of IIT Signatures}
+\section{A Definition of Inductive-Inductive Types}
 \label{sec:theory_of_signatures}
 
-TODO: rewrite.
+In this section we say what it means that a model of ETT supports
+IITs. We first define the notion of IIT signature. Signatures for
+algebraic theories are usually given by inductive definitions. On the
+one hand, we take this even further: our notion of signature is given
+by a small type theory tailer-made to describe signatures, we call
+this the \emph{theory of IIT signatures}. On the other hand we would
+like to avoid using a complicated inductive definition (a type theory
+is a quotient inductive-inductive type) to describe a simpler class of
+inductive types. Hence we use a Church encoding of the theory of IIT
+signatures to define signatures thus avoiding the need for
+pre-existing inductive definitions. Another feature of our signatures
+is that as they can include types inside the model of ETT (such as
+$\N$ in the $\mathsf{isEven}$--$\mathsf{isOdd}$), hence signatures
+arex internal to the particular model of ETT.
 
-In this section we show how to describe signatures for IITs using a
-QIIT, the theory of IIT signatures. We will say what it means for a
-model of ETT to support the theory of IIT signatures. First we need to
-say what a \emph{theory of IIT signature algebra} is and what
-morphisms between such algebras are. Then supporting the theory of IIT
-signatures means that there exists an algebra with a unique morphism
-to every other algebra.
-
-This description is analogous to saying that a model of ETT supports
-natural numbers as follows: Using the internal language of the model,
-a natural number algebra is given by a $\mathsf{Nat}:\Set$,
-$\mathsf{zero}:\mathsf{Nat}$ and
-$\mathsf{suc}:\mathsf{Nat}\ra \mathsf{Nat}$, while a morphism between
-algebras $M$ and $N$ is given by a function
-$f:\mathsf{Nat}^M\ra\mathsf{Nat}^N$ such that
-$f\,\mathsf{zero}^M=\mathsf{zero}^N$ and
-$f\,(\mathsf{suc}^M\,m)=\mathsf{suc}^N\,(f\,m)$. Natural numbers
-themselves are given by an algebra equipped with a unique morphism
-into any other algebra (the recursor).
-
-We will call \emph{theory of IIT signature algebras} simply
-\emph{signature algebras}. A signature algebra can be seen as a model
-of a small type theory which consists of (1) a substitution calculus
-(CwF \cite{Dybjer96internaltype}) equipped with (2) a universe, (3) a
-function space where the domain is small and (3) another function
-space with external domain. We explain the usage of these type
-formers through examples after the definitions.
+We define the theory of IIT signatures by saying what its algebras
+(models) are. We call the \emph{theory of IIT signatures algebras}
+simply \emph{signature algebras}. The theory of signatures is a small
+type theory consisting of a (1) a substitution calculus (category with
+families, CwF \cite{Dybjer96internaltype}) equipped with (2) a
+universe, (3) a function space where the domain is in the universe and
+(4) another function space with external domain. We explain the usage
+of these type formers through examples after the definitions.
 
@@
 \end{definition}
+This can be seen as a Church encoding of $\Con$. For our purposes it
+is not a problem that signatures don't live at the lowest universe
+level.
+
 Note that the notion of IIT signature is relative to a model of ETT:
@@
 \begin{example}[Algebras, displayed algebras and sections for natural
-  numbers]
+  numbers]\label{ex:adsnat}
   For the signature of natural numbers as given in
@@
 terms are functions, the universe $\U$ is given by $\Set$, function
-spaces are given by the external function space. The $^D$ column is a
+spaces are given by the external function space. The $^\D$ column is a
 logical predicate interpretation, $^\A$ and $^\D$ together are a unary
@@
 
-\begin{definition}[A model of ETT supports IITs]
-  A model of ETT supports IITs if for any signature signature
-  $\Omega:(M:\SignAlg)\ra\Con^M$ we have an $$
+\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
+  A model of ETT supports IITs if for any signature
+  $f:(M:\SignAlg)\ra\Con^M$ there is a $\con_f: (f\,\ADS)^\A$ and an
+  $\elim_f : (\gamma^D:(f\,\ADS)^\D\,\con_f)\ra
+  (f\,\ADS)^\S\,\con_f\,\gamma^D$.
 \end{definition}
+In other words, for any signature, we have an algebra called $\con$
+(constructors) and for any displayed algebra over the constructors, we
+have a section (called the eliminator).
 
-Now, given a signature $f:(M:\SignAlg)\ra\Con^M$, we obtain
-$(f\,\ADS)^\A$
+One can check that \Cref{sec:hasiits} gives the right
+notion of constructors and elimination principle for the signatures in
+\Cref{ex:signatures}.
+\begin{example}[An IIT supports natural numbers]
+  For the signature $nat$ of natural numbers in \Cref{ex:signatures},
+  the type of $\con_{nat}$ is
+  \begin{alignat*}{5}
+    & ({nat}\,\ADS)^\A = \\
+    & (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\A = \\
+    & \Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\A = \\
+    & \Big(\gamma'':\big(\gamma':((\gamma:\boldsymbol{\cdot}^\A)\times \U^\A\,\gamma)\big)\times (\El\,(\pi_2\,\id))^\A\,\gamma'\Big)\times\Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\A\,\gamma'' = \\
+    & \Big(\gamma'':\big(\gamma':((\gamma:\top)\times\Set)\big)\times(\proj_2\,\gamma')\Big)\times\big(\proj_2\,(\proj_1\,\gamma'')\ra\proj_2\,(\proj_1\,\gamma'')\big),
+  \end{alignat*}
+  which is a left-nested $\Sigma$ type isomorphic to its
+  right-nested counterpart
+  \[
+    (N:\Set)\times \big(N \times (N\ra N)\big).
+  \]
+  Writing $(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc})$ for
+  $\con_{nat}$, the type of $\elim_{nat}$ computes as follows.
+  \begin{alignat*}{5}
+    & \rlap{$(\gamma^D: ({nat}\,\ADS)^\D\,\con_{nat})\ra ({nat}\,\ADS)^\S\,\con_{nat}\,\gamma^D =$} \\
+    & \rlap{$\bigg(\gamma^D: \Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\D\,\con_{nat}\bigg)\ra$} \\
+    & \rlap{$\Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,\con_{nat}\,\gamma^D =$} \\
+    & \rlap{$\bigg(\gamma^D: \Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\D\,$} \\
+    & \rlap{$\hspace{1em} \big(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc}\big)\bigg)\ra$} \\
+    & \rlap{$\Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,\big(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc}\big)\,\gamma^D = $} \\
+    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\boldsymbol{\cdot}^\D\,\tt)\times \U^\D\,\gamma^D\,\mathsf{Nat})\big)\times (\El\,(\pi_2\,\id))^\D\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
+    & && \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\D\,{\gamma^D}''\,\mathsf{suc}\bigg)\ra \\
+    & \rlap{$ \Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc})\,$} \\
+    & \rlap{$\hspace{1em} \big(((\tt,N^D),z^D),s^D\big) =$} \\
+    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\boldsymbol{\cdot}^\D\,\tt)\times \U^\D\,\gamma^D\,\mathsf{Nat})\big)\times (\El\,(\pi_2\,\id))^\D\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
+    & && \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\D\,{\gamma^D}''\,\mathsf{suc}\bigg)\ra \\
+    & \rlap{$ \Big( {\gamma^S}'':\big({\gamma^S}':((\gamma^S:\boldsymbol{\cdot}^\S\,\tt\,\tt)\times \U^\S\,\gamma^S\,\mathsf{Nat}\,N^D)\big)\times (\El\,(\pi_2\,\id))^\S\,{\gamma^S}'\,\mathsf{zero}\,z^D\Big)\times$} \\
+    & \rlap{$ \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\S\,{\gamma^S}''\,\mathsf{suc}\,s^D = $} \\
+    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\top)\times (\mathsf{Nat}\ra\Set))\big)\times \proj_2\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
+    & && \big(\proj_2\,(\proj_1\,{\gamma^D}'')\,n\ra\proj_2\,(\proj_1\,{\gamma^D}'')\,(\mathsf{suc}\,n)\big)\bigg)\ra \\
+    & \rlap{$ \Big( {\gamma^S}'':\big({\gamma^S}':((\gamma^S:\top)\times ((n:\mathsf{Nat})\ra N^D\,n))\big)\times \proj_2\,{\gamma^S}'\,\mathsf{zero} = z^D\Big)\times$} \\
+    & \rlap{$ \Big(\big(n:\mathsf{Nat}\big)\ra\proj_2\,(\proj_1\,(\proj_1\,{\gamma^S}''))\,(\mathsf{suc}\,n) = s^D\,\big(\proj_2\,(\proj_1\,(\proj_1\,{\gamma^S}''))\,n\big)\Big)$}
+  \end{alignat*}
+  This is again a left-nested version of the expected elimination
+  principle
+  \begin{alignat*}{10}
+    & (N^D:\mathsf{Nat}\ra\Set)(z^D:N^D\,\mathsf{zero})\big(s^D:(n:\mathsf{Nat})\ra N^D\,n\ra N^D\,(\mathsf{suc}\,n)\big)\ra \\
+    & \big(N^S:(n:\mathsf{Nat})\ra N^D\,n\big)\times(N^S\,\mathsf{zero} = z^D)\times \big((n:\mathsf{Nat})\ra N^S\,(\mathsf{suc}\,n) = s^D\,(N^S\,n)\big)
+  \end{alignat*}
+\end{example}
+\begin{remark}
+  Note that as we work in a model of ETT, the computation rules that
+  are expected from the elimination principle are given up to the
+  internal equality type, however by equality reflection we also get
+  them as definitional equalities.
+\end{remark}
 
-  We describe the most important
-  components, a full definition is given in the appendix (TODO). A
-  $\Gamma:\Con$ is given by a $\Sigma$ type with three elements:
-  \[
-    (\Gamma^\A:\Set) \times (\Gamma^\D:\Gamma^\A\ra\Set) \times (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set)
-  \]
-  As a signature is a context in an arbitrary signature algebra, using
-  $\ADS$ a signature provides an algebra $\Gamma^\A$, a displayed
-  algebra $\Gamma^\D$ and a section $\Gamma^\S$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-  A type in $\ADS$ is given by dependent versions of $\Con$:
-  \begin{alignat*}{6}
-    & (A:\Ty\,\Gamma) := \,\,&& (A^\A:\Gamma^A\ra\Set) \times (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) \times \\
-    & && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra A^\D\,\gamma^D\,\alpha\ra\Set)
-  \end{alignat*}
-  Substitutions (and terms) are given by (dependent) functions which
-  respect the predicates $^\D$ and relations $^\S$.
-  \begin{alignat*}{10}
-    & (\sigma:\Sub\,\Gamma\,\Delta) && := \,\, && (\sigma^\A:\Gamma^\A\ra\Delta^\A) \times (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) \times \\
-    & && && \rlap{$(\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra\Delta^\S\,(\sigma^A\,\gamma)\,(\sigma^\D\,\gamma^D))$} \\
-    & (t:\Tm\,\Gamma\,A) && := && (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) \times (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)) \times \\
-    & && && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra A^\S\,(t^A\,\gamma)\,(t^\D\,\gamma^D))
-  \end{alignat*}
-
-  xxx
+\section{Constructing IITs from the Theory of IIT Signatures}
+\label{sec:constructingiits}
 
@@
 \end{definition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Constructing IITs from the Theory of IIT Signatures}
-\label{sec:constructingiits}
 

### 2020-05-01 paper/paper.tex
stuf
using AFS encoding

@@
 like to avoid using a complicated inductive definition (a type theory
-is a quotient inductive-inductive type) to describe a simpler class of
-inductive types. Hence we use a Church encoding of the theory of IIT
-signatures to define signatures thus avoiding the need for
-pre-existing inductive definitions. Another feature of our signatures
-is that as they can include types inside the model of ETT (such as
-$\N$ in the $\mathsf{isEven}$--$\mathsf{isOdd}$), hence signatures
-arex internal to the particular model of ETT.
+is a quotient inductive-inductive type \cite{ttintt}) to describe a
+simpler class of inductive types. Hence we use a Church encoding
+\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
+avoiding the need for pre-existing inductive definitions. Another
+feature of our signatures is that they can include types from the
+model of ETT (such as $\N$ in the
+$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
+specified internally to the particular model of ETT.\footnote{There is
+  another method inspired by Capriotti \cite{paolo} which allows
+  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
+  model of ETT) supports IITs with definitional computation rules. In
+  this method, signatures are described in the internal language of
+  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
+  use this approach because it is more technical, and it would not
+  strengthen our main theorem (TODO: refer) as theorem (TODO: refer)
+  needs $\mathcal{C}$ to be a model of ETT.}
 
@@
 (4) another function space with external domain. We explain the usage
-of these type formers through examples after the definitions.
+of these type formers through examples after the definition.
 
@@
   In a model of ETT, a signature algebra is an iterated $\Sigma$ type
-  consisting of the following components: four (families of) sets, 17
-  operations and 18 equalities.
+  consisting of the following four (families of) sets, 17 operations
+  and 18 equalities.
 \begin{alignat*}{5}
@@
 
+The definition of morphism is determined by the notion of algebra, it
+can be computed following \cite{large_inf_qiit}, but we include it
+here for completeness.
+\begin{definition}[Signature morphism, $\SignMor$]\label{def:morphism}
+  A morphism from signature algebras $M$ to $N$ denoted
+  $\SignMor\,M\,N$ consists of four functions and 17 equalities
+  expressing that the functions preserve the operations of the two
+  algebras. We use the same naming as in \Cref{def:algebra} and use
+  superscripts to denote which algebra is meant.
+\begin{alignat*}{5}
+  & \rlap{(1) Substitution calculus} \\
+  & \Con && : \Con^M && \ra\Con^N \\
+  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
+  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
+  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
+  & \id && : \Sub\,\id^M && = \id^N \\
+  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
+  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
+  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
+  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
+  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
+  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
+  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
+  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
+  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
+  & \rlap{(2) Universe} \\
+  & \U && : \Ty\,\U^M && = \U^N \\
+  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
+  & \rlap{(3) Inductive parameters} \\
+  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
+  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
+  & \rlap{(4) External parameters} \\
+  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
+  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
+\end{alignat*}
+Given an $f:\SignMor\,M\,N$, we denote its first four components just
+by $f_\Con$, $f_\Ty$, $f_\Sub$, $f_\Tm$ or just write $f$ if it is
+clear which one is meant.
+\end{definition}
+
 \begin{definition}[IIT signature]
-  An IIT signature is a context in an arbitrary signature algebra,
-  that is, a function $(M:\SignAlg)\ra\Con^M$.
+  An IIT signature is a context in an arbitrary signature algebra
+  which is compatible with morphisms:
+  \begin{alignat*}{6}
+    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
+    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
+  \end{alignat*}
 \end{definition}
-This can be seen as a Church encoding of $\Con$. For our purposes it
-is not a problem that signatures don't live at the lowest universe
-level.
+The first component is the Church encoding of $\Con$, the second is
+the compatibility condition introduced by Awodey, Frey and Speight
+\cite{DBLP:conf/lics/AwodeyFS18}. We don't need impredicativity, as we
+don't require that $\Sign$ is the initial signature algebra. The
+compatibility condition says that if we obtain an $M$-context using
+${sig}$ in model $M$ and then we transport it to $N$ using $f$, we get
+the same $N$-context as directly applying ${sig}$ to $N$.
 
@@
 \begin{example}[Example signatures]\label{ex:signatures}
-  The signature for natural numbers is given by the function
+  Using informal notation with variable names, the signature for
+  natural numbers would be
   \[
-    nat := \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
+    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N.
   \]
-  For readability we write the same signature without the $\lambda$
-  and the superscripts below on the left and we also write an informal
-  version with variable names on the right for comparison.
+  This signature is a context of length three. We start with the empty
+  context $\boldsymbol{\cdot}$, then we declare a sort $N:\U$, then we
+  declare an operator producing an element of $N$ denoted by $\El\,N$,
+  and finally we declare an operator which takes as input an element
+  of the sort $N$ and produces an element of the same sort. Note the
+  assymetry of the function type $\Ra$: on the left there needs to be
+  an element of $\U$, while on the right any type can appear
+  (including another function type). This ensures strict positivity of
+  the operators.
+
+  Our theory of signatures uses de Bruijn indices instead of variable
+  names, so we rewrite the above signature to
   \[
-  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
-  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
+    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1,
   \]
-  This signature is given by a context of length three. We start with
-  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
-  $\U$, then we declare an operator producing an element of $\El\,0$
-  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
-  we declare an operator which takes as input an element of the sort
-  $1$ and produces an element of the same sort. Note the assymetry of
-  the function type $\Ra$: on the left there needs to be an element of
-  $\U$, while on the right any type can appear (including another
-  function type). This ensures strict positivity of the operators.
+  and as we $\Sign$ is defined formally as a pair of a Church encoding
+  together with a compatibility condition, our final definition for
+  the signature for natural numbers is the following. The
+  compatibility condition is written using equational reasoning.
+  \begin{alignat*}{5}
+    & \rlap{$(nat,natc) := $} \\
+    & \hspace{1em} && \big( && \rlap{$\lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M),$} \\
+    & && && \lambda M\,N\,f\,.\, && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M) = \\
+    & && && && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M)\ext^N f_\Ty\,(1^N\Ra^N \El^N\,1^N) = \\
+    & && && && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M)\,\ext^N\,f_\Ty\,(\El^M\,0^M)\ext^N f_\Tm\,1^N\Ra^M f_\Ty\,(\El^N\,1^N) = \\
+    & && && && f_\Con\,\boldsymbol{\cdot}^M\,\ext^N\,f_\Ty\,\U^M\,\ext^N\,\El^N\,(f_\Tm\,0^M)\ext^N 1^M\Ra^M \El^M\,(f_\Tm\,1^N) = \\
+    & && && && \boldsymbol{\cdot}^N\,\ext^N\,\U^N\,\ext^N\,\El^N\,0^N\,\ext^N\,1^N \Ra^N \El^N\,1^N \big)
+  \end{alignat*}
+  The first component builds the context describing natural numbers in
+  $M$, the second one uses the fact that $f$ is a morphism, that is,
+  it preserves all operations.
 
@@
   following signature. Here we use the function space with external
-  domain $\Ram$ to include a $T$ in the signature. Again, we omit the
-  $\lambda$ and the superscripts for readability.
+  domain $\Ram$ to include a $T$ in the signature. For readability, we
+  omit the $\lambda$ and the superscripts and we don't write the
+  compatibility condition. On the right we list the same signature
+  with variable names.
   \begin{alignat*}{10}
@@
 
-We first explain it by columns: the first column ($^A$ components)
+We first explain it by columns: the first column ($^\A$ components)
 corresponds to the standard model (set model, metacircular
@@
 predicates. For $\Pim$, the predicate is defined pointwise. The last
-column ($^\S$) is a modified dependent logical relation which refers
-to both $^A$ and $^\D$. Contexts are binary relations where the second
+column $^\S$ is a modified dependent logical relation which refers to
+both $^A$ and $^\D$. Contexts are binary relations where the second
 parameter depends on the first one, types are dependent variants of
@@
 7.4]{Kaposi:2019:CQI:3302515.3290315}. In the CwF model, a context is
-given by a CwF. From the category, we only have objects ($\Gamma^\A$),
-and from the families, we have the families for types $\Gamma^\D$ and
-terms $\Gamma^\S$. Types are the corresponding parts of displayed
-CwFs, substitutions are parts of CwF morphisms, terms are parts of CwF
-sections. $\U$ is part of the CwF of sets, $\El\,a$ is the part of the
-discrete displayed CwF coming from $a$ (which is a CwF-morphism from
-$\Gamma$ to the CwF of sets). $\Pi$ is given as by a dependent product
-of displayed CwFs where it is essential that the domain is discrete,
-$\Pim$ is the pointwise direct product.
+given by a CwF. Now, from the category part of the CwF, we only have
+objects ($\Gamma^\A$), and from the families, we have the families for
+types $\Gamma^\D$ and terms $\Gamma^\S$. Types are the corresponding
+parts of displayed CwFs, substitutions are parts of CwF morphisms,
+terms are parts of CwF sections. $\U$ is part of the CwF of sets,
+$\El\,a$ is the part of the discrete displayed CwF coming from $a$
+(which is a CwF-morphism from $\Gamma$ to the CwF of sets). $\Pi$ is
+given by a dependent product of displayed CwFs where it is essential
+that the domain is discrete, $\Pim$ is the pointwise direct product.
 
 \begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
-  A model of ETT supports IITs if for any signature
-  $f:(M:\SignAlg)\ra\Con^M$ there is a $\con_f: (f\,\ADS)^\A$ and an
-  $\elim_f : (\gamma^D:(f\,\ADS)^\D\,\con_f)\ra
-  (f\,\ADS)^\S\,\con_f\,\gamma^D$.
+  A model of ETT supports IITs if for any signature $(sig,nat):\Sign$
+  there is a
+  \[
+    \con_{sig}: ({sig}\,\ADS)^\A
+  \]
+  and an
+  \[
+    \elim_{sig} : (\gamma^D:({sig}\,\ADS)^\D\,\con_{sig})\ra ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
+  \]
 \end{definition}
@@
 \begin{example}[An IIT supports natural numbers]
-  For the signature $nat$ of natural numbers in \Cref{ex:signatures},
+  For the signature $({nat},{natc})$ of natural numbers in \Cref{ex:signatures},
   the type of $\con_{nat}$ is
@@
 \begin{remark}
-  Note that as we work in a model of ETT, the computation rules that
-  are expected from the elimination principle are given up to the
-  internal equality type, however by equality reflection we also get
-  them as definitional equalities.
+  The computation rules of the elimination principle are only expected
+  up to the internal equality type, but as we work with a model of
+  ETT, by equality reflection we also get them as definitional
+  equalities.
 \end{remark}
@@
 
-The definition of morphism is determined by the notion of algebra, it
-can be computed following \cite{large_inf_qiit}, but we include it
-here for completeness.
-\begin{definition}[Signature morphism]\label{def:morphism}
-  A morphism between signature algebras $M$ and $N$ consists of four
-  functions and 17 equalities expressing that the functions preserve
-  the operations of the two algebras. We use the same naming as in
-  \Cref{def:algebra} and use superscripts to denote which algebra is
-  meant.
-\begin{alignat*}{5}
-  & \rlap{(1) Substitution calculus} \\
-  & \Con && : \Con^M && \ra\Con^N \\
-  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
-  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
-  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
-  & \id && : \Sub\,\id^M && = \id^N \\
-  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
-  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
-  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
-  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
-  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
-  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
-  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
-  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
-  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
-  & \rlap{(2) Universe} \\
-  & \U && : \Ty\,\U^M && = \U^N \\
-  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
-  & \rlap{(3) Inductive parameters} \\
-  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
-  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
-  & \rlap{(4) External parameters} \\
-  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
-  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
-\end{alignat*}
-\end{definition}
+In the previous section, using the notions of signature algebras and
+signature morphisms, we defined signatures for IITs and what it means
+for a model of ETT to support all IITs. In this section we show that
+if a model of ETT supports the theory of IIT signatures, then it
+supports all IITs.
+
+The contents of this section are an adjustment of \cite[Sections 4 and
+6]{Kaposi:2019:CQI:3302515.3290315} to our setting.
+
 \begin{definition}\label{def:theoryofsignatures}
   A model of ETT supports the theory of IIT signatures if there is a
-  signature algebra equipped with a unique morphism into any
-  algebra. We call this algebra the syntax or initial algebra, the
-  morphism is called recursor.
+  signature algebra $\I : \SignAlg$ equipped with a unique morphism
+  into any algebra $\rec : (M:\SignAlg)\ra\SignMor\,\I\,M$. We call
+  this algebra the syntax or initial algebra, the morphism is called
+  recursor.
 \end{definition}
-
-In this section, we first specify what it means that a model of ETT
-supports all IITs. In order to even say this, the model already has to
-support the theory of IIT signatures as given in
-\Cref{def:theoryofsignatures}.
 

### 2020-05-01 paper/paper.tex
stuff

@@
   We define an element of $\SignAlg$ by listing all its components
-  $\Con$, $\Ty$, $\Sub$, and so on, one per row. Each such components
-  has three parts denoted by $^\A$, $^\D$ and $^\S$. The equality
-  components of $\SignAlg$ are omitted as they are all reflexivity.
+  $\Con$, $\Ty$, $\Sub$, and so on, one per row. Each such component
+  has three parts denoted by $^\A$, $^\D$ and $^\S$, respectively. The
+  equality components of $\SignAlg$ are omitted as they are all
+  reflexivity.
   \begin{alignat*}{10}
@@
 that the domain is discrete, $\Pim$ is the pointwise direct product.
+
+\begin{definition}[The set signature algebra $\A$]\label{def:a}
+  $\A : \SignAlg$ is given by the first $^\A$ components of $\ADS$
+  (\Cref{def:ads}), that is, $\Con^\A := \Set$,
+  $\Ty^\A\,\Gamma := \Gamma\ra\Set$,
+  $\Sub^\A\,\Gamma\,\Delta := \Gamma\ra\Delta$, and so on. Note that
+  there is a morphism from $\ADS$ to $\A$ defined by $\blank^\A$ at
+  each component, which we also denote by
+  $\blank^\A : \SignMor\,\ADS\,\A$.
+\end{definition}
 
@@
   signature algebra $\I : \SignAlg$ equipped with a unique morphism
-  into any algebra $\rec : (M:\SignAlg)\ra\SignMor\,\I\,M$. We call
-  this algebra the syntax or initial algebra, the morphism is called
-  recursor.
+  $\ll\blank\rr_{M} : \SignMor\,\I\,M$ into any algebra $M$. Sometimes
+  we omit the subscript $_M$. We call $\I$ the syntax or initial
+  algebra, the morphism $\ll\blank\rr$ is called recursor.
 \end{definition}
+
+First we define the term signature algebra using which we obtain the
+constructors $\con$ for any IIT signature. Then we will define another
+signature algebra which provides the eliminator. Before doing these,
+we illustrate the idea on natural numbers.
+\begin{example}
+  For natural numbers, we will define the constructors $\con$ as the
+  following natural number algebra
+  $(\mathsf{Nat},\mathsf{zero},\mathsf{suc})$. We write variable names
+  instead of de Bruijn indices for readability.
+  \begin{alignat*}{5}
+    & \mathsf{Nat} && := \Tm^\I\,(\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N)\,(\El\,N) \\
+    & \mathsf{zero} && := z \\
+    & \mathsf{suc} && := \lambda t. (s\oldapp t)
+  \end{alignat*}
+  Natural numbers are simply $\I$-terms of type $\El\,N$ in the
+  context representing the signature for natural numbers. In this
+  context, the only way to define a term of type $\El\,N$ is to use
+  $z$ and $s$, corresponding to the $\mathsf{zero}$ and $\mathsf{suc}$
+  constructors.
+
+  To define the action of the eliminator on a natural number
+  $n : \mathsf{Nat}$, let's look at the type of the displayed algebra
+  interpretation of the number:
+  \[
+    {\ll n \rr_\ADS}^\D : (\gamma^D : \ll\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N\rr^\D\,\con) \ra \ll\El\,N\rr^\D\,(\ll n\rr^\A\,\con)
+  \]
+  This says that for a displayed algebra $(N^D,z^D,s^D)$ over $\con$
+  (i.e.\ the motives and methods of the eliminator), we get a witness
+  of the predicate $\ll\El\,N\rr^\D = N^D$ at the algebra
+  interpretation of $n$. This is not yet good, as we would like to get
+  $N^D\,n$ instead of $N^D\,(\ll n\rr^\A\,\con)$ as a result. However,
+  interpretation into the term signature algebra will imply that
+  $n = \ll n\rr^\A\,\con$.
+\end{example}
+
+\begin{definition}[Term signature algebra $\A\blank\C_{\blank}$]
+  For an $M:\SignAlg$ and $\Omega:\Con^M$, we define ${\A M\C}_\Omega$
+  which we call the term signature algebra. It is equipped with
+  morphisms $\blank^\A : \SignMor\,({\A M\C}_\Omega)\,\A$ and
+  $\blank^M : \SignMor\,({\A M\C}_\Omega)\,M$.
+
+  We define ${\A M\C}_\Omega$ by listing all its components $\Con$,
+  $\Ty$, $\Sub$, and so on, one per row. Each component has three
+  parts denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are
+  just reusing the corresponding parts from $\A$ and $M$,
+  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
+  defined as the obvious projections. We omit the equality components,
+  as they come from UIP or are trivial. We also omit the components
+  for terms and substitutions as their $^\C$ parts are uninformative
+  equational reasonings.
+  \begin{alignat*}{10}
+    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
+    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
+    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
+    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
+    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
+    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
+    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
+    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
+    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
+    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
+    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
+    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
+    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
+    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
+    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
+  \end{alignat*}
+\end{definition}
+
+xxx
 

### 2020-05-01 paper/paper.tex
stuff

@@
 
-\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by Thematic Excellence Programme, Industry and Digitization Subprogramme (NRDI Office, 2019) and by the European Union, co-financed by the European Social Fund (EFOP-3.6.2-16-2017-00013, Thematic Fundamental Research Collaborations Grounding Innovation in Informatics and Infocommunication).}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
+Development and Innovation Fund of Hungary, financed under the
+Thematic Excellence Programme funding scheme, Project
+no. ED18-1-2019-0030 (Application-specific highly reliable IT
+solutions), by the New National Excellence Program of the Ministry
+for Innovation and Technology, Project no. ÚNKP-19-4-ELTE-874, and
+by the Bolyai Fellowship of the Hungarian Academy of Sciences,
+Project no.
+BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
+\begin{remark}
+  In a model of ETT supporting the theory of ITT signatures, we can
+  view elements of $\Con^\I$ as syntactic signatures. We can turn a
+  $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$ and an
+  $\Omega:\Con^\I$ into a $\Sign$ by
+  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
+  \ll\Omega\rr_N)$ where the equality in the second component comes
+  from uniqueness of the recursor.
+\end{remark}
+
 First we define the term signature algebra using which we obtain the
@@
 we illustrate the idea on natural numbers.
-\begin{example}
+\begin{example}\label{eg:nat_con}
   For natural numbers, we will define the constructors $\con$ as the
@@
   Natural numbers are simply $\I$-terms of type $\El\,N$ in the
-  context representing the signature for natural numbers. In this
-  context, the only way to define a term of type $\El\,N$ is to use
-  $z$ and $s$, corresponding to the $\mathsf{zero}$ and $\mathsf{suc}$
-  constructors.
+  context which is the syntactic signature for natural numbers. In
+  this context, the only way to define a term of type $\El\,N$ is to
+  use $z$ and $s$, corresponding to the $\mathsf{zero}$ and
+  $\mathsf{suc}$ constructors.
 
@@
 
-  We define ${\A M\C}_\Omega$ by listing all its components $\Con$,
-  $\Ty$, $\Sub$, and so on, one per row. Each component has three
-  parts denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are
-  just reusing the corresponding parts from $\A$ and $M$,
+  We define ${\A M\C}_\Omega$ by listing its components $\Con$, $\Ty$,
+  $\Sub$, and so on, one per row. Each component has three parts
+  denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are just
+  reusing the corresponding components from $\A$ and $M$,
   respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
@@
 \end{definition}
+\begin{example}
+  Now given a syntactic signature $\Omega:\Con^\I$, we get the
+  constructors as an $\Omega$-algebra by
+  $(\ll\Omega\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \ll\Omega\rr^\A$. If
+  $\Omega$ is the syntactic signature for natural numbers, we get the
+  constructors as in \Cref{eg:nat_con}.
+
+  An $a : \Tm\,\Omega\,\U$ is a sort term for a signature $\Omega$. If
+  $\Omega$ is the syntactic signature for natural numbers, $a$ can
+  only be $N$ ($1$ as a de Bruijn index). If $\Omega$ is the syntactic
+  signature for $\Con$--$\Ty$ (\Cref{ex:signatures}), $a$ can be
+  $Con$, $Ty\oldapp{empty}$,
+  $Ty\oldapp({ext}\oldapp {empty}\oldapp(U\oldapp{empty}))$, and so
+  on. In any case, for such an $a$, we obtain
+  $(\ll a\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a)
+  = \ll a\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, the algebra
+  interpretation of a sort term at the constructors is equal to terms
+  of that sort.
+
+  A $t : \Tm\,\Omega\,(\El\,a)$ is a term of a sort type $a$
+  constructed using the constructors in $\Omega$. For natural numbers,
+  such a $t$ can only be $s$ applied iteratively to $z$. For such a
+  $t$, we obtain
+  $\ll t\rr^\C\,\id^\I : t = \ll
+  t\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, a constructor term is
+  equal to its algebra interpretation at the constructors. This is
+  exactly the equation needed at the end of \Cref{eg:nat_con}.
+\end{example}
+\begin{remark}
+  Given a syntactic signature $\Omega : \Con^\I$, we can use the term
+  signature algebra with choices other than $M = \I$. For example, if
+  $M = \A$, we usual get the Church encoding by
+  $(\ll\Omega\rr_{{\A\A\C}_{\ll\Omega\rr^\A}})^\C\,\id^\A$. In the
+  case of natural numbers, this computes (up to curry-uncurry) to
+  $(N:\Set)(z:N)(s:N\ra N)\ra N$. Using $M = {\A\M}$ (the signature
+  algebra $\A$ extended with the binary version of $\D$ also called
+  logical relations, see \cite[Section
+  5]{Kaposi:2019:CQI:3302515.3290315}), we obtain the Church encoding
+  of Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}.
+\end{remark}
+
+\begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
+  Given an $\Omega:\Con^\I$,
+  $\omega^D : \ll\Omega\rr^\D\,(\ll\Omega\rr^\C\,\id^\I)$, we define a
+  the signature algebra $\IE_{\omega^\D}$ as follows. It is equipped
+  with a morphism $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$.
+
+  We define $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$,
+  $\Sub$, and so on, one per row. Each component has two parts denoted
+  by $^\I$, and $^\E$. The $^\I$ part just reuses the corresponding
+  components of $\I$, thus the morphism $\blank^\I$ is defined as the
+  obvious projection. We omit the equality components, as they come
+  from UIP or are trivial. We also omit the components for terms and
+  substitutions as their $^\E$ parts are uninformative equational
+  reasonings.
 
 xxx
+\end{definition}
 

### 2020-05-02 paper/paper.tex
sec2

@@
 Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
-$\Sub^M$, $\Tm^M$, $\id^M$, and so on.
+$\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
+only one signature algebra in scope (e.g.\ in
+\Cref{def:abbrevs,ex:contexts}).
 \end{definition}
@@
 \begin{definition}[Abbreviations]\label{def:abbrevs}
-  For any signature algebra, we use
-  $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$ to mean $\pi_1\,\id$. We
-  recover de Bruijn indices by setting $0:= \pi_2 \,\id$ and
-  $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
-  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
-\end{definition}
-
-The definition of morphism is determined by the notion of algebra, it
-can be computed following \cite{large_inf_qiit}, but we include it
-here for completeness.
+  For a signature algebra, we use $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$
+  to mean $\pi_1\,\id$. We recover de Bruijn indices by setting
+  $0:= \pi_2 \,\id$ and $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is
+  abbreviated by $a\Ra B$, $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
+\end{definition}
+
+\begin{example}[Example contexts in a signature algebra]\label{ex:contexts}
+  Given a signature algebra, we can define a context in which
+  specifies natural numbers. For readability, the an informal version
+  of the same context is displayed on the right using variable names.
+  \[
+    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1
+    \hspace{4.9em}
+    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
+  \]
+  We start with the empty context $\boldsymbol{\cdot}$, then we
+  declare a sort $\U$, then we declare an operator producing an
+  element of the sort denoted by $\El\,0$ where $0$ is the de Bruijn
+  index referring to the sort. Finally, we declare an operator which
+  takes as input an element of the sort (now it became de Bruijn index
+  $1$) and produces an element of the same sort. Note the assymetry of
+  the function type $\Ra$: on the left there needs to be an element of
+  $\U$, while on the right any type can appear (including another
+  function type). This ensures strict positivity of the operators.
+
+  Lists with elements of a given $T : \Set$ type are given by the
+  following context. Here we use the function space with external
+  domain $\Ram$ to include a $T$ in the signature. For readability, we
+  omit the $\lambda$ and the superscripts and we don't write the
+  compatibility condition. On the right we list the same signature
+  with variable names.
+  \begin{alignat*}{10}
+    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
+  \end{alignat*}
+
+  The $\Con$--$\Ty$ example from \Cref{sec:intro} is given by the
+  following context.
+  \begin{alignat*}{5}
+    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
+    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
+    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
+    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
+    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
+    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
+    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
+  \end{alignat*}
+\end{example}
+
+The above examples are contexts in any signature algebra, and we could
+take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
+usual Church-encoding of contexts. However (as we will see in TODO)
+the notion of constructor for such signatures would be too
+strong. Another approach would be to assume that there is a syntax for
+signature algebras (an initial signature algebra), and then a
+signature would be a context in this signature algebra. We will define
+syntactic signatures using this approach in the next section
+(\Cref{def:syntacticsignature}), but for now we don't want to assume
+the existence of any inductive type. Instead, we will use a restricted
+Church encoding which needs the notion of morphism of signatures.
+
+The notion of morphism is determined by the notion of algebra
+\cite{large_inf_qiit}, but we include it here for completeness.
 \begin{definition}[Signature morphism, $\SignMor$]\label{def:morphism}
@@
 
+We define IIT signatures using the Church encoding introduced by
+Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
+difference is that we avoid the usage of impredicativity as we don't
+require that signatures form a signature algebra. It is also not a
+problem for our use case that signatures are not in the lowest
+universe.
 \begin{definition}[IIT signature]
@@
 \end{definition}
-The first component is the Church encoding of $\Con$, the second is
-the compatibility condition introduced by Awodey, Frey and Speight
-\cite{DBLP:conf/lics/AwodeyFS18}. We don't need impredicativity, as we
-don't require that $\Sign$ is the initial signature algebra. The
-compatibility condition says that if we obtain an $M$-context using
-${sig}$ in model $M$ and then we transport it to $N$ using $f$, we get
-the same $N$-context as directly applying ${sig}$ to $N$.
+Note that as $\Sign$ refers to $\SignAlg$, it lives in a higher
+universe level than $\SignAlg$. The compatibility condition says that
+if we obtain an $M$-context using ${sig}$ in model $M$ and then we
+transport it to $N$ using $f$, we get the same $N$-context as directly
+applying ${sig}$ to $N$.
 
@@
 
-\begin{example}[Example signatures]\label{ex:signatures}
-  Using informal notation with variable names, the signature for
-  natural numbers would be
-  \[
-    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N.
-  \]
-  This signature is a context of length three. We start with the empty
-  context $\boldsymbol{\cdot}$, then we declare a sort $N:\U$, then we
-  declare an operator producing an element of $N$ denoted by $\El\,N$,
-  and finally we declare an operator which takes as input an element
-  of the sort $N$ and produces an element of the same sort. Note the
-  assymetry of the function type $\Ra$: on the left there needs to be
-  an element of $\U$, while on the right any type can appear
-  (including another function type). This ensures strict positivity of
-  the operators.
-
-  Our theory of signatures uses de Bruijn indices instead of variable
-  names, so we rewrite the above signature to
-  \[
-    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1,
-  \]
-  and as we $\Sign$ is defined formally as a pair of a Church encoding
-  together with a compatibility condition, our final definition for
-  the signature for natural numbers is the following. The
-  compatibility condition is written using equational reasoning.
+\begin{example}[Example signature]\label{ex:signatures}
+  Now we can formally describe the contexts given in
+  \Cref{ex:contexts} as signatures. For natural numbers, we have the
+  following pair of functions. The second function returns an equality
+  proof which we describe using equational reasoning.
   \begin{alignat*}{5}
@@
 
-  Lists with elements of a given $T : \Set$ type are given by the
-  following signature. Here we use the function space with external
-  domain $\Ram$ to include a $T$ in the signature. For readability, we
-  omit the $\lambda$ and the superscripts and we don't write the
-  compatibility condition. On the right we list the same signature
-  with variable names.
-  \begin{alignat*}{10}
-    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
-  \end{alignat*}
-
-  The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
-  follows.
-  \begin{alignat*}{5}
-    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
-    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
-    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
-    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
-    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
-    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
-    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
-  \end{alignat*}
+  The signatures for lists and $\Con$--$\Ty$ can be given analogously.
 \end{example}
@@
   $\Ty^\A\,\Gamma := \Gamma\ra\Set$,
-  $\Sub^\A\,\Gamma\,\Delta := \Gamma\ra\Delta$, and so on. Note that
-  there is a morphism from $\ADS$ to $\A$ defined by $\blank^\A$ at
-  each component, which we also denote by
-  $\blank^\A : \SignMor\,\ADS\,\A$.
+  $\Sub^\A\,\Gamma\,\Delta := \Gamma\ra\Delta$, and so on. There is a
+  morphism from $\ADS$ to $\A$ defined by $\blank^\A$ at each
+  component, which we also denote by $\blank^\A : \SignMor\,\ADS\,\A$.
 \end{definition}
@@
 \begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
-  A model of ETT supports IITs if for any signature $(sig,nat):\Sign$
+  A model of ETT supports IITs if for any signature $(sig,sigc):\Sign$
   there is a
@@
 \Cref{ex:signatures}.
-\begin{example}[An IIT supports natural numbers]
-  For the signature $({nat},{natc})$ of natural numbers in \Cref{ex:signatures},
-  the type of $\con_{nat}$ is
+\begin{example}[A model of ETT supports natural numbers]
+  For the signature $({nat},{natc})$ of natural numbers in
+  \Cref{ex:signatures}, the type of $\con_{nat}$ is
   \begin{alignat*}{5}
@@
   up to the internal equality type, but as we work with a model of
-  ETT, by equality reflection we also get them as definitional
-  equalities.
+  ETT, we also get them as definitional equalities by equality
+  reflection.
 \end{remark}
@@
 
-\begin{remark}
-  In a model of ETT supporting the theory of ITT signatures, we can
-  view elements of $\Con^\I$ as syntactic signatures. We can turn a
+\begin{definition}[Syntactic signatures]\label{def:syntacticsignature}
+  In a model of ETT supporting the theory of ITT signatures, elements
+  of $\Con^\I$ are called syntactic signatures. We can turn a
   $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$ and an
@@
   from uniqueness of the recursor.
-\end{remark}
+\end{definition}
 
@@
   only be $N$ ($1$ as a de Bruijn index). If $\Omega$ is the syntactic
-  signature for $\Con$--$\Ty$ (\Cref{ex:signatures}), $a$ can be
+  signature for $\Con$--$\Ty$ (\Cref{ex:contexts}), $a$ can be
   $Con$, $Ty\oldapp{empty}$,

### 2020-05-02 paper/paper.tex
stuff

@@
 In the previous section, using the notions of signature algebras and
-signature morphisms, we defined signatures for IITs and what it means
-for a model of ETT to support all IITs. In this section we show that
-if a model of ETT supports the theory of IIT signatures, then it
-supports all IITs.
+signature morphisms, we defined IIT signatures and what it means for a
+model of ETT to support all IITs. In this section we show that if a
+model of ETT supports the theory of IIT signatures, then it supports
+all IITs.
 
@@
 signature algebra which provides the eliminator. Before doing these,
-we illustrate the idea on natural numbers.
+we illustrate the idea of both constructions on natural numbers.
 \begin{example}\label{eg:nat_con}
@@
   \]
-  This says that for a displayed algebra $(N^D,z^D,s^D)$ over $\con$
-  (i.e.\ the motives and methods of the eliminator), we get a witness
-  of the predicate $\ll\El\,N\rr^\D = N^D$ at the algebra
+  This says that for a displayed algebra $\gamma^D=(N^D,z^D,s^D)$ over
+  $\con$ (i.e.\ the motives and methods of the eliminator), we get a
+  witness of the predicate $\ll\El\,N\rr^\D = N^D$ at the algebra
   interpretation of $n$. This is not yet good, as we would like to get

### 2020-05-03 paper/paper.tex
stuff

@@
 \end{alignat*}
-Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
+Given an $M:\SignAlg$, we denote its components by $\Con^M$, $\Ty^M$,
 $\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
@@
 Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
-difference is that we avoid the usage of impredicativity as we don't
-require that signatures form a signature algebra. It is also not a
-problem for our use case that signatures are not in the lowest
+difference is that we avoid the usage of impredicativity because it is
+not a problem for our use case that signatures are not in the lowest
 universe.
-\begin{definition}[IIT signature]
-  An IIT signature is a context in an arbitrary signature algebra
-  which is compatible with morphisms:
+\begin{definition}[IIT signature]\label{defn:sign}
+  An IIT signature is a context in an arbitrary signature algebra that
+  is compatible with morphisms:
   \begin{alignat*}{6}
@@
 \end{definition}
-Note that as $\Sign$ refers to $\SignAlg$, it lives in a higher
-universe level than $\SignAlg$. The compatibility condition says that
-if we obtain an $M$-context using ${sig}$ in model $M$ and then we
-transport it to $N$ using $f$, we get the same $N$-context as directly
-applying ${sig}$ to $N$.
+The compatibility condition says that if we obtain an $M$-context
+using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
+we get the same $N$-context as directly applying ${sig}$ to $N$.
+
+Our notion of signatures don't form a signature algebra.
+\begin{lemma}
+  There is no $M : \SignAlg$, in which $\Con^M = \Sign$.
+\end{lemma}
+\begin{proof}
+  If the $\Con$ component in $\SignAlg$ is $\Set_i$, then $\SignAlg$
+  is in $\Set_{i+1}$, but as $\Sign$ is defined as
+  $(\SignAlg\ra\dots)\times\dots$, it is at least in $\Set_{i+1}$, so
+  we can't choose $\Con^M : \Set_i$ to be $\Sign : \Set_{i+1}$.
+\end{proof}
 
@@
 model of ETT supports the theory of IIT signatures, then it supports
-all IITs.
+all IITs. In the sense of the Church encoding of \Cref{defn:sign},
+every model of ETT can describe ITT signatures. In contrast, in
+\Cref{def:theoryofsignatures}, we will require existence of an initial
+signature algebra.
 
@@
 \begin{definition}[Syntactic signatures]\label{def:syntacticsignature}
-  In a model of ETT supporting the theory of ITT signatures, elements
-  of $\Con^\I$ are called syntactic signatures. We can turn a
-  $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$ and an
-  $\Omega:\Con^\I$ into a $\Sign$ by
+  In a model of ETT supporting the theory of ITT signatures, we call
+  elements of $\Con^\I$ syntactic signatures.
+\end{definition}
+
+\begin{lemma}
+  In a model of ETT supporting the theory of ITT signatures,
+  signatures and syntactic signatures are isomorphic.
+\end{lemma}
+\begin{proof}
+  We can turn a $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$
+  and an $\Omega:\Con^\I$ into a $\Sign$ by
   $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
-  \ll\Omega\rr_N)$ where the equality in the second component comes
-  from uniqueness of the recursor.
-\end{definition}
+  (f\circ\ll\rr_M)\,\Omega = \ll\Omega\rr_N)$ where the equality in
+  the second component comes from uniqueness of the recursor (we have
+  to define composition of morphisms $\circ$ for this). The
+  compositions of these two maps are the identities: $({sig},{sigc})$
+  is mapped to
+  $(\lambda M.\ll{sig}\,I\rr_M,\dots) = (\lambda
+  M.\ll\blank\rr_M\,({sig}\,I),\dots)$ which is equal to
+  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\Omega$ is mapped
+  to $\ll\Omega\rr_\I = \Omega$ by uniqueness of $\ll\blank\rr$.
+\end{proof}
 

### 2020-05-03 paper/paper.tex
stuff

@@
   the same $=$ sign for definitional equality. We occasionally
-  indicate by $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the
+  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to the
   equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
@@
     & && && \hspace{1em}  B^\D\,(\gamma^D,\alpha^D)\,(f\,\alpha) && && \hspace{1em}\hphantom{B^\S\,}(f^D\,(a^\S\,\gamma^S\,\alpha)) \\
-    & (t\oldapp u)^\A\,\gamma := t^\A\,\gamma\,(u^\A\,\gamma) && &&  (t\oldapp u)^\D\,\gamma^D := t^\D\,\gamma^D\,(u^\D\,\gamma^D) && &&  (t\oldapp u)^\S\,\gamma^S := u^\S\,\gamma^S \# t^\S\,\gamma^\S\,(u^\A\,\gamma) \\
+    & (t\oldapp u)^\A\,\gamma := t^\A\,\gamma\,(u^\A\,\gamma) && &&  (t\oldapp u)^\D\,\gamma^D := t^\D\,\gamma^D\,(u^\D\,\gamma^D) && &&  (t\oldapp u)^\S\,\gamma^S := _{u^\S\,\gamma^S \#} t^\S\,\gamma^\S\,(u^\A\,\gamma) \\
     & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^\D\,\gamma^D\,f := && && (\Pim\,T\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:T)\ra \\
@@
 
+One may wonder what is the relationship between the two notion of
+signatures.
 \begin{lemma}
@@
   and an $\Omega:\Con^\I$ into a $\Sign$ by
-  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
-  (f\circ\ll\rr_M)\,\Omega = \ll\Omega\rr_N)$ where the equality in
-  the second component comes from uniqueness of the recursor (we have
-  to define composition of morphisms $\circ$ for this). The
+  $\Big(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.\big(f\,\ll\Omega\rr_M =
+  (f\circ\ll\blank\rr_M)\,\Omega = \ll\Omega\rr_N\big)\Big)$ where the equality
+  proof in the second component comes from uniqueness of the recursor
+  (we have to define composition of morphisms $\circ$ for this). The
   compositions of these two maps are the identities: $({sig},{sigc})$
@@
 
-First we define the term signature algebra using which we obtain the
+We will define the term signature algebra using which we obtain the
 constructors $\con$ for any IIT signature. Then we will define another
@@
 
-\begin{definition}[Term signature algebra $\A\blank\C_{\blank}$]
-  For an $M:\SignAlg$ and $\Omega:\Con^M$, we define ${\A M\C}_\Omega$
-  which we call the term signature algebra. It is equipped with
-  morphisms $\blank^\A : \SignMor\,({\A M\C}_\Omega)\,\A$ and
-  $\blank^M : \SignMor\,({\A M\C}_\Omega)\,M$.
-
-  We define ${\A M\C}_\Omega$ by listing its components $\Con$, $\Ty$,
-  $\Sub$, and so on, one per row. Each component has three parts
-  denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are just
-  reusing the corresponding components from $\A$ and $M$,
-  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
-  defined as the obvious projections. We omit the equality components,
-  as they come from UIP or are trivial. We also omit the components
-  for terms and substitutions as their $^\C$ parts are uninformative
-  equational reasonings.
+\begin{definition}[Term signature algebra $\IC_{\blank}$]
+  For an $\Omega:\Con^I$, we define $\IC_\Omega:\SignAlg$ which we
+  call the term signature algebra. It is equipped with a morphism
+  $\blank^\I : \SignMor\,(\IC_\Omega)\,\I$. We define $\IC_{\Omega}$
+  by listing its components $\Con$, $\Ty$, $\Sub$, and so on, one per
+  row. Each component has two parts denoted by $^\I$ and $^\C$. The
+  $^\I$ part just reuses the corresponding components from $\I$, and
+  thus the morphism $\blank^\I$ is defined as the obvious
+  projection. We omit the equality components, as they come from UIP
+  or are trivial. We also omit the components for terms and
+  substitutions as their $^\C$ parts are uninformative equational
+  reasonings.
   \begin{alignat*}{10}
-    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
-    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
-    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
-    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
-    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
-    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
-    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
-    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
-    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
-    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
-    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
-    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
-    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
-    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
-    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
+    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\C: \Sub^\I\,\Omega\,\Gamma^\I\ra \ll\Gamma\rr_\A \\
+    & A^\I:\Ty^\I\,\Gamma^\I && && A^\C:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra\Tm^\I\,\Omega\,(A^\I[\nu])\ra \ll A\rr_\A\,(\Gamma^\C\,\nu) \\
+    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\C:\Delta^\C\,(\sigma^\I\circ\nu)=\ll\sigma\rr_\A\,(\Gamma^\C\,\nu) \\
+    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\C : A^\C\,\nu\,(t^\I[\nu])= \ll t\rr_\A\,(\Gamma^\C\,\nu) \\
+    & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^\I\circ\nu)\,t \\
+    & \cdot^\I := \cdot^\I && && \cdot^\C\,\nu := \tt \\
+    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\C\,\nu := (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
+    & \U^\I := \U^\I && && \U^\C\nu\,a := \Tm^\I\,\Omega\,(\El^\I\,a) \\
+    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\C\,\nu\,t := _{a^\C\,\nu\#}t \\
+    & (\Pi\,a\,B)^\I := \Pi^\I\,a^\I\,B^\I && && (\Pi\,a\,B)^\C\,\nu\,t := \lambda\alpha.B^\C\,(\nu, _{a^\C\,\nu\#}\alpha)\,(t\oldapp {_{a^\C\,\nu\#}\alpha}) \\
+    & (\Pim\,T\,B)^\I := \Pim^\I\,T\,B^\I && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha)
   \end{alignat*}
 \end{definition}
-\begin{example}
+\begin{example}\label{ex:constructors}
   Now given a syntactic signature $\Omega:\Con^\I$, we get the
   constructors as an $\Omega$-algebra by
-  $(\ll\Omega\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \ll\Omega\rr^\A$. If
-  $\Omega$ is the syntactic signature for natural numbers, we get the
-  constructors as in \Cref{eg:nat_con}.
-
-  An $a : \Tm\,\Omega\,\U$ is a sort term for a signature $\Omega$. If
-  $\Omega$ is the syntactic signature for natural numbers, $a$ can
-  only be $N$ ($1$ as a de Bruijn index). If $\Omega$ is the syntactic
-  signature for $\Con$--$\Ty$ (\Cref{ex:contexts}), $a$ can be
-  $Con$, $Ty\oldapp{empty}$,
+  $\omega:=(\ll\Omega\rr_{\IC_\Omega})^\C\,\id^\I :
+  \ll\Omega\rr_\A$. If $\Omega$ is the syntactic signature for natural
+  numbers, we get the constructors as in \Cref{eg:nat_con}.
+
+  An $a : \Tm^\I\,\Omega\,\U$ is a sort term for the syntactic
+  signature $\Omega$. If $\Omega$ is the syntactic signature for
+  natural numbers, $a$ can only be $N$ ($1$ as a de Bruijn index). If
+  $\Omega$ is the syntactic signature for $\Con$--$\Ty$
+  (\Cref{ex:contexts}), $a$ can be $Con$, $Ty\oldapp{empty}$,
   $Ty\oldapp({ext}\oldapp {empty}\oldapp(U\oldapp{empty}))$, and so
   on. In any case, for such an $a$, we obtain
-  $(\ll a\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a)
-  = \ll a\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, the algebra
-  interpretation of a sort term at the constructors is equal to terms
-  of that sort.
-
-  A $t : \Tm\,\Omega\,(\El\,a)$ is a term of a sort type $a$
+  $(\ll a\rr_{\IC_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a) = \ll
+  a\rr_\A\,\omega$. That is, the algebra interpretation of a sort term
+  at the constructors is equal to terms of that sort.
+
+  A $t : \Tm^\I\,\Omega\,(\El\,a)$ is a term of a sort type $a$
   constructed using the constructors in $\Omega$. For natural numbers,
@@
   $t$, we obtain
-  $\ll t\rr^\C\,\id^\I : t = \ll
-  t\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, a constructor term is
-  equal to its algebra interpretation at the constructors. This is
-  exactly the equation needed at the end of \Cref{eg:nat_con}.
+  $(\ll t\rr_{\IC_\Omega})^\C\,\id^\I : (t = \ll
+  t\rr_\A\,\omega)$. That is, a constructor term is equal to its
+  algebra interpretation at the constructors. This is exactly the
+  equation needed at the end of \Cref{eg:nat_con}.
 \end{example}
-\begin{remark}
-  Given a syntactic signature $\Omega : \Con^\I$, we can use the term
-  signature algebra with choices other than $M = \I$. For example, if
-  $M = \A$, we usual get the Church encoding by
-  $(\ll\Omega\rr_{{\A\A\C}_{\ll\Omega\rr^\A}})^\C\,\id^\A$. In the
-  case of natural numbers, this computes (up to curry-uncurry) to
-  $(N:\Set)(z:N)(s:N\ra N)\ra N$. Using $M = {\A\M}$ (the signature
-  algebra $\A$ extended with the binary version of $\D$ also called
-  logical relations, see \cite[Section
-  5]{Kaposi:2019:CQI:3302515.3290315}), we obtain the Church encoding
-  of Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}.
-\end{remark}
 
 \begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
-  Given an $\Omega:\Con^\I$,
-  $\omega^D : \ll\Omega\rr^\D\,(\ll\Omega\rr^\C\,\id^\I)$, we define a
-  the signature algebra $\IE_{\omega^\D}$ as follows. It is equipped
-  with a morphism $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$.
-
-  We define $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$,
-  $\Sub$, and so on, one per row. Each component has two parts denoted
-  by $^\I$, and $^\E$. The $^\I$ part just reuses the corresponding
+  Given an $\Omega:\Con^\I$, we use the abbreviation
+  $\omega := \ll\Omega\rr_{\IC_\Omega}\,\id^\I$ as in
+  \Cref{ex:constructors}. Assuming an
+  $\omega^D : (\ll\Omega\rr_{\ADS})^\D\,\omega$, we define the
+  signature algebra $\IE_{\omega^\D}$. It is equipped with a morphism
+  $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$. We define
+  $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$, $\Sub$,
+  and so on, one per row. Each component has two parts denoted by
+  $^\I$ and $^\E$. The $^\I$ part just reuses the corresponding
   components of $\I$, thus the morphism $\blank^\I$ is defined as the
@@
   reasonings.
+  \begin{alignat*}{10}
+    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra \ll\Gamma\rr^\S\,(\ll\nu\rr^\A\,\omega)\,(\ll\nu\rr^\D\,\omega^D) \\
+    & A^\I:\Ty^\I\,\Gamma^\I && && A^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)(t:\Tm^\I\,\Omega\,(A^\I[\nu]))\ra \\
+    & && && \hspace{1em} \ll A\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega)\,(\ll t\rr^\D\,\omega^D) \\
+    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\E:\Delta^\E\,(\sigma^\I\circ\nu)=\ll\sigma\rr^\S\,(\Gamma^\E\,\nu) \\
+    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\E : A^\E\,\nu\,(t^\I[\nu])= \ll t\rr^\S\,(\Gamma^\E\,\nu) \\
+    & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\E\,\nu\,t := A^\E\,(\sigma^\I\circ\nu)\,t \\
+    & \cdot^\I := \cdot^\I && && \cdot^\E\,\nu := \tt \\
+    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\E\,\nu := (\Gamma^\E\,(\pi_1\,\nu), A^\E\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
+    & \U^\I := \U^\I && && \U^\E\nu\,a := \lambda\alpha . _{\ll\alpha\rr^\C\,\id\#}\big(\ll_{\ll\alpha\rr^\C\,\id\#}\alpha\rr^\D\,\omega^D\big) \\
+    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\E\,\nu\,t := \big(\ll a\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega) \overset{\ll t\rr^\C\,\id}{=} \ll a\rr^\S\,(\Gamma^\E\,\nu)\,t \overset{a^\E\,\nu}{=} \ll t\rr^\D\,\omega^D\big) \\
+    & (\Pi\,a\,B)^\I := \Pi^\I\,a^\I\,B^\I && && (\Pi\,a\,B)^\E\,\nu\,t := \\
+    & && && \hspace{1em} \lambda\alpha.{_{\ll\alpha\rr^\C\,\id\#}}\big(B^\E\,(\nu,_{\ll a\rr^\C\,\id,\ll\nu\rr^\C\,\id\#}\alpha)\,(t\oldapp {_{\ll a\rr^\C\,\id,\ll\nu\rr^\C\,\id\#}u})\big) \\
+    & (\Pim\,T\,B)^\I := \Pim^\I\,T\,B^\I && && (\Pim\,T\,B)^\E\,\nu\,t := \lambda\alpha.(B\,\alpha)^\E\,\nu\,(t\appm\alpha)
+  \end{alignat*}
+\end{definition}
+\begin{example}
+  Given the assumptions $\Omega$, $\omega^D$ of $\IE$, we obtain the
+  eliminator by
+  $\ll\Omega\rr_{\IE_{\omega^\D}}\,\id^\I :
+  \ll\Omega\rr^\S\,\omega\,\omega^D$. The eliminator is a section of
+  the displayed algebra $\omega^D$, that is, a dependent function
+  together with equalities witnessing that all the operations are
+  preserved. If $\Omega$ is the syntactic signature for natural
+  numbers, we get the eliminator of \Cref{eg:nat_con}.
+
+  For a sort term $a:\Tm^\I\,\Omega\,\U$, the interpretation
+  $(\ll a\rr_{\IE_{\omega^D}})^\E\,\id$ says that
+  $(\lambda\alpha.\ll\alpha\rr^\D\,\omega^D) = \ll
+  a\rr^\S\,(\ll\Omega\rr^\E\,\id)$, that is, the function for the sort
+  $a$ in the eliminator section is the displayed algebra
+  interpretation at $\omega^D$ (motives and methods). For natural
+  numbers, this is the same as
+  $\big(\lambda n.\ll n\rr^D\,(N^D,z^D,s^D)\big) = \big(\lambda
+  n.\mathsf{elimNat}\,(N^D,z^D,s^D)\,n)\big)$.
+
+  The interpretation of a constructor term
+  $t:\Tm^\I\,\Omega\,(\El\,a)$ is uninteresting as it provides an
+  equality between two different equality proofs of the computation
+  ($\beta$) rule for $t$.
+\end{example}
 
 xxx
-\end{definition}
 

### 2020-05-03 paper/universaliit.tex
stuff

@@
 defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\Gamma
-= \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
+= \model{\Gamma})\times (\rec\,A = _{e\#}\model{A})$.
 % whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
@@
   &&
- := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = e_E \#
+ := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = _{e_E \#}
  \model{\epsilon})
@@
     & && \qquad
-    (\delta = e_E \# \model\sigma \model{\cons} \model{t} )
+    (\delta = _{e_E \#} \model\sigma \model{\cons} \model{t} )
     \\
@@
     & && \qquad
-     (e_B : \model{B} = e_\Delta \# \model{\wk}\,\model{A}) \times
-     % \\
-     % & && \qquad
-     (\model{t} = e_\Delta,e_B \# \model{\vz})
+     (e_B : \model{B} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times
+     % \\
+     % & && \qquad
+     (\model{t} = _{e_\Delta,e_B \#} \model{\vz})
   \\
@@
     & && \qquad
-     (e_C : \model{C} = e_\Delta \# \model{\wk}\,\model{A}) \times
+     (e_C : \model{C} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times
      \\
      & && \qquad
-     (\model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n})
+     (\model{t} = _{e_\Delta,e_C \#} \model{\vs} \, \model{n})
   \\
@@
       \qquad
-      (\model{x} = e_C \# \model{t} \model{\oldapp}\model{u})
+      (\model{x} = _{e_C \#} \model{t} \model{\oldapp}\model{u})
     \\
@@
       \times
-      (\model{x} = e_B \# \model{t}\model{\hat{\oldapp}}u)
+      (\model{x} = _{e_B \#} \model{t}\model{\hat{\oldapp}}u)
 \end{alignat*}

### 2020-05-03 paper/paper.tex
sec2,3

@@
 
-xxx
-
-We construct all IITs describable by the theory of IIT signatures through the
-term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
-shown that from the syntax of the theory of signatures for finitary QIITs, one
-can construct any particular finitary QIIT. The idea is that for a signature
-$\Gamma$, the initial algebra can be built from sets of terms of the form
-$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
-numbers
-\[
-  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N\big),
-\]
-we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
-since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.
-
-In this section, we only need to check that the signatures and term model
-construction in \cite{Kaposi:2019:CQI:3302515.3290315} restrict to IITs,
-i.e.\ that from a syntax for the theory of IIT signatures all finitary IITs are
-constructible.
-
-\begin{enumerate}
-\item
-  Finitary IIT signatures are obtained exactly by dropping equality constructors from
-  finitary QIIT signatures. It follows that any model for the theory of QIIT signatures
-  restricts to IITs, hence we inherit the categorical semantics of QIITs.
-\item
-  The term model construction also restricts in a straightforward way. For QIITs,
-  term algebras are constructed by induction on the syntax of signatures, and
-  then another induction constructs the eliminator (i.e.\ dependent induction
-  principle). In both cases, simply dropping equality constructors from the
-  induction yields restriction to IITs.
-\end{enumerate}
-
-However, we shall mention that the Agda formalisation for the current paper
-cannot be directly plugged into the Agda code for
-\cite{Kaposi:2019:CQI:3302515.3290315} (which includes most of the term model
-construction). One reason is that the QIIT formalisation uses strict computation
-rules (given by Agda rewrite rules) for induction over signatures, while here we show propositional
-computation. This mismatch can be in principle remedied by noting that both
-formalisations use UIP and function extensionality, hence we can switch between
-strict and propositional equalities, via the known translations between
-extensional and intensional type
-theories \cite{hofmann95extensional,winterhalter2019eliminating}.
-
-Also, the current paper proves initiality, i.e.\ unique recursion for
-signatures, while \cite{Kaposi:2019:CQI:3302515.3290315} uses dependent
-elimination. We expect that the two notions are equivalent. An extension of
-\cite{Kaposi:2019:CQI:3302515.3290315} to large infinitary QIITs would entail
-this equivalence, since that would cover the (large, infinitary) theory of IIT
-signatures.
+\begin{theorem}
+  If a model of type theory supports the theory of IIT signatures,
+  then it supports all IITs.
+\end{theorem}
+\begin{proof}
+  For a signature $({sig},{sigc})$, we define constructors as
+  \[
+    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
+  \]
+  This typechecks as
+  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
+  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
+  eliminator by
+  and an
+  \[
+    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
+  \]
+  This typechecks firstly because the type of $\gamma^D$ matches the
+  type of the parameter of $\IE$:
+  \[
+    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
+    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
+    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
+  \]
+  and the result also has the correct type:
+  \[
+    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
+  \]
+\end{proof}
 

### 2020-05-03 paper/paper.tex
stuff

@@
 Thematic Excellence Programme funding scheme, Project
-no. ED18-1-2019-0030 (Application-specific highly reliable IT
+no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
 solutions), by the New National Excellence Program of the Ministry
-for Innovation and Technology, Project no. ÚNKP-19-4-ELTE-874, and
+for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
 by the Bolyai Fellowship of the Hungarian Academy of Sciences,
-Project no.
-BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\acknowledgements{The authors would like to thank Thorsten Altenkirch, Simon Boulier, Fredrik Nordvall-Forsberg and Jakob von Raumer for discussions on the topics of this paper.}%optional
+\acknowledgements{The authors would like to thank Thorsten Altenkirch, Rafa{\"e}l Bocquet, Simon Boulier, Fredrik Nordvall-Forsberg and Jakob von Raumer for discussions on the topics of this paper.
+We also thank the anonymous reviewers for their helpful comments and suggestions.}%optional
 
@@
 \begin{abstract}
-  Inductive-inductive types (IITs) are a generalisation of inductive types in
+  TODO: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
   type theory. They allow the mutual definition of types with multiple sorts
@@
 theory itself, and the main example is still the Chapman-style
-\cite{chapman09eatitself} syntax for a type theory. A minimsed version
+\cite{chapman09eatitself} syntax for a type theory. A minimised version
 is the IIT of contexts and types given by the following constructors.
@@
 $\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
-itself. Another unique feature of IITs (which also shows up in higher inductive
+itself. Another unique feature of IITs (which also holds for higher inductive
 types \cite{HoTTbook}) is that later constructors can refer to previous
@@
 %  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
-\end{alignat*}
+\end{alignat*} xxx
 Above we used implicit quantifications for $\Gamma:\Con$ and
@@
 
-\section{Constructing IITs from the Theory of IIT Signatures}
+\section{Constructing all IITs from the Theory of IIT Signatures}
 \label{sec:constructingiits}
@@
 
-\begin{theorem}
+\begin{theorem}\label{th:tosToIITs}
   If a model of type theory supports the theory of IIT signatures,
@@
 \label{sec:ambroise}
+
+In this section we show that any model of ETT which supports indexed
+W-types also supports the theory of signatures, and as a consequence
+of \Cref{th:tosToIITs}, all IITs. For this, we work in the internal
+language of a model of ETT supporting indexed W-types
+\cite{indexedcont}. Indexed W-types correspond to the usual notion of
+(possibly mutual) indexed inductive types. We will use Agda-style
+notation for defining such inductive families: we list the sorts and
+constructors and use pattern matching when eliminating from them. qFor
+an encoding of mutual inductive families as families as indexed
+W-types, see e.g.\ \cite{mutual}.
 

### 2020-05-03 paper/paper.tex
make uppercase greek letters italic

@@
   & \mathsf{empty} && : \Con \\
-  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
-  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
-  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
-%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
+  & \mathsf{ext} && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
+  & \mathsf{U} && : (\GG:\Con)\ra\Ty\,\GG \\
+  & \mathsf{El} && : (\GG:\Con)\ra\Ty\,(\mathsf{ext}\,\GG\,(\mathsf{U}\,\GG))
+%  & \mathsf{Pi} && : (\GG:\Con)(A:\Ty\,\GG)\ra\Ty\,(\mathsf{ext}\,\GG\,A)\ra\,\Ty\,\GG
 \end{alignat*}
@@
   & Con^D && : \Con\ra\Set \\
-  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
+  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
   & {empty}^D && : Con^D\,\mathsf{empty} \\
-  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
-  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
-  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
-%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
+  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
+  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
+  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
+%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
 \end{alignat*} xxx
-Above we used implicit quantifications for $\Gamma:\Con$ and
-$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
-parameter $\Gamma$ before its explicit parameter of type
-$Con^D\,\Gamma$.
+Above we used implicit quantifications for $\GG:\Con$ and
+$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
+parameter $\GG$ before its explicit parameter of type
+$Con^D\,\GG$.
 
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
-  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)\,A
+  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
+  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
 \end{alignat*}
@@
   & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
-  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = {ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
-  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
-  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
 \end{alignat*}
@@
   & Con^S && : \Con\ra\Set \\
-  & Ty^S && : \Ty\,\Gamma\ra\Set \\
+  & Ty^S && : \Ty\,\GG\ra\Set \\
   & {empty}^S && : Con^S\,\mathsf{empty} \\
-  & {ext}^S && : Con^S\,\Gamma\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\Gamma\,A) \\
-  & U^S && : Con^S\,\Gamma\ra Ty^S\,(\mathsf{U}\,\Gamma) \\
-  & El^S && : Con^S\,\Gamma\ra\,Ty^S\,(\mathsf{El}\,\Gamma) \\
-  & \mathsf{selimCon} && : (\Gamma:\Con)\ra Con^S\,\Gamma \\
-  & \mathsf{selimTy} && : (A:\Ty\,\Gamma)\ra Ty^S\,A
+  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
+  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
+  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
+  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
+  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
 \end{alignat*}
@@
   & \ll\blank\rr && : \Con\ra\Set_1 \\
-  & \ll\blank\rr && : \ll\Gamma\rr\ra\Set_1 \\
+  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
   & \ll\mathsf{empty}\rr && := \top \\
-  & \ll\mathsf{ext}\,\Gamma\,A\rr && := (\gamma:\ll\Gamma\rr)\times\ll A\rr\,\gamma \\
-  & \ll\mathsf{U}\,\Gamma\rr\,\gamma && := \Set \\
-  & \ll\mathsf{El}\,\Gamma\rr\,(\gamma,X) && := X
+  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
+  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
+  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
 \end{alignat*}
@@
 \[
-  \Pi_\infty : (\Gamma : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\Gamma) \rightarrow \Ty\,\Gamma
+  \Pi_\infty : (\GG : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\GG) \rightarrow \Ty\,\GG
 \]
@@
   & \Sub  && : \Con\ra\Con\ra\Set \\
-  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
-  & \id && : \Sub\,\Gamma\,\Gamma \\
-  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
+  & \Tm  && : (\GG:\Con)\ra\Ty\,\GG\ra\Set \\
+  & \id && : \Sub\,\GG\,\GG \\
+  & \blank\circ\blank && : \Sub\,\GT\,\GD\ra\Sub\,\GG\,\GT\ra\Sub\,\GG\,\GD \\
   & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
@@
   & \idr && : \sigma\circ\id = \sigma \\
-  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
-  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
+  & \blank[\blank] && : \Ty\,\GD\ra\Sub\,\GG\,\GD\ra\Ty\,\GG \\
+  & \blank[\blank] && : \Tm\,\GD\,A\ra(\sigma:\Sub\,\GG\,\GD)\ra\Tm\,\GG\,(A[\sigma]) \\
   & [\id] && : A[\id] = A \\
@@
   & \boldsymbol{\cdot} && : \Con \\
-  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
-  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
-  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
-  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
-  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
-  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
+  & \epsilon && : \Sub\,\GG\,\boldsymbol{\cdot} \\
+  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\GG\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
+  & \blank\ext\blank && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
+  & \blank,\blank && : (\sigma:\Sub\,\GG\,\GD)\ra\Tm\,\GG\,(A[\sigma])\ra\Sub\,\GG\,(\GD\ext A) \\
+  & \pi_1 && : \Sub\,\GG\,(\GD\ext A)\ra \Sub\,\GG\,\GD \\
+  & \pi_2 && : (\sigma : \Sub\,\GG\,(\GD\ext A))\ra \Tm\,\GG\,(A[\pi_1 \sigma]) \\
   & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
@@
   & \rlap{(2) Universe} \\
-  & \U && : \Ty\,\Gamma \\
-  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
+  & \U && : \Ty\,\GG \\
+  & \El && : \Tm\,\GG\,\U\ra\Ty\,\GG \\
   & {\U[]} && : \U[\sigma] = \U \\
@@
   & \rlap{(3) Inductive parameters} \\
-  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
-  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
-  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
+  & \Pi && : (a:\Tm\,\GG\,\U)\ra\Ty\,(\GG\ext\El\,a)\ra\Ty\,\GG \\
+  & \blank\oldapp\blank && : \Tm\,\GG\,(\Pi\,a\,B)\ra (u : \Tm\,\GG\,(\El\,a))
+  \ra \Tm\,\GG\,(\El\,(B[\id,\,u])) \\
   & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
@@
   & \rlap{(4) External parameters} \\
-  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
-  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
+  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\GG)\ra\Ty\,\GG \\
+  & \blank\appm \blank && : \Tm\,\GG\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\GG\,(B\,\alpha) \\
   & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
@@
 \begin{definition}[Abbreviations]\label{def:abbrevs}
-  For a signature algebra, we use $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$
+  For a signature algebra, we use $\wk : \Sub\,(\GG\ext A)\,\GG$
   to mean $\pi_1\,\id$. We recover de Bruijn indices by setting
@@
     & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
-    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
-    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
-    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
+    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\GG : Con)\,(Ty\oldapp\GG \Ra \El\,Con)\ext                         \\
+    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\GG:Con)\,(\El\,(Ty\oldapp\GG))\,\ext                                 \\
+    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\GG : Con)\,(\El\,(Ty\oldapp(ext\oldapp\GG\oldapp (U\oldapp\GG))))
   \end{alignat*}
@@
   & \Con && : \Con^M && \ra\Con^N \\
-  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
-  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
-  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
+  & \Ty  && : \Ty^M\,\GG && \ra\Ty^N\,(\Con\,\GG) \\
+  & \Sub  && : \Sub^M\,\GG\,\GD && \ra\Sub^N(\Con\,\GG)\,(\Con\,\GD) \\
+  & \Tm  && : \Tm^M\,\GG\,A && \ra\Tm^N\,(\Con\,\GG)\,(\Ty\,A) \\
   & \id && : \Sub\,\id^M && = \id^N \\
@@
   & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
-  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
+  & \ext && : \Con\,(\GG\ext^M A) && = \Con\,\GG\ext^N\Ty\,A \\
   & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
@@
   \begin{alignat*}{10}
-    & (\Gamma^\A:\Set) && \hspace{-0.3em}\times && (\Gamma^\D:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set) \\
-    & (A^\A:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) && \hspace{-0.3em}\times && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra \\
+    & (\GG^\A:\Set) && \hspace{-0.3em}\times && (\GG^\D:\GG^\A\ra\Set) && \hspace{-0.3em}\times && (\GG^\S:(\gamma:\GG^\A)\ra\GG^\D\,\gamma\ra\Set) \\
+    & (A^\A:\GG^\A\ra\Set) && \hspace{-0.3em}\times && (A^\D:\GG^\D\,\gamma\ra A^\A\,\gamma\ra\Set) && \hspace{-0.3em}\times && (A^\S:\GG^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra \\
     & && && && && \hspace{1em} A^\D\,\gamma^D\,\alpha\ra\Set) \\
-    & (\sigma^\A:\Gamma^\A\ra\Delta^\A) && \hspace{-0.3em}\times && (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) && \hspace{-0.3em}\times && (\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra \\
-    & && && && && \hspace{1em} \Delta^\S\,(\sigma^\A\,\gamma)\,(\sigma^\D\,\gamma^D)) \\
-    & (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) && \hspace{-0.3em}\times && (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra && \hspace{-0.3em}\times && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra \\
+    & (\sigma^\A:\GG^\A\ra\GD^\A) && \hspace{-0.3em}\times && (\sigma^\D:\GG^\D\,\gamma\ra\GD^\D\,(\sigma^\A\,\gamma)) && \hspace{-0.3em}\times && (\sigma^\S:\GG^\S\,\gamma\,\gamma^D\ra \\
+    & && && && && \hspace{1em} \GD^\S\,(\sigma^\A\,\gamma)\,(\sigma^\D\,\gamma^D)) \\
+    & (t^\A:(\gamma:\GG^\A)\ra A^\A\,\gamma) && \hspace{-0.3em}\times && (t^\D:(\gamma^D:\GG^\D\,\gamma)\ra && \hspace{-0.3em}\times && (t^\S:(\gamma^S:\GG^\S\,\gamma\,\gamma^D)\ra \\
     & && && \hspace{1em} A^\D\,\gamma^D\,(t^\A\,\gamma)) && && \hspace{1em}  A^\S\,(t^\A\,\gamma)\,(t^\D\,\gamma^D)) \\
@@
     & \epsilon^\A\,\_ := \tt && && \epsilon^\D\,\_ := \tt && && \epsilon^\S\,\_ := \tt \\
-    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^\D\,(\gamma,\alpha) :=  && && (\Gamma\rhd A)^\S\,(\gamma,\alpha)\,(\gamma^D,\alpha^D) := \\
-    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} (\gamma^D:\Gamma^\D\,\gamma)\times A^\D\,\gamma^D\,\alpha && && \hspace{1em} (\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\times A^\S\,\gamma^S\,\alpha\,\alpha^D \\
+    & (\GG\rhd A)^\A := && && (\GG\rhd A)^\D\,(\gamma,\alpha) :=  && && (\GG\rhd A)^\S\,(\gamma,\alpha)\,(\gamma^D,\alpha^D) := \\
+    & \hspace{1em} (\gamma:\GG^\A)\times A^\A\,\gamma && && \hspace{1em} (\gamma^D:\GG^\D\,\gamma)\times A^\D\,\gamma^D\,\alpha && && \hspace{1em} (\gamma^S:\GG^\S\,\gamma\,\gamma^D)\times A^\S\,\gamma^S\,\alpha\,\alpha^D \\
     & (\sigma,t)^\A\,\gamma := (\sigma^\A\,\gamma,t^\A\,\gamma) && && (\sigma,t)^\D\,\gamma^D := (\sigma^\D\,\gamma^D,t^\D\,\gamma^D) && && (\sigma,t)^\S\,\gamma^S := (\sigma^\S\,\gamma^S,t^\S\,\gamma^S) \\
@@
 given by a CwF. Now, from the category part of the CwF, we only have
-objects ($\Gamma^\A$), and from the families, we have the families for
-types $\Gamma^\D$ and terms $\Gamma^\S$. Types are the corresponding
+objects ($\GG^\A$), and from the families, we have the families for
+types $\GG^\D$ and terms $\GG^\S$. Types are the corresponding
 parts of displayed CwFs, substitutions are parts of CwF morphisms,
@@
 $\El\,a$ is the part of the discrete displayed CwF coming from $a$
-(which is a CwF-morphism from $\Gamma$ to the CwF of sets). $\Pi$ is
+(which is a CwF-morphism from $\GG$ to the CwF of sets). $\Pi$ is
 given by a dependent product of displayed CwFs where it is essential
@@
   (\Cref{def:ads}), that is, $\Con^\A := \Set$,
-  $\Ty^\A\,\Gamma := \Gamma\ra\Set$,
-  $\Sub^\A\,\Gamma\,\Delta := \Gamma\ra\Delta$, and so on. There is a
+  $\Ty^\A\,\GG := \GG\ra\Set$,
+  $\Sub^\A\,\GG\,\GD := \GG\ra\GD$, and so on. There is a
   morphism from $\ADS$ to $\A$ defined by $\blank^\A$ at each
@@
   We can turn a $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$
-  and an $\Omega:\Con^\I$ into a $\Sign$ by
-  $\Big(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.\big(f\,\ll\Omega\rr_M =
-  (f\circ\ll\blank\rr_M)\,\Omega = \ll\Omega\rr_N\big)\Big)$ where the equality
+  and an $\GO:\Con^\I$ into a $\Sign$ by
+  $\Big(\lambda M.\ll\GO\rr_M, \lambda M\,N\,f.\big(f\,\ll\GO\rr_M =
+  (f\circ\ll\blank\rr_M)\,\GO = \ll\GO\rr_N\big)\Big)$ where the equality
   proof in the second component comes from uniqueness of the recursor
@@
   M.\ll\blank\rr_M\,({sig}\,I),\dots)$ which is equal to
-  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\Omega$ is mapped
-  to $\ll\Omega\rr_\I = \Omega$ by uniqueness of $\ll\blank\rr$.
+  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\GO$ is mapped
+  to $\ll\GO\rr_\I = \GO$ by uniqueness of $\ll\blank\rr$.
 \end{proof}
@@
 \begin{definition}[Term signature algebra $\IC_{\blank}$]
-  For an $\Omega:\Con^I$, we define $\IC_\Omega:\SignAlg$ which we
+  For an $\GO:\Con^I$, we define $\IC_\GO:\SignAlg$ which we
   call the term signature algebra. It is equipped with a morphism
-  $\blank^\I : \SignMor\,(\IC_\Omega)\,\I$. We define $\IC_{\Omega}$
+  $\blank^\I : \SignMor\,(\IC_\GO)\,\I$. We define $\IC_{\GO}$
   by listing its components $\Con$, $\Ty$, $\Sub$, and so on, one per
@@
   \begin{alignat*}{10}
-    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\C: \Sub^\I\,\Omega\,\Gamma^\I\ra \ll\Gamma\rr_\A \\
-    & A^\I:\Ty^\I\,\Gamma^\I && && A^\C:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra\Tm^\I\,\Omega\,(A^\I[\nu])\ra \ll A\rr_\A\,(\Gamma^\C\,\nu) \\
-    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\C:\Delta^\C\,(\sigma^\I\circ\nu)=\ll\sigma\rr_\A\,(\Gamma^\C\,\nu) \\
-    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\C : A^\C\,\nu\,(t^\I[\nu])= \ll t\rr_\A\,(\Gamma^\C\,\nu) \\
+    & \GG^\I:\Con^\I && \hspace{1em} && \GG^\C: \Sub^\I\,\GO\,\GG^\I\ra \ll\GG\rr_\A \\
+    & A^\I:\Ty^\I\,\GG^\I && && A^\C:(\nu:\Sub^\I\,\GO\,\GG^\I)\ra\Tm^\I\,\GO\,(A^\I[\nu])\ra \ll A\rr_\A\,(\GG^\C\,\nu) \\
+    & \sigma^\I:\Sub^\I\,\GG^\I\,\GD^\I && && \sigma^\C:\GD^\C\,(\sigma^\I\circ\nu)=\ll\sigma\rr_\A\,(\GG^\C\,\nu) \\
+    & t^\I:\Tm^\I\,\GG^\I\,A^\I && && t^\C : A^\C\,\nu\,(t^\I[\nu])= \ll t\rr_\A\,(\GG^\C\,\nu) \\
     & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^\I\circ\nu)\,t \\
     & \cdot^\I := \cdot^\I && && \cdot^\C\,\nu := \tt \\
-    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\C\,\nu := (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
-    & \U^\I := \U^\I && && \U^\C\nu\,a := \Tm^\I\,\Omega\,(\El^\I\,a) \\
+    & (\GG\rhd A)^\I := \GG^\I\rhd^\I A^\I && && (\GG\rhd A)^\C\,\nu := (\GG^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
+    & \U^\I := \U^\I && && \U^\C\nu\,a := \Tm^\I\,\GO\,(\El^\I\,a) \\
     & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\C\,\nu\,t := _{a^\C\,\nu\#}t \\
@@
 \begin{example}\label{ex:constructors}
-  Now given a syntactic signature $\Omega:\Con^\I$, we get the
-  constructors as an $\Omega$-algebra by
-  $\omega:=(\ll\Omega\rr_{\IC_\Omega})^\C\,\id^\I :
-  \ll\Omega\rr_\A$. If $\Omega$ is the syntactic signature for natural
+  Now given a syntactic signature $\GO:\Con^\I$, we get the
+  constructors as an $\GO$-algebra by
+  $\omega:=(\ll\GO\rr_{\IC_\GO})^\C\,\id^\I :
+  \ll\GO\rr_\A$. If $\GO$ is the syntactic signature for natural
   numbers, we get the constructors as in \Cref{eg:nat_con}.
 
-  An $a : \Tm^\I\,\Omega\,\U$ is a sort term for the syntactic
-  signature $\Omega$. If $\Omega$ is the syntactic signature for
+  An $a : \Tm^\I\,\GO\,\U$ is a sort term for the syntactic
+  signature $\GO$. If $\GO$ is the syntactic signature for
   natural numbers, $a$ can only be $N$ ($1$ as a de Bruijn index). If
-  $\Omega$ is the syntactic signature for $\Con$--$\Ty$
+  $\GO$ is the syntactic signature for $\Con$--$\Ty$
   (\Cref{ex:contexts}), $a$ can be $Con$, $Ty\oldapp{empty}$,
@@
   on. In any case, for such an $a$, we obtain
-  $(\ll a\rr_{\IC_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a) = \ll
+  $(\ll a\rr_{\IC_\GO})^\C\,\id^\I : \Tm^\I\,\GO\,(\El\,a) = \ll
   a\rr_\A\,\omega$. That is, the algebra interpretation of a sort term
@@
 
-  A $t : \Tm^\I\,\Omega\,(\El\,a)$ is a term of a sort type $a$
-  constructed using the constructors in $\Omega$. For natural numbers,
+  A $t : \Tm^\I\,\GO\,(\El\,a)$ is a term of a sort type $a$
+  constructed using the constructors in $\GO$. For natural numbers,
   such a $t$ can only be $s$ applied iteratively to $z$. For such a
   $t$, we obtain
-  $(\ll t\rr_{\IC_\Omega})^\C\,\id^\I : (t = \ll
+  $(\ll t\rr_{\IC_\GO})^\C\,\id^\I : (t = \ll
   t\rr_\A\,\omega)$. That is, a constructor term is equal to its
@@
 \begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
-  Given an $\Omega:\Con^\I$, we use the abbreviation
-  $\omega := \ll\Omega\rr_{\IC_\Omega}\,\id^\I$ as in
+  Given an $\GO:\Con^\I$, we use the abbreviation
+  $\omega := \ll\GO\rr_{\IC_\GO}\,\id^\I$ as in
   \Cref{ex:constructors}. Assuming an
-  $\omega^D : (\ll\Omega\rr_{\ADS})^\D\,\omega$, we define the
+  $\omega^D : (\ll\GO\rr_{\ADS})^\D\,\omega$, we define the
   signature algebra $\IE_{\omega^\D}$. It is equipped with a morphism
@@
   \begin{alignat*}{10}
-    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra \ll\Gamma\rr^\S\,(\ll\nu\rr^\A\,\omega)\,(\ll\nu\rr^\D\,\omega^D) \\
-    & A^\I:\Ty^\I\,\Gamma^\I && && A^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)(t:\Tm^\I\,\Omega\,(A^\I[\nu]))\ra \\
-    & && && \hspace{1em} \ll A\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega)\,(\ll t\rr^\D\,\omega^D) \\
-    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\E:\Delta^\E\,(\sigma^\I\circ\nu)=\ll\sigma\rr^\S\,(\Gamma^\E\,\nu) \\
-    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\E : A^\E\,\nu\,(t^\I[\nu])= \ll t\rr^\S\,(\Gamma^\E\,\nu) \\
+    & \GG^\I:\Con^\I && \hspace{1em} && \GG^\E:(\nu:\Sub^\I\,\GO\,\GG^\I)\ra \ll\GG\rr^\S\,(\ll\nu\rr^\A\,\omega)\,(\ll\nu\rr^\D\,\omega^D) \\
+    & A^\I:\Ty^\I\,\GG^\I && && A^\E:(\nu:\Sub^\I\,\GO\,\GG^\I)(t:\Tm^\I\,\GO\,(A^\I[\nu]))\ra \\
+    & && && \hspace{1em} \ll A\rr^\S\,(\GG^\E\,\nu)\,(\ll t\rr^\A\,\omega)\,(\ll t\rr^\D\,\omega^D) \\
+    & \sigma^\I:\Sub^\I\,\GG^\I\,\GD^\I && && \sigma^\E:\GD^\E\,(\sigma^\I\circ\nu)=\ll\sigma\rr^\S\,(\GG^\E\,\nu) \\
+    & t^\I:\Tm^\I\,\GG^\I\,A^\I && && t^\E : A^\E\,\nu\,(t^\I[\nu])= \ll t\rr^\S\,(\GG^\E\,\nu) \\
     & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\E\,\nu\,t := A^\E\,(\sigma^\I\circ\nu)\,t \\
     & \cdot^\I := \cdot^\I && && \cdot^\E\,\nu := \tt \\
-    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\E\,\nu := (\Gamma^\E\,(\pi_1\,\nu), A^\E\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
+    & (\GG\rhd A)^\I := \GG^\I\rhd^\I A^\I && && (\GG\rhd A)^\E\,\nu := (\GG^\E\,(\pi_1\,\nu), A^\E\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
     & \U^\I := \U^\I && && \U^\E\nu\,a := \lambda\alpha . _{\ll\alpha\rr^\C\,\id\#}\big(\ll_{\ll\alpha\rr^\C\,\id\#}\alpha\rr^\D\,\omega^D\big) \\
-    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\E\,\nu\,t := \big(\ll a\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega) \overset{\ll t\rr^\C\,\id}{=} \ll a\rr^\S\,(\Gamma^\E\,\nu)\,t \overset{a^\E\,\nu}{=} \ll t\rr^\D\,\omega^D\big) \\
+    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\E\,\nu\,t := \big(\ll a\rr^\S\,(\GG^\E\,\nu)\,(\ll t\rr^\A\,\omega) \overset{\ll t\rr^\C\,\id}{=} \ll a\rr^\S\,(\GG^\E\,\nu)\,t \overset{a^\E\,\nu}{=} \ll t\rr^\D\,\omega^D\big) \\
     & (\Pi\,a\,B)^\I := \Pi^\I\,a^\I\,B^\I && && (\Pi\,a\,B)^\E\,\nu\,t := \\
@@
 \begin{example}
-  Given the assumptions $\Omega$, $\omega^D$ of $\IE$, we obtain the
+  Given the assumptions $\GO$, $\omega^D$ of $\IE$, we obtain the
   eliminator by
-  $\ll\Omega\rr_{\IE_{\omega^\D}}\,\id^\I :
-  \ll\Omega\rr^\S\,\omega\,\omega^D$. The eliminator is a section of
+  $\ll\GO\rr_{\IE_{\omega^\D}}\,\id^\I :
+  \ll\GO\rr^\S\,\omega\,\omega^D$. The eliminator is a section of
   the displayed algebra $\omega^D$, that is, a dependent function
   together with equalities witnessing that all the operations are
-  preserved. If $\Omega$ is the syntactic signature for natural
+  preserved. If $\GO$ is the syntactic signature for natural
   numbers, we get the eliminator of \Cref{eg:nat_con}.
 
-  For a sort term $a:\Tm^\I\,\Omega\,\U$, the interpretation
+  For a sort term $a:\Tm^\I\,\GO\,\U$, the interpretation
   $(\ll a\rr_{\IE_{\omega^D}})^\E\,\id$ says that
   $(\lambda\alpha.\ll\alpha\rr^\D\,\omega^D) = \ll
-  a\rr^\S\,(\ll\Omega\rr^\E\,\id)$, that is, the function for the sort
+  a\rr^\S\,(\ll\GO\rr^\E\,\id)$, that is, the function for the sort
   $a$ in the eliminator section is the displayed algebra
@@
   The interpretation of a constructor term
-  $t:\Tm^\I\,\Omega\,(\El\,a)$ is uninteresting as it provides an
+  $t:\Tm^\I\,\GO\,(\El\,a)$ is uninteresting as it provides an
   equality between two different equality proofs of the computation

### 2020-05-03 paper/universaliit.tex
make uppercase greek letters italic

@@
   \item defining a relation $\blank \sim \blank$ between the well-typed syntax
-    and a given model. The idea is that given a syntactic context $\Gamma$ and a
-    semantic context $\model{\Gamma}$ of the model $M$, we have
-    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
-    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
-    % semantic context $\model{\Gamma}$ of
-    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
+    and a given model. The idea is that given a syntactic context $\GG$ and a
+    semantic context $\model{\GG}$ of the model $M$, we have
+    $\GG \sim \model{\GG}$ if and only if $\rec\,\,\GG
+    = \model{\GG}$, and similarly for types, terms, and substitutions;
+    % semantic context $\model{\GG}$ of
+    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
     % enjoyed by the initial morphism from the
@@
     $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
-    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
+    $\GG$ we have $\GG\sim f\,\GG$.
       % TODO donner composantes sur les contextes
@@
 % (which appears in the typing rule of a weakening occuring in the middle of a context),
-% and the length $\length{\Gamma}$ of a
-% context $\Gamma$:
+% and the length $\length{\GG}$ of a
+% context $\GG$:
 \begin{alignat*}{5}
   & \blank \merge \blank && :  \Conu\ra\Conu\ra\Conu \\
-  & \Gamma \merge \cdot && :=  \Gamma \\
-  & \Gamma \merge (\Delta \extu A) && :=  (\Gamma \merge \Delta)\extu A
+  & \GG \merge \cdot && :=  \GG \\
+  & \GG \merge (\GD \extu A) && :=  (\GG \merge \GD)\extu A
 \end{alignat*}
@@
   \\
-  & \wkO\,(\Delta\,\extu A) && := \wkO\,\Delta\,\extu \wk_{\length{\Delta}}\,A
-  \end{alignat*}
-$\length{\Gamma}$ denotes the length of a the context $\Gamma$.
+  & \wkO\,(\GD\,\extu A) && := \wkO\,\GD\,\extu \wk_{\length{\GD}}\,A
+  \end{alignat*}
+$\length{\GG}$ denotes the length of a the context $\GG$.
 
@@
   \end{alignat*}
-  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
-  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
-  and $\Delta\ext A$ for any type $A$. This occurs when defining
+  The idea is that if $\sigma$ is a substitution between contexts $\GG$ and
+  $\GD$, then $\keep\,\sigma$ is a substitution between contexts $\GG\ext A[\sigma]$
+  and $\GD\ext A$ for any type $A$. This occurs when defining
   $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.
 
-  We can define the identity substitution on a context $\Gamma$ as
-  follows, where $\keep^{\length{\Gamma}}$ is $\keep$ iterated
-  $\length{\Gamma}$ times:
+  We can define the identity substitution on a context $\GG$ as
+  follows, where $\keep^{\length{\GG}}$ is $\keep$ iterated
+  $\length{\GG}$ times:
 \begin{alignat*}{5}
   & \idu && : \Conu \ra \Subu \\
-  & && := \lambda \Gamma.\keep^{\length{\Gamma}} \epsilonu
+  & && := \lambda \GG.\keep^{\length{\GG}} \epsilonu
   \end{alignat*}
@@
   & \emptyw && : \emptyu \vdash \\
-  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
-  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
-  \Gamma\extu A \vdash \\
+  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
+  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
+  \GG\extu A \vdash \\
   & \consw && :
-    (\Delta \vdash) \ra
-    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
-    ( \Delta \vdash A) \ra
-    ( \Gamma \vdash t \in A[\sigma]) \ra
-    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
-   \\
-  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
-  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash \mathsf{0} \invar \wku \, A \\
-  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
+    (\GD \vdash) \ra
+    (\GG\vdash \sigma \Rightarrow\GD)\ra
+    ( \GD \vdash A) \ra
+    ( \GG \vdash t \in A[\sigma]) \ra
+    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
+   \\
+  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
+  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
+  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
   \vdash \mathsf{S}\,n \invar \wku \, A \\
   & \rlap{(2) Universe} \\[0.5em]
-  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
-  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
+  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
+  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
   & \rlap{(3) Inductive parameters} \\[0.5em]
-  & \Piw && :
-    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
-  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
+\  & \Piw && :
+    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
+  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
   & \appw && :
-    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
+    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
     \Elu\,a \vdash B)
@@
     & &&
-    \ra(\Gamma \vdash t \in \Piu \, a \, B )
+    \ra(\GG \vdash t \in \Piu \, a \, B )
     \ra
-    (\Gamma \vdash u \in \Elu\,a)
+    (\GG \vdash u \in \Elu\,a)
     \ra
-    \Gamma \vdash t \appu u \in  B [0 := u] \\
+    \GG \vdash t \appu u \in  B [0 := u] \\
   & \rlap{(4) Inductive parameters} \\[0.5em]
   & \Pimw && :
-    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
-    ((t : T)\ra\Gamma\vdash A\,t) \ra
-    \Gamma \vdash \Pimu \,T\,A
+    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
+    ((t : T)\ra\GG\vdash A\,t) \ra
+    \GG \vdash \Pimu \,T\,A
     \\
     & \appmw  && :
-    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
-    ((t : T)\ra\Gamma\vdash A\,t)
+    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
+    ((t : T)\ra\GG\vdash A\,t)
     \\
     & &&
-    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
-    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
+    \ra(\GG \vdash t \in \Pimu \,T\,A)
+    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash) \ra
-  \Gamma \extu A \merge \wkO\, \Delta \vdash \\
+  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash) \ra
+  \GG \extu A \merge \wkO\, \GD \vdash \\
   % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
-  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash B) \ra
-  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, B  \\
-  & \wf{\wk} && : (\Gamma \vdash A)\ra(\Gamma \merge \Delta\vdash t\in B) \ra
-  \Gamma \extu A \merge \wkO\,\Delta \vdash \wk_{\length{\Delta}}\, t \in \wk_{\length{\Delta}}\,B  \\
-  & \wf{\wkO} && : (\Gamma \vdash A)\ra(\Gamma \vdash \sigma
-  \Rightarrow \Delta) \ra
-  \Gamma \extu A \vdash \wkO\, \sigma \Rightarrow \Delta
+  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash B) \ra
+  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, B  \\
+  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash t\in B) \ra
+  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, t \in \wk_{\length{\GD}}\,B  \\
+  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \vdash \sigma
+  \Rightarrow \GD) \ra
+  \GG \extu A \vdash \wkO\, \sigma \Rightarrow \GD
   \end{alignat*}
@@
   % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
-  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
-  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  \Gamma \vdash A [ \sigma ]
-  \\
-  & \wf{[]} && : (\Gamma \vdash)\ra(\Delta \vdash t \in A)\ra
-  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  \Gamma \vdash t [\sigma] \in A [ \sigma ]
-  \\
-  & \wf{[]} && : (\Delta \vdash x \invar A)\ra
-  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  \Gamma \vdash x [\sigma] \in A [ \sigma ]
+  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash A)\ra
+  (\GG \vdash \sigma \Rightarrow \GD) \ra
+  \GG \vdash A [ \sigma ]
+  \\
+  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash t \in A)\ra
+  (\GG \vdash \sigma \Rightarrow \GD) \ra
+  \GG \vdash t [\sigma] \in A [ \sigma ]
+  \\
+  & \wf{[]} && : (\GD \vdash x \invar A)\ra
+  (\GG \vdash \sigma \Rightarrow \GD) \ra
+  \GG \vdash x [\sigma] \in A [ \sigma ]
   % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
@@
   & \wf{{\circ}} && :
-  (\Gamma \vdash) \ra
-  (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  (\Delta \vdash \tau \Rightarrow E) \ra
-  \Gamma \vdash \tau \circ \sigma \Rightarrow E
+  (\GG \vdash) \ra
+  (\GG \vdash \sigma \Rightarrow \GD) \ra
+  (\GD \vdash \tau \Rightarrow E) \ra
+  \GG \vdash \tau \circ \sigma \Rightarrow E
   % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
-  % & \wf{keep-\El} && : (\Gamma \vdash)\ra(\Delta \vdash A)\ra
-  % (\Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  % \Gamma \vdash A [ \sigma ]
+  % & \wf{keep-\El} && : (\GG \vdash)\ra(\GD \vdash A)\ra
+  % (\GG \vdash \sigma \Rightarrow \GD) \ra
+  % \GG \vdash A [ \sigma ]
   % \\
@@
   % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
-  & [\idu] && : (\Gamma \vdash A)\ra A [ \idu\,\Gamma]  \\
-  & [\idu] && : (\Gamma \vdash x \invar A)\ra x [ \idu\,\Gamma] = V x \\
-  & [\idu] && : (\Gamma \vdash t \in A)\ra t [ \idu\,\Gamma] = t \\
-  & \idru && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \sigma \circ \idu\,\Gamma = \sigma \\
-  & \idlu && : (\Gamma \vdash \sigma \Rightarrow\Delta)\ra \idu\,\Delta\circ \sigma = \sigma
+  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG]  \\
+  & [\idu] && : (\GG \vdash x \invar A)\ra x [ \idu\,\GG] = V x \\
+  & [\idu] && : (\GG \vdash t \in A)\ra t [ \idu\,\GG] = t \\
+  & \idru && : (\GG \vdash \sigma \Rightarrow\GD)\ra \sigma \circ \idu\,\GG = \sigma \\
+  & \idlu && : (\GG \vdash \sigma \Rightarrow\GD)\ra \idu\,\GD\circ \sigma = \sigma
   \end{alignat*}
@@
   \begin{alignat*}{5}
-  & \idw && : (\Gamma \vdash)\ra \Gamma \vdash \idu\,\Gamma \Rightarrow \Gamma
+  & \idw && : (\GG \vdash)\ra \GG \vdash \idu\,\GG \Rightarrow \GG
   \end{alignat*}
@@
 \begin{alignat*}{5}
- & \isp{\Conw} && : \isaprop\, (\Gamma\vdash) \\
- & \isp{\Tyw} && : \isaprop\, (\Gamma\vdash A) \\
- & \isp{\wf\Var} && : \isaprop\, (\Gamma\vdash x \invar A) \\
- & \isp{\Tmw} && : \isaprop\, (\Gamma\vdash t \in A) \\
- & \isp{\wf\Sub} && : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
+ & \isp{\Conw} && : \isaprop\, (\GG\vdash) \\
+ & \isp{\Tyw} && : \isaprop\, (\GG\vdash A) \\
+ & \isp{\wf\Var} && : \isaprop\, (\GG\vdash x \invar A) \\
+ & \isp{\Tmw} && : \isaprop\, (\GG\vdash t \in A) \\
+ & \isp{\wf\Sub} && : \isaprop\, (\GG\vdash \sigma \Rightarrow \GD)
 \end{alignat*}
@@
   \Tmw{=}\Ty & &:
-  (\Gamma \vdash t \in A) \ra
-  (\Gamma \vdash t \in B) \ra A = B
+  (\GG \vdash t \in A) \ra
+  (\GG \vdash t \in B) \ra A = B
   \\
@@
   \wf\Var{=}\Ty & &:
-  (\Gamma \vdash x \invar A) \ra
-  (\Gamma \vdash x \invar B) \ra A = B
+  (\GG \vdash x \invar A) \ra
+  (\GG \vdash x \invar B) \ra A = B
   \end{alignat*}
@@
 \begin{alignat*}{5}
- & \syn\Con && := \sum_\Gamma \Gamma\vdash
- \\
- & \syn\Ty \,(\Gamma,\wf\Gamma)&& := \sum_A \Gamma\vdash A
- \\
- & \syn\Tm \,(\Gamma,\wf\Gamma)(A,\wf A)&& := \sum_t \Gamma\vdash t \in A
- \\
- & \syn\Sub \,(\Gamma,\wf\Gamma)(\Delta,\wf \Delta)&& := \sum_\sigma \Gamma\vdash \sigma \Rightarrow \Delta
+ & \syn\Con && := \sum_\GG \GG\vdash
+ \\
+ & \syn\Ty \,(\GG,\wf\GG)&& := \sum_A \GG\vdash A
+ \\
+ & \syn\Tm \,(\GG,\wf\GG)(A,\wf A)&& := \sum_t \GG\vdash t \in A
+ \\
+ & \syn\Sub \,(\GG,\wf\GG)(\GD,\wf \GD)&& := \sum_\sigma \GG\vdash \sigma \Rightarrow \GD
 \end{alignat*}
@@
 % As these judgments are propositions, we will sometimes
-If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic context
+If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
 (i.e.\ a context of the model $M$), we want to define a type
-$\Gamma\sim \model{\Gamma}$ equivalent to $\rec\,\,\Gamma=\model{\Gamma}$. Of
+$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
 course, at this stage, $\rec$ is not available yet since the point of
@@
 
-For a type $A$ in a context $\Gamma$, we want to define a relation
+For a type $A$ in a context $\GG$, we want to define a relation
 $A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
 % where $\model{A}$ is a
-% semantic type in a semantic context $\model{\Gamma}$.
+% semantic type in a semantic context $\model{\GG}$.
 For this equality to make sense, the semantic type $\model{A}$ must live
-in the semantic context $\rec\,\,\Gamma$. But again, $\rec$ is not
+in the semantic context $\rec\,\,\GG$. But again, $\rec$ is not
 yet available at this stage. Exploiting the expected equivalence between
-$\Gamma\sim\model\Gamma$ and $\rec\,\,\Gamma=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
-hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
-related to $\Gamma$. Then, the type of the relation for types is
+$\GG\sim\model\GG$ and $\rec\,\,\GG=\model\GG$, we may consider defining $A\sim\model{A}$ under the
+hypotheses that $\model{A}$ lies in a semantics context $\model\GG$ which is
+related to $\GG$. Then, the type of the relation for types is
 \[
-  (\Gamma : \syn\Con)\ra (A : \syn\Ty\,\syn\Gamma) \ra
-  (\model\Gamma : \model\Con)
+  (\GG : \syn\Con)\ra (A : \syn\Ty\,\syn\GG) \ra
+  (\model\GG : \model\Con)
   \ra
-  (\Gamma \sim \model\Gamma)
+  (\GG \sim \model\GG)
   \ra
-  (\model{A}:\model\Ty\,\model\Gamma)
+  (\model{A}:\model\Ty\,\model\GG)
   \ra
@@
 definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
-$\Gamma\sim\model\Gamma$ from the list of arguments.  We proceed similarly for
+$\GG\sim\model\GG$ from the list of arguments.  We proceed similarly for
 terms and substitutions. Actually, this removal is not without harm. For
 example, consider relating the empty substitution
-$\Gamma\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
-$\model\sigma : \model\Sub\,\model\Gamma\,\model\Delta$. We would like to assert
+$\GG\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
+$\model\sigma : \model\Sub\,\model\GG\,\model\GD$. We would like to assert
 that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
-this is not possible because typechecking requires that $\model\Delta$ is the
+this is not possible because typechecking requires that $\model\GD$ is the
 empty semantic context. This is precisely what was ensured by the hypothesis
-$\syn\cdot \sim \model\Delta$ we removed.  Our way out here is to state that $\model\sigma$
-is related to the empty substitution if the target semantic context $\model\Delta$
+$\syn\cdot \sim \model\GD$ we removed.  Our way out here is to state that $\model\sigma$
+is related to the empty substitution if the target semantic context $\model\GD$
 is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
@@
 Let us mention another possible solution for avoiding recursion-recursion:
-defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\Gamma
-= \model{\Gamma})\times (\rec\,A = _{e\#}\model{A})$.
-% whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
-% For a substitution $\sigma$ between contexts $\Gamma$ and $\Delta$,
-% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \Gamma
-% \sim \model\Gamma \to \Set $ where $\model{A}$ is a semantic type in the
-% semantic context $\model{\Gamma}$. We define the relation on types under the hypothesis
+defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\GG
+= \model{\GG})\times (\rec\,A = _{e\#}\model{A})$.
+% whereas the right hand side is of type $\model\Ty\model{\GG}$.
+% For a substitution $\sigma$ between contexts $\GG$ and $\GD$,
+% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \GG
+% \sim \model\GG \to \Set $ where $\model{A}$ is a semantic type in the
+% semantic context $\model{\GG}$. We define the relation on types under the hypothesis
 % that the contexts are already related.
 % it is natural to define the relation with a semantic type $\model{A}$ in a
-% semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
+% semantic context $\model{\GG}$ such that it is equivalent to  $(\rec\,\GG = \model{\GG})\times (\rec\,A = \model{A})$ (note that the first equality is
 % required for the second one to be well-typed).
@@
 % Then, the ultimate goal is to prove that
-% $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
-% $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
+% $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
+% $\sum_{\model{\GG}}({\model{A}:\model{\Ty}\,\model{\GG}})\times(A\sim\model{A})$
 % are contractible, i.e., have a unique inhabitant.
@@
 to the definition
-  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
+  $  \Uw\,\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
   instead of
   our definition
-  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
+  $  \Uw\,\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
 % Recall that our adapted goal, for types, is to prove that
 % $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
-% is inhabited if $\Gamma \sim \model{\Gamma}$ is.
+% is inhabited if $\GG \sim \model{\GG}$ is.
 %  This means that in the definition of the relation, we assume
@@
 We define the relation by recursion on the typing judgments.
-In the following, we abbreviate $\relT{\wf{A}}{\model\Gamma}{\model{A}}$ by
-$\wf{A}\sim\model{A}$ when $\model{\Gamma}$ can be inferred, and similarly for
+In the following, we abbreviate $\relT{\wf{A}}{\model\GG}{\model{A}}$ by
+$\wf{A}\sim\model{A}$ when $\model{\GG}$ can be inferred, and similarly for
 terms and substitutions.
@@
   & \rlap{(1) Substitution calculus} \\[0.5em]
-  & \blank \sim{ \blank} && :  \Gamma \vdash \, \ra \model{\Con} \ra \Set \\
-  & \relT{\blank}{\model{\Gamma}}{\blank} && :  \Gamma \vdash A \, \ra \model{\Ty}\,\model{\Gamma} \ra \Set \\
-  & \relt{\blank}{\model{\Gamma}}{\model{A}}{\blank} && :  \Gamma \vdash t \in A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
-  & \relV{\blank}{\model{\Gamma}}{\model{A}}{\blank}&& :  \Gamma \vdash x \invar A \, \ra \model{\Tm}\,\model{\Gamma}\,\model{A} \ra \Set \\
-  & \relS{\blank}{\model{\Gamma}}{\model{\Delta}}{ \blank} && :  \Gamma \vdash \sigma \Rightarrow \Delta \, \ra \model{\Sub}\,\model{\Gamma}\,\model{\Delta} \ra \Set \\
-  \\
-  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\boldsymbol{\cdot}} \\
-  & \relS{\epsilonw}{\model{\Gamma}}{\model{E}}{\model{\delta}}
+  & \blank \sim{ \blank} && :  \GG \vdash \, \ra \model{\Con} \ra \Set \\
+  & \relT{\blank}{\model{\GG}}{\blank} && :  \GG \vdash A \, \ra \model{\Ty}\,\model{\GG} \ra \Set \\
+  & \relt{\blank}{\model{\GG}}{\model{A}}{\blank} && :  \GG \vdash t \in A \, \ra \model{\Tm}\,\model{\GG}\,\model{A} \ra \Set \\
+  & \relV{\blank}{\model{\GG}}{\model{A}}{\blank}&& :  \GG \vdash x \invar A \, \ra \model{\Tm}\,\model{\GG}\,\model{A} \ra \Set \\
+  & \relS{\blank}{\model{\GG}}{\model{\GD}}{ \blank} && :  \GG \vdash \sigma \Rightarrow \GD \, \ra \model{\Sub}\,\model{\GG}\,\model{\GD} \ra \Set \\
+  \\
+  & \emptyw\sim \model{\GG} && := \model{\GG}=\model{\boldsymbol{\cdot}} \\
+  & \relS{\epsilonw}{\model{\GG}}{\model{E}}{\model{\delta}}
   % (  \vdash \model{\delta}\Rightarrow \model{E})
@@
   \\
-  & (\wf{\Gamma}\extw \wf{A})\sim \model{\Delta} &&
+  & (\wf{\GG}\extw \wf{A})\sim \model{\GD} &&
    :=
-    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
+    \sum_{\model{\GG}} (\wf{\GG}\sim\model{\GG}) \times
     \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
@@
     & && \qquad
-    (\model{\Delta} = \model{\Gamma} \model{\ext} \model{A})
-    \\
-    &
-    \relS{(\consw\wf{\Delta}\wf{\sigma}\wf{A}\wf{t})}{\model{\Gamma}}{\model{E}}{\model{\delta}}
-    % (\consw \wf{\Delta}\wf{\sigma}\wf{A}\wf{t})\sim
+    (\model{\GD} = \model{\GG} \model{\ext} \model{A})
+    \\
+    &
+    \relS{(\consw\wf{\GD}\wf{\sigma}\wf{A}\wf{t})}{\model{\GG}}{\model{E}}{\model{\delta}}
+    % (\consw \wf{\GD}\wf{\sigma}\wf{A}\wf{t})\sim
     % \\
     % & \qquad
-    % ( \model{\Gamma} \vdash \model{\delta}\Rightarrow \model{E})
+    % ( \model{\GG} \vdash \model{\delta}\Rightarrow \model{E})
     &&
    :=
-    \sum_{\model{\Delta}} (\wf{\Delta}\sim\model{\Delta}) \times
+    \sum_{\model{\GD}} (\wf{\GD}\sim\model{\GD}) \times
     \sum_{\model{\sigma}} (\wf{\sigma}\sim\model{\sigma}) \times
@@
     & && \qquad
-    (e_E : \model{E} = \model{\Delta} \model{\ext} \model{A}) \times
+    (e_E : \model{E} = \model{\GD} \model{\ext} \model{A}) \times
     \\
@@
    && := \wf{x}\sim \model{t} \\
-   & \relV{\vzw\wf{\Gamma}\wf{A}}{\model{\Delta}}{\model{B}}{\model{t}}
-  %  \vzw\wf{\Gamma}\wf{A}
+   & \relV{\vzw\wf{\GG}\wf{A}}{\model{\GD}}{\model{B}}{\model{t}}
+  %  \vzw\wf{\GG}\wf{A}
   % \sim \\
-  % & \qquad (\model{\Delta}\vdash \model{t}\in \model{B} )
+  % & \qquad (\model{\GD}\vdash \model{t}\in \model{B} )
    && :=
-    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
+    \sum_{\model{\GG}} (\wf{\GG}\sim\model{\GG}) \times
     \sum_{\model{A}} (\wf{A}\sim\model{A}) \times \\
@@
     \qquad
-  (e_\Delta : \model{\Delta} = \model{\Gamma} \model{\ext} \model{A})
+  (e_\GD : \model{\GD} = \model{\GG} \model{\ext} \model{A})
      \times
@@
     & && \qquad
-     (e_B : \model{B} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times
-     % \\
-     % & && \qquad
-     (\model{t} = _{e_\Delta,e_B \#} \model{\vz})
-  \\
-  &
-   \relV{\vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B}}{\model\Delta}{\model{C}}{\model{t}}
-  % \vsw \wf{\Gamma} \wf{A} \wf{n} \wf{B} \sim
+     (e_B : \model{B} = _{e_\GD \#} \model{\wk}\,\model{A}) \times
+     % \\
+     % & && \qquad
+     (\model{t} = _{e_\GD,e_B \#} \model{\vz})
+  \\
+  &
+   \relV{\vsw \wf{\GG} \wf{A} \wf{n} \wf{B}}{\model\GD}{\model{C}}{\model{t}}
+  % \vsw \wf{\GG} \wf{A} \wf{n} \wf{B} \sim
   % \\
   % & \qquad
-  %    (\model{\Delta}\vdash \model{t}\in \model{C} )
+  %    (\model{\GD}\vdash \model{t}\in \model{C} )
    && :=
-    \sum_{\model{\Gamma}} (\wf{\Gamma}\sim\model{\Gamma}) \times
+    \sum_{\model{\GG}} (\wf{\GG}\sim\model{\GG}) \times
     \sum_{\model{A}} (\wf{A}\sim\model{A}) \times
@@
     \qquad
-  (e_\Delta : \model{\Delta} = \model{\Gamma} \model{\ext} \model{B})
+  (e_\GD : \model{\GD} = \model{\GG} \model{\ext} \model{B})
      \times
@@
     & && \qquad
-     (e_C : \model{C} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times
+     (e_C : \model{C} = _{e_\GD \#} \model{\wk}\,\model{A}) \times
      \\
      & && \qquad
-     (\model{t} = _{e_\Delta,e_C \#} \model{\vs} \, \model{n})
+     (\model{t} = _{e_\GD,e_C \#} \model{\vs} \, \model{n})
   \\
   & \rlap{(2) Universe} \\[0.5em]
-  & \Uw\, \wf{\Gamma} \wf{A} \sim \model{A} &&
+  & \Uw\, \wf{\GG} \wf{A} \sim \model{A} &&
    := \model{A} = \model{\U}
   \\
-  & \Elw\, \wf{\Gamma} \wf{a} \sim \model{A}
+  & \Elw\, \wf{\GG} \wf{a} \sim \model{A}
   && :=
@@
   & \rlap{(3) Inductive parameters} \\[0.5em]
-  & \Piw\,\wf{\Gamma} \wf{a} \wf{B} \sim \model{C} &&
+  & \Piw\,\wf{\GG} \wf{a} \wf{B} \sim \model{C} &&
    :=
@@
   &
-  \relt{\appw\, \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u}}{\model{\Gamma}}{\model{C}}{\model{x}}  &&
-  % & \appw \wf{\Gamma} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
-  % & \qquad (\model{\Gamma}\vdash x\in \model{C})  &&
+  \relt{\appw\, \wf{\GG} \wf{a} \wf{B} \wf{t} \wf{u}}{\model{\GG}}{\model{C}}{\model{x}}  &&
+  % & \appw \wf{\GG} \wf{a} \wf{B} \wf{t} \wf{u} \sim \\
+  % & \qquad (\model{\GG}\vdash x\in \model{C})  &&
     :=
@@
   & \rlap{(4) Metatheoretic parameters} \\[0.5em]
-  & \Pimw T\,A\,\wf{\Gamma}\wf{A} \sim \model{B} &&
+  & \Pimw T\,A\,\wf{\GG}\wf{A} \sim \model{B} &&
     :=
@@
     \\
-    & \relt{\appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u}{\model{\Gamma}}{\model{B}}{\model{x}}
-    % & \appmw T\,A\,\wf{\Gamma}\wf{A} \wf{t} u \sim
+    & \relt{\appmw T\,A\,\wf{\GG}\wf{A} \wf{t} u}{\model{\GG}}{\model{B}}{\model{x}}
+    % & \appmw T\,A\,\wf{\GG}\wf{A} \wf{t} u \sim
     % \\
     % & \qquad
-    % (\model{\Gamma}\vdash x\in \model{B})
+    % (\model{\GG}\vdash x\in \model{B})
      &&
@@
 % to the definition
-%   $  \Uw\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
+%   $  \Uw\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
 %   instead of
 %   the current definition
-%   $  \Uw\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
+%   $  \Uw\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
 %   Our choice makes the definitions more concise, and similarly
@@
   % approach:
-  % the relation gives a semantic context $\model\Gamma$ that must be related
-  % to the syntactic $\Gamma$, and a semantic
+  % the relation gives a semantic context $\model\GG$ that must be related
+  % to the syntactic $\GG$, and a semantic
   % type $\model{A}$ related to the syntactic $A$. The input semantic context is
-  % then required to equal $\model{\Gamma}\model{\ext}\model{A}$.   As
+  % then required to equal $\model{\GG}\model{\ext}\model{A}$.   As
   % $\blank \model{\ext}\blank$ is not guaranteed to be injective, these
@@
     \isp{\Sigma{\sim}}
-    && : (\wf\Gamma : \Gamma \vdash ) && \ra
-    \isaprop\, (\sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma})
+    && : (\wf\GG : \GG \vdash ) && \ra
+    \isaprop\, (\sum_{\model{\GG}} \wf{\GG} \sim \model{\GG})
     \\
@@
     \isp{\Sigma{\sim}}
-    && : (\wf A : \Gamma \vdash A) && \ra
+    && : (\wf A : \GG \vdash A) && \ra
     \isaprop\, (\sum_{\model{A}} \wf{A} \sim \model{A})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf t : \Gamma \vdash t \in A) && \ra
+    && : (\wf t : \GG \vdash t \in A) && \ra
     \isaprop\, (\sum_{\model{t}} \wf{t} \sim \model{t})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf x : \Gamma \vdash x \invar A) && \ra
+    && : (\wf x : \GG \vdash x \invar A) && \ra
     \isaprop\, (\sum_{\model{x}} \wf{x} \sim \model{x})
@@
     \isp{\Sigma{\sim}}
-    && : (\wf \sigma : \Gamma \vdash \sigma \Rightarrow \Delta) && \ra
+    && : (\wf \sigma : \GG \vdash \sigma \Rightarrow \GD) && \ra
     \isaprop\, (\sum_{\model{\sigma}} \wf{\sigma} \sim \model{\sigma})
@@
   equalities. Because of this, we need UIP to show that
-  $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
+  $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
   $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
@@
   suggested before, expecting that
-  $\sum_{\model{\Gamma}}\sum_{\model{A}}A\sim\model{A}$, rather than
+  $\sum_{\model{\GG}}\sum_{\model{A}}A\sim\model{A}$, rather than
   $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
   obvious. For example, we were not able to define
-  $\Elw\,\wf{\Gamma}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
+  $\Elw\,\wf{\GG}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
   Hugunin investigated constructing IITs without
@@
    % & && \qquad
-  (\wf{\Gamma} : \Gamma \vdash) \ra
-  (\wf\Gamma \sim \model{\Gamma}) \ra
+  (\wf{\GG} : \GG \vdash) \ra
+  (\wf\GG \sim \model{\GG}) \ra
   \\
   & && \qquad
-  (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  (\wf\Delta \sim \model{\Delta}) \ra
+  (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  (\wf\GD \sim \model{\GD}) \ra
   \\ &&& \qquad
-  (\model{A} : \model\Ty\model\Gamma)\ra
-  (\model{\Delta'} : \model{\Con}) \times (\model{\Sub}\model{\Delta'}{\model{\Delta}})
-\end{alignat*}
-Here, $\model{\Delta'}$ should be thought of as the context $\model\Delta$ where
+  (\model{A} : \model\Ty\model\GG)\ra
+  (\model{\GD'} : \model{\Con}) \times (\model{\Sub}\model{\GD'}{\model{\GD}})
+\end{alignat*}
+Here, $\model{\GD'}$ should be thought of as the context $\model\GD$ where
 the weakening has happened in the middle of the context, by inserting the type
-$\model{A}$ after the prefix $\model{\Gamma}$. Indeed, we expect that
-$\model{\Gamma}$ is a prefix of $\model{\Delta}$, as $\model{\Gamma}$ relates to
-$\Gamma$ and $\model{\Delta}$ to $\Gamma\merge\Delta$.
-% $\wf{\Gamma}\sim\model\Gamma$
-% and $\wf\Delta\sim\model\Delta$, and $\Gamma$ is a prefix of $\Gamma\merge\Delta$.
+$\model{A}$ after the prefix $\model{\GG}$. Indeed, we expect that
+$\model{\GG}$ is a prefix of $\model{\GD}$, as $\model{\GG}$ relates to
+$\GG$ and $\model{\GD}$ to $\GG\merge\GD$.
+% $\wf{\GG}\sim\model\GG$
+% and $\wf\GD\sim\model\GD$, and $\GG$ is a prefix of $\GG\merge\GD$.
 The substitution from the weakened context to the original one must
@@
 \begin{alignat*}{5}
-  & \model{\wkO}\,\model{A}\,\model\Delta  && :
-  % (\model{A} : \model\Ty\model\Gamma)\ra
-  % (\wf{\Gamma} : \Gamma \vdash) \ra
-  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  % \\
-  % & && \qquad
-  % (\wf\Gamma \sim \model{\Gamma}) \ra
-  % (\wf\Delta \sim \model{\Delta}) \ra
+  & \model{\wkO}\,\model{A}\,\model\GD  && :
+  % (\model{A} : \model\Ty\model\GG)\ra
+  % (\wf{\GG} : \GG \vdash) \ra
+  % (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  % \\
+  % & && \qquad
+  % (\wf\GG \sim \model{\GG}) \ra
+  % (\wf\GD \sim \model{\GD}) \ra
   % \\ &&& \qquad
@@
    \\
-  & \model{\wk{\Rightarrow}}\,\model{A}\,\model\Delta  && :
-  % (\model{A} : \model\Ty\model\Gamma)\ra
-  % (\wf{\Gamma} : \Gamma \vdash) \ra
-  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  % \\
-  % & && \qquad
-  % (\wf\Gamma \sim \model{\Gamma}) \ra
-  % (\wf\Delta \sim \model{\Delta}) \ra
+  & \model{\wk{\Rightarrow}}\,\model{A}\,\model\GD  && :
+  % (\model{A} : \model\Ty\model\GG)\ra
+  % (\wf{\GG} : \GG \vdash) \ra
+  % (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  % \\
+  % & && \qquad
+  % (\wf\GG \sim \model{\GG}) \ra
+  % (\wf\GD \sim \model{\GD}) \ra
   % \\ &&& \qquad
-   \model{\Sub}(\model{\wkO}\model{A}\,\model\Delta)\model\Delta
+   \model{\Sub}(\model{\wkO}\model{A}\,\model\GD)\model\GD
 \end{alignat*}
@@
 hypotheses
-  $(\wf{\Gamma} : \Gamma \vdash)$,
-  $(\wf\Gamma \sim \model{\Gamma})$,
-  $(\wf{\Delta} : \Gamma\merge\Delta \vdash)$,
-  $(\wf\Delta \sim \model{\Delta})$,
-  $(\wf{A} : \Gamma \vdash A)$,
+  $(\wf{\GG} : \GG \vdash)$,
+  $(\wf\GG \sim \model{\GG})$,
+  $(\wf{\GD} : \GG\merge\GD \vdash)$,
+  $(\wf\GD \sim \model{\GD})$,
+  $(\wf{A} : \GG \vdash A)$,
   and
@@
   % , except the last one about weakening substitutions,
-  % which do not require that $\model\Delta$ is related to any well-formed context.
+  % which do not require that $\model\GD$ is related to any well-formed context.
 \begin{alignat*}{5}
@@
    :
-  % (\wf{\Gamma} : \Gamma \vdash) \ra
-  % (\wf\Gamma \sim \model{\Gamma}) \ra
-  % \\
-  %  & && \qquad
-  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  % (\wf\Delta \sim \model{\Delta}) \ra
-  % \\
-  % & && \qquad
-  % (\wf{A} : \Gamma \vdash A) \ra
+  % (\wf{\GG} : \GG \vdash) \ra
+  % (\wf\GG \sim \model{\GG}) \ra
+  % \\
+  %  & && \qquad
+  % (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  % (\wf\GD \sim \model{\GD}) \ra
+  % \\
+  % & && \qquad
+  % (\wf{A} : \GG \vdash A) \ra
   % (\wf{A} \sim \model{A}) \ra
@@
   % & && \qquad
-  \wf{\wkO}\,\wf{A}\,\wf{\Delta}\sim\model{\wkO}\model{A}\model{\Delta}
+  \wf{\wkO}\,\wf{A}\,\wf{\GD}\sim\model{\wkO}\model{A}\model{\GD}
   \\
@@
   & \wk{\sim} && :
-  % (\wf{\Gamma} : \Gamma \vdash) \ra
-  % (\wf\Gamma \sim \model{\Gamma}) \ra
-  % \\
-  %  & && \qquad
-  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  % (\wf\Delta \sim \model{\Delta}) \ra
-  % \\
-  % & && \qquad
-  % (\wf{A} : \Gamma \vdash A) \ra
+  % (\wf{\GG} : \GG \vdash) \ra
+  % (\wf\GG \sim \model{\GG}) \ra
+  % \\
+  %  & && \qquad
+  % (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  % (\wf\GD \sim \model{\GD}) \ra
+  % \\
+  % & && \qquad
+  % (\wf{A} : \GG \vdash A) \ra
   % (\wf{A} \sim \model{A}) \ra
@@
   % &&& \qquad
-  (\wf{T} : \Gamma\merge\Delta \vdash T) \ra
+  (\wf{T} : \GG\merge\GD \vdash T) \ra
   (\wf{T} \sim \model{T}) \ra
   % \\ &&& \qquad
-  \wf{\wk}\,\wf{A}\,\wf{T}\sim\model{T}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
+  \wf{\wk}\,\wf{A}\,\wf{T}\sim\model{T}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\GD}]}
   \\
@@
   % &&& \qquad
-  (\wf{t} : \Gamma\merge\Delta \vdash t\in T) \ra
+  (\wf{t} : \GG\merge\GD \vdash t\in T) \ra
   (\wf{t} \sim \model{t}) \ra
   % \\ &&& \qquad
-  \wf{\wk}\,\wf{A}\,\wf{t}\sim\model{t}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
+  \wf{\wk}\,\wf{A}\,\wf{t}\sim\model{t}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\GD}]}
   \\
@@
   % &&& \qquad
-  (\wf{x} : \Gamma\merge\Delta \vdash t\invar T) \ra
+  (\wf{x} : \GG\merge\GD \vdash t\invar T) \ra
   (\wf{x} \sim \model{x}) \ra
   % \\ &&& \qquad
-  \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
+  \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\GD}]}
   % \\ &&& \qquad
-  % \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\Delta}]}
+  % \wf{\wk}\,\wf{A}\,\wf{x}\sim\model{x}\model{[\model{\wkO{\Rightarrow}}\model{A}\model{\GD}]}
 \end{alignat*}
@@
   & \wkO{\sim} && :
-  (\wf{\Gamma} : \Gamma \vdash) \ra
-  (\wf\Gamma \sim \model{\Gamma}) \ra
-  % \\
-  %  & && \qquad
-  % (\wf{\Delta} : \Gamma\merge\Delta \vdash) \ra
-  % (\wf\Delta \sim \model{\Delta}) \ra
-  % \\
-  % & && \qquad
-  (\wf{A} : \Gamma \vdash A) \ra
+  (\wf{\GG} : \GG \vdash) \ra
+  (\wf\GG \sim \model{\GG}) \ra
+  % \\
+  %  & && \qquad
+  % (\wf{\GD} : \GG\merge\GD \vdash) \ra
+  % (\wf\GD \sim \model{\GD}) \ra
+  % \\
+  % & && \qquad
+  (\wf{A} : \GG \vdash A) \ra
   (\wf{A} \sim \model{A}) \ra
@@
   &&& \qquad
-  (\wf{\sigma} : \Gamma\vdash \sigma \Rightarrow \Delta) \ra
+  (\wf{\sigma} : \GG\vdash \sigma \Rightarrow \GD) \ra
   (\wf{\sigma} \sim \model{\sigma}) \ra
@@
 %   & \blank\leq\blank  && : \model\Con\ra \model\Con\ra\Set \\
-%   & {\leq}\boldsymbol{\cdot}  && : \model{\Gamma} \leq \model{\Gamma} \\
-%   & \blank{\leq}{\ext}\blank  && : \model{\Gamma} \leq \model{\Delta} \ra (A :
-%   \model{\Ty}\Delta) \ra \model{\Gamma}\leq\model{\Delta}\model{\ext}\model{A}
+%   & {\leq}\boldsymbol{\cdot}  && : \model{\GG} \leq \model{\GG} \\
+%   & \blank{\leq}{\ext}\blank  && : \model{\GG} \leq \model{\GD} \ra (A :
+%   \model{\Ty}\GD) \ra \model{\GG}\leq\model{\GD}\model{\ext}\model{A}
 % \end{alignat*}
-% Now, a telescope on a semantic context $\model{\Gamma}$ is a semantic context
-% $\model{\Delta}$ such that $\model{\Gamma}\leq \model{\Delta}$.
+% Now, a telescope on a semantic context $\model{\GG}$ is a semantic context
+% $\model{\GD}$ such that $\model{\GG}\leq \model{\GD}$.
 
 % \begin{alignat*}{5}
-%   & \Tel\,\model{\Gamma}  && := (\model{\Delta} : \model{\Con})\times (\model{\Gamma}\leq \model{\Delta})
+%   & \Tel\,\model{\GG}  && := (\model{\GD} : \model{\Con})\times (\model{\GG}\leq \model{\GD})
 % \end{alignat*}
@@
 % \begin{alignat*}{5}
-%   & \blank\model{\merge}\blank  && := (\model{\Gamma} : \model{\Con}) \ra
-%   (\model{\Delta} : \Tel\,\model\Gamma) \ra \model{\Con} \\
-%   & \model\Gamma \merge (\model\Gamma,{\leq}\boldsymbol{\cdot}) && :=  \model\Gamma \\
-%   & \model\Gamma \merge (\model\Delta\model{\ext}\model{A}, \Delta_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\Gamma \model\merge \model\Delta)\model\ext \model{A}\\
+%   & \blank\model{\merge}\blank  && := (\model{\GG} : \model{\Con}) \ra
+%   (\model{\GD} : \Tel\,\model\GG) \ra \model{\Con} \\
+%   & \model\GG \merge (\model\GG,{\leq}\boldsymbol{\cdot}) && :=  \model\GG \\
+%   & \model\GG \merge (\model\GD\model{\ext}\model{A}, \GD_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\GG \model\merge \model\GD)\model\ext \model{A}\\
 % \end{alignat*}
@@
 % list of types.
-% A syntactic telescope $\Delta$ is then well formed in a context $\Gamma$ if
-% $\Gamma\merge\Delta$ is well formed.
+% A syntactic telescope $\GD$ is then well formed in a context $\GG$ if
+% $\GG\merge\GD$ is well formed.
 % The relation is defined by recursion on the untyped syntactic telescope:
 % \begin{alignat*}{5}
-%   & \blank\sim\blank  && := ( \Gamma \merge \Delta \vdash) \ra (\model\Gamma\leq\model\Delta)
+%   & \blank\sim\blank  && := ( \GG \merge \GD \vdash) \ra (\model\GG\leq\model\GD)
 %   \ra \Set \\
-%   & (\wf\Gamma : \Gamma \merge \emptyu \vdash)\sim\model\Delta  && :=
-%   \model\Delta =  {\leq}\boldsymbol{\cdot}\\
-%   & (\wf\Delta \extw \wf{A} : \Gamma \merge (\Delta \extu A) \vdash)\sim\model{E}  && :=
-%   \sum_{\model\Delta} (\wf\Delta\sim\model\Delta)
+%   & (\wf\GG : \GG \merge \emptyu \vdash)\sim\model\GD  && :=
+%   \model\GD =  {\leq}\boldsymbol{\cdot}\\
+%   & (\wf\GD \extw \wf{A} : \GG \merge (\GD \extu A) \vdash)\sim\model{E}  && :=
+%   \sum_{\model\GD} (\wf\GD\sim\model\GD)
 %   \times
@@
 %   & && \qquad
-%   \model{E} = \model\Delta {\leq}{\ext} \model{A}
+%   \model{E} = \model\GD {\leq}{\ext} \model{A}
 % \end{alignat*}
@@
  We want to prove that given any well-typed substitution $\sigma
-: \Sub\,\Gamma\,\Delta$ and semantic contexts $\model{\Gamma}$ and
-$\model{\Delta}$, respectively related to $\Gamma$ and $\Delta$, there exists a
+: \Sub\,\GG\,\GD$ and semantic contexts $\model{\GG}$ and
+$\model{\GD}$, respectively related to $\GG$ and $\GD$, there exists a
 semantic substitution which is related to $\sigma$.  In the extension case
-$\Gamma\vdash \sigma \consu\,t\Rightarrow \Delta\,\extu\,A$, the induction
-hypothesis provides $\model{\sigma}$, $\model{\Delta}$, $\model{A}$ related to
+$\GG\vdash \sigma \consu\,t\Rightarrow \GD\,\extu\,A$, the induction
+hypothesis provides $\model{\sigma}$, $\model{\GD}$, $\model{A}$ related to
 their syntactic counterpart. However, the premises of the induction hypothesis
@@
   & []{\sim} && :
-  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
+  (\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD) \ra
   (\wf\sigma \sim \model{\sigma}) \ra
-  (\wf{x} : \Delta \vdash x \invar A) \ra
+  (\wf{x} : \GD \vdash x \invar A) \ra
   (\wf{x} \sim \model{x}) \ra
@@
 under the common hypotheses
-  $(\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta)$,
+  $(\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD)$,
   $(\wf\sigma \sim \model{\sigma})$,
-  $(\wf{\Gamma} : \Gamma \vdash )$,
-  $(\wf{\Gamma} \sim \model{\Gamma})$,
+  $(\wf{\GG} : \GG \vdash )$,
+  $(\wf{\GG} \sim \model{\GG})$,
   % \\
   % &&&\qquad
-    $(\wf{\Delta} : \Delta \vdash )$,
+    $(\wf{\GD} : \GD \vdash )$,
     and
-  $(\wf{\Delta} \sim \model{\Delta})$:
+  $(\wf{\GD} \sim \model{\GD})$:
 \begin{alignat*}{5}
      & []{\sim} && :
-    (\wf{A} : \Delta \vdash A) \ra
+    (\wf{A} : \GD \vdash A) \ra
   (\wf{A} \sim \model{A}) \ra
@@
   % & && \qquad
-   \wf{[]}\wf{\Gamma}\wf{A}\wf{\sigma} \sim \model{A}\model{[\model{\sigma}]}
+   \wf{[]}\wf{\GG}\wf{A}\wf{\sigma} \sim \model{A}\model{[\model{\sigma}]}
    \\
      & []{\sim} && :
-    (\wf{t} : \Delta \vdash t \in A ) \ra
+    (\wf{t} : \GD \vdash t \in A ) \ra
   (\wf{t} \sim \model{t}) \ra
@@
   % & && \qquad
-   \wf{[]}\wf{\Gamma}\wf{t}\wf{\sigma} \sim \model{t}\model{[\model{\sigma}]}
+   \wf{[]}\wf{\GG}\wf{t}\wf{\sigma} \sim \model{t}\model{[\model{\sigma}]}
   \end{alignat*}
@@
   % &&&  \qquad
-    (\wf{\delta} : \Delta \vdash \delta \Rightarrow E) \ra
+    (\wf{\delta} : \GD \vdash \delta \Rightarrow E) \ra
   (\wf{\delta} \sim \model{\delta}) \ra
@@
   & && \qquad
-   \wf{{\circ}}\,\wf{\Gamma}\,\wf{\delta}\,\wf{\sigma} \sim \model{\delta}\mathbin{\model{\circ}}\model{\sigma}
+   \wf{{\circ}}\,\wf{\GG}\,\wf{\delta}\,\wf{\sigma} \sim \model{\delta}\mathbin{\model{\circ}}\model{\sigma}
   \end{alignat*}
@@
      & {\id}{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-  (\wf{\Gamma} \sim \model{\Gamma}) \ra
-   \wf{{\id}}\,\wf{\Gamma} \sim {\id}_{\model{\Gamma}}
+    (\wf{\GG} : \GG \vdash ) \ra
+  (\wf{\GG} \sim \model{\GG}) \ra
+   \wf{{\id}}\,\wf{\GG} \sim {\id}_{\model{\GG}}
   \end{alignat*}
@@
      & \Sigma\Con{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
+    (\wf{\GG} : \GG \vdash ) \ra
+    \sum_{\model{\GG}} \wf{\GG} \sim \model{\GG}
     \\
      & \Sigma\Ty{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-    (\wf{\Gamma}\sim \model\Gamma) \ra
-    (\wf{A} : \Gamma \vdash A ) \ra
-    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
+    (\wf{\GG} : \GG \vdash ) \ra
+    (\wf{\GG}\sim \model\GG) \ra
+    (\wf{A} : \GG \vdash A ) \ra
+    ({\model{A}:\model{\Ty}\model\GG})\times ( \wf{A} \sim \model{A})
     \\
      & \Sigma\Tm{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-    (\wf{\Gamma}\sim \model\Gamma) \ra
-    (\wf{A} : \Gamma \vdash A ) \ra
+    (\wf{\GG} : \GG \vdash ) \ra
+    (\wf{\GG}\sim \model\GG) \ra
+    (\wf{A} : \GG \vdash A ) \ra
     ( \wf{A} \sim \model{A}) \ra
     \\ & && \qquad
-    (\wf{t} : \Gamma \vdash t \in A ) \ra
-    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
+    (\wf{t} : \GG \vdash t \in A ) \ra
+    ({\model{t}:\model{\Tm}\model\GG}\model{A})\times ( \wf{t} \sim \model{t})
     \\
      & \Sigma\Var{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-    (\wf{\Gamma}\sim \model\Gamma) \ra
-    (\wf{A} : \Gamma \vdash A ) \ra
+    (\wf{\GG} : \GG \vdash ) \ra
+    (\wf{\GG}\sim \model\GG) \ra
+    (\wf{A} : \GG \vdash A ) \ra
     ( \wf{A} \sim \model{A}) \ra
     \\ & && \qquad
-    (\wf{x} : \Gamma \vdash x \invar A ) \ra
-    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
+    (\wf{x} : \GG \vdash x \invar A ) \ra
+    ({\model{x}:\model{\Tm}\model\GG}\model{A})\times ( \wf{x} \sim \model{x})
     \\
      & \Sigma\Sub{\sim} && :
-    (\wf{\Gamma} : \Gamma \vdash ) \ra
-    (\wf{\Gamma}\sim \model\Gamma) \ra
-    (\wf{\Delta} : \Delta \vdash  ) \ra
-    ( \wf{\Delta} \sim \model{\Delta}) \ra
+    (\wf{\GG} : \GG \vdash ) \ra
+    (\wf{\GG}\sim \model\GG) \ra
+    (\wf{\GD} : \GD \vdash  ) \ra
+    ( \wf{\GD} \sim \model{\GD}) \ra
     \\ & && \qquad
-    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
-    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
+    (\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD ) \ra
+    ({\model{\sigma}:\model{\Sub}\model\GG}\model{\GD})\times ( \wf{\sigma} \sim \model{\sigma})
   \end{alignat*}
@@
   && :
-   (\Gamma \vdash) \ra \model\Con
+   (\GG \vdash) \ra \model\Con
    \\
@@
   && :
-   (\wf\Gamma:\Gamma \vdash) \ra (\Gamma\vdash A)\ra\model\Ty\,(\mor\Con\wf\Gamma)
+   (\wf\GG:\GG \vdash) \ra (\GG\vdash A)\ra\model\Ty\,(\mor\Con\wf\GG)
    \\
@@
   && :
-  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
-  (\Gamma\vdash t \in A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
-  (\mor\Ty\wf\Gamma\,\wf{A})
+  (\wf\GG:\GG \vdash) \ra (\wf{A}:\GG\vdash A)\ra
+  (\GG\vdash t \in A)\ra\model\Tm\,(\mor\Con\wf\GG)\,
+  (\mor\Ty\wf\GG\,\wf{A})
    \\
@@
   && :
-  (\wf\Gamma:\Gamma \vdash) \ra (\wf{A}:\Gamma\vdash A)\ra
-  (\Gamma\vdash x \invar A)\ra\model\Tm\,(\mor\Con\wf\Gamma)\,
-  (\mor\Ty\wf\Gamma\,\wf{A})
+  (\wf\GG:\GG \vdash) \ra (\wf{A}:\GG\vdash A)\ra
+  (\GG\vdash x \invar A)\ra\model\Tm\,(\mor\Con\wf\GG)\,
+  (\mor\Ty\wf\GG\,\wf{A})
    \\
@@
   && :
-  (\wf\Gamma:\Gamma \vdash) \ra
-  (\wf\Delta:\Delta \vdash) \ra
-  (\Gamma\vdash \sigma \Rightarrow \Delta)\ra\model\Sub\,(\mor\Con\wf\Gamma)\,
-  (\mor\Con\wf\Delta)
+  (\wf\GG:\GG \vdash) \ra
+  (\wf\GD:\GD \vdash) \ra
+  (\GG\vdash \sigma \Rightarrow \GD)\ra\model\Sub\,(\mor\Con\wf\GG)\,
+  (\mor\Con\wf\GD)
 \end{alignat*}
@@
   && :
-  (\wf\Gamma : \Gamma \vdash) \ra \wf\Gamma\sim \mor\Con\,\wf\Gamma
+  (\wf\GG : \GG \vdash) \ra \wf\GG\sim \mor\Con\,\wf\GG
   \\
@@
   && :
-  (\wf\Gamma : \Gamma \vdash) \ra
-  (\wf{A} : \Gamma \vdash A) \ra
-  \wf\Gamma\sim \mor\Ty\,\wf\Gamma\,\wf{A}
+  (\wf\GG : \GG \vdash) \ra
+  (\wf{A} : \GG \vdash A) \ra
+  \wf\GG\sim \mor\Ty\,\wf\GG\,\wf{A}
   \\
@@
   && :
-  (\wf\Gamma : \Gamma \vdash) \ra
-  (\wf{A} : \Gamma \vdash A) \ra
-  (\wf{t} : \Gamma \vdash t \in A) \ra
-  \wf\Gamma\sim \mor\Tm\,\wf\Gamma\,\wf{A}\,\wf{t}
+  (\wf\GG : \GG \vdash) \ra
+  (\wf{A} : \GG \vdash A) \ra
+  (\wf{t} : \GG \vdash t \in A) \ra
+  \wf\GG\sim \mor\Tm\,\wf\GG\,\wf{A}\,\wf{t}
   \\
@@
   && :
-  (\wf\Gamma : \Gamma \vdash) \ra
-  (\wf{A} : \Gamma \vdash A) \ra
-  (\wf{x} : \Gamma \vdash x \invar A) \ra
-  \wf\Gamma\sim \mor\Var\,\wf\Gamma\,\wf{A}\,\wf{x}
+  (\wf\GG : \GG \vdash) \ra
+  (\wf{A} : \GG \vdash A) \ra
+  (\wf{x} : \GG \vdash x \invar A) \ra
+  \wf\GG\sim \mor\Var\,\wf\GG\,\wf{A}\,\wf{x}
   \\
@@
   && :
-  (\wf\Gamma : \Gamma \vdash) \ra
-  (\wf{\Delta} : \Delta \vdash) \ra
-  (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta) \ra
-  \wf\Gamma\sim \mor\Sub\,\wf\Gamma\,\wf{\Delta}\,\wf{\sigma}
+  (\wf\GG : \GG \vdash) \ra
+  (\wf{\GD} : \GD \vdash) \ra
+  (\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD) \ra
+  \wf\GG\sim \mor\Sub\,\wf\GG\,\wf{\GD}\,\wf{\sigma}
 \end{alignat*}

### 2020-05-03 paper/paper.tex
stuff

@@
 %  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
-\end{alignat*} xxx
+\end{alignat*}
 Above we used implicit quantifications for $\GG:\Con$ and
@@
 \end{alignat*}
-The reason that we need the general elimination principle is that
-$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
-this function is recursive-recursive.
+The reason that we need the general elimination principle to define
+$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
+$\ll\blank\rr$ for contexts, hence this function is
+recursive-recursive.
 
@@
 In this paper we show that any model of extensional type theory with
-indexed W-types supports the theory of IIT signatures, hence all
-IITs. The difficulty is that the theory of IIT signatures is a QIIT,
-it is both inductive-inductive and has equality constructors. However
-it can be seen as the well-typed syntax of a small type theory without
-any computation rules. Hence we can represent the syntax of normal
-forms without quotienting. We construct this well-typed normal syntax
-using preterms and typing relations from indexed W-types. Finally, we
-prove the elimination principle in the style of the initiality proof
-of Streicher \cite{streichersemantics}.
-
-Just as ibid., we only consider finitary IITs, that is, constructors
-can only have a finite number of recursive arguments. An example
-constructor for $\Con$--$\Ty$ which is not allowed is the following:
+indexed W-types supports the theory of IIT signatures, and as a
+consequence all IITs. The difficulty in this construction is that the
+theory of IIT signatures is itself a QIIT, it is both
+inductive-inductive and has equality constructors. However it can be
+seen as the well-typed syntax of a small type theory without any
+computation rules. Hence we can represent the syntax of normal forms
+without quotienting. We construct this well-typed normal syntax using
+preterms and typing relations from indexed W-types. Finally, we prove
+the elimination principle in the style of the initiality proof of
+Streicher \cite{streichersemantics}.
+
+Just as \cite{Kaposi:2019:CQI:3302515.3290315}, we only consider
+finitary IITs, that is, constructors can only have a finite number of
+recursive arguments. An example constructor for $\Con$--$\Ty$ which is
+not allowed is the following:
 \[
@@
 
-TODO: list of contributions here. Notion of signature free from encoding.
+\paragraph*{Structure of paper and list of contributions}
+
+We describe related work in \Cref{sec:related}, and explain our
+notation and Agda formalisaton in \Cref{sec:notation}. Then the
+following three sections describe our three contributions:
+\begin{itemize}
+\item \Cref{sec:theory_of_signatures}. We define what it means for a
+  model of extensional type theory (ETT, \Cref{def:ett}) to support
+  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
+  contribution here is a (predicative) Church encoding of signatures
+  following \cite{DBLP:conf/lics/AwodeyFS18}.
+\item \Cref{sec:constructingiits}. We show that if a model of ETT
+  supports the theory of IIT signatures
+  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
+  sense: \Cref{th:tosToIITs}. This is an adaptation of a proof in
+  \cite{Kaposi:2019:CQI:3302515.3290315}.
+\item \Cref{sec:ambroise}. Our main contribution is showing that if a
+  model of ETT supports indexed W-types, then it supports the theory
+  of IIT signatures (\Cref{TODO}), and hence, all IITs (\Cref{TODO}).
+\end{itemize}
+We conclude in \Cref{sec:conclusions}.
 
@@
 
-\subsection{Structure of the Paper}
-
-We first define a theory of IIT signatures in which every context is a signature
-(Section \ref{sec:theory_of_signatures}). We present this as an algebraic
-\cite{ttintt} theory, specifically as a quotient inductive-inductive type
-(QIIT). In Section \ref{sec:ambroise} we show that this QIIT can be constructed
-using inductive types, UIP and function extensionality. In Section
-\ref{sec:constructingiits} we describe the construction of IITs from the syntax
-of IIT signatures.
-
 \subsection{Related Work}
+\label{sec:related}
 
@@
 infinitary QIITs \cite{large_inf_qiit} to derive the specification of
-the elimination principle for the theory of IIT signatures. TODO: also
-for specifying signatures.
+the elimination principle for the theory of IIT signatures.
 
@@
 
+Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18} construct
+inductive types using a restricted Church encoding in a type theory
+with an impredicative universe. We use the predicative version of
+their encoding to define IIT signatures.
+
 Our reduction of IITs to indexed inductive types goes through two
@@
 \subsection{Notation and Formalisation}
-
-\begin{definition}[Model of extensional type theory (ETT)]
+\label{sec:notation}
+
+\begin{definition}[Model of extensional type theory (ETT)]\label{def:ett}
   By a model of ETT we mean a category with families (CwF)
@@
   predicative hierarchy of universes closed under the following type
-  formers: $\Pi$, $\Sigma$, $\bot$, $\top$, $\Bool$, indexed W-types
-  \cite{indexedcont} and an identity type with uniqueness of identity
-  proofs and equality reflection. TODO: remove indexed W-types, bot
-  and bool.
+  formers: $\Pi$, $\Sigma$, $\top$ and an identity type with
+  uniqueness of identity proofs and equality reflection.
 \end{definition}
@@
   pairs.
-\item We will define indexed W-types \cite{indexedcont} by listing
-  their constructors and using pattern matching to eliminate from
-  them.
+\item The unit type $\top$ has the constructor $\tt$ which is
+  definitionally equal to all elements of $\top$.
 \item The equality (identity) type is written $t = u$, it has a
@@
   the same $=$ sign for definitional equality. We occasionally
-  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to the
-  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
-  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
-  function extensionality, expressing
-  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
+  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to
+  the equalities $e_1$,\dots,$e_n$. To construct proofs, sometimes we
+  write equational reasoning, e.g. $f a \overset{e}{=} f b$ where
+  $e : a = b$. We also have uniqueness of identity proofs (UIP),
+  expressing $(e : t = t)\ra e = \refl$. Note that function
+  extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$ is
+  derivable.
 \end{itemize}
 
-The contents of Sections \ref{sec:theory_of_signatures} and
-\ref{sec:ambroise} were formalised in Agda, the formalisation is
-available at \url{https://github.com/amblafont/UniversalII}. Agda's
-pattern matching mechanism implies uniqueness of identity proofs, we
-assumed function extensionality as an axiom and used rewrite rules
+The contents of Section \ref{sec:ambroise} were formalised in Agda,
+the formalisation is available at
+\url{https://github.com/amblafont/UniversalII}. Agda's pattern
+matching mechanism implies uniqueness of identity proofs, we assumed
+function extensionality as an axiom and used rewrite rules
 \cite{cockxsprinkles} to obtain limited equality reflection.
@@
 
-\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
+\begin{definition}[A model of ETT supports IITs]\label{def:hasiits}
   A model of ETT supports IITs if for any signature $(sig,sigc):\Sign$
@@
 
-One can check that \Cref{sec:hasiits} gives the right
+One can check that \Cref{def:hasiits} gives the right
 notion of constructors and elimination principle for the signatures in
@@
 
-\section{Further Work}
+\section{Conclusions}
+\label{sec:conclusions}
+
+TODO
 

### 2020-05-03 paper/paper.tex
stuff

@@
   & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
-  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
-  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
+  & \pi_1 && : \Sub\,({\pi_1}^M\,\sigma) && = {\pi_1}^N\,(\Sub\,\sigma) \\
+  & \pi_2 && : \Tm\,({\pi_2}^M\,\sigma) && = {\pi_2}^N\,(\Sub\,\sigma) \\
   & \rlap{(2) Universe} \\
@@
 notation for defining such inductive families: we list the sorts and
-constructors and use pattern matching when eliminating from them. qFor
+constructors and use pattern matching when eliminating from them. For
 an encoding of mutual inductive families as families as indexed

### 2020-05-03 paper/universaliit.tex
stuff

@@
 
-The construction consists of the following steps:
-
+We construct the theory of IIT signatures in the following steps:
 \begin{enumerate}
-\item Definition of untyped syntax (as a family of inductive datatypes) together
-  with typing judgments (as inductive relations on the untyped syntax), and
-  construction of a model of the theory of IIT signatures from well-formed
-  terms, denoted $\mathsf{S}$.
-\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
+\item We view the theory of signatures as a type theory, and we define
+  its untyped syntax as mutual inductive types together with typing
+  judgments given by inductive predicates on the untyped syntax. Then
+  the syntax $\I : \SignAlg$ is constructed using those untyped terms
+  for which the typing predicate holds.
+\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
+  $M : \SignAlg$, by:
   \begin{enumerate}
@@
     semantic context $\model{\GG}$ of the model $M$, we have
-    $\GG \sim \model{\GG}$ if and only if $\rec\,\,\GG
+    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
     = \model{\GG}$, and similarly for types, terms, and substitutions;
@@
   \item Proving the uniqueness of this morphism by showing that any morphism
-    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
+    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
     $\GG$ we have $\GG\sim f\,\GG$.
@@
 The next sections detail each of these steps.
-\subsection{Syntactic Model}
-
-The goal is to define the syntactic model where contexts are pairs of a
-precontext together with a well-formedness proof, and similarly for types, terms
-and substitutions.
-
-Crucially, we do not have conversion relations for typed syntax, nor do we need
-to use quotients when giving the syntactic model. This is possible because there
-are no $\beta$-rules in the theory of signatures. Hence, we consider only normal
-terms in the untyped syntax, and define weakening and substitution by recursion.
-Avoiding quotients is important for two reasons. First, it greatly simplifies
-formalisation. Second, we aim to reduce IITs to a minimal feature set, and we
-get a stronger result if we do not use quotients.
-
-The next sections present the definition of the untyped syntax and the
+
+\subsection{Syntax}
+
+The goal is to define the syntactic signature algebra where contexts
+are pairs of a precontext together with a well-formedness proof, and
+similarly for types, terms and substitutions.
+
+Crucially, we do not have conversion relations for typed syntax, nor
+do we need to use quotients when giving the syntactic model. This is
+possible because there are no $\beta$-rules in the theory of
+signatures. Hence, we consider only normal terms in the untyped
+syntax, and define weakening and substitution by recursion.  Avoiding
+quotients is important for two reasons. First, it greatly simplifies
+formalisation. Second, we aim to reduce the theory of signatures only
+to inductive types, thus making \Cref{TODO} stronger.
+
+Now we present the definition of the untyped syntax and the
 associated typing judgments.
 
-\subsubsection{Untyped syntax}
-The untyped syntax is defined as the following inductive datatype. Variables are
-modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
-the context. We use the additional default constructor $\erru:\Tmu$ in case of
-error (ill-scoped substitution). The typing judgments will not mention $\erru$.
-The main interest of $\erru$ is that it behaves like a closed term (which the
-theory of signatures lacks), in the sense that it is
-invariant under substitution. This makes expected equalities about 
-substitution true even in the ill-typed case,
-thus reducing the number of hypotheses for the corresponding lemmas
-(see Section~\ref{ss:exch-laws}).
-\begin{alignat*}{5}
-  & \rlap{(1) Substitution calculus} \\[0.5em]
+\subsubsection{Untyped syntax and its properties}
+
+\begin{definition}[Untyped syntax]
+The untyped syntax is defined as the following inductive datatype.
+\begin{alignat*}{5}
+  & \rlap{(1) Substitution calculus} \\
   & \Conu && : \Set \\
@@
   & \varu  && : \N \ra \Tmu \\
-  & \rlap{(2) Universe} \\[0.5em]
+  & \rlap{(2) Universe} \\
   & \Uu && : \Tyu \\
   & \Elu && : \Tmu\ra\Tyu \\
-  & \rlap{(3) Inductive parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\
   & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
   & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
-  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
+  & \rlap{(4) Metatheoretic parameters} \\
   & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
@@
   & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
-  & \rlap{(6) Default value} \\[0.5em]
+  & \rlap{(5) Default value} \\
   & \erru && : \Tmu
 \end{alignat*}
+\end{definition}
+Variables are modeled as de Bruijn indices, i.e.\ as natural numbers
+pointing to a position in the context. We use the additional default
+constructor $\erru:\Tmu$ in case of error (ill-scoped
+substitution). The typing judgments will not mention $\erru$.  The
+main interest of $\erru$ is that it behaves like a closed term (which
+the theory of signatures lacks), in the sense that it is invariant
+under substitution. This makes expected equalities about substitution
+true even in the ill-typed case, thus reducing the number of
+hypotheses for the corresponding lemmas (see
+\Cref{lem:exch-laws}).
 
@@
 
-\subsubsection{Untyped weakening}
-
-% Substitutions of types and terms are defined recursively.
-
-Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
-\,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$, and thus we need to define
-$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
-Bruijn indices of all the variables.  Actually, this is not so simple because of
-the $\Piu$ type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
-of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
-the second last variable of the context, rather than the last one.  For this
-reason, we need to generalise the weakening as occuring anywhere in the context.
+We will define substitutions $\blank[\blank]$ of types and terms
+recursively.
+
+Note that $(\Piu \,A\,B)[\sigma]$ should be defined as
+$\Piu \,(A[\sigma])\,(B[\wkO\,\sigma \,\consu\, \varu\,\mathsf{0}])$,
+and thus we need to define $\wkO$, the weakening of substitutions.
+The basic idea is to increment the de Bruijn indices of all the
+variables.  Actually, this is not so simple because of the $\Piu$
+type: we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type of the
+weakening of $A$ and $B$, but here, $B$ must be weakened with respect
+to the second last variable of the context, rather than the last one.
+For this reason, we need to generalise the weakening as occuring
+anywhere in the context.
+
+\begin{definition}[Untyped weakening]
+  We define untyped weaking recursively on terms by the following
+  functions.
   \begin{alignat*}{5}
@@
   Here, $\wkO$ weakens with respect to the last variable.
-
-Later, in Section~\ref{ss:weaken-typing}, we show that weakening preserves
-typing. Stating a typing rule for this operation requires weakening at the
-middle of a context. We consider pairs of untyped contexts, which should be
-thought of as a splitting of a context at some position. The full context is
-recovered by merging the two components:
+\end{definition}
+
+Later, in \Cref{lem:weaken-typing}, we show that weakening preserves
+typing.  Stating a typing rule for this operation requires weakening
+at the middle of a context. This is why we define pairs of untyped
+contexts, which should be thought of as a splitting of a context at
+some position. We call the second context a telescope over the first
+one.
 % (which appears in the typing rule of a weakening occuring in the middle of a context),
@@
 % context $\GG$:
+\begin{definition}[Untyped telescopes]
+  An untyped telescope is given simply by a $\Conu$.
+\end{definition}
+\begin{definition}[Merging of a context and a telescope]
 \begin{alignat*}{5}
@@
 \end{alignat*}
-We think of the second context as a telescope over the first one. We also define
-weakening for telescopes, which will be used to give typing rules for telescopes
-in Section \ref{sec:typing_judgments}:
+\end{definition}
+\begin{definition}[Weakening for telescopes]
+Weakening for telescopes is defined pointwise. $\length{\GG}$ denotes
+the length of the context $\GG$.
 \begin{alignat*}{5}
@@
   \end{alignat*}
-$\length{\GG}$ denotes the length of a the context $\GG$.
-
-\subsubsection{Untyped substitution}
-  We define single substitution by recursion on presyntax:
+\end{definition}
+This will be used to give typing rules for telescopes
+in \Cref{def:typing_judgments}.
+
+\begin{definition}[Untyped unary substitution]
+  We define single substitution by recursion on the presyntax:
   % & \rlap{(1) Substitution calculus} \\[0.5em]
@@
 \end{alignat*}
-  This is enough to define the typing judgments: indeed, the typing rule for application
+\end{definition}
+  This is enough to state the typing judgments: indeed, the typing rule for application
   involves only a unary substitution.
 
-  However, to construct the initial model of the QIIT, we need to define
-  the full substitution calculus:
+  However, to construct the syntax as a signature algebra, we need to define
+  parallel substitutions:
+\begin{definition}[Untyped substitution calculus]
   \begin{alignat*}{5}
@@
 on untyped syntax: the two ways yield provably equal definitions. In the
-following, we assume that it is defined by recursion. We also make use of the
+following, we assume that they are defined by recursion. We also make use of the
 following definition:
@@
   & \keep && : \Subu \ra \Subu \\
-   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
-  \end{alignat*}
-  The idea is that if $\sigma$ is a substitution between contexts $\GG$ and
+   &&& := \lambda \sigma. (\wkO\,\sigma \,\consu\, \varu\, \mathsf{0})
+  \end{alignat*}
+  The idea is that if $\sigma$ is a substitution from $\GG$ to
   $\GD$, then $\keep\,\sigma$ is a substitution between contexts $\GG\ext A[\sigma]$
-  and $\GD\ext A$ for any type $A$. This occurs when defining
+  and $\GD\ext A$ for any type $A$ where the last term is just a de Bruijn index $0$. This occurs when defining
   $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.
 
-  We can define the identity substitution on a context $\GG$ as
+  We define the identity substitution on a context $\GG$ as
   follows, where $\keep^{\length{\GG}}$ is $\keep$ iterated
@@
   \end{alignat*}
-
-  \subsubsection{Exchange laws: weakening and substitution}
-  \label{ss:exch-laws}
-  Many lemmas for types and terms are shown by induction on the untyped syntax.
-  Below, $Z$ denotes either a term or a type.
+\end{definition}
+
+\begin{lemma}[Exchange laws for weakening and substitution]\label{lem:exch-laws}
+  Below, $Z$ denotes either a term or a type and $\keep^n$ denotes
+  the $n$ times iteration of $\keep$.
 \begin{alignat*}{5}
@@
   % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
-  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])]
-  \end{alignat*}
-  Below, $\keep^n$ denotes $\keep$ iterated $n$ times.
-\begin{alignat*}{5}
+  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])] \\
   % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
@@
   :
-  (\wk_n\,Z) [\keep^n\, (u \consu \sigma)] = Z [ \keep^n \sigma]
+  (\wk_n\,Z) [\keep^n\, (\sigma \consu u)] = Z [ \keep^n \sigma]
   \\
@@
   \end{alignat*}
+\end{lemma}
+\begin{proof}
+By induction on the untyped syntax.
+\end{proof}
+
+\begin{corollary}
 As particular cases for $n=0$, we get
@@
   & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
-  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau \\
+  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (\tau \consu t) = \sigma \circ \tau \\
   & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
-  & \wk_0[\cons] && : (\wk_0 Z)[u \consu \sigma] = Z[\sigma] \\
-  & [0:=][] && : Z [ 0:= u][ \sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
-\end{alignat*}
-Finally, we show the following:
+  & \wk_0[\cons] && : (\wk_0\,Z)[\sigma \consu u] = Z[\sigma] \\
+  & [0:=][] && : Z [ 0:= u][\sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
+\end{alignat*}
+\end{corollary}
+
+\begin{lemma}[Composition functor law and associativity]
 \begin{alignat*}{5}
@@
 \end{alignat*}
+\end{lemma}
+
 We defer laws for identity substitutions after the definition of the typing
@@
 
-  \subsubsection{Typing judgments}
-  \label{sec:typing_judgments}
-  The typing judgments are defined as the following inductive datatype indexed over the
+\subsubsection{Typing judgments and their properties}
+
+\begin{definition}[Typing judgements]\label{def:typing_judgments}
+  The typing judgments are defined as the following inductive type indexed over the
   untyped syntax:
@@
   \end{alignat*}
+\end{definition}
 There is possibility of redundancy in the arguments of the constructors. Here,
-we are ``{paranoid}'', so that we get more inductive hypotheses when performing
+we are ``{paranoid}'' (following nomenclature from xxx), so that we get more inductive hypotheses when performing
 recursion.
@@
 \subsubsection{Weakening preserves typing}
-\label{ss:weaken-typing}
+\label{lem:weaken-typing}
 We prove by mutual induction that typing judgments are stable under weakening,

### 2020-05-03 paper/universaliit.tex
stuff

@@
   its untyped syntax as mutual inductive types together with typing
-  judgments given by inductive predicates on the untyped syntax. Then
+  judgments given by inductive relations on the untyped syntax. Then
   the syntax $\I : \SignAlg$ is constructed using those untyped terms
-  for which the typing predicate holds.
+  for which the typing relation holds.
 \item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
@@
   & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
-  & \rlap{(4) Metatheoretic parameters} \\
+  & \rlap{(4) External parameters} \\
   & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
@@
     ( \GG \vdash t \in A[\sigma]) \ra
-    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
+    \GG \vdash \sigma\consu t \Rightarrow \GD\extu A
    \\
@@
     \GG \vdash t \appu u \in  B [0 := u] \\
-  & \rlap{(4) Inductive parameters} \\[0.5em]
+  & \rlap{(4) External parameters} \\[0.5em]
   & \Pimw && :
@@
 \end{definition}
-There is possibility of redundancy in the arguments of the constructors. Here,
-we are ``{paranoid}'' (following nomenclature from xxx), so that we get more inductive hypotheses when performing
-recursion.
+There is possibility of redundancy in the arguments of the
+constructors. Here, we are ``{paranoid}'' (nomenclature from
+\cite{paranoid}), so that we get more inductive hypotheses when
+performing recursion.
 
@@
 
-\subsubsection{Weakening preserves typing}
-\label{lem:weaken-typing}
-We prove by mutual induction that typing judgments are stable under weakening,
-for contexts, types, terms, and substitutions.
+\begin{lemma}[Weakening preserves typing]\label{lem:weaken-typing}
 % wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
@@
   \end{alignat*}
-\subsubsection{Substitution preserves typing}
+\end{lemma}
+\begin{proof}
+By mutual induction on the typing relations.
+\end{proof}
+
 We show that judgments are stable under substitution.
+\begin{lemma}[Substitution preserves typing]
 \begin{alignat*}{5}
@@
 \end{alignat*}
-\subsubsection{Laws for identity substitutions}
-We show category and functor laws involving identity substitution for
-well-formed types, terms and substitutions.
+\end{lemma}
+\begin{proof}
+By mutual induction on the typing relations.
+\end{proof}
+
+We show the category and functor laws involving identity substitution
+for well-formed types, terms and substitutions.
+\begin{lemma}[Identity laws]\label{lem:identity}
 \begin{alignat*}{5}
   % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
-  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG]  \\
+  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG] = A \\
   & [\idu] && : (\GG \vdash x \invar A)\ra x [ \idu\,\GG] = V x \\
@@
   \end{alignat*}
+\end{lemma}
 Finally, we show that the identity substitution itself is well-typed:
+\begin{lemma}[Typing for the identity substitution]
   \begin{alignat*}{5}
@@
   \end{alignat*}
-
-\subsubsection{Proof irrelevance and unicity of typing}
-\label{ss:uniq-typing-types}
+\end{lemma}
+
+\begin{definition}[Proposition]
   A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
@@
   \]
-We show that each of the typing judgments is unique in the
-following sense:
+\end{definition}
+
+\begin{lemma}[Proof irrelevance of typing relations]\label{lem:uniq-typing-types}
 \begin{alignat*}{5}
@@
 \end{alignat*}
-We also show unicity of typing:
+\end{lemma}
+
+\begin{lemma}[Unicity of typing]
 \begin{alignat*}{5}
@@
 \end{alignat*}
+\end{lemma}
 Let us consider for instance the application constructor $\appw$: for a codomain
@@
 
- \subsubsection{Syntactic model}
- The syntactic model is obtained by packing the untyped syntax with the typing
- judgments:
+\subsubsection{The syntax as a signature algebra}
+
+\begin{definition}[Syntax for the theory of signatures]\label{def:syntax}
+  We define the syntax as an element of $\SignAlg$ by pairs of untyped
+  syntax and typing relations:
 % % \subsection{QIIT model}
@@
 enough to prove them only for the untyped syntactic part: as we argued in
-Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are
+\Cref{lem:uniq-typing-types}, the proofs of typing judgments are
 automatically equal.
-
-Up until the construction of the syntactic model, UIP is not used. Function
-extensionality on the other hand is necessary because the untyped metatheoretic
-$\Pi$ takes a metatheoretic function as an argument. An example induction step
-that uses function extensionality is in the type preservation proof for identity substitutions (this
-is an equation satisfied by a model of the QIIT), in particular in the case
-$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
-equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
-hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.
+\end{definition}
+
+\begin{remark}
+  Up until \Cref{def:syntax}, UIP is not used. Function extensionality
+  on the other hand is necessary because the untyped metatheoretic
+  $\Pi$ takes a metatheoretic function as an argument. An example
+  induction step that uses function extensionality is in
+  \Cref{lem:identity}, in particular in the case
+  $(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this
+  equation is equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by
+  definition, whereas the induction hypothesis states that
+  $(t:T)\ra (A\,t)[\id]=A\,t$.
+\end{remark}
 
@@
 \label{sec:streicher}
-It remains to show that the constructed syntactic model is initial. To achieve
-this, we first define a relation between the syntactic model and an arbitrary
-model, then show that the relation is functional, which lets us extract a
-homomorphism from the relation.
-
-This approach is an alternative version of Streicher's method for interpreting
-preterms in an arbitrary model \cite{streichersemantics}. Streicher first defines
-a family of partial maps from the presyntax to a model, then shows that the maps
-are total on well-formed input. We have found that our approach is significantly
-easier to formalise. To see why, note that the right notion of partial map in
-type theory, which does not presume decidable definedness, is fairly
-heavyweight:
+
+It remains to show that the constructed syntax $\I$ is the initial
+signature algebra. To achieve this, we first define a relation between
+the syntax and any signature algebra, then show that the relation is
+functional, which lets us extract a signature morphism from the relation.
+
+This approach is an alternative presentation of Streicher's method for
+interpreting preterms in an arbitrary model of type theory
+\cite{streichersemantics}. Streicher first defines a family of partial
+maps from the presyntax to a model, then shows that the maps are total
+on well-formed input. We have found that our approach is significantly
+easier to formalise. To see why, note that the right notion of partial
+map in type theory, which does not presume decidable definedness, is
+fairly heavyweight:
 \[
@@
 
-Given a model $M$ of the QIIT, we define the functional relation satisfied by
+xxx
+
+Given an $M : \SignAlg$, we define the functional relation satisfied by
 the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.

### 2020-05-03 paper/paper.tex
stuff

@@
 \begin{abstract}
-  TODO: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
+  xxx: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
   type theory. They allow the mutual definition of types with multiple sorts
@@
   on function extensionality and uniqueness of identity proofs. We formalised
-  the construction of the theory of IIT signatures in Agda. TODO: rewrite
+  the construction of the theory of IIT signatures in Agda.
 \end{abstract}
@@
   (\Cref{def:theoryofsignatures}), then it supports IITs in the above
-  sense: \Cref{th:tosToIITs}. This is an adaptation of a proof in
+  sense: \Cref{thm:ToSToIITs}. This is an adaptation of a proof in
   \cite{Kaposi:2019:CQI:3302515.3290315}.
@@
   model of ETT supports indexed W-types, then it supports the theory
-  of IIT signatures (\Cref{TODO}), and hence, all IITs (\Cref{TODO}).
+  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
 \end{itemize}
@@
   use this approach because it is more technical, and it would not
-  strengthen our main theorem (TODO: refer) as theorem (TODO: refer)
-  needs $\mathcal{C}$ to be a model of ETT.}
+  strengthen our main result \Cref{thm:WToIITs} as the proof of
+  \Cref{thm:WToToS} needs $\mathcal{C}$ to be a model of ETT.}
 
@@
 take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
-usual Church-encoding of contexts. However (as we will see in TODO)
-the notion of constructor for such signatures would be too
-strong. Another approach would be to assume that there is a syntax for
-signature algebras (an initial signature algebra), and then a
-signature would be a context in this signature algebra. We will define
-syntactic signatures using this approach in the next section
+usual Church-encoding of contexts. However (as we will see in
+\Cref{remark:sigc}) the notion of constructor for such signatures
+would be too strong. Another approach would be to assume that there is
+a syntax for signature algebras (an initial signature algebra), and
+then a signature would be a context in this signature algebra. We will
+define syntactic signatures using this approach in the next section
 (\Cref{def:syntacticsignature}), but for now we don't want to assume
@@
 
-\begin{theorem}\label{th:tosToIITs}
+\begin{theorem}\label{thm:ToSToIITs}
   If a model of type theory supports the theory of IIT signatures,
@@
 
+\begin{remark}\label{remark:sigc}
+  In the above proof, we crucially relied on the ${sigc}$ property to
+  define the constructors (and the eliminator). This is why the simple
+  Church encoding of signatures is not sufficient.
+\end{remark}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 W-types also supports the theory of signatures, and as a consequence
-of \Cref{th:tosToIITs}, all IITs. For this, we work in the internal
+of \Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal
 language of a model of ETT supporting indexed W-types

### 2020-05-03 paper/universaliit.tex
stuff

@@
     % syntax to a given model;
-  \item showing that this relation is functional.
+  \item showing that this relation is functional and thus obtaining a function.
     \end{enumerate}
@@
 formalisation. Second, we aim to reduce the theory of signatures only
-to inductive types, thus making \Cref{TODO} stronger.
+to inductive types, thus making \Cref{thm:WToToS} stronger.
 
@@
 
-\subsubsection{Untyped syntax and its properties}
+\subsubsection{Untyped Syntax and its Properties}
 
@@
 
-\subsubsection{Typing judgments and their properties}
-
-\begin{definition}[Typing judgements]\label{def:typing_judgments}
-  The typing judgments are defined as the following inductive type indexed over the
+\subsubsection{Typing Relations and Their Properties}
+
+\begin{definition}[Typing relations]\label{def:typing_judgments}
+  The typing relations are defined as the following inductive type indexed over the
   untyped syntax:
@@
 
-\subsubsection{The syntax as a signature algebra}
+\subsubsection{The Syntax as a Signature Algebra}
 
@@
 
-\subsection{Relating the Syntax to a Model}
+\subsection{Relating the Syntax to a Signature Algebra}
 \label{sec:streicher}
@@
 
-xxx
-
 Given an $M : \SignAlg$, we define the functional relation satisfied by
-the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
+the  $\ll\blank\rr : \SignMor\,\I\, M$ by recursion on the typing judgments.
 % As these judgments are propositions, we will sometimes
-If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
-(i.e.\ a context of the model $M$), we want to define a type
-$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
-course, at this stage, $\rec$ is not available yet since the point of
-defining this relation is to construct $\rec$ in the end.
+If $\GG$ is a context in $\I$ and $\model{\GG}$ is a semantic context
+(i.e.\ a context in the signature algebra $M$), we want to define a type
+$\GG\sim \model{\GG}$ equivalent to $\ll\GG\rr =\model{\GG}$. Of
+course, at this stage, $\ll\blank\rr$ is not available yet since the point of
+defining this relation is to construct $\ll\blank\rr$ in the end.
 
 For a type $A$ in a context $\GG$, we want to define a relation
-$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
+$A\sim \model{A}$ that is equivalent to $\ll A\rr=\model{A}$.
 % where $\model{A}$ is a
@@
 For this equality to make sense, the semantic type $\model{A}$ must live
-in the semantic context $\rec\,\,\GG$. But again, $\rec$ is not
+in the semantic context $\ll\GG\rr$. But again, $\ll\blank\rr$ is not
 yet available at this stage. Exploiting the expected equivalence between
-$\GG\sim\model\GG$ and $\rec\,\,\GG=\model\GG$, we may consider defining $A\sim\model{A}$ under the
+$\GG\sim\model\GG$ and $\ll\GG\rr=\model\GG$, we may consider defining $A\sim\model{A}$ under the
 hypotheses that $\model{A}$ lies in a semantics context $\model\GG$ which is
@@
 \[
-  (\GG : \syn\Con)\ra (A : \syn\Ty\,\syn\GG) \ra
+  (\GG : \syn\Con)\ra (A : \syn\Ty\,\GG) \ra
   (\model\GG : \model\Con)
@@
 Let us mention another possible solution for avoiding recursion-recursion:
-defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\GG
-= \model{\GG})\times (\rec\,A = _{e\#}\model{A})$.
+defining $A\sim\model{A}$ so that it is equivalent to $(e:\ll\GG\rr
+= \model{\GG})\times (\ll A\rr = _{e\#}\model{A})$.
 % whereas the right hand side is of type $\model\Ty\model{\GG}$.
@@
 
+\begin{definition}[Relation $\blank\sim\blank$]
 We define the relation by recursion on the typing judgments.
@@
 \begin{alignat*}{5}
-  & \rlap{(1) Substitution calculus} \\[0.5em]
+  & \rlap{(1) Substitution calculus} \\
   & \blank \sim{ \blank} && :  \GG \vdash \, \ra \model{\Con} \ra \Set \\
@@
   \\
-  & \rlap{(2) Universe} \\[0.5em]
+  & \rlap{(2) Universe} \\
   & \Uw\, \wf{\GG} \wf{A} \sim \model{A} &&
@@
   \\
-  & \rlap{(3) Inductive parameters} \\[0.5em]
+  & \rlap{(3) Inductive parameters} \\
   & \Piw\,\wf{\GG} \wf{a} \wf{B} \sim \model{C} &&
@@
     \\
-  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
+  & \rlap{(4) Metatheoretic parameters} \\
   & \Pimw T\,A\,\wf{\GG}\wf{A} \sim \model{B} &&
@@
 \end{alignat*}
+\end{definition}
 
@@
   % relations are required to show that the relation is right unique.
+
+\subsubsection{Right Uniqueness}
+
 Next, we prove that this relation is right unique.  Then, we show that the
@@
 judgments.
-  \subsubsection{Right uniqueness}
-  \label{sec:right_uniqueness}
-  We show by induction that the relation is right unique in the following sense:
+
+\begin{lemma}[Right uniqueness]\label{lem:right_uniqueness}
+The relation is right unique in the following sense:
   \begin{alignat*}{5}
@@
   \end{alignat*}
+\end{lemma}
+
+\begin{remark}
   We mentioned that in order to avoid a recursive-recursive definition, we
@@
   that well-formedness predicates used in syntactic models can subtly break
-  in that setting. Also, while Hugunin does not use UIP, he only shows a
-  weak version of dependent elimination for the constructed IITs. Hence, the
+  in that setting. Also, while Hugunin does not use UIP, he only shows the simple version
+  version of dependent elimination for the constructed IITs. Hence, the
   question whether IITs are reducible to inductive types in a UIP-free setting
   remains open.
+\end{remark}
+
+\subsubsection{Stability under Weakening and Substitution}
 
@@
   %    well-formedness relations implies that UIP is required.}
-  \subsubsection{Stability under weakening}
   % Let us argue that it is necessary to show s
@@
   We would like to apply the induction hypothesis, so we need to show that
-  $\keep\,\sigma=\varu\,\mathsf{0}\consu\wkO\,\sigma$ is related to
+  $\keep\,\sigma=\wkO\,\sigma\consu \varu\,\mathsf{0}$ is related to
   $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
   related to $\model{\sigma}$.
-  As $\keep\,\sigma=\varu\, \mathsf{0}\consu \wkO\,\sigma$, we are left with showing that
+  As $\keep\,\sigma=\wkO\,\sigma\consu \varu\, \mathsf{0}$, we are left with showing that
   $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO\,\id$)
@@
 The substitution from the weakened context to the original one must
-be computed at the same time otherwise the recursion hypothesis is not strong enough.
-Then, we seperate the two components under the same (implicit) hypotheses:
+be computed at the same time otherwise the induction hypothesis is not strong enough.
+Then, we separate the two components under the same (implicit) hypotheses:
 \begin{alignat*}{5}
@@
 
-Now, we are ready to prove by mutual recursion on well-typed judgments that
-weakening preserves typing. The following statements are all under the
+\begin{lemma}[Weakening preserves typing]
+The following statements are all under the
 hypotheses
@@
 \end{alignat*}
-Then we deduce, still by induction, that weakening of substitution preserves the
-relation:
+\end{lemma}
+\begin{proof}
+By mutual induction on the typing judgments.
+\end{proof}
+
+\begin{lemma}[Weakening of substitution preserves $\blank\sim\blank$]
 \begin{alignat*}{5}
@@
   \end{alignat*}
+\end{lemma}
+\begin{proof}
+By induction on the typing judgments.
+\end{proof}
 
@@
 
-\subsubsection{Stability under substitution}
-
- We want to prove that given any well-typed substitution $\sigma
+Next, we want to prove that given any well-typed substitution $\sigma
 : \Sub\,\GG\,\GD$ and semantic contexts $\model{\GG}$ and
-$\model{\GD}$, respectively related to $\GG$ and $\GD$, there exists a
-semantic substitution which is related to $\sigma$.  In the extension case
+$\model{\GD}$, related to $\GG$ and $\GD$, respectively, there is a
+semantic substitution related to $\sigma$.  In the extension case
 $\GG\vdash \sigma \consu\,t\Rightarrow \GD\,\extu\,A$, the induction
@@
 $\model{A}\model{[\model{\sigma}]}$ is related to the syntactic type
-$A[\sigma]$. We first establish preservation of the relation by substitution for
-variables:
+$A[\sigma]$.
+\begin{lemma}[Preservation of the relation by substitution for variables]
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Then we show it for terms and types  by mutual induction
-under the common hypotheses
+\end{lemma}
+\begin{proof}
+  Induction on typing.
+\end{proof}
+
+\begin{lemma}[Preservation of the relation by substitution for types and terms]
+We assume
   $(\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD)$,
@@
  \end{alignat*}
-  Eventually, we show under the same hypotheses the following statement
+\end{lemma}
+\begin{proof}
+  Mutual induction on typing.
+\end{proof}
+\begin{lemma}[The relation is preserved by composition and identity]
+  We have the same hypotheses as in the previous lemma.
   \begin{alignat*}{5}
@@
   & && \qquad
-   \wf{{\circ}}\,\wf{\GG}\,\wf{\delta}\,\wf{\sigma} \sim \model{\delta}\mathbin{\model{\circ}}\model{\sigma}
-  \end{alignat*}
-  and the fact that identity preserves the relation:
-  \begin{alignat*}{5}
+   \wf{{\circ}}\,\wf{\GG}\,\wf{\delta}\,\wf{\sigma} \sim \model{\delta}\mathbin{\model{\circ}}\model{\sigma} \\
      & {\id}{\sim} && :
@@
  \end{alignat*}
+ \end{lemma}
 % $\model{t}\model\oldapp\model{u}$, but it has type
@@
 
-\subsection{Left-Totality and the Recursor}
-For the recursor, we build a morphism from the syntactic model to the semantic
-one. The image of a syntactic context is a unique semantic context which is
-related to it, and similarly for types and terms.
+\subsubsection{Left-Totality and the Recursor}
+
+Before defining the recursor $\ll\blank\rr$, we show left totality of
+the relation: that is, the image of a syntactic context is a unique
+semantic context which is related to it, and similarly for types and
+terms.
 % We construct it as the semantic
@@
 % semantic counter-part.
-Thus, as a first step, we use induction on well-formedness judgments to construct
-semantic counterparts:
+\begin{lemma}[Left totality of $\blank\sim\blank$]
 \begin{alignat*}{5}
@@
   \end{alignat*}
-  The right uniqueness of the relation is used in this induction. It is then
-straightforward to show (without induction, but using right uniqueness) that the
-first projection of these constructions yields a model morphism from the syntax
-to the model.
+\end{lemma}
+\begin{proof}
+  By induction on well-formedness judgments. The right uniqueness of
+  the relation is used in this induction.
+\end{proof}
+
+\begin{lemma}[Existence of the recursor]\label{lem:rec}
+  For any $M:\SignAlg$ there is a $\ll\blank\rr : \SignMor\,\I\,M$
+  where $\I$ is given in \Cref{def:syntax}.
+\end{lemma}
+\begin{proof}
+  Using the first projections in the construction of the left-totality
+  construction and right uniqueness.
+\end{proof}
 
 \subsection{Uniqueness}
-Our goal is to show that the syntactic model is initial. Thus, it remains to
-show that the morphism constructed in the previous section is unique. We exploit
-right uniqueness of the relation: it is enough to show that any such morphism
-maps a syntactic context to a related semantic context, and similarly for types
-and terms.
-
-More formally, we assume an arbitrary morphism $f$ from the
-syntax to the model, inducing the following maps:
+
+It remains to show that the morphism constructed in \Cref{lem:rec} is
+unique. We exploit right uniqueness of the relation: it is enough to
+show that any such morphism maps a syntactic context to a related
+semantic context, and similarly for types and terms.
+
+\begin{lemma}
+  We assume an arbitrary signature morphism $f$ from $\I$ to $M$. This
+  induces the following maps:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Then, we show by induction on typing judgments that
-the image is related:
+The images of the above maps are related by $\blank\sim\blank$:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-This justifies the uniqueness of the morphism, by right uniqueness of
-$\blank\sim\blank$.
+\end{lemma}
+\begin{proof}
+  By induction on typing relations.
+\end{proof}
+
+\begin{corollary}[Uniqueness of the recursor]\label{lem:uniq}
+  By right uniqueness of $\blank\sim\blank$, there is only one
+  morphism $\SignMor\,\I\,M$ for any $M$.
+\end{corollary}
+
+\begin{theorem}\label{thm:WToToS}
+  If a model of ETT supports indexed W-types, it supports the theory
+  of IIT signatures.
+\end{theorem}
+\begin{proof}
+  We define the syntax $\I$ by \Cref{def:syntax} which only used
+  indexed W-types, the recursor by \Cref{lem:rec} and we prove its
+  uniqueness property by \Cref{lem:uniq}.
+\end{proof}
+
+\begin{corollary}\label{thm:WToIITs}
+  If a model of ETT supports indexed W-types, it supports all IITs.
+\end{corollary}
+\begin{proof}
+  Combining \Cref{thm:WToToS} and \Cref{thm:ToSToIITs}.
+\end{proof}
 

### 2020-05-03 paper/paper.tex
abstract

@@
 \begin{abstract}
-  xxx: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
-  type theory. They allow the mutual definition of types with multiple sorts
-  where later sorts can be indexed by previous ones. An example is the
-  Chapman-style syntax of type theory with conversion relations for each sort
-  where e.g.\ the sort of types is indexed by contexts. It follows from previous
-  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
-  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
-  a small domain-specific type theory where a context is a signature for an
-  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
-  thereby showing a reduction of all finitary IITs to inductive types.  We rely
-  on function extensionality and uniqueness of identity proofs. We formalised
-  the construction of the theory of IIT signatures in Agda.
+  Inductive-inductive types (IITs) are a generalisation of inductive
+  types in type theory. They allow the mutual definition of types with
+  multiple sorts where later sorts can be indexed by previous ones. An
+  example is the Chapman-style syntax of type theory with conversion
+  relations for each sort where e.g.\ the sort of types is indexed by
+  contexts. In this paper we show that if a model of extensional type
+  theory (ETT) supports indexed W-types, then it supports finitely
+  branching IITs. We use a small internal type theory called the
+  theory of signatures to specify signatures for IITs. We show that if
+  a model of ETT supports the syntax for the theory of signatures,
+  then it supports all IITs. We construct this syntax from indexed
+  W-types using preterms and typing relations and prove its initiality
+  following Streicher. The construction of the syntax and its
+  initiality proof were formalised in Agda.
 \end{abstract}

### 2020-05-03 paper/paper.tex
finish

@@
 \end{itemize}
-We conclude in \Cref{sec:conclusions}.
+We list further work in \Cref{sec:conclusions}.
 
@@
 
-TODO
-
-The current work only concerns finitary IITs. An extension would be to also
-allow infinitely branching inductive types such as W-types. This would first
-require giving semantics for infinitary IITs (to our knowledge there is no
-previously published semantics that we can borrow), and also giving a term model
-construction analogously to finitary QIITs. These steps seem feasible. However,
-it seems to be more difficult to construct the syntax of infinitary IIT
-signatures without using quotients. The reason is that such syntax would not be
-strictly restricted to neutral terms: we would need $\lambda$-abstraction and
-$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
-construction for infinitary IITs. A definition of normal preterms and typing
-judgments on them may still be possible, but it appears to be much more
-complicated than before (the current authors have attempted this without
-conclusive success).
+The current work only concerns finitary IITs. An extension would be to
+also allow infinitely branching inductive types such as W-types. This
+would first require giving semantics for infinitary IITs and adapting
+the term model construction. These would be straightforward following
+\cite{large_inf_qiit}. However, it seems to be more difficult to
+construct the syntax of infinitary IIT signatures without using
+quotients. The reason is that such syntax would not be strictly
+restricted to neutral terms: the term model construction for
+infinitary IITs requires $\lambda$-abstraction and $\beta\eta$-rules
+for infinitary $\Pi$ types. A definition of normal preterms and typing
+judgments on them may still be possible, but it appears to be much
+more complicated than before (the current authors have attempted this
+without conclusive success).
 

### 2020-05-03 paper/paper.tex
fix

--- 2020-05-03 paper/paper.tex
@@
 \begin{document}
-\maketitle
+%% \maketitle
 
 \begin{abstract}
-  xxx: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
+  TODO: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
   type theory. They allow the mutual definition of types with multiple sorts
@@
 \end{alignat*}
-Above we used implicit quantifications for $\GG:\Con$ and
+Above we use implicit quantifications for $\GG:\Con$ and
 $A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
@@
 Reducing IITs to inductive types (more precisely, to indexed W-types)
-is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
-extensional type theory however this only provides a simpler,
+is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
+extensional type theory, however, this only provides a simpler,
 non-recursive-recursive elimination principle. Hugunin \cite{jasper}
@@
 Kaposi, Kov{\'a}cs, and Altenkirch
-\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory
-called the theory of signatures to describe quotient
+\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory,
+called the theory of signatures, to describe quotient
 inductive-inductive types (QIIT). QIITs are generalisations of IITs
@@
   following \cite{DBLP:conf/lics/AwodeyFS18}.
-\item \Cref{sec:constructingiits}. We show that if a model of ETT
-  supports the theory of IIT signatures
+\item \Cref{sec:constructingiits}. In \Cref{th:tosToIITs}, we show that if a
+  model of ETT supports the theory of IIT signatures
   (\Cref{def:theoryofsignatures}), then it supports IITs in the above
-  sense: \Cref{thm:ToSToIITs}. This is an adaptation of a proof in
+  sense. This is an adaptation of a proof in
   \cite{Kaposi:2019:CQI:3302515.3290315}.
@@
   model of ETT supports indexed W-types, then it supports the theory
-  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
+  of IIT signatures (\Cref{TODO}), and hence, all IITs (\Cref{TODO}).
 \end{itemize}
@@
 
-IITs (however not by this name) were first used to describe the
+IITs (although not by this name) were first used to describe the
 well-typed syntax of type theory \cite{nisse,chapman09eatitself}. Agda
@@
 semantics, however without proving existence of initial
-algebras. Their notion of signature, just as Nordvall Forsberg's
+algebras. Their notion of signature, like Nordvall Forsberg's,
 involves more encoding overhead than ours.
@@
 Cartmell \cite{gat} introduced generalised algebraic theories using a
-type-theoretic syntax. Removing equations from his signatures, we
-obtain finitary IIT signatures similar to ours. He does not consider
-constructing initial algebras using simpler classes of inductive
-types.
+type-theoretic syntax. Removing equations from his signatures and only
+considering finite signatures, we obtain finitary IIT signatures similar to
+ours. He does not consider constructing initial algebras using simpler classes
+of inductive types.
 
@@
 $\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
-universes. He used UIP, function extensionality and quotient types in
-his formalisation. In this paper we construct a type theory without
+universes. They used UIP, function extensionality and quotient types in
+the formalisation. In this paper we construct a type theory without
 computation rules, hence we avoid using quotients.
@@
 one hand, we take this even further: our notion of signature is given
-by a small type theory tailer-made to describe signatures, we call
-this the \emph{theory of IIT signatures}. On the other hand we would
+by a small type theory tailor-made to describe signatures, which we call
+ the \emph{theory of IIT signatures}. On the other hand we would
 like to avoid using a complicated inductive definition (a type theory
 is a quotient inductive-inductive type \cite{ttintt}) to describe a
-simpler class of inductive types. Hence we use a Church encoding
-\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
+simpler class of inductive types. Hence we use a Church-encoding
+\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures, thereby
 avoiding the need for pre-existing inductive definitions. Another
@@
   this method, signatures are described in the internal language of
-  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
+  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We do not
   use this approach because it is more technical, and it would not
-  strengthen our main result \Cref{thm:WToIITs} as the proof of
-  \Cref{thm:WToToS} needs $\mathcal{C}$ to be a model of ETT.}
+  strengthen our main theorem (TODO: refer) as theorem (TODO: refer)
+  needs $\mathcal{C}$ to be a model of ETT.}
 
@@
 \begin{example}[Example contexts in a signature algebra]\label{ex:contexts}
-  Given a signature algebra, we can define a context in which
-  specifies natural numbers. For readability, the an informal version
+  Given a signature algebra, we can define a context which
+  specifies natural numbers. For readability, an informal version
   of the same context is displayed on the right using variable names.
@@
   \]
-  We start with the empty context $\boldsymbol{\cdot}$, then we
-  declare a sort $\U$, then we declare an operator producing an
-  element of the sort denoted by $\El\,0$ where $0$ is the de Bruijn
-  index referring to the sort. Finally, we declare an operator which
-  takes as input an element of the sort (now it became de Bruijn index
-  $1$) and produces an element of the same sort. Note the assymetry of
-  the function type $\Ra$: on the left there needs to be an element of
-  $\U$, while on the right any type can appear (including another
-  function type). This ensures strict positivity of the operators.
+  We start with the empty context $\boldsymbol{\cdot}$, then we declare a sort
+  $\U$, then we declare an operator producing an element of the sort denoted by
+  $\El\,0$ where $0$ is the de Bruijn index referring to the sort. Finally, we
+  declare an operator which takes as input an element of the sort (now it became
+  de Bruijn index $1$) and produces an element of the same sort. Note the
+  asymmetry of the function type $\Ra$: the domain element of $\U$, while the
+  codomain can be any type (including another function type). This ensures
+  strict positivity of the operators.
 
@@
   following context. Here we use the function space with external
-  domain $\Ram$ to include a $T$ in the signature. For readability, we
-  omit the $\lambda$ and the superscripts and we don't write the
+  domain $\Ram$ to include $T$ in the signature. For readability, we
+  omit the $\lambda$ and the superscripts and we do not write the
   compatibility condition. On the right we list the same signature
@@
 take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
-usual Church-encoding of contexts. However (as we will see in
-\Cref{remark:sigc}) the notion of constructor for such signatures
-would be too strong. Another approach would be to assume that there is
-a syntax for signature algebras (an initial signature algebra), and
-then a signature would be a context in this signature algebra. We will
-define syntactic signatures using this approach in the next section
-(\Cref{def:syntacticsignature}), but for now we don't want to assume
+usual Church encoding of contexts. However (as we will see in TODO)
+the notion of constructor for such signatures would be too
+strong. Another approach would be to assume that there is a syntax for
+signature algebras (an initial signature algebra), and then a
+signature would be a context in this signature algebra. We will define
+syntactic signatures using this approach in the next section
+(\Cref{def:syntacticsignature}), but for now we do not want to assume
 the existence of any inductive type. Instead, we will use a restricted
@@
 
-\begin{theorem}\label{thm:ToSToIITs}
+\begin{theorem}\label{th:tosToIITs}
   If a model of type theory supports the theory of IIT signatures,
@@
 
-\begin{remark}\label{remark:sigc}
-  In the above proof, we crucially relied on the ${sigc}$ property to
-  define the constructors (and the eliminator). This is why the simple
-  Church encoding of signatures is not sufficient.
-\end{remark}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 W-types also supports the theory of signatures, and as a consequence
-of \Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal
+of \Cref{th:tosToIITs}, all IITs. For this, we work in the internal
 language of a model of ETT supporting indexed W-types

### 2020-05-03 paper/paper.tex
fixes

--- 2020-05-03 paper/paper.tex
@@
 were originally introduced to represent the well-typed syntax of type
-theory itself, and the main example is still the Chapman-style
+theory itself, and a prominent example is still Chapman's
 \cite{chapman09eatitself} syntax for a type theory. A minimised version
@@
 \end{alignat*}
-This simple elimination principle is too weak e.g.\ to define the
+This simple elimination principle is too weak, e.g.\ to define the
 standard (metacircular) interpretation of our small syntax
@@
 Kaposi, Kov{\'a}cs, and Altenkirch
-\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory
-called the theory of signatures to describe quotient
+\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory,
+ called the theory of signatures, to describe quotient
 inductive-inductive types (QIIT). QIITs are generalisations of IITs
@@
 theory of IIT signatures is itself a QIIT, it is both
-inductive-inductive and has equality constructors. However it can be
+inductive-inductive and has equality constructors. However, it can be
 seen as the well-typed syntax of a small type theory without any
@@
   following \cite{DBLP:conf/lics/AwodeyFS18}.
-\item \Cref{sec:constructingiits}. We show that if a model of ETT
-  supports the theory of IIT signatures
-  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
-  sense: \Cref{thm:ToSToIITs}. This is an adaptation of a proof in
+\item \Cref{sec:constructingiits}. In \Cref{thm:ToSToIITs}, we show that if a
+  model of ETT supports the theory of IIT signatures
+  (\Cref{def:theoryofsignatures}), then it supports IITs in the previously
+  given. This is an adaptation of a proof in
   \cite{Kaposi:2019:CQI:3302515.3290315}.
@@
 semantics by restricting to IITs, and we reuse the term model
-construction of QIITs as well. We also make use of the extesion to
+construction of QIITs as well. We also make use of the extension to
 infinitary QIITs \cite{large_inf_qiit} to derive the specification of
@@
 works on specific examples (although the examples vary greatly). He
-also only constructs the simple elimination principle.
+also only constructs simple elimination principles.
 
@@
 $\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
-universes. He used UIP, function extensionality and quotient types in
-his formalisation. In this paper we construct a type theory without
+universes. They used UIP, function extensionality and quotient types in
+the formalisation. In this paper we construct a type theory without
 computation rules, hence we avoid using quotients.
 
-Intrinsic (well-typed) syntax for type theories were constructed using
+Intrinsic (well-typed) syntaxes for type theories were constructed using
 IITs \cite{chapman09eatitself}, inductive-recursive types
@@
   By a model of ETT we mean a category with families (CwF)
-  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
+  \cite{Dybjer96internaltype,Hofmann97syntaxand} with a countable
   predicative hierarchy of universes closed under the following type
@@
 
-In this section we say what it means that a model of ETT supports
+In this section we specify what it means that a model of ETT supports
 IITs. We first define the notion of IIT signature. Signatures for
@@
 one hand, we take this even further: our notion of signature is given
-by a small type theory tailer-made to describe signatures, we call
-this the \emph{theory of IIT signatures}. On the other hand we would
+by a small type theory tailor-made to describe signatures, which we call
+the \emph{theory of IIT signatures}. On the other hand we would
 like to avoid using a complicated inductive definition (a type theory
@@
 simpler class of inductive types. Hence we use a Church encoding
-\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
+\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures, thereby
 avoiding the need for pre-existing inductive definitions. Another
@@
   this method, signatures are described in the internal language of
-  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
+  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We do not
   use this approach because it is more technical, and it would not
@@
   takes as input an element of the sort (now it became de Bruijn index
-  $1$) and produces an element of the same sort. Note the assymetry of
-  the function type $\Ra$: on the left there needs to be an element of
-  $\U$, while on the right any type can appear (including another
+  $1$) and produces an element of the same sort. Note the asymmetry of
+  the function type $\Ra$: the domain needs to be an element of
+  $\U$, while the codomain can be any type (including another
   function type). This ensures strict positivity of the operators.
@@
   domain $\Ram$ to include a $T$ in the signature. For readability, we
-  omit the $\lambda$ and the superscripts and we don't write the
+  omit the $\lambda$ and the superscripts and we do not write the
   compatibility condition. On the right we list the same signature
@@
 define syntactic signatures using this approach in the next section
-(\Cref{def:syntacticsignature}), but for now we don't want to assume
+(\Cref{def:syntacticsignature}), but for now we do not want to assume
 the existence of any inductive type. Instead, we will use a restricted
-Church encoding which needs the notion of morphism of signatures.
+Church encoding. This requires the notion of morphism of signatures.
 
@@
 
-We define IIT signatures using the Church encoding introduced by
-Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
-difference is that we avoid the usage of impredicativity because it is
-not a problem for our use case that signatures are not in the lowest
-universe.
+We define IIT signatures using the Church encoding introduced by Awodey, Frey
+and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A difference is that we avoid
+impredicativity. This restricts the possible eliminations on signatures: we can
+only eliminate into a universe which is smaller than the level of
+signatures. However, this still covers all eliminations in this paper, and it is
+also not an issue for us that signatures do not live in the smallest universe.
+
+%% We define IIT signatures using the Church encoding introduced by
+%% Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
+%% difference is that we avoid the usage of impredicativity because it is
+%% not a problem for our use case that signatures are not in the lowest
+%% universe.
 \begin{definition}[IIT signature]\label{defn:sign}
-  An IIT signature is a context in an arbitrary signature algebra that
-  is compatible with morphisms:
+  An IIT signature is a context in an arbitrary signature algebra, which is also
+  compatible with morphisms:
   \begin{alignat*}{6}
@@
 
-Our notion of signatures don't form a signature algebra.
+The lack of impredicativity implies that our notion of signatures do not form a
+signature algebra.
 \begin{lemma}
@@
 it is expressed as a term (of a function type) in the model. This is
-necessary because of the function space $\Pim$ which has as domain an
+necessary because of the function space $\Pim$, which has as domain an
 arbitrary type in the model. We make use of $\Pim$ in signatures with
@@
 function again says that the function respects the relation, however
-we don't simply say
+we do not simply say
 \[
@@
 model of ETT supports the theory of IIT signatures, then it supports
-all IITs. In the sense of the Church encoding of \Cref{defn:sign},
+all IITs. Using the Church encoding of \Cref{defn:sign},
 every model of ETT can describe ITT signatures. In contrast, in
@@
 
-We will define the term signature algebra using which we obtain the
-constructors $\con$ for any IIT signature. Then we will define another
-signature algebra which provides the eliminator. Before doing these,
-we illustrate the idea of both constructions on natural numbers.
+We will define the term signature algebra by which we obtain the constructors
+$\con$ for any IIT signature. Then we will define another signature algebra
+which provides the eliminator. Before doing these, we illustrate the idea of
+both constructions on natural numbers.
 \begin{example}\label{eg:nat_con}
@@
   or are trivial. We also omit the components for terms and
-  substitutions as their $^\C$ parts are uninformative equational
-  reasonings.
+  substitutions as their $^\C$ parts consist of uninformative equational
+  reasoning.
   \begin{alignat*}{10}
@@
 \begin{example}\label{ex:constructors}
-  Now given a syntactic signature $\GO:\Con^\I$, we get the
+  Now, given a syntactic signature $\GO:\Con^\I$, we get the
   constructors as an $\GO$-algebra by
@@
 
-In this section we show that any model of ETT which supports indexed
-W-types also supports the theory of signatures, and as a consequence
-of \Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal
-language of a model of ETT supporting indexed W-types
-\cite{indexedcont}. Indexed W-types correspond to the usual notion of
-(possibly mutual) indexed inductive types. We will use Agda-style
-notation for defining such inductive families: we list the sorts and
-constructors and use pattern matching when eliminating from them. For
-an encoding of mutual inductive families as families as indexed
-W-types, see e.g.\ \cite{mutual}.
+In this section we show that any model of ETT which supports indexed W-types
+also supports the theory of signatures, and as a consequence of
+\Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal language of a
+model of ETT supporting indexed W-types \cite{indexedcont}. Indexed W-types
+correspond to the usual notion of (possibly mutual) indexed inductive types. We
+use Agda-style notation to define such inductive families: we list the sorts and
+constructors and use pattern matching when eliminating from them. For an
+encoding of mutual inductive families as indexed W-types, see
+e.g.\ \cite{mutual}.
 
@@
 
-\section{Conclusions}
+\section{Further Work}
 \label{sec:conclusions}

### 2020-05-04 paper/paper.tex
identifiers^S/^D (Con,Ty,..) in mathsf

@@
 \author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
-\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}
+\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@gmail.com}{https://orcid.org/0000-0002-9299-641X}{this author was supported by the CoqHoTT ERC Grant 637339.}
 
@@
 \begin{alignat*}{5}
-  & Con^D && : \Con\ra\Set \\
-  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
-  & {empty}^D && : Con^D\,\mathsf{empty} \\
-  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
-  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
-  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
-%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
+  & \Con^D && : \Con\ra\Set \\
+  & \Ty^D && : \Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
+  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
+  & \mathsf{ext}^D && : (\GG^D:\Con^D\,\GG)\ra \Ty^D\,\GG^D\,A\ra \Con^D\,(\mathsf{ext}\,\GG\,A) \\
+  & \U^D && : (\GG^D:\Con^D\,\GG)\ra \Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
+  & \El^D && : (\GG^D:\Con^D\,\GG)\ra\,\Ty^D\,(\mathsf{ext}^D\,\GG^D\,(\U^D\,\GG^D))\,(\mathsf{El}\,\GG)
+%  & Pi^D && : (\GG^D:\Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
 \end{alignat*}
 Above we used implicit quantifications for $\GG:\Con$ and
-$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
+$A:\Ty\,\GG$ to ease readability, e.g.\ $\Ty^D$ has an implicit
 parameter $\GG$ before its explicit parameter of type
-$Con^D\,\GG$.
+$\Con^D\,\GG$.
 
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
-  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
+  & \mathsf{elimCon} && : (\GG:\Con)\ra \Con^D\,\GG \\
+  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra \Ty^D\,(\mathsf{elimCon}\,\GG)\,A
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
-  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
-  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
-  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
+  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
+  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
+  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = \U^D\,(\mathsf{elimCon}\,\GG) \\
+  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = \El^D\,(\mathsf{elimCon}\,\GG)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & Con^S && : \Con\ra\Set \\
-  & Ty^S && : \Ty\,\GG\ra\Set \\
-  & {empty}^S && : Con^S\,\mathsf{empty} \\
-  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
-  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
-  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
-  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
-  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
+  & \Con^S && : \Con\ra\Set \\
+  & \Ty^S && : \Ty\,\GG\ra\Set \\
+  & \mathsf{empty}^S && : \Con^S\,\mathsf{empty} \\
+  & \mathsf{ext}^S && : \Con^S\,\GG\ra \Ty^S\,\,A\ra \Con^S\,(\mathsf{ext}\,\GG\,A) \\
+  & \U^S && : \Con^S\,\GG\ra \Ty^S\,(\mathsf{U}\,\GG) \\
+  & \El^S && : \Con^S\,\GG\ra\,\Ty^S\,(\mathsf{El}\,\GG) \\
+  & \mathsf{selimCon} && : (\GG:\Con)\ra \Con^S\,\GG \\
+  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra \Ty^S\,A
 \end{alignat*}

### 2020-05-04 paper/paper.tex
use signature algebra instead of model

@@
 The compatibility condition says that if we obtain an $M$-context
-using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
+using ${sig}$ at signature algebra $M$ and then we transport it to $N$ using $f$,
 we get the same $N$-context as directly applying ${sig}$ to $N$.
@@
 \begin{theorem}\label{thm:ToSToIITs}
-  If a model of type theory supports the theory of IIT signatures,
+  If a model of ETT supports the theory of IIT signatures,
   then it supports all IITs.

### 2020-05-04 paper/universaliit.tex
finish
small fixes
fixes
use signature algebra instead of model

@@
   \item defining a relation $\blank \sim \blank$ between the well-typed syntax
-    and a given model. The idea is that given a syntactic context $\GG$ and a
-    semantic context $\model{\GG}$ of the model $M$, we have
+    and a given signature algebra. The idea is that given a syntactic context $\GG$ and a
+    semantic context $\model{\GG}$ of the signature algebra $M$, we have
     $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
@@
     % syntax to a given model;
-  \item showing that this relation is functional and thus obtaining a function.
+  \item showing that this relation is functional and thus obtaining a morphism.
     \end{enumerate}
@@
 Crucially, we do not have conversion relations for typed syntax, nor
-do we need to use quotients when giving the syntactic model. This is
+do we need to use quotients when constructing the syntax. This is
 possible because there are no $\beta$-rules in the theory of
@@
 
-\subsubsection{The functional relation}
+\subsubsection{The Functional Relation}
 
@@
 
-\subsubsection{Right Uniqueness}
+\subsubsection{Right Uniqueness}\label{sec:right_uniqueness}
 
@@
   UIP \cite{jasper} in cubical type theory, and demonstrated
-  that well-formedness predicates used in syntactic models can subtly break
+  that well-formedness predicates used in syntactic algebras can subtly break
   in that setting. Also, while Hugunin does not use UIP, he only shows the simple version

### 2020-07-08 paper/paper.tex
newest lipics style and some small fixes in the paper

@@
 Thematic Excellence Programme funding scheme, Project
-no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
-solutions), by the New National Excellence Program of the Ministry
+no.\ ED\_18-1-2019-0030, by the New National Excellence Program of the Ministry
 for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
@@
 
-\Copyright{A. Kaposi, A. Kov{\'a}cs and A. Lafont}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Ambrus Kaposi, Andr{\'a}s Kov{\'a}cs and Ambroise Lafont}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\category{}%optional, e.g. invited paper
+%\category{}%optional, e.g. invited paper
 
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
+\EventEditors{Marc Bezem and Assia Mahboubi}
 \EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventLongTitle{25th International Conference on Types for Proofs and Programs (TYPES 2019)}
+\EventShortTitle{TYPES 2019}
+\EventAcronym{TYPES}
+\EventYear{2019}
+\EventDate{June 11--14, 2019}
+\EventLocation{Oslo, Norway}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{175}
+\ArticleNo{6}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   branching IITs. We use a small internal type theory called the
-  theory of signatures to specify signatures for IITs. We show that if
-  a model of ETT supports the syntax for the theory of signatures,
-  then it supports all IITs. We construct this syntax from indexed
-  W-types using preterms and typing relations and prove its initiality
-  following Streicher. The construction of the syntax and its
-  initiality proof were formalised in Agda.
+  theory of signatures to specify IITs. We show that if a model of ETT
+  supports the syntax for the theory of signatures, then it supports
+  all IITs. We construct this syntax from indexed W-types using
+  preterms and typing relations and prove its initiality following
+  Streicher. The construction of the syntax and its initiality proof
+  were formalised in Agda.
 \end{abstract}
@@
 proof assistant Agda \cite{norell07thesis} allows defining such
-functions (even from non-IITs) and it is currently the only proof
+functions (even from non-IITs) and is currently the only proof
 assistant supporting IITs\footnote{An experimental version of Coq with
@@
 theory, but he also only constructed the simple eliminator. To
-illustrate the difference, we list the motives and methods and the
+illustrate the difference, we list the motives, methods and the
 simple elimination principle for the $\Con$--$\Ty$ example. Again, we
@@
 \end{alignat*}
-This simple elimination principle is too weak, e.g.\ to define the
-standard (metacircular) interpretation of our small syntax
-\cite{ttintt}. Using pattern matching notation, this interpretation is
-the following:
+This simple elimination principle is not capable of defining standard
+(metacircular) interpretation \cite{ttintt} of our small syntax. Using
+pattern matching notation, this interpretation is the following:
 \begin{alignat*}{5}
@@
 the elimination principle in the style of the initiality proof of
-Streicher \cite{streichersemantics}.
+Streicher.
+
+Streicher \cite{streichersemantics} constructs the syntactic model of
+type theory using well-typed preterms and then shows initiality of
+this model by (1) defining a partial map to any other model by
+induction on preterms and (2) showing that whenever this partial
+function receives a well-typed preterm on its input it actually gives
+an output. Instead of defining a partial function, we define the graph
+of the same function as a relation and then show that it is functional
+as a second step. This can be seen as an indexed variant of the
+construction using partial functions.
 
@@
   model of ETT supports the theory of IIT signatures
-  (\Cref{def:theoryofsignatures}), then it supports IITs in the previously
-  given. This is an adaptation of a proof in
+  (\Cref{def:theoryofsignatures}), then it supports IITs. This is an adaptation of a proof in
   \cite{Kaposi:2019:CQI:3302515.3290315}.

### 2020-10-21 paper/generalTerm.tex
a note on the most general version of the term model

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{csquotes} %for \llbracket and \rrbracket
\usepackage{stmaryrd}
% \usepackage{pdflscape} % if we want \begin{landscape} ... \end{landscape}

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\begin{document}

\input{abbrevs.tex}

\begin{definition}[Term signature algebra $\A\blank\C_{\blank}$]
  For an $M:\SignAlg$ and $\Omega:\Con^M$, we define ${\A M\C}_\Omega$
  which we call the term signature algebra. It is equipped with
  morphisms $\blank^\A : \SignMor\,({\A M\C}_\Omega)\,\A$ and
  $\blank^M : \SignMor\,({\A M\C}_\Omega)\,M$.

  We define ${\A M\C}_\Omega$ by listing its components $\Con$, $\Ty$,
  $\Sub$, and so on, one per row. Each component has three parts
  denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are just
  reusing the corresponding components from $\A$ and $M$,
  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
  defined as the obvious projections. We omit the equality components,
  as they come from UIP or are trivial. We also omit the components
  for terms and substitutions as their $^\C$ parts are uninformative
  equational reasonings.
  \begin{alignat*}{10}
    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
  \end{alignat*}
\end{definition}

\begin{remark}
  Given a syntactic signature $\Omega : \Con^\I$, we can use the term
  signature algebra with choices other than $M = \I$. For example, if
  $M = \A$, we usual get the Church encoding by
  $(\ll\Omega\rr_{{\A\A\C}_{\ll\Omega\rr^\A}})^\C\,\id^\A$. In the
  case of natural numbers, this computes (up to curry-uncurry) to
  $(N:\Set)(z:N)(s:N\ra N)\ra N$. Using $M = {\A\M}$ (the signature
  algebra $\A$ extended with the binary version of $\D$ also called
  logical relations, see \cite[Section
  5]{Kaposi:2019:CQI:3302515.3290315}), we obtain the Church encoding
  of Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}.
\end{remark}

\end{document}
